---
title: "Replicates Analysis for QC of Biomarker Data"
author: "ERM"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---
# A. Overview

This protocol describes a workflow for analysis of the replicate and control samples (HIHG replicates, SC, IPC, NC) for QC purposes. Goals to identify problems with NC or IPCs and any problem plates or batch effects.

Look at CVs within plates, within Runs and overall

12/23/2025 This version corrects the CV calculation to work with the ratio before taking log2 for the NPQ.

**For a new QC run, change the title in YAML and this RMarkdown file name (where .html output will go)**


## A.1 Input Data

Input files are auto-detected from `input_files/`. If no files are found, the pipeline will attempt to copy from `Primary_QC/input_files/` (the raw data files, not the processed outputs).

**Required Input Data Files:** Place all files in a folder called `input_files` in the working directory (or let the pipeline copy them from Primary_QC/input_files).

1. **Biomarker NPQ data** (`NPQ_xxxxxxxx.csv`) – Rows: biomarkers; Columns: samples; downloaded from NAS. Assign the filename to the variable `NPQ_data` below.
2. **Raw count data** (`Raw_counts.csv`) – Rows: biomarkers; Columns: samples; exported from NPQ_Values.xlsx.
3. **Replicate list** (`Replicate_list.csv`) – List of HIHG-replicate sample IDs for assay precision checks. Created by analyst.


**Output Data Files:** 

1. RMarkdown html file – named after this file. Saved in working directory.



## A.2 Thresholds

Adjustable QC parameters include:

| Parameter | Default | Description |
|-----------|---------|-------------|
| cv_threshold | 25 | CV (%) above which a biomarker is flagged at plate / Run / Bay / overall level |

## A.3 Initial Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# ==============================================================================
# INPUT FILE SETUP
# ==============================================================================

# Create input/output directories if they don't exist
if (!dir.exists("input_files")) dir.create("input_files")
if (!dir.exists("output_files")) dir.create("output_files")

# Check if input files exist, if not copy from Primary_QC/input_files
# Note: Replicate_Analysis uses the RAW NPQ file (biomarkers as rows, samples as columns)
#       NOT the processed post-QC file from Primary_QC/output_files
primary_qc_input_dir <- "../../Primary_QC/input_files"

# --- NPQ file ---
# Look for NPQ file in input_files (any file starting with NPQ and ending in .csv)
npq_files <- list.files("input_files", pattern = "^NPQ.*\\.csv$", full.names = TRUE)

if (length(npq_files) == 0 && dir.exists(primary_qc_input_dir)) {
  # No NPQ file found - try to copy from Primary_QC/input_files
  primary_npq <- list.files(primary_qc_input_dir, pattern = "^NPQ.*\\.csv$", full.names = TRUE)

  if (length(primary_npq) > 0) {
    # Use most recent file if multiple exist
    if (length(primary_npq) > 1) {
      primary_npq <- primary_npq[which.max(file.mtime(primary_npq))]
    }
    file.copy(primary_npq, "input_files/")
    message("Copied NPQ file from Primary_QC/input_files: ", basename(primary_npq))
    npq_files <- list.files("input_files", pattern = "^NPQ.*\\.csv$", full.names = TRUE)
  }
}

# Set NPQ filename (use most recent if multiple)
if (length(npq_files) > 0) {
  if (length(npq_files) > 1) {
    npq_files <- npq_files[which.max(file.mtime(npq_files))]
  }
  NPQ_data <- basename(npq_files)
} else {
  stop("No NPQ file found in input_files/ and could not copy from Primary_QC/input_files/")
}

# --- Raw counts file ---
if (!file.exists("input_files/Raw_counts.csv") && dir.exists(primary_qc_input_dir)) {
  primary_raw <- file.path(primary_qc_input_dir, "Raw_counts.csv")
  if (file.exists(primary_raw)) {
    file.copy(primary_raw, "input_files/")
    message("Copied Raw_counts.csv from Primary_QC/input_files")
  }
}

# --- Sample_QC.csv file ---
if (!file.exists("input_files/Sample_QC.csv") && dir.exists(primary_qc_input_dir)) {
  primary_qc_file <- file.path(primary_qc_input_dir, "Sample_QC.csv")
  if (file.exists(primary_qc_file)) {
    file.copy(primary_qc_file, "input_files/")
    message("Copied Sample_QC.csv from Primary_QC/input_files")
  }
}

# --- Replicate list ---
# Try to copy from Primary_QC/input_files if it exists there
if (!file.exists("input_files/Replicate_list.csv")) {
  primary_rep <- file.path(primary_qc_input_dir, "Replicate_list.csv")
  if (file.exists(primary_rep)) {
    file.copy(primary_rep, "input_files/")
    message("Copied Replicate_list.csv from Primary_QC/input_files")
  } else {
    warning("Replicate_list.csv not found. This file must be created manually.")
  }
}

# ==============================================================================
# PARAMETERS
# ==============================================================================

cv_threshold = 25

```

**Input files detected:**
- NPQ file: `r NPQ_data`
- Raw counts: `r if(file.exists("input_files/Raw_counts.csv")) "Found" else "MISSING"`
- Sample QC: `r if(file.exists("input_files/Sample_QC.csv")) "Found" else "MISSING"`
- Replicate list: `r if(file.exists("input_files/Replicate_list.csv")) "Found" else "MISSING - must create manually"`
---
  
# B. Data Formatting
This section formats the data and produces initial summaries.

## B.1 Get Data 
Input file is: `r NPQ_data`

```{r load-and-format, message=FALSE, warning=FALSE, echo=TRUE}
library(tidyverse)

# --- Load NPQ data & reshape ----------------------------------------------
combined_df <- read_csv(
  file.path("input_files", NPQ_data),
  show_col_types = FALSE
) %>%
  rename(Biomarker = 1) %>%            # first column = biomarker names
  pivot_longer(-Biomarker, names_to = "SampleID", values_to = "Value") %>%
  pivot_wider(names_from = Biomarker, values_from = Value)

# --- Load QC metadata ------------------------------------------------------
meta_df <- read_csv(
  "input_files/Sample_QC.csv",
  skip = 12,
  show_col_types = FALSE
) %>%
  mutate(
    Run = str_extract(plateID, "^[0-9]{8}-[0-9]{4}"),
    Bay = str_extract(plateID, "Bay[0-9]+")
  ) %>%
  transmute(SampleID = `Sample Name`, Run, Bay)

# --- Attach metadata -------------------------------------------------------
combined_df <- combined_df %>%
  inner_join(meta_df, by = "SampleID") %>%
  relocate(SampleID, Run, Bay)

# --- Load raw counts & standardize format ---------------------------------
raw_df <- read_csv("input_files/Raw_counts.csv", show_col_types = FALSE) %>%
  rename(Biomarker = 1) %>%
  pivot_longer(-Biomarker, names_to = "SampleID", values_to = "Raw") %>%
  pivot_wider(names_from = Biomarker, values_from = Raw) %>%
  inner_join(meta_df, by = "SampleID") %>%
  relocate(SampleID, Run, Bay)

# --- Internal control (mCherry) ------------------------------------------------
# The old NAS format carried the mCherry IC as a row in Raw-counts; the new CNS Disease Panel V2
# export drops that row but keeps the per-sample IC read count in Sample QC as `IC Reads` (verified
# byte-identical to the legacy mCherry Raw-counts row). Inject it as an `mCherry` column when absent
# so the IC sections (B.2 IC reps / C.2 IC CV / normalization) work on both formats.
if (!"mCherry" %in% names(raw_df)) {
  ic_lookup <- read_csv("input_files/Sample_QC.csv", skip = 12, show_col_types = FALSE) %>%
    transmute(SampleID = `Sample Name`, mCherry = suppressWarnings(as.numeric(`IC Reads`))) %>%
    filter(!is.na(SampleID), !is.na(mCherry))
  if (nrow(ic_lookup) > 0) {
    raw_df <- raw_df %>% left_join(ic_lookup, by = "SampleID")
    message(sprintf("Injected mCherry IC from Sample QC 'IC Reads' for %d wells (new-panel format).",
                    sum(!is.na(raw_df$mCherry))))
  }
}
```
  

## B.2 Pull Reps and Controls
Pull out Alamar control and HIHG replicate samples (NPQs and Reads)

```{r pull_samples, message=FALSE, warning=FALSE, echo=TRUE}
# ==============================================================================
# B.2 PULL REPS AND CONTROLS
# ==============================================================================

library(dplyr)
library(tidyr)

# --- Read HIHG replicate list ---
rep <- read_csv("input_files/Replicate_list.csv", show_col_types = FALSE)

# --- EXTRACT REPS (Wide format) ---
# NPQs
nc_samples_npq   <- combined_df[grepl("^NC", combined_df$SampleID), ]
sc_samples_npq   <- combined_df[grepl("^SC", combined_df$SampleID), ]
ipc_samples_npq  <- combined_df[grepl("^IPC", combined_df$SampleID), ]
hihg_reps_npq    <- combined_df[combined_df$SampleID %in% rep[[1]], ]

# Raw reads
nc_samples_reads   <- raw_df[grepl("^NC", raw_df$SampleID), ]
sc_samples_reads   <- raw_df[grepl("^SC", raw_df$SampleID), ]
ipc_samples_reads  <- raw_df[grepl("^IPC", raw_df$SampleID), ]
hihg_reps_reads    <- raw_df[raw_df$SampleID %in% rep[[1]], ]

# --- GET ALL REP SAMPLE IDS ---
rep_sample_ids <- unique(c(
  nc_samples_npq$SampleID,
  ipc_samples_npq$SampleID,
  sc_samples_npq$SampleID,
  hihg_reps_npq$SampleID
))

# --- CONVERT TO 2^NPQ FOR CV CALCULATIONS ---
nc_samples   <- nc_samples_npq %>% mutate(across(where(is.numeric), ~ 2^.x))
sc_samples   <- sc_samples_npq %>% mutate(across(where(is.numeric), ~ 2^.x))
ipc_samples  <- ipc_samples_npq %>% mutate(across(where(is.numeric), ~ 2^.x))
hihg_reps    <- hihg_reps_npq %>% mutate(across(where(is.numeric), ~ 2^.x))

# --- SUMMARY TABLE ---
summarize_counts <- function(df, type_name) {
  df %>%
    group_by(Run, Bay) %>%
    summarise(!!type_name := n(), .groups = "drop")
}

control_summary <- full_join(
  summarize_counts(nc_samples, "NC"),
  summarize_counts(ipc_samples, "IPC"),
  by = c("Run","Bay")
) %>%
  full_join(summarize_counts(sc_samples, "SC"), by = c("Run","Bay")) %>%
  full_join(summarize_counts(hihg_reps, "HIHG_Replicate"), by = c("Run","Bay")) %>%
  replace_na(list(NC = 0, IPC = 0, SC = 0, HIHG_Replicate = 0)) %>%
  arrange(Run, Bay)

print(as.data.frame(control_summary))

# --- WRITE OUTPUT FILES ---
write_csv(hihg_reps_npq, "output_files/hihg_reps_NPQ.csv")
write_csv(sc_samples_npq, "output_files/sc_reps_NPQ.csv")
write_csv(nc_samples_npq, "output_files/nc_reps_NPQ.csv")
write_csv(ipc_samples_npq, "output_files/ipc_reps_NPQ.csv")
write_csv(hihg_reps_reads, "output_files/hihg_reps_READS.csv")
write_csv(sc_samples_reads, "output_files/sc_reps_READS.csv")
write_csv(nc_samples_reads, "output_files/nc_reps_READS.csv")
write_csv(ipc_samples_reads, "output_files/ipc_reps_READS.csv")

```
 

# C. Analysis

## C.1 Coefficient of Variation
Compute CVs for each biomarker comparing replicates at different levels (within Plates, within Runs and within Bays). CVs are already on the % scale. Bad CVs would be ~ >20.

Alamar excludes values below LOD from CV calculations. I did not exclude anything.

I need to change NPQ to 2^NPQ

In general, the largest CV's for raw reads for each of the rep types are in Run 0804 when we collapse over Bays. Below we see how this makes sense for mCherry because the reads have so much variability over bays in that run.

```{r biomarker-cv, message=FALSE, warning=FALSE, echo=TRUE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

cv_plots <- function(df, top_n = 10, cv_threshold = 25){

  # Long format
  df_long <- df %>%
    pivot_longer(
      cols = -c(SampleID, Run, Bay),
      names_to = "Biomarker",
      values_to = "Value"
    )

  # CV per Plate (Run × Bay)
  cv_by_run_bay <- df_long %>%
    group_by(Run, Bay, Biomarker) %>%
    summarize(
      n    = n(),
      mean = mean(Value, na.rm = TRUE),
      sd   = sd(Value, na.rm = TRUE),
      CV   = 100 * sd / mean,
      .groups = "drop"
    )

  # CV per Run
  cv_by_run <- df_long %>%
    group_by(Run, Biomarker) %>%
    summarize(
      n    = n(),
      mean = mean(Value, na.rm = TRUE),
      sd   = sd(Value, na.rm = TRUE),
      CV_run   = 100 * sd / mean,
      .groups = "drop"
    )

  # CV per Bay
  cv_by_bay <- df_long %>%
    group_by(Bay, Biomarker) %>%
    summarize(
      n    = n(),
      mean = mean(Value, na.rm = TRUE),
      sd   = sd(Value, na.rm = TRUE),
      CV_bay   = 100 * sd / mean,
      .groups = "drop"
    )

  # CV overall (all data)
cv_overall <- df_long %>%
  group_by(Biomarker) %>%
  summarize(
    mean = mean(Value, na.rm = TRUE),
    sd   = sd(Value, na.rm = TRUE),
    CV_overall = 100 * sd / mean,
    .groups = "drop"
  )

  # ---- HIGH CV TRACKING ----
  # Plate level
  high_cv_plate <- cv_by_run_bay %>%
    filter(CV > cv_threshold) %>%
    mutate(Level = "Plate",
           Location = paste0(Run, "_", Bay)) %>%
    select(Biomarker, Level, Location, CV, n, mean, sd)
  
  # Run level
  high_cv_run <- cv_by_run %>%
    filter(CV_run > cv_threshold) %>%
    mutate(Level = "Run",
           Location = Run,
           CV = CV_run) %>%
    select(Biomarker, Level, Location, CV, n, mean, sd)
  
  # Bay level
  high_cv_bay <- cv_by_bay %>%
    filter(CV_bay > cv_threshold) %>%
    mutate(Level = "Bay",
           Location = Bay,
           CV = CV_bay) %>%
    select(Biomarker, Level, Location, CV, n, mean, sd)
  
  # Overall level
  high_cv_overall <- cv_overall %>%
    filter(CV_overall > cv_threshold) %>%
    mutate(Level = "Overall",
           Location = "All",
           CV = CV_overall,
           n = NA_integer_) %>%
    select(Biomarker, Level, Location, CV, n, mean, sd)
  
  # Combine all high CV instances
  high_cv_all <- bind_rows(
    high_cv_plate,
    high_cv_run,
    high_cv_bay,
    high_cv_overall
  ) %>%
    arrange(Biomarker, Level, desc(CV))
  
 # Summary: unique biomarkers with high CV at any level
high_cv_biomarkers <- high_cv_all %>%
  group_by(Biomarker) %>%
  summarize(
    n_instances = n(),
    max_CV = max(CV, na.rm = TRUE),
    levels_affected = paste(unique(Level), collapse = ", "),
    worst_location = Location[which.max(CV)],
    .groups = "drop"
  ) %>%
  arrange(desc(max_CV))
  
  # Top worst biomarkers for each category
  top_plate   <- cv_by_run_bay %>% group_by(Biomarker) %>% summarize(max_CV = max(CV, na.rm = TRUE), .groups = "drop") %>% slice_max(max_CV, n = top_n) %>% pull(Biomarker)
  top_run     <- cv_by_run %>% group_by(Biomarker) %>% summarize(max_CV = max(CV_run, na.rm = TRUE), .groups = "drop") %>% slice_max(max_CV, n = top_n) %>% pull(Biomarker)
  top_bay     <- cv_by_bay %>% group_by(Biomarker) %>% summarize(max_CV = max(CV_bay, na.rm = TRUE), .groups = "drop") %>% slice_max(max_CV, n = top_n) %>% pull(Biomarker)
  top_overall <- cv_overall %>% slice_max(CV_overall, n = top_n) %>% pull(Biomarker)

  # Wide tables for Markdown
  plate_table <- cv_by_run_bay %>%
    filter(Biomarker %in% top_plate) %>%
    mutate(Biomarker = factor(Biomarker, levels = top_plate)) %>%
    arrange(Biomarker) %>%
    mutate(Run_Bay = paste0(Run, "_", Bay)) %>%
    select(Biomarker, Run_Bay, CV) %>%
    distinct(Biomarker, Run_Bay, .keep_all = TRUE) %>%
    pivot_wider(names_from = Run_Bay, values_from = CV)

  run_table <- cv_by_run %>%
    filter(Biomarker %in% top_run) %>%
    mutate(Biomarker = factor(Biomarker, levels = top_run)) %>%
    arrange(Biomarker) %>%
    select(Biomarker, Run, CV_run) %>%
    distinct(Biomarker, Run, .keep_all = TRUE) %>%
    pivot_wider(names_from = Run, values_from = CV_run)

  bay_table <- cv_by_bay %>%
    filter(Biomarker %in% top_bay) %>%
    mutate(
      Biomarker = factor(Biomarker, levels = top_bay),
      stat_string = sprintf(
        "CV=%.1f%%; mean=%.1f; sd=%.1f",
        CV_bay, mean, sd
      )
    ) %>%
    arrange(Biomarker) %>%
    select(Biomarker, Bay, stat_string) %>%
    distinct(Biomarker, Bay, .keep_all = TRUE) %>%
    pivot_wider(
      names_from = Bay,
      values_from = stat_string
    )

  overall_table <- cv_overall %>%
  filter(Biomarker %in% top_overall) %>%
  mutate(
    Biomarker = factor(Biomarker, levels = top_overall)
  ) %>%
  arrange(Biomarker) %>%
  select(Biomarker, CV_overall, mean, sd)  # MODIFY THIS LINE
  
  # Plots
  plot_plate <- ggplot(cv_by_run_bay, aes(x = Bay, y = CV)) +
    geom_boxplot() +
    geom_hline(yintercept = cv_threshold, linetype = "dashed", color = "red", alpha = 0.7) +
    facet_wrap(~ Run) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1)
    ) +
    labs(title = "Per-plate CV by Bay and Run", x = "Bay", y = "CV (%)",
         caption = paste0("Each point represents one Plate (Run × Bay × Biomarker). Red line = CV threshold (", cv_threshold, "%)."))

  plot_run <- ggplot(cv_by_run, aes(x = Run, y = CV_run)) +
    geom_boxplot() +
    geom_hline(yintercept = cv_threshold, linetype = "dashed", color = "red", alpha = 0.7) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1)
    ) +
    labs(title = "CV per Run over Bays", x = "Run", y = "CV (%)",
         caption = paste0("CV per biomarker across Bays within each Run. Red line = CV threshold (", cv_threshold, "%)."))

  plot_bay <- ggplot(cv_by_bay, aes(x = Bay, y = CV_bay)) +
    geom_boxplot() +
    geom_hline(yintercept = cv_threshold, linetype = "dashed", color = "red", alpha = 0.7) +
    theme_bw() +
    labs(title = "CV per Bay over Runs", x = "Bay", y = "CV (%)",
         caption = paste0("CV per biomarker across Runs within each Bay. Red line = CV threshold (", cv_threshold, "%)."))

  plot_overall <- ggplot(cv_overall, aes(x = "", y = CV_overall)) +
    geom_boxplot() +
    geom_hline(yintercept = cv_threshold, linetype = "dashed", color = "red", alpha = 0.7) +
    theme_bw() +
    labs(title = "Overall CV Across All Plates", x = "", y = "CV (%)",
         caption = paste0("CV for each biomarker across all Runs and Bays. Red line = CV threshold (", cv_threshold, "%)."))

  # Captions for tables
  plate_table_caption   <- "Top plate-level CV biomarkers (worst CVs across Run × Bay)."
  run_table_caption     <- "Top CV biomarkers per Run."
  bay_table_caption     <- "Top CV biomarkers per Bay."
  overall_table_caption <- "Top overall CV biomarkers across all data."

  # Return
  return(list(
    plots = list(
      plate   = plot_plate,
      run     = plot_run,
      bay     = plot_bay,
      overall = plot_overall
    ),
    tables = list(
      plate   = list(data = plate_table, caption = plate_table_caption),
      run     = list(data = run_table, caption = run_table_caption),
      bay     = list(data = bay_table, caption = bay_table_caption),
      overall = list(data = overall_table, caption = overall_table_caption)
    ),
    cv_by_run_bay = cv_by_run_bay,
    cv_overall = cv_overall,
    high_cv = list(
      all_instances = high_cv_all,
      summary = high_cv_biomarkers,
      threshold = cv_threshold
    )
  ))
}
```

### C.1.1 Inter-plate Control Samples (IPC)
3 per plate


```{r ipc_cv, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
res <- cv_plots(ipc_samples_reads, top_n = 10)

# Plots
res$plots$plate
knitr::kable(res$tables$plate$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::add_header_above(c(" " = 1, "Run × Bay Plates" = ncol(res$tables$plate$data) - 1)) %>%
  kableExtra::footnote(res$tables$plate$caption)

res$plots$run
knitr::kable(res$tables$run$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$run$caption)

res$plots$bay
knitr::kable(res$tables$bay$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$bay$caption)

res$plots$overall
knitr::kable(res$tables$overall$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$overall$caption)
```

#### NPQ

```{r ipc_cv_npq, message=FALSE, warning=FALSE, echo=TRUE}

res <- cv_plots(ipc_samples, top_n = 10)

# Plots
res$plots$plate
knitr::kable(res$tables$plate$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::add_header_above(c(" " = 1, "Run × Bay Plates" = ncol(res$tables$plate$data) - 1)) %>%
  kableExtra::footnote(res$tables$plate$caption)

res$plots$run
knitr::kable(res$tables$run$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$run$caption)

res$plots$bay
knitr::kable(res$tables$bay$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$bay$caption)

res$plots$overall
knitr::kable(res$tables$overall$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$overall$caption)


# Calculate raw read stats
raw_stats <- ipc_samples_reads %>%
  pivot_longer(cols = -c(SampleID, Run, Bay), 
               names_to = "Biomarker", 
               values_to = "Value") %>%
  group_by(Biomarker) %>%
  summarize(
    reads_mean = mean(Value, na.rm = TRUE),
    reads_min = min(Value, na.rm = TRUE),
    reads_max = max(Value, na.rm = TRUE),
    .groups = "drop"
  )
# Summary
ipc_hi <- res$high_cv$summary %>%
  left_join(raw_stats, by = "Biomarker")

```

C.1.2 Negative Controls (NC)
4 per plate

Excluding NCs from this evaluation because CVs are not so useful here. We can pull the back in below to look at specifica examples.


NaN happens if the mean is 0 (which means the var is also 0). 200 is the max when exactly one NC is 1 and the others are 0. CVs may not be appropriate for very low counts.

```{r nc_cv, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
res <- cv_plots(nc_samples_reads, top_n = 10)

# Plots
res$plots$plate
knitr::kable(res$tables$plate$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::add_header_above(c(" " = 1, "Run × Bay Plates" = ncol(res$tables$plate$data) - 1)) %>%
  kableExtra::footnote(res$tables$plate$caption)

res$plots$run
knitr::kable(res$tables$run$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$run$caption)

res$plots$bay
knitr::kable(res$tables$bay$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$bay$caption)

res$plots$overall
knitr::kable(res$tables$overall$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$overall$caption)
```

 NPQ

```{r nc_cv_npq, message=FALSE, warning=FALSE, echo=TRUE,eval=FALSE}

res <- cv_plots(nc_samples, top_n = 10)

# Plots
res$plots$plate
knitr::kable(res$tables$plate$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::add_header_above(c(" " = 1, "Run × Bay Plates" = ncol(res$tables$plate$data) - 1)) %>%
  kableExtra::footnote(res$tables$plate$caption)

res$plots$run
knitr::kable(res$tables$run$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$run$caption)

res$plots$bay
knitr::kable(res$tables$bay$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$bay$caption)

res$plots$overall
knitr::kable(res$tables$overall$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$overall$caption)
```

### C.1.3 Sample Controls (SC)
3 per plate


```{r sc_cv, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}

res <- cv_plots(sc_samples_reads, top_n = 10)

# Plots
res$plots$plate
knitr::kable(res$tables$plate$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::add_header_above(c(" " = 1, "Run × Bay Plates" = ncol(res$tables$plate$data) - 1)) %>%
  kableExtra::footnote(res$tables$plate$caption)

res$plots$run
knitr::kable(res$tables$run$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$run$caption)

res$plots$bay
knitr::kable(res$tables$bay$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$bay$caption)

res$plots$overall
knitr::kable(res$tables$overall$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$overall$caption)


```

#### NPQ

```{r sc_cv_npq, message=FALSE, warning=FALSE, echo=TRUE}
res <- cv_plots(sc_samples, top_n = 10)

# Plots
res$plots$plate
knitr::kable(res$tables$plate$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::add_header_above(c(" " = 1, "Run × Bay Plates" = ncol(res$tables$plate$data) - 1)) %>%
  kableExtra::footnote(res$tables$plate$caption)

res$plots$run
knitr::kable(res$tables$run$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$run$caption)

res$plots$bay
knitr::kable(res$tables$bay$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$bay$caption)

res$plots$overall
knitr::kable(res$tables$overall$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$overall$caption)

# Calculate raw read stats
raw_stats <- sc_samples_reads %>%
  pivot_longer(cols = -c(SampleID, Run, Bay), 
               names_to = "Biomarker", 
               values_to = "Value") %>%
  group_by(Biomarker) %>%
  summarize(
    reads_mean = mean(Value, na.rm = TRUE),
    reads_min = min(Value, na.rm = TRUE),
    reads_max = max(Value, na.rm = TRUE),
    .groups = "drop"
  )
# Summary
sc_hi <- res$high_cv$summary %>%
  left_join(raw_stats, by = "Biomarker")

```

### C.1.4 HIHG Reps
2 per plate

Raw Reads

```{r hihg_cv, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}

res <- cv_plots(hihg_reps_reads, top_n = 10)

# Plots
res$plots$plate
knitr::kable(res$tables$plate$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::add_header_above(c(" " = 1, "Run × Bay Plates" = ncol(res$tables$plate$data) - 1)) %>%
  kableExtra::footnote(res$tables$plate$caption)

res$plots$run
knitr::kable(res$tables$run$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$run$caption)

res$plots$bay
knitr::kable(res$tables$bay$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$bay$caption)

res$plots$overall
knitr::kable(res$tables$overall$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$overall$caption)

```

#### NPQ

```{r hihg_cv_npq, message=FALSE, warning=FALSE, echo=TRUE}
res <- cv_plots(hihg_reps, top_n = 10)

# Plots
res$plots$plate
knitr::kable(res$tables$plate$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::add_header_above(c(" " = 1, "Run × Bay Plates" = ncol(res$tables$plate$data) - 1)) %>%
  kableExtra::footnote(res$tables$plate$caption)

res$plots$run
knitr::kable(res$tables$run$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$run$caption)

res$plots$bay
knitr::kable(res$tables$bay$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$bay$caption)

res$plots$overall
knitr::kable(res$tables$overall$data, digits = 2) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::footnote(res$tables$overall$caption)

# Calculate raw read stats
raw_stats <- hihg_reps_reads %>%
  pivot_longer(cols = -c(SampleID, Run, Bay), 
               names_to = "Biomarker", 
               values_to = "Value") %>%
  group_by(Biomarker) %>%
  summarize(
    reads_mean = mean(Value, na.rm = TRUE),
    reads_min = min(Value, na.rm = TRUE),
    reads_max = max(Value, na.rm = TRUE),
    .groups = "drop"
  )
# Summary
hihg_hi <- res$high_cv$summary %>%
  left_join(raw_stats, by = "Biomarker")

```

## C.2 Internal Control (IC = mCherry)

### C.2.1 CV for mCherry 
Compute the cv for mCherry for all samples (reps and non-reps) within each plate.
*(The new CNS Disease Panel V2 export drops the mCherry IC row from Raw-counts, so the IC reads are taken from Sample QC `IC Reads` — verified identical to the legacy mCherry row. Skipped only if neither source is present.)*
We did already look for IC outliers in the Sample QC, but that did not include the replicates.

This shows a problem with 0804-Bay1 and also 0711-Bay1. The plots show extreme outliers. Our Sample QC already picked these up.

I can try to drop outliers (40% of median) and rerun cv. Are the low reads in 0804-Bay1 really the problem? Also see if any of the reps are outliers (Sample QC did not look at that).

```{r ic_cv, message=FALSE, warning=FALSE, echo=TRUE, eval = ("mCherry" %in% names(raw_df))}

cv_mCherry_plate <- raw_df %>%
  pivot_longer(
    cols = -c(SampleID, Run, Bay),
    names_to = "Biomarker",
    values_to = "Value"
  ) %>%
  filter(Biomarker == "mCherry") %>%
  group_by(Run, Bay) %>%
  summarize(
    n           = n(),
    median      = median(Value, na.rm = TRUE),
    mean        = mean(Value, na.rm = TRUE),
    sd          = sd(Value, na.rm = TRUE),
    CV_mCherry  = 100 * sd / mean,
    .groups = "drop"
  ) %>%
  mutate(Plate = paste0(Run, "_", Bay)) %>%
  relocate(Plate)
  
knitr::kable(cv_mCherry_plate, digits = 2, caption = "CV for mCherry reads for all samples (including reps)") %>%
  kableExtra::kable_styling(full_width = FALSE)

#Plot reads for each plate including all samples
mCherry_raw <- raw_df %>%
  pivot_longer(
    cols = -c(SampleID, Run, Bay),
    names_to = "Biomarker",
    values_to = "Value"
  ) %>%
  filter(Biomarker == "mCherry") %>%
  mutate(Plate = paste0(Run, "_", Bay))

ggplot(mCherry_raw, aes(x = Bay, y = Value, fill = Bay)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(width = 0.15, size = 2.5, alpha = 0.6) +
  facet_wrap(~ Run, scales = "fixed") +
  theme_bw() +
  labs(
    title = "Raw mCherry Reads per Plate (Run × Bay)",
    x = "Bay",
    y = "mCherry Reads"
  ) +
  theme(
    axis.text.x = element_text(face = "bold"),
    strip.background = element_rect(fill = "lightgray"),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )

```

These are the plate-specific outliers. There are no reps in here. See what happens to cvs if we remove these samples.

```{r ic_outliers, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)

mCherry_flagged <- raw_df %>%
  pivot_longer(
    cols = -c(SampleID, Run, Bay),
    names_to = "Biomarker",
    values_to = "Value"
  ) %>%
  filter(Biomarker == "mCherry") %>%
  left_join(
    cv_mCherry_plate %>%
      select(Run, Bay, median),
    by = c("Run", "Bay")
  ) %>%
  mutate(
    lower_cut = 0.6 * median,
    upper_cut = 1.4 * median,
    Flag_Outlier = Value < lower_cut |
                   Value > upper_cut |
                   Value < 1000,
    Plate = paste0(Run, "_", Bay)
  ) %>%
  select(
    SampleID, Run, Bay, Plate,
    Value,
    median, lower_cut, upper_cut,
    Flag_Outlier
  )


outlier_table <- mCherry_flagged %>%
  filter(Flag_Outlier == TRUE) %>%
  arrange(Run, Bay, Value)

outlier_table %>%
  kable(format = "html", caption = "Flagged mCherry Outliers") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

### C.2.2 IC Raw Reads

#### C.2.2.1 IC Reads in Reps
IC (mCherry) raw read values for each of the replicate sets.

```{r ic_reps, message=FALSE, warning=FALSE, echo=TRUE, eval = ("mCherry" %in% names(raw_df))}
#For all sets of reads from reps(ipc_samples_reads,nc_samples_reads,sc_samples_reads,hihg_reps_reads) make plots showing all values for the biomarker mCherry for each Run xBay, colored by each type of rep.

library(dplyr)
library(tidyr)
library(ggplot2)

combined_long <- bind_rows(
  ipc_samples_reads %>% mutate(RepType = "IPC"),
  nc_samples_reads  %>% mutate(RepType = "NC"),
  sc_samples_reads  %>% mutate(RepType = "SC"),
  hihg_reps_reads   %>% mutate(RepType = "HIHG")
) %>%
  pivot_longer(
    cols = -c(SampleID, Run, Bay, RepType),
    names_to = "Biomarker",
    values_to = "Value"
  ) %>%
  filter(Biomarker == "mCherry")

# Plot: x = Bay, colored by RepType, one panel per Run
ggplot(combined_long, aes(x = Bay, y = Value, color = RepType)) +
  geom_jitter(width = 0.2, height = 0, size = 2, alpha = 0.8) +
  facet_wrap(~ Run, scales = "fixed") +   # <- SAME Y-AXIS IN ALL PANELS
  theme_bw() +
  labs(
    title = "mCherry values by Bay and Run",
    x = "Bay",
    y = "mCherry value",
    color = "Replicate type"
  ) +
  theme(
    axis.text.x = element_text(face = "bold"),
    strip.background = element_rect(fill = "lightgray"),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )

 med_table <- combined_long %>%
    group_by(Run, Bay) %>%
    summarize(Median = median(Value, na.rm = TRUE), .groups = "drop") %>%
    arrange(Run, Bay) %>%
    tidyr::pivot_wider(names_from = Bay, values_from = Median)
 knitr::kable(med_table, digits = 2, caption = "Median Values of mCherry for all reps") %>%
  kableExtra::kable_styling(full_width = FALSE)
 
```

#### C.2.2.2 Non-IC Reads in Reps

mCherry in reps does not track with non-mCherry Biomarkers for 0804 Bay1. For example, For non-IC biomarkers in the reps, Bay1 has similar median numbers of reads across the first three runs. But for mCherry, in Run 0804-Bay1 has a lot fewer than Bay1's in the other runs.

This is not the case with Bay2 which has a consistent pattern in mCherry and other biomarkers across the first 3 runs.

```{r non-ic_reps, message=FALSE, warning=FALSE, echo=TRUE, eval = ("mCherry" %in% names(raw_df))}
library(dplyr)
library(ggplot2)

combined_long <- bind_rows(
  ipc_samples_reads %>% mutate(RepType = "IPC"),
  nc_samples_reads  %>% mutate(RepType = "NC"),
  sc_samples_reads  %>% mutate(RepType = "SC"),
  hihg_reps_reads   %>% mutate(RepType = "HIHG")
) %>%
  pivot_longer(
    cols = -c(SampleID, Run, Bay, RepType),
    names_to = "Biomarker",
    values_to = "Value"
  )
  # <-- NO FILTER(Biomarker == "mCherry") so we keep ALL biomarkers

# Plot all biomarkers together
# Function: return plot + median table for a RepType
plot_and_medians <- function(data, rep_name) {
  
  df_rep <- data %>% filter(RepType == rep_name)

  # ---- PLOT ----
  p <- ggplot(df_rep, aes(x = Bay, y = Value, fill = Bay, color = Bay)) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA) +
    geom_jitter(width = 0.15, height = 0, size = 2.5, alpha = 0.6) +
    facet_wrap(~ Run, scales = "fixed") +
    theme_bw() +
    labs(
      title = paste("All Biomarkers —", rep_name, "Samples"),
      subtitle = "Boxplots + Points by Bay and Run",
      x = "Bay",
      y = "Value"
    ) +
    theme(
      axis.text.x = element_text(face = "bold"),
      strip.background = element_rect(fill = "lightgray"),
      strip.text = element_text(face = "bold")
    ) +
    guides(fill = "none", color = "none")

  # ---- MEDIAN TABLE ----
  med_table <- df_rep %>%
    group_by(Run, Bay) %>%
    summarize(Median = median(Value, na.rm = TRUE), .groups = "drop") %>%
    arrange(Run, Bay) %>%
    tidyr::pivot_wider(names_from = Bay, values_from = Median)

  return(list(plot = p, medians = med_table))
}

# ---------------------------------------------------------
# Generate all objects
res_IPC  <- plot_and_medians(combined_long, "IPC")
res_NC   <- plot_and_medians(combined_long, "NC")
res_SC   <- plot_and_medians(combined_long, "SC")
res_HIHG <- plot_and_medians(combined_long, "HIHG")


res_IPC$plot
knitr::kable(res_IPC$medians, digits = 2, caption = "Median Value per Bay × Run (IPC)") %>%
  kableExtra::kable_styling(full_width = FALSE)
res_NC$plot
knitr::kable(res_NC$medians, digits = 2, caption = "Median Value per Bay × Run (NC)") %>%
  kableExtra::kable_styling(full_width = FALSE)

res_SC$plot
knitr::kable(res_SC$medians, digits = 2, caption = "Median Value per Bay × Run (SC)") %>%
  kableExtra::kable_styling(full_width = FALSE)

res_HIHG$plot
knitr::kable(res_HIHG$medians, digits = 2, caption = "Median Value per Bay × Run (HIHG)") %>%
  kableExtra::kable_styling(full_width = FALSE)

```


## C.3 High CV Biomarkers
Let's walk through the Alamar normalization process and inspect biomarkers with high CVs.
(Cross-plate normalization itself is validated separately in the CrossPlate_Normalization project; here we
just inspect the high-CV replicate biomarkers.)

The starting level of mCherry (the IC) should be the same in each well. Divide each biomarker raw read count for the sample in the well by the read count of mCherry. Below shows this for each of the replicate pairs.

```{r plot_functions, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}

# Function to make violin plots for all samples
plot_biomarker_violins <- function(data, 
                                   biomarkers,
                                   rep_types = NULL,
                                   value_col = "Value",
                                   facet_by_biomarker = TRUE) {
  
  # Filter to specified biomarkers
  plot_data <- data %>%
    filter(Biomarker %in% biomarkers)
  
  if (nrow(plot_data) == 0) {
    stop("No data found for specified biomarkers")
  }
  
  # Filter by rep_types if specified
  if (!is.null(rep_types)) {
    if (!"RepType" %in% names(plot_data)) {
      warning("RepType column not found, ignoring rep_types filter")
    } else {
      plot_data <- plot_data %>%
        filter(RepType %in% rep_types)
    }
  }
  
  # Rename value column if needed
  if (value_col != "Value" && value_col %in% names(plot_data)) {
    names(plot_data)[names(plot_data) == value_col] <- "Value"
  }
  
  # Check for required columns
  required_cols <- c("Bay", "Run", "Biomarker", "Value")
  missing_cols <- setdiff(required_cols, names(plot_data))
  if (length(missing_cols) > 0) {
    stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
  }
  
  # Abbreviate Run names if long
  plot_data <- plot_data %>%
    mutate(Run_short = case_when(
      nchar(as.character(Run)) > 15 ~ paste0(substr(Run, 1, 12), "..."),
      TRUE ~ as.character(Run)
    ))
  
  # Determine if we have RepType for coloring
  has_reptype <- "RepType" %in% names(plot_data)
  
  # Create plots
  if (facet_by_biomarker && length(biomarkers) > 1) {
    # Multiple biomarkers - create separate plots
    plots <- lapply(biomarkers, function(biom) {
      data_biom <- plot_data %>% filter(Biomarker == biom)
      
      p <- ggplot(data_biom, aes(x = Bay, y = Value))
      
      if (has_reptype && is.null(rep_types)) {
  p <- p + 
    geom_violin(aes(fill = RepType), alpha = 0.7, position = position_dodge(0.9)) +
    geom_jitter(aes(color = RepType), alpha = 0.5, size = 2,
               position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9)) +
    labs(fill = "Replicate Type", color = "Replicate Type")
} else {
  p <- p + 
    geom_violin(fill = "steelblue", alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.5, size = 2, color = "darkblue")
}
      
      p <- p +
        facet_wrap(~ Run_short) +
        theme_bw() +
        labs(
          title = paste("Biomarker:", biom),
          x = "Bay",
          y = "Value"
        ) +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, face = "bold"),
          strip.background = element_rect(fill = "lightgray"),
          strip.text = element_text(face = "bold", size = 9),
          legend.position = "top"
        )
      
      p
    })
    names(plots) <- biomarkers
    
    return(plots)
    
  } else {
    # Single biomarker or combined plot
    p <- ggplot(plot_data, aes(x = Bay, y = Value))
    
    if (has_reptype && is.null(rep_types)) {
      p <- p + 
        geom_violin(aes(fill = RepType), alpha = 0.7, position = position_dodge(0.9)) +
        geom_jitter(aes(color = RepType), alpha = 0.5, size = 2,
                   position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9)) +
        labs(fill = "Replicate Type", color = "Replicate Type")
    } else {
      p <- p + 
        geom_violin(fill = "steelblue", alpha = 0.7) +
        geom_jitter(width = 0.2, alpha = 0.5, size = 2, color = "darkblue")
    }
    
    if (length(biomarkers) > 1) {
      p <- p + facet_grid(Biomarker ~ Run_short)
    } else {
      p <- p + facet_wrap(~ Run_short)
    }
    
    p <- p +
      theme_bw() +
      labs(
        title = if(length(biomarkers) == 1) paste("Biomarker:", biomarkers[1]) else "Biomarker Distribution by Bay and Run",
        x = "Bay",
        y = "Value"
      ) +
      theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, face = "bold"),
        strip.background = element_rect(fill = "lightgray"),
        strip.text = element_text(face = "bold", size = 9),
        legend.position = "top"
      )
    
    return(p)
  }
}

# Function to plot replicates only
plot_and_tables <- function(data, 
                           value_col = NULL,
                           top = 1,
                           return_medians = FALSE,
                           return_top = FALSE,
                           return_all_reps = FALSE) {
  
  # Ensure all arguments are proper types
  if (!is.null(value_col) && (!is.character(value_col) || length(value_col) != 1)) {
    stop("value_col must be a single character string or NULL")
  }
  
  # Determine value column
  if (is.null(value_col)) {
    # Auto-detect value column
    value_candidates <- names(data)[grepl("Value", names(data), ignore.case = TRUE)]
    if (length(value_candidates) > 0) {
      value_col <- value_candidates[1]
    } else {
      stop("No value column found. Please specify 'value_col' parameter.")
    }
  }
  
  # Rename value column if needed
  if (is.character(value_col) && nchar(value_col) > 0 && value_col %in% names(data)) {
    if (value_col != "Value") {
      # Remove existing "Value" column if it exists
      if ("Value" %in% names(data)) {
        data <- data %>% select(-Value)
      }
      names(data)[names(data) == value_col] <- "Value"
    }
  } else if (!"Value" %in% names(data)) {
    stop("Could not find or rename value column to 'Value'")
  }
  
  # Check for required columns
  required_cols <- c("Bay", "RepType", "Value")
  missing_cols <- setdiff(required_cols, names(data))
  if (length(missing_cols) > 0) {
    stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
  }
  
  # Determine faceting based on available columns
  has_biomarker <- "Biomarker" %in% names(data) && length(unique(data$Biomarker)) > 0
  has_run <- "Run" %in% names(data) && length(unique(data$Run)) > 0
  
  # Get unique biomarkers if column exists
  biomarkers <- if(has_biomarker) unique(data$Biomarker) else NULL
  
  # Abbreviate Run names if they're long
  if (has_run) {
    data <- data %>%
      mutate(Run_short = case_when(
        nchar(as.character(Run)) > 15 ~ paste0(substr(Run, 1, 12), "..."),
        TRUE ~ as.character(Run)
      ))
  }
  
  # Initialize results list
  results <- list()
  
  # Create plots for each biomarker
  if (has_biomarker && length(biomarkers) > 1) {
    # Multiple biomarkers - create individual plots
    results$plots <- lapply(biomarkers, function(biom) {
      data_biom <- data %>% filter(Biomarker == biom)
      
      p <- ggplot(data_biom, aes(x = Bay, y = Value, color = RepType)) +
        geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.8) +
        theme_bw() +
        labs(
          title = paste("Biomarker:", biom),
          x = "Bay",
          y = "Value",
          color = "Replicate type"
        ) +
        theme(
          axis.text.x = element_text(face = "bold", angle = 90, vjust = 0.5, hjust = 1),
          strip.background = element_rect(fill = "lightgray"),
          strip.text = element_text(face = "bold", size = 8),
          strip.text.x = element_text(angle = 0),
          legend.position = "top"
        )
      
      if (has_run) {
        p <- p + facet_wrap(~ Run_short)
      }
      
      p
    })
    names(results$plots) <- biomarkers
    
    # Also create a combined plot
    p_combined <- ggplot(data, aes(x = Bay, y = Value, color = RepType)) +
      geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.8) +
      theme_bw() +
      labs(
        title = "Biomarker Values by Bay and Run",
        x = "Bay",
        y = "Value",
        color = "Replicate type"
      ) +
      theme(
        axis.text.x = element_text(face = "bold", angle = 90, vjust = 0.5, hjust = 1),
        strip.background = element_rect(fill = "lightgray"),
        strip.text = element_text(face = "bold", size = 8),
        strip.text.x = element_text(angle = 0),
        legend.position = "top"
      )
    
    if (has_run) {
      p_combined <- p_combined + facet_grid(Biomarker ~ Run_short)
    } else {
      p_combined <- p_combined + facet_wrap(~ Biomarker)
    }
    
    results$plot_combined <- p_combined
    
  } else {
    # Single biomarker or no biomarker column - create single plot
    p <- ggplot(data, aes(x = Bay, y = Value, color = RepType)) +
      geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.8) +
      theme_bw() +
      labs(
        title = if(has_biomarker) paste("Biomarker:", biomarkers[1]) else "Biomarker Values by Bay and Run",
        x = "Bay",
        y = "Value",
        color = "Replicate type"
      ) +
      theme(
        axis.text.x = element_text(face = "bold", angle = 90, vjust = 0.5, hjust = 1),
        strip.background = element_rect(fill = "lightgray"),
        strip.text = element_text(face = "bold", size = 8),
        strip.text.x = element_text(angle = 0),
        legend.position = "top"
      )
    
    if (has_run) {
      p <- p + facet_wrap(~ Run_short)
    }
    
    results$plot <- p
  }
  
  # ---- MEDIAN TABLE ----
  if (isTRUE(return_medians)) {
    group_cols <- c(if(has_biomarker) "Biomarker", if(has_run) "Run", "Bay")
    
    med_table <- data %>%
      group_by(across(all_of(group_cols))) %>%
      summarize(Median = median(Value, na.rm = TRUE), .groups = "drop") %>%
      arrange(across(all_of(group_cols))) %>%
      tidyr::pivot_wider(names_from = Bay, values_from = Median)
    
    results$medians <- med_table
  }
  
  # ---- TOP VALUES TABLE ----
  if (isTRUE(return_top)) {
    group_cols <- c(if(has_biomarker) "Biomarker", if(has_run) "Run", "Bay")
    select_cols <- c(group_cols, "SampleID", "Value")
    
    top_table <- data %>%
      group_by(across(all_of(group_cols))) %>%
      arrange(desc(Value)) %>%
      slice_head(n = top) %>%
      ungroup() %>%
      select(all_of(select_cols)) %>%
      arrange(across(all_of(group_cols)), desc(Value))
    
    results$top_values <- top_table
  }
  
  # ---- ALL REPS TABLE ----
  if (isTRUE(return_all_reps)) {
    select_cols <- c(if(has_biomarker) "Biomarker", 
                     if(has_run) "Run", 
                     "Bay", "SampleID", "RepType", "Value")
    
    all_reps_table <- data %>%
      select(all_of(intersect(select_cols, names(data)))) %>%
      arrange(across(all_of(intersect(select_cols[1:(length(select_cols)-2)], 
                                      names(data)))), SampleID)
    
    results$all_reps <- all_reps_table
  }
  
  return(results)
}
```

### Summary

```{r normalize, message=FALSE, warning=FALSE, echo=TRUE, eval = ("mCherry" %in% names(raw_df)), results='asis', fig.width=12, fig.height=10}
# ==============================================================================
# C.3 HIGH CV BIOMARKERS - DATA PREPARATION
# ==============================================================================

# --- COMBINE ALL REPS IN LONG FORMAT (READS) ---
combined_reps_long <- bind_rows(
  nc_samples_reads %>%
    pivot_longer(cols = -c(SampleID, Run, Bay),
                 names_to = "Biomarker", values_to = "Value") %>%
    mutate(RepType = "NC"),
  ipc_samples_reads %>%
    pivot_longer(cols = -c(SampleID, Run, Bay),
                 names_to = "Biomarker", values_to = "Value") %>%
    mutate(RepType = "IPC"),
  sc_samples_reads %>%
    pivot_longer(cols = -c(SampleID, Run, Bay),
                 names_to = "Biomarker", values_to = "Value") %>%
    mutate(RepType = "SC"),
  hihg_reps_reads %>%
    pivot_longer(cols = -c(SampleID, Run, Bay),
                 names_to = "Biomarker", values_to = "Value") %>%
    mutate(RepType = "HIHG")
)

# --- COMBINE ALL REPS IN LONG FORMAT (NPQ) ---
combined_reps_npq_long <- bind_rows(
  nc_samples_npq %>%
    pivot_longer(cols = -c(SampleID, Run, Bay),
                 names_to = "Biomarker", values_to = "NPQ") %>%
    mutate(RepType = "NC"),
  ipc_samples_npq %>%
    pivot_longer(cols = -c(SampleID, Run, Bay),
                 names_to = "Biomarker", values_to = "NPQ") %>%
    mutate(RepType = "IPC"),
  sc_samples_npq %>%
    pivot_longer(cols = -c(SampleID, Run, Bay),
                 names_to = "Biomarker", values_to = "NPQ") %>%
    mutate(RepType = "SC"),
  hihg_reps_npq %>%
    pivot_longer(cols = -c(SampleID, Run, Bay),
                 names_to = "Biomarker", values_to = "NPQ") %>%
    mutate(RepType = "HIHG")
)

# --- SAMPLES ONLY (non-reps) IN LONG FORMAT (READS) ---
combined_samples_long <- raw_df %>%
  filter(!SampleID %in% rep_sample_ids) %>%
  pivot_longer(
    cols = -c(SampleID, Run, Bay),
    names_to = "Biomarker",
    values_to = "Value"
  ) %>%
  mutate(RepType = "Sample")

# --- SAMPLES ONLY (non-reps) IN LONG FORMAT (NPQ) ---
combined_samples_npq_long <- combined_df %>%
  filter(!SampleID %in% rep_sample_ids) %>%
  pivot_longer(
    cols = -c(SampleID, Run, Bay),
    names_to = "Biomarker",
    values_to = "NPQ"
  ) %>%
  mutate(RepType = "Sample")

# --- COMBINED REPS + SAMPLES (READS) ---
combined_all_long <- bind_rows(combined_reps_long, combined_samples_long)

# --- COMBINED REPS + SAMPLES (NPQ) ---
combined_all_npq_long <- bind_rows(combined_reps_npq_long, combined_samples_npq_long)

# --- IC NORMALIZATION ---
# Extract mCherry values
mCherry_vals <- combined_all_long %>%
  filter(Biomarker == "mCherry") %>%
  select(SampleID, mCherry_Value = Value)

# IC normalize
combined_norm <- combined_all_long %>%
  left_join(mCherry_vals, by = "SampleID") %>%
  mutate(Value_norm = Value / mCherry_Value)

# Remove mCherry from normalized data
combined_norm_no_mCherry <- combined_norm %>%
  filter(Biomarker != "mCherry")

# For ALZ123, add transformed normalized value for CV calculations
combined_norm <- combined_norm %>%
  mutate(Value_norm2 = Value_norm * 10000 + 1)

# ==============================================================================
# BIOMARKER SELECTION AND PLOTTING
# ==============================================================================

# --- SELECT BIOMARKERS TO EXAMINE ---
cat("**Biomarkers with HIGH CVs for 2^NPQ**\n\n")

knitr::kable(ipc_hi, digits = 3, caption = "High CV for IPC") %>% 
  kableExtra::kable_styling(full_width = TRUE)
knitr::kable(sc_hi, digits = 3, caption = "High CV for SC") %>% 
  kableExtra::kable_styling(full_width = TRUE)
knitr::kable(hihg_hi, digits = 3, caption = "High CV for HIHG") %>% 
  kableExtra::kable_styling(full_width = TRUE)

cat("\n\n**Biomarkers with HIGH CVs for IPC or SC:**\n\n")
biom_hi <- unique(c(ipc_hi$Biomarker, sc_hi$Biomarker))
cat(paste(biom_hi, collapse = ", "), "\n\n")

cat("\n\n**Biomarkers with HIGH CVs only for HIHG:**\n\n")
hihg_only_all <- setdiff(hihg_hi$Biomarker, biom_hi)
cat(paste(hihg_only_all, collapse = ", "), "\n\n")

cat("\n\n**Biomarkers with HIGH CVs only for HIHG with min_reads>=500:**\n\n")
hihg_only_low <- hihg_hi %>%
  filter(Biomarker %in% hihg_only_all, reads_min < 500) %>%
  pull(Biomarker)
hihg_only <- setdiff(hihg_only_all, hihg_only_low)
cat(paste(hihg_only, collapse = ", "), "\n\n")

# --- PREPARE DATA FOR PLOTTING ---
my_biomarkers <- c(biom_hi, hihg_only)
#my_biomarkers=c("IL17A", "NEFH")

# Filter to biomarkers of interest
combined_reps_short <- combined_reps_long %>%
  filter(Biomarker %in% my_biomarkers)

combined_norm_short <- combined_norm %>%
  filter(Biomarker %in% my_biomarkers)

# --- GENERATE PLOTS ---
cat("\n\n**mCherry Reads**\n\n")
mCherry_data <- combined_reps_long %>% filter(Biomarker == "mCherry")
res_mCherry <- plot_and_tables(mCherry_data)
print(res_mCherry$plot)

# Generate results for each normalization
res_reads <- plot_and_tables(combined_reps_short)
res_norm <- plot_and_tables(combined_norm_short, value_col = "Value_norm2")

# --- IPC NORMALIZATION ---
ipc_medians_long <- combined_norm_short %>%
  filter(RepType == "IPC") %>%
  group_by(Run, Bay, Biomarker) %>%
  summarize(IPC_plate_median = median(Value_norm, na.rm = TRUE), .groups = "drop")

combined_ipc_norm <- combined_norm_short %>%
  filter(RepType != "IPC") %>%
  left_join(ipc_medians_long, by = c("Run", "Bay", "Biomarker")) %>%
  mutate(
    Value_IPC_norm = log2(1 + (Value_norm / IPC_plate_median) * 10000),
    Value_IPC_norm2 = 2^Value_IPC_norm
  )

combined_ipc_norm <- combined_ipc_norm %>%  #These are IPC-normalized values for NC, SC and HIHG
  filter(RepType != "Sample")

res_npq <- plot_and_tables(combined_ipc_norm, value_col = "Value_IPC_norm2")

# --- PLOT EACH BIOMARKER ---
for (biom in my_biomarkers) {
  cat("\n### ", biom, "\n\n")
  
  cat("**Raw Reads**\n\n")
  if (!is.null(res_reads$plots[[biom]])) {
    print(res_reads$plots[[biom]])
  }
  cat("\n\n")
  
  cat("**IPC-normalized**\n\n")
  if (!is.null(res_npq$plots[[biom]])) {
    print(res_npq$plots[[biom]])
  }
  cat("\n\n")
  
  cat("**Non-rep NPQ**\n\n")
  print(plot_biomarker_violins(combined_all_npq_long, biomarkers = biom, value_col= "NPQ"))
  cat("\n\n")
}


```

# Dataset Notes

```{r notes, echo=FALSE, results='asis'}
# Read notes from external file (not tracked in git)
# Create notes.md in this directory to add dataset-specific notes
notes_file <- "notes.md"
if (file.exists(notes_file)) {
  cat(paste(readLines(notes_file, warn = FALSE), collapse = "\n"))
} else {
  cat("*No notes file found. Create `notes.md` in this directory to add dataset-specific notes.*\n")
}
```
