---
title: "Metadata Merge Pipeline"
subtitle: "Merge Post-QC Biomarker Data with Metadata"
date: today
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
params:
  filter_only: false
  no_copy: true
  include_derived_covariates: true
  include_derived_biomarkers: false
  filter_na_covars: true
  output_dir: "output_files_n2851_2026May"
  review_dir: "review"   # optional dir holding {CDX,Ancestry}_review.csv (data-driven Step 6b)
editor: visual
---

<!--
CHANGELOG — Metadata_Merge_Pipeline.qmd
  2026-06-12 (propagated from the 8-plate run copy):
   * Step 6b is data-driven from `<review_dir>/{CDX,Ancestry}_review.csv` when present (edit the
     FINAL_* columns to regroup); falls back to the built-in config below when the CSVs are absent.
   * Step 6d reads APOE_WGS from the auto-detected metadata. NO WGS plate-swap correction is applied;
     swap-affected AD-Hispanic samples are masked to Sanger via
     ../../../May_2026_WGS_QC/WGS_chr1_corrected_idmap_2026Jun11.csv (replaces the archived
     2026-06-08 ..._plateswapcorrected_2026Jun.csv).
   * Step-6 chunks suppress covariate_explorer.R console side-effects and render clean kable tables.
  PROVENANCE: older rendered reports (Metadata_Merge_Pipeline.html dated 2026-06-08, output_files_n2851_*,
  output_files_n2850_*, etc.) were produced by EARLIER versions of this file on EARLIER metadata.
  They are preserved, not deleted — see README.md "Report / output provenance".
-->


```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)
library(kableExtra)

# Configuration
PRIMARY_QC_OUTPUT <- "../Primary_QC/output_files"
INPUT_DIR <- "input_files"
OUTPUT_DIR <- params$output_dir
NPQ_STANDARD_FILE <- "NPQ_post_QC.csv"
NPQ_LOW_FILE <- "NPQ_Low_post_QC.csv"
NPQ_DATE_PATTERN <- NULL  # Set to specific date or NULL for auto-detect

# Get parameters (set in YAML)
FILTER_ONLY              <- params$filter_only
NO_COPY                  <- params$no_copy
INCLUDE_DERIVED_COVARS   <- params$include_derived_covariates
INCLUDE_DERIVED_BIOMARKS <- params$include_derived_biomarkers
FILTER_NA_COVARS         <- params$filter_na_covars

# Covariate explorer (config-driven grouping and exploration utilities)
source("covariate_explorer.R")

# Create output directory if needed
if (!dir.exists(OUTPUT_DIR)) dir.create(OUTPUT_DIR)
if (!dir.exists(file.path(OUTPUT_DIR, "filtered"))) dir.create(file.path(OUTPUT_DIR, "filtered"))
```

# Overview

This pipeline merges post-QC biomarker data with sample metadata and applies exclusion filters to create analysis-ready datasets.

**Pipeline Steps:**

1.  **Input Files** - Verify/copy post-QC files from Primary_QC
2.  **Merge Data** - Merge biomarker data with metadata (APOE.geno normalized to `"a/b"` format)
3.  **Quality Control** - Identify duplicates, implausible values, and issues
4.  **Apply Filters** - Create final analysis-ready datasets
5.  **Exclusion Summary** - Review filtered sample counts
6.  **Covariate Groupings** *(optional: `include_derived_covariates`, default TRUE)* — explore distributions (6a), define configs (6b), preview decisions (6c), apply + validate + filter (6d); joins WGS APOE genotypes, produces CDX_collapsed, Ancestry, APOE_WGS, APOE_WGS_norm, APOE_WGS_changed, APOE.geno_final (WGS preferred), APOE4_carrier, APOE4_count; rows excluded by default when either covariate is NA (`filter_na_covars`)
7.  **Derived Biomarkers** *(optional: `include_derived_biomarkers`, default FALSE)* - APOE4_minus_APOE, ABeta42_minus_ABeta40
8.  **Longitudinal Handling** - Flag repeat-subject visits; write the cross-sectional **default** (`filtered_combined_post_QC.csv`, one-per-subject = most recent) and a separate **longitudinal** dataset (`filtered_combined_longitudinal.csv`)
9.  **Save RDS** - Write `preprocessed_data.rds` (cross-sectional default) with data + biomarker_cols

```{r}
#| label: mode-display
#| echo: false
#| results: asis

if (FILTER_ONLY) {
  cat("::: {.callout-note}\n")
  cat("## Filter-Only Mode\n")
  cat("Running in **filter-only mode** - using existing merged files and exclusion report.\n")
  cat(":::\n")
}
```

```{r}
#| label: helper-functions
#| include: false

find_metadata_file <- function(input_dir) {
  all_files <- list.files(input_dir, full.names = TRUE)
  metadata_files <- all_files[grepl("metadata", basename(all_files), ignore.case = TRUE)]

  if (length(metadata_files) == 0) {
    stop("No metadata file found in ", input_dir, "\n",
         "  Expected: a file with 'metadata' in the name")
  }

  if (length(metadata_files) > 1) {
    metadata_files <- metadata_files[which.max(file.mtime(metadata_files))]
  }

  return(metadata_files)
}

find_latest_npq_files <- function(output_dir) {
  standard_files <- list.files(output_dir, pattern = "NPQ_.*_post_QC\\.csv$", full.names = TRUE)
  standard_files <- standard_files[!grepl("Low|triage", standard_files)]

  low_files <- list.files(output_dir, pattern = "NPQ_.*_Low_post_QC\\.csv$", full.names = TRUE)
  low_files <- low_files[!grepl("triage", low_files)]

  if (length(standard_files) == 0 || length(low_files) == 0) {
    stop("Could not find NPQ post-QC files in ", output_dir)
  }

  list(
    standard = standard_files[which.max(file.mtime(standard_files))],
    low = low_files[which.max(file.mtime(low_files))]
  )
}
```

# Step 1: Input Files {#step1}

```{r}
#| label: step1-input-files
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  # Check if NPQ files already exist
  existing_npq <- list.files(INPUT_DIR, pattern = "NPQ", ignore.case = TRUE)

  if (!NO_COPY && length(existing_npq) == 0) {
    cat("## Copying Files from Primary_QC\n\n")

    npq_files <- find_latest_npq_files(PRIMARY_QC_OUTPUT)
    standard_src <- npq_files$standard
    low_src <- npq_files$low

    cat("No NPQ files found in `input_files/` - copying from Primary_QC:\n\n")
    cat("- **Standard:** `", basename(standard_src), "`\n", sep = "")
    cat("- **Low:** `", basename(low_src), "`\n\n", sep = "")

    file.copy(standard_src, file.path(INPUT_DIR, NPQ_STANDARD_FILE), overwrite = TRUE)
    file.copy(low_src, file.path(INPUT_DIR, NPQ_LOW_FILE), overwrite = TRUE)

    cat("Files copied successfully.\n\n")
    existing_npq <- list.files(INPUT_DIR, pattern = "NPQ", ignore.case = TRUE)
  } else {
    cat("## Using Existing Input Files\n\n")
  }

  # Auto-detect metadata file
  metadata_file <- find_metadata_file(INPUT_DIR)

  cat("**Input files:**\n\n")
  cat("| File | Status |\n")
  cat("|------|--------|\n")
  for (f in existing_npq) {
    cat("| `", f, "` | ✓ Found |\n", sep = "")
  }
  cat("| `", basename(metadata_file), "` | ✓ Metadata (auto-detected) |\n", sep = "")

} else {
  cat("*Skipped in filter-only mode*\n\n")
  cat("Using existing files in `output_files/`\n")
  metadata_file <- find_metadata_file(INPUT_DIR)
}
```

```{r}
#| label: load-data
#| include: false

if (!FILTER_ONLY) {
  # Load metadata
  metadata <- read_csv(metadata_file, show_col_types = FALSE)
  metadata <- metadata %>%
    rename(APOE.geno = APOE) %>%
    mutate(
      # Normalize APOE.geno to "a/b" digit format (e.g. "E3    E4" -> "3/4")
      APOE.geno = case_when(
        is.na(APOE.geno) ~ NA_character_,
        TRUE ~ paste0(str_extract(APOE.geno, "\\d"),
                      "/",
                      str_extract(APOE.geno, "\\d$"))
      ),
      Group_Race_Ethnicity = paste(Group, Race, Ethnicity, sep = "_")
    )

  # Capture metadata column names before merging (used later to identify biomarker cols)
  metadata_cols_captured <- names(metadata)

  # Load biomarker files
  biomarkers_postQC <- read_csv(file.path(INPUT_DIR, NPQ_STANDARD_FILE), show_col_types = FALSE)
  biomarkers_low_postQC <- read_csv(file.path(INPUT_DIR, NPQ_LOW_FILE), show_col_types = FALSE)

  # Clean column names
  biomarkers_postQC <- biomarkers_postQC %>%
    rename_with(~str_replace_all(., "[^A-Za-z0-9_]", "_"))
  biomarkers_low_postQC <- biomarkers_low_postQC %>%
    rename_with(~str_replace_all(., "[^A-Za-z0-9_]", "_"))
} else {
  # In filter_only mode: capture metadata cols from the metadata file directly
  meta_tmp <- read_csv(metadata_file, show_col_types = FALSE)
  meta_tmp <- meta_tmp %>% rename(APOE.geno = APOE)
  metadata_cols_captured <- c(names(meta_tmp), "Group_Race_Ethnicity")
  rm(meta_tmp)
}
```

# Step 2: Merge Data {#step2}

```{r}
#| label: step2-merge
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("## Merging Biomarker Data with Metadata\n\n")

  # Merge standard
  merged_standard <- metadata %>%
    inner_join(biomarkers_postQC, by = c("SAMPLE_ALIQUOT" = "SampleID"), suffix = c("_meta", "_bio")) %>%
    select(-any_of(c("Run", "Bay")))

  # Merge low
  merged_low <- metadata %>%
    inner_join(biomarkers_low_postQC, by = c("SAMPLE_ALIQUOT" = "SampleID"), suffix = c("_meta", "_bio")) %>%
    select(-any_of(c("Run", "Bay")))

  # Combined biomarkers
  standard_biomarkers <- setdiff(names(biomarkers_postQC), c("SampleID", "Run", "Bay"))
  low_biomarkers <- setdiff(names(biomarkers_low_postQC), c("SampleID", "Run", "Bay"))
  unique_to_low <- setdiff(low_biomarkers, standard_biomarkers)

  combined_biomarkers <- biomarkers_postQC %>%
    full_join(
      biomarkers_low_postQC %>% select(SampleID, all_of(unique_to_low)),
      by = "SampleID"
    )

  merged_combined <- metadata %>%
    inner_join(combined_biomarkers, by = c("SAMPLE_ALIQUOT" = "SampleID"), suffix = c("_meta", "_bio"))

  cat("| Dataset | Samples | Columns |\n")
  cat("|---------|---------|--------|\n")
  cat("| Metadata | ", nrow(metadata), " | ", ncol(metadata), " |\n", sep = "")
  cat("| Standard Biomarkers | ", nrow(biomarkers_postQC), " | ", ncol(biomarkers_postQC), " |\n", sep = "")
  cat("| Low Biomarkers | ", nrow(biomarkers_low_postQC), " | ", ncol(biomarkers_low_postQC), " |\n", sep = "")
  cat("| **Merged Standard** | ", nrow(merged_standard), " | ", ncol(merged_standard), " |\n", sep = "")
  cat("| **Merged Low** | ", nrow(merged_low), " | ", ncol(merged_low), " |\n", sep = "")
  cat("| **Merged Combined** | ", nrow(merged_combined), " | ", ncol(merged_combined), " |\n", sep = "")

} else {
  cat("*Skipped in filter-only mode*\n\n")

  # Load existing merged files
  merged_combined <- read_csv(file.path(OUTPUT_DIR, "merged_combined_post_QC.csv"), show_col_types = FALSE)
  merged_standard <- read_csv(file.path(OUTPUT_DIR, "merged_standard_post_QC.csv"), show_col_types = FALSE)
  merged_low <- read_csv(file.path(OUTPUT_DIR, "merged_low_post_QC.csv"), show_col_types = FALSE)

  cat("**Using existing merged files:**\n\n")
  cat("| Dataset | Samples | Columns |\n")
  cat("|---------|---------|--------|\n")
  cat("| Merged Standard | ", nrow(merged_standard), " | ", ncol(merged_standard), " |\n", sep = "")
  cat("| Merged Low | ", nrow(merged_low), " | ", ncol(merged_low), " |\n", sep = "")
  cat("| Merged Combined | ", nrow(merged_combined), " | ", ncol(merged_combined), " |\n", sep = "")
}
```

```{r}
#| label: samples-not-in-metadata
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("\n## Samples Not in Metadata\n\n")

  all_npq_samples <- combined_biomarkers %>% select(SampleID) %>% distinct()

  technical_rep_pattern <- "^(SC|NC|IPC|HIHG)|_(SC|NC|IPC|HIHG)_|_(SC|NC|IPC|HIHG)$|HIHG"

  samples_not_in_metadata <- all_npq_samples %>%
    filter(!SampleID %in% metadata$SAMPLE_ALIQUOT) %>%
    mutate(is_technical_rep = grepl(technical_rep_pattern, SampleID, ignore.case = TRUE))

  technical_reps <- samples_not_in_metadata %>% filter(is_technical_rep)
  unmatched_samples <- samples_not_in_metadata %>% filter(!is_technical_rep)

  if (nrow(unmatched_samples) > 0) {
    cat("::: {.callout-warning}\n")
    cat("### ", nrow(unmatched_samples), " Sample(s) Not in Metadata\n\n", sep = "")
    cat("These samples are in NPQ files but NOT in metadata:\n\n")
    for (s in unmatched_samples$SampleID) {
      cat("- `", s, "`\n", sep = "")
    }
    cat("\nThese samples will NOT be included in merged output files.\n")
    cat(":::\n\n")

    write_csv(unmatched_samples %>% select(SampleID),
              file.path(OUTPUT_DIR, "samples_not_in_metadata.csv"))
  } else {
    cat("All non-technical samples in NPQ files are present in metadata. ✓\n\n")
  }

  if (nrow(technical_reps) > 0) {
    cat("*Technical replicates excluded (expected):* ", nrow(technical_reps), " samples (SC, NC, IPC, HIHG)\n", sep = "")
  }
}
```

```{r}
#| label: write-merged-files
#| include: false

if (!FILTER_ONLY) {
  write_csv(merged_standard, file.path(OUTPUT_DIR, "merged_standard_post_QC.csv"))
  write_csv(merged_low, file.path(OUTPUT_DIR, "merged_low_post_QC.csv"))
  write_csv(merged_combined, file.path(OUTPUT_DIR, "merged_combined_post_QC.csv"))
}
```

# Step 3: Quality Control {#step3}

```{r}
#| label: step3-qc-duplicates
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("## Duplicate Samples\n\n")

  duplicate_report <- merged_combined %>%
    group_by(SAMPLE) %>%
    filter(n() > 1) %>%
    arrange(SAMPLE, RUN) %>%
    mutate(
      run_order = row_number(),
      max_run = max(RUN),
      exclude = RUN != max_run,
      exclude_reason = if_else(exclude, paste0("Duplicate - keeping most recent run (", max_run, ")"), NA_character_)
    ) %>%
    ungroup()

  if (nrow(duplicate_report) > 0) {
    n_dup_samples <- n_distinct(duplicate_report$SAMPLE)
    n_to_exclude <- sum(duplicate_report$exclude)

    cat("Found **", n_dup_samples, "** samples with multiple runs (", nrow(duplicate_report), " total records)\n\n", sep = "")
    cat("- **", n_to_exclude, "** older runs will be excluded\n", sep = "")
    cat("- **", nrow(duplicate_report) - n_to_exclude, "** most recent runs will be kept\n\n", sep = "")

    dup_summary <- duplicate_report %>%
      group_by(SAMPLE) %>%
      summarise(
        n_runs = n(),
        runs = paste(RUN, collapse = ", "),
        kept_run = max(RUN),
        .groups = "drop"
      )

    kable(dup_summary, caption = "Duplicate Samples") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      print()

  } else {
    cat("No duplicate samples found. ✓\n")
  }
}
```

```{r}
#| label: step3-qc-implausible
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("\n## Implausible Values\n\n")

  plausible_ranges <- list(
    age_at_subject = c(18, 110),
    BMI = c(10, 80),
    weight = c(30, 300),
    height = c(100, 250)
  )

  implausible_list <- list()

  for (var in names(plausible_ranges)) {
    if (var %in% names(merged_combined)) {
      implausible <- merged_combined %>%
        filter(!is.na(.data[[var]]) &
               (.data[[var]] < plausible_ranges[[var]][1] |
                .data[[var]] > plausible_ranges[[var]][2])) %>%
        mutate(
          variable = var,
          value = .data[[var]],
          range = paste0("[", plausible_ranges[[var]][1], ", ", plausible_ranges[[var]][2], "]"),
          exclude_reason = paste0("Implausible ", var, ": ", value, " (expected ", range, ")")
        ) %>%
        select(SAMPLE_ALIQUOT, SAMPLE, variable, value, range, exclude_reason)

      if (nrow(implausible) > 0) {
        implausible_list[[var]] <- implausible
      }
    }
  }

  if (length(implausible_list) > 0) {
    implausible_report <- bind_rows(implausible_list)
    cat("Found **", nrow(implausible_report), "** samples with implausible values:\n\n", sep = "")

    kable(implausible_report %>% select(-exclude_reason),
          caption = "Implausible Values") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      print()

  } else {
    implausible_report <- data.frame()
    cat("No implausible values found. ✓\n")
  }
}
```

```{r}
#| label: step3-qc-race-ethnicity
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("\n## Unexpected Race/Ethnicity Combinations\n\n")

  unexpected_race_eth <- merged_combined %>%
    filter(
      (grepl("BL", Group_Race_Ethnicity) & grepl("WH", Group_Race_Ethnicity)) |
      (grepl("HI", Group_Race_Ethnicity) & grepl("NH", Group_Race_Ethnicity)) |
      (grepl("AA", Group_Race_Ethnicity) & grepl("WH", Group_Race_Ethnicity)) |
      (grepl("AFDC", Group_Race_Ethnicity) & grepl("WH", Group_Race_Ethnicity)) |
      (grepl("AFDC", Group_Race_Ethnicity) & grepl("HI", Group_Race_Ethnicity)) |
      (grepl("AA", Group_Race_Ethnicity) & grepl("HI", Group_Race_Ethnicity))
    ) %>%
    mutate(
      exclude_reason = paste0("Review: Unexpected Race/Ethnicity combination: ", Group_Race_Ethnicity),
      exclude = FALSE  # Flagged for review but NOT auto-excluded
    )

  if (nrow(unexpected_race_eth) > 0) {
    cat("::: {.callout-note}\n")
    cat("### Flagged for Review (Not Auto-Excluded)\n\n")
    cat("Found **", nrow(unexpected_race_eth), "** samples with potentially unexpected Race/Ethnicity combinations.\n\n", sep = "")
    cat("These are flagged in the exclusion report but **not automatically excluded**.\n")
    cat("Review and manually set `exclude=TRUE` if needed.\n")
    cat(":::\n\n")

    kable(unexpected_race_eth %>%
            select(SAMPLE_ALIQUOT, SAMPLE, Group, Race, Ethnicity, Group_Race_Ethnicity) %>%
            distinct(),
          caption = "Unexpected Race/Ethnicity Combinations") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      print()

  } else {
    cat("No unexpected combinations found. ✓\n")
  }
}
```

```{r}
#| label: build-exclusion-report
#| include: false

if (!FILTER_ONLY) {
  # Build exclusion report
  exclusion_report <- merged_combined %>%
    select(SAMPLE_ALIQUOT, SAMPLE, RUN, Group, Race, Ethnicity, Group_Race_Ethnicity,
           Case_Control, sex, age_at_subject, CDX, APOE.geno, Site) %>%
    mutate(exclude = FALSE, exclude_reason = NA_character_)

  # Mark duplicates
  if (nrow(duplicate_report) > 0) {
    exclusion_report <- exclusion_report %>%
      left_join(
        duplicate_report %>% select(SAMPLE_ALIQUOT, exclude, exclude_reason) %>% filter(exclude),
        by = "SAMPLE_ALIQUOT", suffix = c("", "_dup")
      ) %>%
      mutate(
        exclude = coalesce(exclude_dup, exclude),
        exclude_reason = coalesce(exclude_reason_dup, exclude_reason)
      ) %>%
      select(-exclude_dup, -exclude_reason_dup)
  }

  # Mark implausible values
  if (nrow(implausible_report) > 0) {
    implausible_combined <- implausible_report %>%
      group_by(SAMPLE_ALIQUOT) %>%
      summarise(exclude_reason_implausible = paste(exclude_reason, collapse = "; "), .groups = "drop") %>%
      mutate(exclude_implausible = TRUE)

    exclusion_report <- exclusion_report %>%
      left_join(implausible_combined, by = "SAMPLE_ALIQUOT") %>%
      mutate(
        exclude_reason = case_when(
          !is.na(exclude_reason) & !is.na(exclude_reason_implausible) ~
            paste(exclude_reason, exclude_reason_implausible, sep = "; "),
          !is.na(exclude_reason_implausible) ~ exclude_reason_implausible,
          TRUE ~ exclude_reason
        ),
        exclude = exclude | coalesce(exclude_implausible, FALSE)
      ) %>%
      select(-exclude_reason_implausible, -exclude_implausible)
  }

  # Mark unexpected race/ethnicity (for review, not auto-exclude)
  if (nrow(unexpected_race_eth) > 0) {
    race_eth_combined <- unexpected_race_eth %>%
      group_by(SAMPLE_ALIQUOT) %>%
      summarise(exclude_reason_race_eth = first(exclude_reason), .groups = "drop")

    exclusion_report <- exclusion_report %>%
      left_join(race_eth_combined, by = "SAMPLE_ALIQUOT") %>%
      mutate(
        exclude_reason = case_when(
          !is.na(exclude_reason) & !is.na(exclude_reason_race_eth) ~
            paste(exclude_reason, exclude_reason_race_eth, sep = "; "),
          !is.na(exclude_reason_race_eth) ~ exclude_reason_race_eth,
          TRUE ~ exclude_reason
        )
      ) %>%
      select(-exclude_reason_race_eth)
  }

  write_csv(exclusion_report, file.path(OUTPUT_DIR, "sample_exclusion_report.csv"))
}
```

# Step 4: Exclusion Summary {#step4}

```{r}
#| label: step4-exclusion-summary
#| echo: false
#| results: asis

exclusion_report <- read_csv(file.path(OUTPUT_DIR, "sample_exclusion_report.csv"), show_col_types = FALSE)

n_total <- nrow(exclusion_report)
n_exclude <- sum(exclusion_report$exclude)
n_keep <- sum(!exclusion_report$exclude)

cat("## Summary\n\n")

cat("| Metric | Count | Percent |\n")
cat("|--------|-------|--------|\n")
cat("| Total samples | ", n_total, " | 100% |\n", sep = "")
cat("| Samples to **EXCLUDE** | ", n_exclude, " | ", round(100*n_exclude/n_total, 1), "% |\n", sep = "")
cat("| Samples to **KEEP** | ", n_keep, " | ", round(100*n_keep/n_total, 1), "% |\n", sep = "")
cat("\n")

if (n_exclude > 0) {
  cat("## Exclusion Reasons\n\n")

  reasons <- exclusion_report %>%
    filter(exclude) %>%
    count(exclude_reason, sort = TRUE)

  kable(reasons, col.names = c("Reason", "Count"),
        caption = "Samples Excluded") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
    print()
}

# Show flagged but not excluded
review_items <- exclusion_report %>%
  filter(!exclude & !is.na(exclude_reason) & exclude_reason != "")

if (nrow(review_items) > 0) {
  cat("\n## Flagged for Review (Not Excluded)\n\n")

  review_reasons <- review_items %>% count(exclude_reason, sort = TRUE)

  kable(review_reasons, col.names = c("Reason", "Count"),
        caption = "Samples Flagged for Review") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
    print()
}
```

# Step 5: Apply Filters {#step5}

```{r}
#| label: step5-apply-filters
#| echo: false
#| results: asis

cat("## Applying Exclusion Filters\n\n")

# Get samples to exclude
samples_to_exclude <- exclusion_report %>%
  filter(exclude) %>%
  pull(SAMPLE_ALIQUOT)

cat("Excluding **", length(samples_to_exclude), "** samples from merged datasets...\n\n", sep = "")

# Filter datasets
filtered_standard <- merged_standard %>%
  filter(!SAMPLE_ALIQUOT %in% samples_to_exclude)

filtered_low <- merged_low %>%
  filter(!SAMPLE_ALIQUOT %in% samples_to_exclude)

filtered_combined <- merged_combined %>%
  filter(!SAMPLE_ALIQUOT %in% samples_to_exclude)

# Write filtered files
write_csv(filtered_standard, file.path(OUTPUT_DIR, "filtered", "filtered_standard_post_QC.csv"))
write_csv(filtered_low, file.path(OUTPUT_DIR, "filtered", "filtered_low_post_QC.csv"))
write_csv(filtered_combined, file.path(OUTPUT_DIR, "filtered", "filtered_combined_post_QC.csv"))

cat("| Dataset | Before | After | Excluded |\n")
cat("|---------|--------|-------|----------|\n")
cat("| Standard | ", nrow(merged_standard), " | ", nrow(filtered_standard), " | ", nrow(merged_standard) - nrow(filtered_standard), " |\n", sep = "")
cat("| Low | ", nrow(merged_low), " | ", nrow(filtered_low), " | ", nrow(merged_low) - nrow(filtered_low), " |\n", sep = "")
cat("| Combined | ", nrow(merged_combined), " | ", nrow(filtered_combined), " | ", nrow(merged_combined) - nrow(filtered_combined), " |\n", sep = "")
```

# Step 6: Covariate Groupings {#step6}

```{r}
#| label: step6-normalize
#| include: false

if (INCLUDE_DERIVED_COVARS) {
  # Normalize ancestry columns before rule matching:
  # 1. Trim whitespace — Excel/CSV round-trips can leave trailing spaces on
  #    multi-word values like "BL WH" or "WH MU", causing silent rule mismatches.
  # 2. Group NA → "NA" string — datasets without a Group assignment (e.g.
  #    OtherProjects) store actual NA here; the rules match on the string "NA"
  #    to classify those samples by Race + Ethnicity alone.
  filtered_combined <- filtered_combined %>%
    mutate(
      across(c(Group, Race, Ethnicity), ~str_trim(as.character(.))),
      Group = replace_na(Group, "NA")
    )
}
```

## Step 6a: Explore Distributions

```{r}
#| label: step6a-explore
#| echo: false
#| results: asis
#| message: false
#| warning: false

if (INCLUDE_DERIVED_COVARS) {
  cat("Review these distributions to inform your grouping decisions in Step 6b.\n\n")

  # NOTE: explore_*() print a summary to the console as a side effect; under results='asis'
  # that pollutes the report, so capture/suppress it and render via kable instead.
  cat("### CDX (Diagnosis)\n\n")
  invisible(capture.output(suppressMessages(
    .cdx_tab <- explore_covariate(filtered_combined, "CDX"))))
  print(kable_styling(kable(.cdx_tab, col.names = c("CDX", "N", "Pct")),
                      bootstrap_options = c("striped", "condensed"), full_width = FALSE))

  cat("\n### Group × Race × Ethnicity Combinations\n\n")
  invisible(capture.output(suppressMessages(
    .gre_tab <- explore_combinations(filtered_combined, "Group", "Race", "Ethnicity"))))
  print(kable_styling(kable(.gre_tab),
                      bootstrap_options = c("striped", "condensed"), full_width = FALSE))
} else {
  cat("*Skipped (include_derived_covariates = FALSE)*\n\n")
}
```

## Step 6b: Define Grouping Configs

\*\* Pedro is checking on "Dementia" and "Cognitively Impaired" from PER. For now I am calling them Other Dementia

```{r}
#| label: step6b-configs
#| message: false
#| warning: false

if (INCLUDE_DERIVED_COVARS) {

  # ── EDIT THIS BLOCK FOR EACH NEW DATASET ───────────────────────────────────

  REVIEW_DIR <- if (is.null(params$review_dir)) "review" else params$review_dir
  .cdx_csv <- file.path(REVIEW_DIR, "CDX_review.csv")
  .anc_csv <- file.path(REVIEW_DIR, "Ancestry_review.csv")
  if (file.exists(.cdx_csv) && file.exists(.anc_csv)) {
    # DATA-DRIVEN: decisions live in the review tables' FINAL_* columns (blank = exclude)
    .cdx_rev <- read_csv(.cdx_csv, show_col_types = FALSE) %>%
      mutate(FINAL_CDX_collapsed = na_if(trimws(FINAL_CDX_collapsed), ""))
    .anc_rev <- read_csv(.anc_csv, show_col_types = FALSE) %>%
      mutate(FINAL_Ancestry = na_if(trimws(FINAL_Ancestry), ""))
    .coll <- .cdx_rev %>% filter(!is.na(FINAL_CDX_collapsed), !is.na(CDX))
    dx_config <- make_category_config(
      source_col = "CDX", target_col = "CDX_collapsed", keep = character(0),
      collapse = split(.coll$CDX, .coll$FINAL_CDX_collapsed),
      exclude  = .cdx_rev %>% filter(is.na(FINAL_CDX_collapsed), !is.na(CDX)) %>% pull(CDX) %>% as.character()
    )
    .anc_keep <- .anc_rev %>% filter(!is.na(FINAL_Ancestry))
    .anc_rules <- list()
    for (.i in seq_len(nrow(.anc_keep))) {
      .cond <- list(Group = .anc_keep$Group[.i], Race = .anc_keep$Race[.i], Ethnicity = .anc_keep$Ethnicity[.i])
      .cond <- .cond[!vapply(.cond, function(v) is.na(v) || v == "", logical(1))]
      .anc_rules <- c(.anc_rules, setNames(list(.cond), .anc_keep$FINAL_Ancestry[.i]))
    }
    ancestry_config <- make_grouping_config(
      target_col = "Ancestry", other_label = "Other", exclude_other = FILTER_NA_COVARS, rules = .anc_rules
    )
    message("Step 6b: data-driven config from ", REVIEW_DIR)

  } else {
  # BUILT-IN DEFAULTS (used when no review CSVs are present) — edit per dataset:
  dx_config <- make_category_config(
    source_col = "CDX",
    target_col = "CDX_collapsed",
    keep     = c("MCI"),
    collapse = list(
      "AD" = c("AD", "Alzheimers Disease"),
      "NCI" = c("NCI", "Non - Cognitively Impaired"),
      "Dementia_Other" = c("Dementia Undetermined","Dementia – Undetermined", "Dementia Vascular", "Dementia – Vascular", "Dementia Other", "Cognitively Impaired", "Dementia")
    ),
    exclude  = c("0", "Other", "Insufficient Data", "NA")
  )

  ancestry_config <- make_grouping_config(
    target_col    = "Ancestry",
    other_label   = "Other",
    exclude_other = FILTER_NA_COVARS,
    rules = list(
      "NH_WH" = list(Group = "NA", Race = "WH", Ethnicity= "NH"),
      "HI_WH" = list(Group = "HI", Race = "WH"),
      "HI_WH" = list(Group = "NA", Race = "WH", Ethnicity= "HI"),
      "HI_MU" = list(Group = "HI", Race = "MU"),
      "HI_MU" = list(Race = "BL WH", Ethnicity= "HI"),
      "HI_MU" = list(Race = "BL  WH", Ethnicity= "HI"),
      "HI_MU" = list(Race = "WH MU", Ethnicity= "HI"),
      "HI_MU" = list(Group = "NA", Race = "MU", Ethnicity= "HI"),
      "HI_MU" = list(Group = "NA", Race = "WH  MU", Ethnicity= "HI"),
      "HI_BL" = list(Group = "HI", Race = "BL"),
      "HI_BL" = list(Group = "AA", Ethnicity = "HI"),
      "HI_BL" = list(Group = "NA", Race = "BL", Ethnicity= "HI"),
      "AA"    = list(Group = "AA", Ethnicity = "NH", Race = "BL"),
      "AA"    = list(Group = "AA", Ethnicity = "NH", Race = "MU"),
      "AA"    = list(Group = "NA", Race = "BL", Ethnicity= "NH"),
      "AA"    = list(Group = "AA", Ethnicity = "NH", Race = "WH"),
      "AFDC"  = list(Group = "AFDC")
    )
  )
  }  # end built-in-defaults fallback (the data-driven branch sets the configs above)

  # ───────────────────────────────────────────────────────────────────────────

  invisible(capture.output(suppressMessages({
    print_config(dx_config)
    print_config(ancestry_config)
  })))
}
```

## Step 6c: Preview Decisions

```{r}
#| label: step6c-preview
#| echo: false
#| results: asis
#| message: false
#| warning: false

if (INCLUDE_DERIVED_COVARS) {
  cat("### Diagnosis (CDX → CDX_collapsed)\n\n")
  dx_summary <- summarize_category_decisions(filtered_combined, dx_config)
  kable(dx_summary$all %>% arrange(maps_to, desc(n)), caption = "CDX decision summary (sorted by maps_to)") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
    print()

  cat(sprintf("\n*Samples retained: %d of %d*\n\n",
              dx_summary$total_after, dx_summary$total_before))

  cat("### Ancestry (Group × Race × Ethnicity → Ancestry)\n\n")
  anc_summary <- summarize_grouping_decisions(filtered_combined, ancestry_config)
  kable(anc_summary$group_summary, caption = "Ancestry group summary") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
    print()

  # Explicit combo -> Ancestry mapping (parallel to the CDX raw -> collapsed table)
  cat("\n**Ancestry mapping — Group × Race × Ethnicity → Ancestry**\n\n")
  .anc_map_csv <- file.path(if (exists("REVIEW_DIR")) REVIEW_DIR else "review", "Ancestry_review.csv")
  if (file.exists(.anc_map_csv)) {
    print(kable_styling(kable(
      suppressMessages(read_csv(.anc_map_csv, show_col_types = FALSE)) %>%
        select(any_of(c("Group","Race","Ethnicity","n_in_batch","FINAL_Ancestry","unexpected_race_eth"))) %>%
        arrange(FINAL_Ancestry, desc(n_in_batch)),
      caption = "Ancestry mapping (source: Ancestry_review.csv FINAL_Ancestry; blank = excluded)"),
      bootstrap_options = c("striped", "condensed"), full_width = FALSE))
  } else {
    print(kable_styling(kable(
      filtered_combined %>% count(Group, Race, Ethnicity, Ancestry) %>% arrange(Ancestry, desc(n)),
      caption = "Ancestry mapping (applied: Group × Race × Ethnicity → Ancestry)"),
      bootstrap_options = c("striped", "condensed"), full_width = FALSE))
  }

  if (nrow(anc_summary$unmatched) > 0) {
    cat("\n::: {.callout-warning}\n")
    cat("#### Unmatched rows — will be excluded\n\n")
    kable(anc_summary$unmatched) %>%
      kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
      print()
    cat(":::\n\n")
  } else {
    cat("\n*All rows matched an ancestry rule.* ✓\n\n")
  }
}
```

## Step 6d: Apply, Validate, and Save

```{r}
#| label: step6d-apply
#| echo: false
#| results: asis
#| message: false
#| warning: false

if (INCLUDE_DERIVED_COVARS) {
  n_before <- nrow(filtered_combined)

  # Apply CDX grouping (drops excluded and unaccounted rows internally)
  filtered_combined <- apply_category_config(filtered_combined, dx_config)

  # Apply ancestry grouping (drops unmatched rows when exclude_other = FILTER_NA_COVARS)
  filtered_combined <- apply_grouping_config(filtered_combined, ancestry_config)

  # Explicit NA safety-net filter
  if (FILTER_NA_COVARS) {
    filtered_combined <- filtered_combined %>%
      filter(!is.na(CDX_collapsed), !is.na(Ancestry))
  }

  n_after <- nrow(filtered_combined)
  cat(sprintf("**Rows before grouping:** %d | **After:** %d | **Removed:** %d\n\n",
              n_before, n_after, n_before - n_after))

  # ---- Join WGS APOE genotypes (NO plate-swap correction applied) ----
  # WGS is read from the same auto-detected metadata; used where available, Sanger APOE.geno as
  # fallback. The 2026-06 AD-Hispanic WGS plate-swap is NOT corrected here (the 2026-06-08
  # ..._plateswapcorrected_2026Jun.csv was wrong and is archived); instead the swap-affected
  # samples are masked below so APOE.geno_final falls back to Sanger for them, pending USUHS
  # confirmation. See ../../../May_2026_WGS_QC/WGS_Plate_Swap_Summary_2026Jun11.md.
  wgs_src <- read_csv(metadata_file, show_col_types = FALSE)
  if ("APOE_WGS" %in% names(wgs_src)) {
    wgs_lookup <- wgs_src %>%
      select(SAMPLE_ALIQUOT, APOE_WGS) %>%
      filter(!is.na(APOE_WGS)) %>%
      distinct() %>%
      mutate(
        APOE_WGS_norm = if_else(
          !is.na(APOE_WGS) & nchar(trimws(APOE_WGS)) == 2,
          paste0(substr(trimws(APOE_WGS), 1, 1), "/", substr(trimws(APOE_WGS), 2, 2)),
          NA_character_
        )
      )
  } else {
    wgs_lookup <- tibble(SAMPLE_ALIQUOT = character(), APOE_WGS_norm = character())
  }

  # Drop any APOE_WGS column already present so the join doesn't create .x/.y duplicates.
  filtered_combined <- filtered_combined %>%
    select(-any_of("APOE_WGS")) %>%
    left_join(wgs_lookup, by = "SAMPLE_ALIQUOT")

  # Mask swap-affected AD-Hispanic samples -> APOE.geno_final falls back to Sanger for them.
  .swap_idmap <- "../../../May_2026_WGS_QC/WGS_chr1_corrected_idmap_2026Jun11.csv"
  if (file.exists(.swap_idmap)) {
    .affected <- read_csv(.swap_idmap, show_col_types = FALSE)$SAMPLE_aslabeled
    filtered_combined <- filtered_combined %>%
      mutate(APOE_WGS_norm = if_else(as.character(SAMPLE) %in% as.character(.affected),
                                     NA_character_, APOE_WGS_norm))
  }

  # APOE4 carrier variables — derived from WGS-corrected genotype where available
  filtered_combined <- filtered_combined %>%
    mutate(
      # Flag samples where WGS genotype differs from original (both non-NA)
      APOE_WGS_changed = !is.na(APOE.geno) & !is.na(APOE_WGS_norm) &
                         (APOE.geno != APOE_WGS_norm),
      # Best available genotype: WGS takes priority, fall back to original
      APOE.geno_final  = coalesce(APOE_WGS_norm, APOE.geno),
      APOE4_count      = case_when(
        grepl("4.*4", APOE.geno_final) ~ 2L,
        grepl("4",    APOE.geno_final) ~ 1L,
        !is.na(APOE.geno_final)        ~ 0L,
        TRUE                           ~ NA_integer_
      ),
      # Derived from APOE4_count so NA propagates consistently
      APOE4_carrier    = APOE4_count >= 1L
    )

  cat(sprintf(
    "**WGS genotypes joined:** %d samples have WGS data | %d genotypes changed by WGS | %d samples gained a genotype via WGS\n\n",
    sum(!is.na(filtered_combined$APOE_WGS_norm)),
    sum(filtered_combined$APOE_WGS_changed, na.rm = TRUE),
    sum(is.na(filtered_combined$APOE.geno) & !is.na(filtered_combined$APOE_WGS_norm))
  ))

  cat("### Final Group Sizes (CDX_collapsed × Ancestry)\n\n")
  # validate_groups() prints to the console (mangled under asis); capture it and render a clean cross-tab.
  invisible(capture.output(suppressMessages(
    validate_groups(filtered_combined, c("CDX_collapsed", "Ancestry"), min_n = 5))))
  print(kable_styling(
    kable(filtered_combined %>%
            count(CDX_collapsed, Ancestry) %>%
            tidyr::pivot_wider(names_from = Ancestry, values_from = n, values_fill = 0),
          caption = "Final group sizes: CDX_collapsed × Ancestry"),
    bootstrap_options = c("striped", "condensed"), full_width = FALSE))

  write_csv(filtered_combined, file.path(OUTPUT_DIR, "filtered", "filtered_combined_post_QC.csv"))
  cat("\n`filtered_combined_post_QC.csv` updated with CDX_collapsed, Ancestry, APOE_WGS, APOE_WGS_norm, APOE_WGS_changed, APOE.geno_final, APOE4_carrier, APOE4_count.\n\n")

  # Sync standard and low CSVs to the final Step-6d sample set.
  # filtered_combined has been further filtered by ancestry/CDX grouping
  # (apply_grouping_config + FILTER_NA_COVARS). Standard and low were written
  # at Step 5 before this filtering; re-filter them now so all three files
  # share the same sample universe.
  passing_ids       <- filtered_combined$SAMPLE_ALIQUOT
  filtered_standard <- filtered_standard %>% filter(SAMPLE_ALIQUOT %in% passing_ids)
  filtered_low      <- filtered_low      %>% filter(SAMPLE_ALIQUOT %in% passing_ids)
  write_csv(filtered_standard, file.path(OUTPUT_DIR, "filtered", "filtered_standard_post_QC.csv"))
  write_csv(filtered_low,      file.path(OUTPUT_DIR, "filtered", "filtered_low_post_QC.csv"))
  cat(sprintf(
    "`filtered_standard_post_QC.csv` and `filtered_low_post_QC.csv` re-synced to Step-6d sample set: standard %d rows, low %d rows.\n\n",
    nrow(filtered_standard), nrow(filtered_low)
  ))

} else {
  cat("*Skipped (include_derived_covariates = FALSE)*\n\n")
}
```

# Step 7: Add Derived Biomarkers {#step7}

```{r}
#| label: step7-derived-biomarkers
#| echo: false
#| results: asis

if (INCLUDE_DERIVED_BIOMARKS) {
  cat("## Creating Derived Biomarker Columns\n\n")

  # Check that required columns exist
  required_cols <- c("APOE", "APOE4", "A_40", "A_42")
  missing_combined <- required_cols[!required_cols %in% names(filtered_combined)]

  if (length(missing_combined) > 0) {
    cat("::: {.callout-warning}\n")
    cat("### Missing Source Columns\n\n")
    cat("Missing in filtered_combined: ", paste(missing_combined, collapse = ", "), "\n")
    cat(":::\n\n")
  } else {
    # Add derived columns to filtered_combined
    filtered_combined <- filtered_combined %>%
      mutate(
        APOE4_minus_APOE = APOE4 - APOE,
        ABeta42_minus_ABeta40 = A_42 - A_40
      )

    cat("Added two derived biomarker columns:\n\n")
    cat("| Derived Biomarker | Formula |\n")
    cat("|-------------------|--------|\n")
    cat("| `APOE4_minus_APOE` | APOE4 - APOE |\n")
    cat("| `ABeta42_minus_ABeta40` | A_42 - A_40 |\n\n")

    # Overwrite filtered files with new columns
    write_csv(filtered_combined, file.path(OUTPUT_DIR, "filtered", "filtered_combined_post_QC.csv"))

    cat("Filtered files updated with derived biomarkers.\n\n")

    # Summary statistics for derived biomarkers
    cat("### Summary Statistics\n\n")

    apoe_stats <- filtered_combined %>%
      summarise(
        N = sum(!is.na(APOE4_minus_APOE)),
        Mean = round(mean(APOE4_minus_APOE, na.rm = TRUE), 3),
        SD = round(sd(APOE4_minus_APOE, na.rm = TRUE), 3),
        Min = round(min(APOE4_minus_APOE, na.rm = TRUE), 3),
        Median = round(median(APOE4_minus_APOE, na.rm = TRUE), 3),
        Max = round(max(APOE4_minus_APOE, na.rm = TRUE), 3)
      ) %>%
      mutate(Biomarker = "APOE4_minus_APOE", .before = 1)

    abeta_stats <- filtered_combined %>%
      summarise(
        N = sum(!is.na(ABeta42_minus_ABeta40)),
        Mean = round(mean(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
        SD = round(sd(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
        Min = round(min(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
        Median = round(median(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
        Max = round(max(ABeta42_minus_ABeta40, na.rm = TRUE), 3)
      ) %>%
      mutate(Biomarker = "ABeta42_minus_ABeta40", .before = 1)

    derived_stats <- bind_rows(apoe_stats, abeta_stats)

    kable(derived_stats, caption = "Derived Biomarker Summary Statistics") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      print()
  }
} else {
  cat("*Skipped (include_derived_biomarkers = FALSE)*\n\n")
}
```

# Step 8: Longitudinal Handling — Cross-Sectional vs Longitudinal Datasets {#step8}

Same-sample **reruns** (same `SAMPLE` & metadata, different biomarker `RUN` — *not* the study's designed technical replicates) are already resolved in Step 2 (most recent `RUN` kept). This step handles **longitudinal repeats** — the same subject (`Record_ID`) sampled at different visits (different `SAMPLE`).

We add per-subject flags (`n_visits`, `is_repeat_subject`, `subject_visit_n` \[1 = earliest by age\], `is_most_recent`) and then write **two datasets**, so the analyst always selects the right one:

-   **`filtered_combined_post_QC.csv` — DEFAULT (cross-sectional):** one row per subject, the **most recent** visit. Use for all cross-sectional analyses (no double-counting).
-   **`filtered_combined_longitudinal.csv` — longitudinal-only:** *every* visit of subjects with ≥2 visits. Use for trajectory / conversion analyses.

`preprocessed_data.rds` is built from the cross-sectional default set.

```{r}
#| label: step8-longitudinal-datasets
#| echo: false
#| results: asis

flagged <- filtered_combined %>%
  group_by(Record_ID) %>%
  mutate(
    n_visits          = dplyr::n(),
    is_repeat_subject = n_visits > 1L,
    subject_visit_n   = rank(age_at_subject, ties.method = "first"),  # 1 = earliest (age axis)
    is_most_recent    = subject_visit_n == n_visits
  ) %>%
  ungroup()

# Two datasets
filtered_cross <- flagged %>% filter(is_most_recent)                                  # one row/subject (most recent)
filtered_long  <- flagged %>% filter(is_repeat_subject) %>% arrange(Record_ID, subject_visit_n)  # all visits, ≥2-visit subjects

write_csv(filtered_cross, file.path(OUTPUT_DIR, "filtered", "filtered_combined_post_QC.csv"))
write_csv(filtered_long,  file.path(OUTPUT_DIR, "filtered", "filtered_combined_longitudinal.csv"))

n_rep_subj <- dplyr::n_distinct(filtered_long$Record_ID)
cat(sprintf(paste0("- All post-QC samples: **%d**\n",
                   "- **Cross-sectional (DEFAULT, one-per-subject):** **%d** → `filtered_combined_post_QC.csv`\n",
                   "- **Longitudinal-only (≥2 visits, all visits):** **%d** rows / **%d** subjects → `filtered_combined_longitudinal.csv`\n\n"),
            nrow(flagged), nrow(filtered_cross), nrow(filtered_long), n_rep_subj))

# DEFAULT downstream object (drives preprocessed_data.rds) = cross-sectional set
filtered_combined <- filtered_cross

if (n_rep_subj > 0) {
  repeat_tbl <- filtered_long %>%
    group_by(Record_ID) %>%
    summarise(
      n_visits = dplyr::first(n_visits),
      ages = paste(age_at_subject[order(subject_visit_n)], collapse = " / "),
      CDX  = paste(CDX_collapsed[order(subject_visit_n)], collapse = " / "),
      cdx_changed = dplyr::n_distinct(CDX_collapsed) > 1,   # diagnosis differs across visits
      RUNs = paste(RUN[order(subject_visit_n)], collapse = " / "),
      .groups = "drop"
    ) %>% arrange(desc(cdx_changed), Record_ID)

  kable(repeat_tbl,
        caption = "Longitudinal subjects (in `filtered_combined_longitudinal.csv`); `cdx_changed` = diagnosis differs across visits") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
    print()

  cat(sprintf("\n%d of %d longitudinal subject(s) have a diagnosis change across visits.\n\n",
              sum(repeat_tbl$cdx_changed), nrow(repeat_tbl)))
}
```

# Save preprocessed_data.rds {#save-rds}

```{r}
#| label: save-rds
#| include: false

# Biomarker cols = everything from NPQ files (not in original metadata, not pipeline-derived,
#   and not NPQ run-level tracking columns)
pipeline_derived_cols <- c("CDX_collapsed", "Ancestry",
                           "APOE_WGS", "APOE_WGS_norm", "APOE_WGS_changed",
                           "APOE.geno_final", "APOE4_carrier", "APOE4_count",
                           "APOE4_minus_APOE", "ABeta42_minus_ABeta40",
                           "n_visits", "is_repeat_subject", "subject_visit_n", "is_most_recent")
npq_run_cols <- c("___1", "Run", "Bay")   # NPQ instrument metadata, not protein measurements
biomarker_cols <- setdiff(
  names(filtered_combined),
  c(metadata_cols_captured, pipeline_derived_cols, npq_run_cols)
)

saveRDS(
  list(
    data           = filtered_combined,
    biomarker_cols = biomarker_cols
  ),
  file.path(OUTPUT_DIR, "preprocessed_data.rds")
)

message(sprintf("Saved preprocessed_data.rds: %d samples, %d biomarker cols",
                nrow(filtered_combined), length(biomarker_cols)))
```

# QC / Error Reports & Mapping Keys {#reports}

Per-issue error reports (with metadata context, for sending to the contributing groups) and the
grouping decisions as shareable keys. CSVs go to `<output_dir>/reports/`. Checks run on the full
batch present in metadata, so issues are surfaced even for samples that are later excluded.

```{r}
#| label: step-qc-reports
#| echo: false
#| results: asis
#| message: false
#| warning: false

if (INCLUDE_DERIVED_COVARS && exists("merged_combined")) {
  REPORTS_DIR <- file.path(OUTPUT_DIR, "reports")
  dir.create(REPORTS_DIR, showWarnings = FALSE, recursive = TRUE)
  .review_dir <- if (exists("REVIEW_DIR")) REVIEW_DIR else if (!is.null(params$review_dir)) params$review_dir else "review"

  .miss <- function(x) is.na(x) | trimws(as.character(x)) %in% c("", "Missing", "NA", "Unknown")
  .ctx  <- intersect(c("SAMPLE_ALIQUOT","SAMPLE","Record_ID","Site","Group","Race","Ethnicity",
                       "sex","age_at_subject","CDX","Case_Control","APOE.geno","APOE_WGS"),
                     names(merged_combined))
  .valid_apoe <- c("2/2","2/3","2/4","3/3","3/4","4/4")
  .written <- character(0)
  .xlsx_ok <- requireNamespace("writexl", quietly = TRUE)
  wr <- function(df, f) {
    base <- sub("\\.csv$", "", f)
    # Reports are written as .xlsx so Excel keeps APOE genotypes ("3/4") as text (no date
    # conversion). Fall back to .csv only if the writexl package is unavailable.
    if (.xlsx_ok) {
      writexl::write_xlsx(df, file.path(REPORTS_DIR, paste0(base, ".xlsx")))
      .written <<- c(.written, sprintf("`%s.xlsx` — %d row(s)", base, nrow(df)))
    } else {
      readr::write_csv(df, file.path(REPORTS_DIR, f), na = "")
      .written <<- c(.written, sprintf("`%s` (install writexl for .xlsx) — %d row(s)", f, nrow(df)))
    }
  }

  # 1. Mapping keys (shareable) -----------------------------------------------
  if (exists("dx_config")) {
    .cdx_lut <- character(0)
    for (k in dx_config$keep) .cdx_lut[k] <- k
    for (tg in names(dx_config$collapse)) for (s in dx_config$collapse[[tg]]) .cdx_lut[s] <- tg
    wr(merged_combined %>% count(CDX, name = "n_in_batch") %>%
         mutate(CDX_collapsed = unname(.cdx_lut[CDX]),
                status = if_else(is.na(CDX_collapsed), "EXCLUDED", "kept")) %>%
         arrange(is.na(CDX_collapsed), desc(n_in_batch)),
       "mapping_key_CDX.csv")
  }
  if (file.exists(file.path(.review_dir, "Ancestry_review.csv"))) {
    wr(suppressMessages(read_csv(file.path(.review_dir, "Ancestry_review.csv"), show_col_types = FALSE)) %>%
         select(any_of(c("Group","Race","Ethnicity","n_in_batch","FINAL_Ancestry","unexpected_race_eth"))),
       "mapping_key_Ancestry.csv")
  }

  # 2. Implausible values ------------------------------------------------------
  plausible <- list(age_at_subject=c(18,110), BMI=c(10,80), weight_lb=c(50,500), height_inches=c(36,90))
  .impl <- list()
  for (v in names(plausible)) if (v %in% names(merged_combined)) {
    x <- suppressWarnings(as.numeric(merged_combined[[v]]))
    hit <- which(!is.na(x) & (x < plausible[[v]][1] | x > plausible[[v]][2]))
    if (length(hit)) .impl[[v]] <- merged_combined[hit, .ctx] %>%
      mutate(variable = v, value = x[hit],
             expected_range = paste0("[", plausible[[v]][1], ", ", plausible[[v]][2], "]"))
  }
  wr(if (length(.impl)) bind_rows(.impl) else merged_combined[0, .ctx], "error_implausible_values.csv")

  # 3. Unexpected Race/Ethnicity ----------------------------------------------
  g <- merged_combined$Group_Race_Ethnicity
  unexp_idx <- which(
    (grepl("BL",g)&grepl("WH",g)) | (grepl("HI",g)&grepl("NH",g)) | (grepl("AA",g)&grepl("WH",g)) |
    (grepl("AFDC",g)&grepl("WH",g)) | (grepl("AFDC",g)&grepl("HI",g)) | (grepl("AA",g)&grepl("HI",g)))
  wr(merged_combined[unexp_idx, c(.ctx, "Group_Race_Ethnicity")] %>% distinct(),
     "error_unexpected_race_ethnicity.csv")

  # 4. Missing data ------------------------------------------------------------
  md <- merged_combined %>% transmute(SAMPLE_ALIQUOT, SAMPLE, Record_ID, Site,
    miss_CDX=.miss(CDX), miss_Group=.miss(Group), miss_Race=.miss(Race), miss_Ethnicity=.miss(Ethnicity),
    miss_sex=.miss(sex), miss_age=.miss(age_at_subject),
    miss_APOE_Sanger=.miss(APOE.geno), miss_APOE_WGS=.miss(APOE_WGS),
    miss_APOE_any=.miss(APOE.geno) & .miss(APOE_WGS))
  # by-sample report flags CRITICAL missingness only (WGS-missing is the norm, kept as a column);
  # the summary below still tallies every field incl. Sanger/WGS coverage.
  .crit <- c("miss_CDX","miss_Race","miss_Ethnicity","miss_sex","miss_age","miss_APOE_any")
  wr(md %>% filter(if_any(all_of(.crit), ~ .x)), "missing_data_by_sample.csv")
  wr(md %>% summarise(across(starts_with("miss_"), sum)) %>%
       pivot_longer(everything(), names_to="field", values_to="n_missing") %>%
       mutate(field=sub("^miss_","",field), pct_missing=round(100*n_missing/nrow(md),1)) %>%
       arrange(desc(n_missing)), "missing_data_summary.csv")

  # 5. APOE integrity (Sanger vs WGS discordance + invalid format) -------------
  .nrm <- function(x){ x<-trimws(as.character(x)); ifelse(!is.na(x)&nchar(x)==2, paste0(substr(x,1,1),"/",substr(x,2,2)), x) }
  ap <- merged_combined %>% mutate(
    wgs_norm = .nrm(APOE_WGS), sanger = as.character(APOE.geno),
    discordant     = !.miss(sanger) & !.miss(wgs_norm) & sanger != wgs_norm,
    invalid_sanger = !.miss(sanger) & !(sanger %in% .valid_apoe),
    invalid_wgs    = !.miss(wgs_norm) & !(wgs_norm %in% .valid_apoe))
  wr(ap %>% filter(discordant|invalid_sanger|invalid_wgs) %>%
       select(all_of(.ctx), wgs_norm, discordant, invalid_sanger, invalid_wgs),
     "error_APOE_integrity.csv")

  # 6. Clinical consistency ----------------------------------------------------
  caselike <- function(x) grepl("AD|Case|Dementia|Lewy|Cog Imp", x, ignore.case=TRUE)
  ctrllike <- function(x) grepl("NCI|Control", x, ignore.case=TRUE)
  cl <- merged_combined %>% mutate(
    .age=suppressWarnings(as.numeric(age_at_subject)), .aoo=suppressWarnings(as.numeric(AOO)),
    aoo_gt_age = !is.na(.aoo) & !is.na(.age) & .aoo > .age,
    cdx_casecontrol_conflict = (caselike(CDX)&ctrllike(Case_Control)) | (ctrllike(CDX)&caselike(Case_Control)))
  wr(cl %>% filter(aoo_gt_age | (cdx_casecontrol_conflict %in% TRUE)) %>%
       select(all_of(.ctx), AOO, Years_Onset, aoo_gt_age, cdx_casecontrol_conflict),
     "error_clinical_consistency.csv")
  wr(merged_combined %>% count(CDX, Case_Control) %>% arrange(desc(n)), "check_CDX_vs_CaseControl_crosstab.csv")

  # 7. Categorical integrity ---------------------------------------------------
  wr(merged_combined %>% mutate(
       bad_sex = !.miss(sex) & !(sex %in% c("M","F")),
       bad_ethnicity = !.miss(Ethnicity) & !(Ethnicity %in% c("HI","NH")),
       bad_race = !.miss(Race) & !grepl(" ", Race) & !(Race %in% c("BL","WH","MU","AI"))) %>%
       filter(bad_sex|bad_ethnicity|bad_race) %>%
       select(all_of(.ctx), bad_sex, bad_ethnicity, bad_race),
     "error_categorical_unexpected_levels.csv")
  wr(merged_combined %>% group_by(SAMPLE_ALIQUOT) %>% filter(n()>1) %>% ungroup() %>% select(all_of(.ctx)),
     "error_duplicate_SAMPLE_ALIQUOT.csv")

  # 8. NPQ-not-in-metadata is NOT duplicated here — it already exists in the output root as
  #    samples_not_in_metadata.csv (Step 2); it is referenced in the "Files to Send" section.

  cat("Reports written to `", REPORTS_DIR, "/`:\n\n", sep = "")
  for (w in .written) cat("- ", w, "\n", sep = "")
  cat("\n")
} else {
  cat("*Reports skipped (covariates off or no merged data).*\n\n")
}
```

# Output Files {#output}

```{r}
#| label: output-summary
#| echo: false
#| results: asis

cat("## Files Generated\n\n")

cat("### Merged Files (`output_files/`)\n\n")
cat("| File | Samples | Columns |\n")
cat("|------|---------|--------|\n")
merged_files <- c("merged_standard_post_QC.csv", "merged_low_post_QC.csv", "merged_combined_post_QC.csv")
for (f in merged_files) {
  fp <- file.path(OUTPUT_DIR, f)
  if (file.exists(fp)) {
    df <- read_csv(fp, show_col_types = FALSE)
    cat("| `", f, "` | ", nrow(df), " | ", ncol(df), " |\n", sep = "")
  }
}

cat("\n### Filtered Files (`output_files/filtered/`)\n\n")
cat("| File | Samples | Columns |\n")
cat("|------|---------|--------|\n")
filtered_files <- list.files(file.path(OUTPUT_DIR, "filtered"), pattern = "\\.csv$")
for (f in filtered_files) {
  fp <- file.path(OUTPUT_DIR, "filtered", f)
  df <- read_csv(fp, show_col_types = FALSE)
  cat("| `", f, "` | ", nrow(df), " | ", ncol(df), " |\n", sep = "")
}

cat("\n### Reports\n\n")
cat("- `output_files/sample_exclusion_report.csv` - Full exclusion report\n")
if (file.exists(file.path(OUTPUT_DIR, "samples_not_in_metadata.csv"))) {
  cat("- `output_files/samples_not_in_metadata.csv` - Samples missing from metadata\n")
}
```

::: callout-tip
## Pipeline Complete!

Filtered datasets are ready for downstream analysis in `output_files/filtered/`.
:::

## Next Steps

To manually edit exclusions and re-run filtering: 1. Open `output_files/sample_exclusion_report.csv` 2. Change `exclude` column as needed 3. Re-render this document with `filter_only: true`:

``` r
quarto::quarto_render("Metadata_Merge_Pipeline.qmd",
                      execute_params = list(filter_only = TRUE))
```

---

# Files to Send for Review & Verification {#send}

```{r}
#| label: files-to-send
#| echo: false
#| results: asis
#| message: false
#| warning: false

if (exists("REPORTS_DIR") && dir.exists(REPORTS_DIR)) {
  .ext  <- if (file.exists(file.path(REPORTS_DIR, "mapping_key_CDX.xlsx"))) "xlsx" else "csv"
  .rows <- function(p) {
    if (!file.exists(p)) return(NA_integer_)
    if (grepl("\\.xlsx$", p)) nrow(suppressMessages(readxl::read_excel(p)))
    else nrow(suppressMessages(readr::read_csv(p, show_col_types = FALSE)))
  }
  .rp <- function(base) file.path(REPORTS_DIR, paste0(base, ".", .ext))

  cat("Outputs below go to the contributing groups for **review / verification**. ",
      "Reports are written as `.", .ext, "`",
      if (.ext == "xlsx") " — open in Excel; APOE genotypes stay text (no date conversion)."
      else " (install the `writexl` package for Excel-safe `.xlsx`).",
      " `sample_exclusion_report.csv` is **internal** (drives filtering) and is not sent.\n\n", sep = "")

  cat("### 1. Mapping keys — confirm the groupings\n\n")
  for (b in c("mapping_key_CDX", "mapping_key_Ancestry")) {
    p <- .rp(b); if (file.exists(p)) cat(sprintf("- `reports/%s.%s` (%d rows)\n", b, .ext, .rows(p)))
  }

  cat("\n### 2. Data-issue reports — correct & verify (⚠ = has flags, ✓ = clean)\n\n")
  for (b in c("error_implausible_values", "error_unexpected_race_ethnicity", "missing_data_by_sample",
              "error_APOE_integrity", "error_clinical_consistency", "error_categorical_unexpected_levels",
              "error_duplicate_SAMPLE_ALIQUOT")) {
    n <- .rows(.rp(b))
    if (!is.na(n)) cat(sprintf("- %s `reports/%s.%s` — %d flagged\n", if (n > 0) "⚠" else "✓", b, .ext, n))
  }
  nn <- .rows(file.path(OUTPUT_DIR, "samples_not_in_metadata.csv"))
  if (!is.na(nn)) cat(sprintf("- %s `samples_not_in_metadata.csv` (output root) — %d NPQ samples lacking metadata\n",
                              if (nn > 0) "⚠" else "✓", nn))

  cat("\n### Supporting context (not action items)\n\n")
  cat(sprintf("- `reports/missing_data_summary.%s` — per-field missingness incl. APOE Sanger / WGS / any\n", .ext))
  cat(sprintf("- `reports/check_CDX_vs_CaseControl_crosstab.%s` — CDX × Case_Control cross-tab\n", .ext))
  cat("\n*Internal, do not send:* `sample_exclusion_report.csv` (drives sample filtering).\n\n")
} else {
  cat("*No reports/ folder (covariates off).*\n\n")
}
```
