---
title: "Metadata Merge Pipeline"
subtitle: "Merge Post-QC Biomarker Data with Metadata"
date: today
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
params:
  filter_only: false
  no_copy: false
editor: visual
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)
library(kableExtra)

# Configuration
PRIMARY_QC_OUTPUT <- "../Primary_QC/output_files"
INPUT_DIR <- "input_files"
OUTPUT_DIR <- "output_files"
NPQ_STANDARD_FILE <- "NPQ_post_QC.csv"
NPQ_LOW_FILE <- "NPQ_Low_post_QC.csv"
NPQ_DATE_PATTERN <- NULL  # Set to specific date or NULL for auto-detect

# Get parameters (can be set when rendering)
FILTER_ONLY <- params$filter_only
NO_COPY <- params$no_copy

# Create output directory if needed
if (!dir.exists(OUTPUT_DIR)) dir.create(OUTPUT_DIR)
if (!dir.exists(file.path(OUTPUT_DIR, "filtered"))) dir.create(file.path(OUTPUT_DIR, "filtered"))
```

# Overview

This pipeline merges post-QC biomarker data with sample metadata and applies exclusion filters to create analysis-ready datasets.

**Pipeline Steps:**

1.  **Input Files** - Verify/copy post-QC files from Primary_QC
2.  **Merge Data** - Merge biomarker data with metadata
3.  **Quality Control** - Identify duplicates, implausible values, and issues
4.  **Apply Filters** - Create final analysis-ready datasets
5.  **Add Derived Biomarkers** - Create APOE4_minus_APOE and ABeta42_minus_ABeta40 columns

```{r}
#| label: mode-display
#| echo: false
#| results: asis

if (FILTER_ONLY) {
  cat("::: {.callout-note}\n")
  cat("## Filter-Only Mode\n")
  cat("Running in **filter-only mode** - using existing merged files and exclusion report.\n")
  cat(":::\n")
}
```

```{r}
#| label: helper-functions
#| include: false

find_metadata_file <- function(input_dir) {
  all_files <- list.files(input_dir, full.names = TRUE)
  metadata_files <- all_files[grepl("metadata", basename(all_files), ignore.case = TRUE)]

  if (length(metadata_files) == 0) {
    stop("No metadata file found in ", input_dir, "\n",
         "  Expected: a file with 'metadata' in the name")
  }

  if (length(metadata_files) > 1) {
    metadata_files <- metadata_files[which.max(file.mtime(metadata_files))]
  }

  return(metadata_files)
}

find_latest_npq_files <- function(output_dir) {
  standard_files <- list.files(output_dir, pattern = "NPQ_.*_post_QC\\.csv$", full.names = TRUE)
  standard_files <- standard_files[!grepl("Low|triage", standard_files)]

  low_files <- list.files(output_dir, pattern = "NPQ_.*_Low_post_QC\\.csv$", full.names = TRUE)
  low_files <- low_files[!grepl("triage", low_files)]

  if (length(standard_files) == 0 || length(low_files) == 0) {
    stop("Could not find NPQ post-QC files in ", output_dir)
  }

  list(
    standard = standard_files[which.max(file.mtime(standard_files))],
    low = low_files[which.max(file.mtime(low_files))]
  )
}
```

# Step 1: Input Files {#step1}

```{r}
#| label: step1-input-files
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  # Check if NPQ files already exist
  existing_npq <- list.files(INPUT_DIR, pattern = "NPQ", ignore.case = TRUE)

  if (!NO_COPY && length(existing_npq) == 0) {
    cat("## Copying Files from Primary_QC\n\n")

    npq_files <- find_latest_npq_files(PRIMARY_QC_OUTPUT)
    standard_src <- npq_files$standard
    low_src <- npq_files$low

    cat("No NPQ files found in `input_files/` - copying from Primary_QC:\n\n")
    cat("- **Standard:** `", basename(standard_src), "`\n", sep = "")
    cat("- **Low:** `", basename(low_src), "`\n\n", sep = "")

    file.copy(standard_src, file.path(INPUT_DIR, NPQ_STANDARD_FILE), overwrite = TRUE)
    file.copy(low_src, file.path(INPUT_DIR, NPQ_LOW_FILE), overwrite = TRUE)

    cat("Files copied successfully.\n\n")
    existing_npq <- list.files(INPUT_DIR, pattern = "NPQ", ignore.case = TRUE)
  } else {
    cat("## Using Existing Input Files\n\n")
  }

  # Auto-detect metadata file
  metadata_file <- find_metadata_file(INPUT_DIR)

  cat("**Input files:**\n\n")
  cat("| File | Status |\n")
  cat("|------|--------|\n")
  for (f in existing_npq) {
    cat("| `", f, "` | ✓ Found |\n", sep = "")
  }
  cat("| `", basename(metadata_file), "` | ✓ Metadata (auto-detected) |\n", sep = "")

} else {
  cat("*Skipped in filter-only mode*\n\n")
  cat("Using existing files in `output_files/`\n")
  metadata_file <- find_metadata_file(INPUT_DIR)
}
```

```{r}
#| label: load-data
#| include: false

if (!FILTER_ONLY) {
  # Load metadata
  metadata <- read_csv(metadata_file, show_col_types = FALSE)
  metadata <- metadata %>%
    rename(APOE.geno = APOE) %>%
    mutate(Group_Race_Ethnicity = paste(Group, Race, Ethnicity, sep = "_"))

  # Load biomarker files
  biomarkers_postQC <- read_csv(file.path(INPUT_DIR, NPQ_STANDARD_FILE), show_col_types = FALSE)
  biomarkers_low_postQC <- read_csv(file.path(INPUT_DIR, NPQ_LOW_FILE), show_col_types = FALSE)

  # Clean column names
  biomarkers_postQC <- biomarkers_postQC %>%
    rename_with(~str_replace_all(., "[^A-Za-z0-9_]", "_"))
  biomarkers_low_postQC <- biomarkers_low_postQC %>%
    rename_with(~str_replace_all(., "[^A-Za-z0-9_]", "_"))
}
```

# Step 2: Merge Data {#step2}

```{r}
#| label: step2-merge
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("## Merging Biomarker Data with Metadata\n\n")

  # Merge standard
  merged_standard <- metadata %>%
    inner_join(biomarkers_postQC, by = c("SAMPLE_ALIQUOT" = "SampleID"), suffix = c("_meta", "_bio")) %>%
    select(-any_of(c("Run", "Bay")))

  # Merge low
  merged_low <- metadata %>%
    inner_join(biomarkers_low_postQC, by = c("SAMPLE_ALIQUOT" = "SampleID"), suffix = c("_meta", "_bio")) %>%
    select(-any_of(c("Run", "Bay")))

  # Combined biomarkers
  standard_biomarkers <- setdiff(names(biomarkers_postQC), c("SampleID", "Run", "Bay"))
  low_biomarkers <- setdiff(names(biomarkers_low_postQC), c("SampleID", "Run", "Bay"))
  unique_to_low <- setdiff(low_biomarkers, standard_biomarkers)

  combined_biomarkers <- biomarkers_postQC %>%
    full_join(
      biomarkers_low_postQC %>% select(SampleID, all_of(unique_to_low)),
      by = "SampleID"
    )

  merged_combined <- metadata %>%
    inner_join(combined_biomarkers, by = c("SAMPLE_ALIQUOT" = "SampleID"), suffix = c("_meta", "_bio"))

  cat("| Dataset | Samples | Columns |\n")
  cat("|---------|---------|--------|\n")
  cat("| Metadata | ", nrow(metadata), " | ", ncol(metadata), " |\n", sep = "")
  cat("| Standard Biomarkers | ", nrow(biomarkers_postQC), " | ", ncol(biomarkers_postQC), " |\n", sep = "")
  cat("| Low Biomarkers | ", nrow(biomarkers_low_postQC), " | ", ncol(biomarkers_low_postQC), " |\n", sep = "")
  cat("| **Merged Standard** | ", nrow(merged_standard), " | ", ncol(merged_standard), " |\n", sep = "")
  cat("| **Merged Low** | ", nrow(merged_low), " | ", ncol(merged_low), " |\n", sep = "")
  cat("| **Merged Combined** | ", nrow(merged_combined), " | ", ncol(merged_combined), " |\n", sep = "")

} else {
  cat("*Skipped in filter-only mode*\n\n")

  # Load existing merged files
  merged_combined <- read_csv(file.path(OUTPUT_DIR, "merged_combined_post_QC.csv"), show_col_types = FALSE)
  merged_standard <- read_csv(file.path(OUTPUT_DIR, "merged_standard_post_QC.csv"), show_col_types = FALSE)
  merged_low <- read_csv(file.path(OUTPUT_DIR, "merged_low_post_QC.csv"), show_col_types = FALSE)

  cat("**Using existing merged files:**\n\n")
  cat("| Dataset | Samples | Columns |\n")
  cat("|---------|---------|--------|\n")
  cat("| Merged Standard | ", nrow(merged_standard), " | ", ncol(merged_standard), " |\n", sep = "")
  cat("| Merged Low | ", nrow(merged_low), " | ", ncol(merged_low), " |\n", sep = "")
  cat("| Merged Combined | ", nrow(merged_combined), " | ", ncol(merged_combined), " |\n", sep = "")
}
```

```{r}
#| label: samples-not-in-metadata
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("\n## Samples Not in Metadata\n\n")

  all_npq_samples <- combined_biomarkers %>% select(SampleID) %>% distinct()

  technical_rep_pattern <- "^(SC|NC|IPC|HIHG)|_(SC|NC|IPC|HIHG)_|_(SC|NC|IPC|HIHG)$|HIHG"

  samples_not_in_metadata <- all_npq_samples %>%
    filter(!SampleID %in% metadata$SAMPLE_ALIQUOT) %>%
    mutate(is_technical_rep = grepl(technical_rep_pattern, SampleID, ignore.case = TRUE))

  technical_reps <- samples_not_in_metadata %>% filter(is_technical_rep)
  unmatched_samples <- samples_not_in_metadata %>% filter(!is_technical_rep)

  if (nrow(unmatched_samples) > 0) {
    cat("::: {.callout-warning}\n")
    cat("### ", nrow(unmatched_samples), " Sample(s) Not in Metadata\n\n", sep = "")
    cat("These samples are in NPQ files but NOT in metadata:\n\n")
    for (s in unmatched_samples$SampleID) {
      cat("- `", s, "`\n", sep = "")
    }
    cat("\nThese samples will NOT be included in merged output files.\n")
    cat(":::\n\n")

    write_csv(unmatched_samples %>% select(SampleID),
              file.path(OUTPUT_DIR, "samples_not_in_metadata.csv"))
  } else {
    cat("All non-technical samples in NPQ files are present in metadata. ✓\n\n")
  }

  if (nrow(technical_reps) > 0) {
    cat("*Technical replicates excluded (expected):* ", nrow(technical_reps), " samples (SC, NC, IPC, HIHG)\n", sep = "")
  }
}
```

```{r}
#| label: write-merged-files
#| include: false

if (!FILTER_ONLY) {
  write_csv(merged_standard, file.path(OUTPUT_DIR, "merged_standard_post_QC.csv"))
  write_csv(merged_low, file.path(OUTPUT_DIR, "merged_low_post_QC.csv"))
  write_csv(merged_combined, file.path(OUTPUT_DIR, "merged_combined_post_QC.csv"))
}
```

# Step 3: Quality Control {#step3}

```{r}
#| label: step3-qc-duplicates
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("## Duplicate Samples\n\n")

  duplicate_report <- merged_combined %>%
    group_by(SAMPLE) %>%
    filter(n() > 1) %>%
    arrange(SAMPLE, RUN) %>%
    mutate(
      run_order = row_number(),
      max_run = max(RUN),
      exclude = RUN != max_run,
      exclude_reason = if_else(exclude, paste0("Duplicate - keeping most recent run (", max_run, ")"), NA_character_)
    ) %>%
    ungroup()

  if (nrow(duplicate_report) > 0) {
    n_dup_samples <- n_distinct(duplicate_report$SAMPLE)
    n_to_exclude <- sum(duplicate_report$exclude)

    cat("Found **", n_dup_samples, "** samples with multiple runs (", nrow(duplicate_report), " total records)\n\n", sep = "")
    cat("- **", n_to_exclude, "** older runs will be excluded\n", sep = "")
    cat("- **", nrow(duplicate_report) - n_to_exclude, "** most recent runs will be kept\n\n", sep = "")

    dup_summary <- duplicate_report %>%
      group_by(SAMPLE) %>%
      summarise(
        n_runs = n(),
        runs = paste(RUN, collapse = ", "),
        kept_run = max(RUN),
        .groups = "drop"
      )

    kable(dup_summary, caption = "Duplicate Samples") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      print()

  } else {
    cat("No duplicate samples found. ✓\n")
  }
}
```

```{r}
#| label: step3-qc-implausible
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("\n## Implausible Values\n\n")

  plausible_ranges <- list(
    age_at_subject = c(18, 110),
    BMI = c(10, 80),
    weight = c(30, 300),
    height = c(100, 250)
  )

  implausible_list <- list()

  for (var in names(plausible_ranges)) {
    if (var %in% names(merged_combined)) {
      implausible <- merged_combined %>%
        filter(!is.na(.data[[var]]) &
               (.data[[var]] < plausible_ranges[[var]][1] |
                .data[[var]] > plausible_ranges[[var]][2])) %>%
        mutate(
          variable = var,
          value = .data[[var]],
          range = paste0("[", plausible_ranges[[var]][1], ", ", plausible_ranges[[var]][2], "]"),
          exclude_reason = paste0("Implausible ", var, ": ", value, " (expected ", range, ")")
        ) %>%
        select(SAMPLE_ALIQUOT, SAMPLE, variable, value, range, exclude_reason)

      if (nrow(implausible) > 0) {
        implausible_list[[var]] <- implausible
      }
    }
  }

  if (length(implausible_list) > 0) {
    implausible_report <- bind_rows(implausible_list)
    cat("Found **", nrow(implausible_report), "** samples with implausible values:\n\n", sep = "")

    kable(implausible_report %>% select(-exclude_reason),
          caption = "Implausible Values") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      print()

  } else {
    implausible_report <- data.frame()
    cat("No implausible values found. ✓\n")
  }
}
```

```{r}
#| label: step3-qc-race-ethnicity
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("\n## Illogical Race/Ethnicity Combinations\n\n")

  illogical_race_eth <- merged_combined %>%
    filter(
      (grepl("BL", Group_Race_Ethnicity) & grepl("WH", Group_Race_Ethnicity)) |
      (grepl("HI", Group_Race_Ethnicity) & grepl("NH", Group_Race_Ethnicity)) |
      (grepl("AA", Group_Race_Ethnicity) & grepl("WH", Group_Race_Ethnicity)) |
      (grepl("AFDC", Group_Race_Ethnicity) & grepl("WH", Group_Race_Ethnicity)) |
      (grepl("AFDC", Group_Race_Ethnicity) & grepl("HI", Group_Race_Ethnicity)) |
      (grepl("AA", Group_Race_Ethnicity) & grepl("HI", Group_Race_Ethnicity))
    ) %>%
    mutate(
      exclude_reason = paste0("Review: Illogical Race/Ethnicity combination: ", Group_Race_Ethnicity),
      exclude = FALSE  # Flagged for review but NOT auto-excluded
    )

  if (nrow(illogical_race_eth) > 0) {
    cat("::: {.callout-note}\n")
    cat("### Flagged for Review (Not Auto-Excluded)\n\n")
    cat("Found **", nrow(illogical_race_eth), "** samples with potentially illogical Race/Ethnicity combinations.\n\n", sep = "")
    cat("These are flagged in the exclusion report but **not automatically excluded**.\n")
    cat("Review and manually set `exclude=TRUE` if needed.\n")
    cat(":::\n\n")

    kable(illogical_race_eth %>%
            select(SAMPLE_ALIQUOT, SAMPLE, Group, Race, Ethnicity, Group_Race_Ethnicity) %>%
            distinct(),
          caption = "Illogical Race/Ethnicity Combinations") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      print()

  } else {
    cat("No illogical combinations found. ✓\n")
  }
}
```

```{r}
#| label: build-exclusion-report
#| include: false

if (!FILTER_ONLY) {
  # Build exclusion report
  exclusion_report <- merged_combined %>%
    select(SAMPLE_ALIQUOT, SAMPLE, RUN, Group, Race, Ethnicity, Group_Race_Ethnicity,
           Case_Control, sex, age_at_subject, CDX, APOE.geno, Site) %>%
    mutate(exclude = FALSE, exclude_reason = NA_character_)

  # Mark duplicates
  if (nrow(duplicate_report) > 0) {
    exclusion_report <- exclusion_report %>%
      left_join(
        duplicate_report %>% select(SAMPLE_ALIQUOT, exclude, exclude_reason) %>% filter(exclude),
        by = "SAMPLE_ALIQUOT", suffix = c("", "_dup")
      ) %>%
      mutate(
        exclude = coalesce(exclude_dup, exclude),
        exclude_reason = coalesce(exclude_reason_dup, exclude_reason)
      ) %>%
      select(-exclude_dup, -exclude_reason_dup)
  }

  # Mark implausible values
  if (nrow(implausible_report) > 0) {
    implausible_combined <- implausible_report %>%
      group_by(SAMPLE_ALIQUOT) %>%
      summarise(exclude_reason_implausible = paste(exclude_reason, collapse = "; "), .groups = "drop") %>%
      mutate(exclude_implausible = TRUE)

    exclusion_report <- exclusion_report %>%
      left_join(implausible_combined, by = "SAMPLE_ALIQUOT") %>%
      mutate(
        exclude_reason = case_when(
          !is.na(exclude_reason) & !is.na(exclude_reason_implausible) ~
            paste(exclude_reason, exclude_reason_implausible, sep = "; "),
          !is.na(exclude_reason_implausible) ~ exclude_reason_implausible,
          TRUE ~ exclude_reason
        ),
        exclude = exclude | coalesce(exclude_implausible, FALSE)
      ) %>%
      select(-exclude_reason_implausible, -exclude_implausible)
  }

  # Mark illogical race/ethnicity (for review, not auto-exclude)
  if (nrow(illogical_race_eth) > 0) {
    race_eth_combined <- illogical_race_eth %>%
      group_by(SAMPLE_ALIQUOT) %>%
      summarise(exclude_reason_race_eth = first(exclude_reason), .groups = "drop")

    exclusion_report <- exclusion_report %>%
      left_join(race_eth_combined, by = "SAMPLE_ALIQUOT") %>%
      mutate(
        exclude_reason = case_when(
          !is.na(exclude_reason) & !is.na(exclude_reason_race_eth) ~
            paste(exclude_reason, exclude_reason_race_eth, sep = "; "),
          !is.na(exclude_reason_race_eth) ~ exclude_reason_race_eth,
          TRUE ~ exclude_reason
        )
      ) %>%
      select(-exclude_reason_race_eth)
  }

  write_csv(exclusion_report, file.path(OUTPUT_DIR, "sample_exclusion_report.csv"))
}
```

# Step 4: Exclusion Summary {#step4}

```{r}
#| label: step4-exclusion-summary
#| echo: false
#| results: asis

exclusion_report <- read_csv(file.path(OUTPUT_DIR, "sample_exclusion_report.csv"), show_col_types = FALSE)

n_total <- nrow(exclusion_report)
n_exclude <- sum(exclusion_report$exclude)
n_keep <- sum(!exclusion_report$exclude)

cat("## Summary\n\n")

cat("| Metric | Count | Percent |\n")
cat("|--------|-------|--------|\n")
cat("| Total samples | ", n_total, " | 100% |\n", sep = "")
cat("| Samples to **EXCLUDE** | ", n_exclude, " | ", round(100*n_exclude/n_total, 1), "% |\n", sep = "")
cat("| Samples to **KEEP** | ", n_keep, " | ", round(100*n_keep/n_total, 1), "% |\n", sep = "")
cat("\n")

if (n_exclude > 0) {
  cat("## Exclusion Reasons\n\n")

  reasons <- exclusion_report %>%
    filter(exclude) %>%
    count(exclude_reason, sort = TRUE)

  kable(reasons, col.names = c("Reason", "Count"),
        caption = "Samples Excluded") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
    print()
}

# Show flagged but not excluded
review_items <- exclusion_report %>%
  filter(!exclude & !is.na(exclude_reason) & exclude_reason != "")

if (nrow(review_items) > 0) {
  cat("\n## Flagged for Review (Not Excluded)\n\n")

  review_reasons <- review_items %>% count(exclude_reason, sort = TRUE)

  kable(review_reasons, col.names = c("Reason", "Count"),
        caption = "Samples Flagged for Review") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
    print()
}
```

# Step 5: Apply Filters {#step5}

```{r}
#| label: step5-apply-filters
#| echo: false
#| results: asis

cat("## Applying Exclusion Filters\n\n")

# Get samples to exclude
samples_to_exclude <- exclusion_report %>%
  filter(exclude) %>%
  pull(SAMPLE_ALIQUOT)

cat("Excluding **", length(samples_to_exclude), "** samples from merged datasets...\n\n", sep = "")

# Filter datasets
filtered_standard <- merged_standard %>%
  filter(!SAMPLE_ALIQUOT %in% samples_to_exclude)

filtered_low <- merged_low %>%
  filter(!SAMPLE_ALIQUOT %in% samples_to_exclude)

filtered_combined <- merged_combined %>%
  filter(!SAMPLE_ALIQUOT %in% samples_to_exclude)

# Write filtered files
write_csv(filtered_standard, file.path(OUTPUT_DIR, "filtered", "filtered_standard_post_QC.csv"))
write_csv(filtered_low, file.path(OUTPUT_DIR, "filtered", "filtered_low_post_QC.csv"))
write_csv(filtered_combined, file.path(OUTPUT_DIR, "filtered", "filtered_combined_post_QC.csv"))

cat("| Dataset | Before | After | Excluded |\n")
cat("|---------|--------|-------|----------|\n")
cat("| Standard | ", nrow(merged_standard), " | ", nrow(filtered_standard), " | ", nrow(merged_standard) - nrow(filtered_standard), " |\n", sep = "")
cat("| Low | ", nrow(merged_low), " | ", nrow(filtered_low), " | ", nrow(merged_low) - nrow(filtered_low), " |\n", sep = "")
cat("| Combined | ", nrow(merged_combined), " | ", nrow(filtered_combined), " | ", nrow(merged_combined) - nrow(filtered_combined), " |\n", sep = "")
```

# Step 6: Add Derived Biomarkers {#step6}

```{r}
#| label: step6-derived-biomarkers
#| echo: false
#| results: asis

cat("## Creating Derived Biomarker Columns\n\n")

# Check that required columns exist
required_cols <- c("APOE", "APOE4", "A_40", "A_42")
missing_combined <- required_cols[!required_cols %in% names(filtered_combined)]

if (length(missing_combined) > 0) {
  cat("::: {.callout-warning}\n")
  cat("### Missing Source Columns\n\n")
  if (length(missing_combined) > 0) {
    cat("Missing in filtered_combined: ", paste(missing_combined, collapse = ", "), "\n")
  }
  cat(":::\n\n")
} else {
  # Add derived columns to filtered_combined
  filtered_combined <- filtered_combined %>%
    mutate(
      APOE4_minus_APOE = APOE4 - APOE,
      ABeta42_minus_ABeta40 = A_42 - A_40
    )


  cat("Added two derived biomarker columns:\n\n")
  cat("| Derived Biomarker | Formula |\n")
  cat("|-------------------|--------|\n")
  cat("| `APOE4_minus_APOE` | APOE4 - APOE |\n")
  cat("| `ABeta42_minus_ABeta40` | A_42 - A_40 |\n\n")

  # Overwrite filtered files with new columns
  write_csv(filtered_combined, file.path(OUTPUT_DIR, "filtered", "filtered_combined_post_QC.csv"))

  cat("Filtered files updated with derived biomarkers.\n\n")

  # Summary statistics for derived biomarkers
  cat("### Summary Statistics\n\n")

  apoe_stats <- filtered_combined %>%
    summarise(
      N = sum(!is.na(APOE4_minus_APOE)),
      Mean = round(mean(APOE4_minus_APOE, na.rm = TRUE), 3),
      SD = round(sd(APOE4_minus_APOE, na.rm = TRUE), 3),
      Min = round(min(APOE4_minus_APOE, na.rm = TRUE), 3),
      Median = round(median(APOE4_minus_APOE, na.rm = TRUE), 3),
      Max = round(max(APOE4_minus_APOE, na.rm = TRUE), 3)
    ) %>%
    mutate(Biomarker = "APOE4_minus_APOE", .before = 1)

  abeta_stats <- filtered_combined %>%
    summarise(
      N = sum(!is.na(ABeta42_minus_ABeta40)),
      Mean = round(mean(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
      SD = round(sd(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
      Min = round(min(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
      Median = round(median(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
      Max = round(max(ABeta42_minus_ABeta40, na.rm = TRUE), 3)
    ) %>%
    mutate(Biomarker = "ABeta42_minus_ABeta40", .before = 1)

  derived_stats <- bind_rows(apoe_stats, abeta_stats)

  kable(derived_stats, caption = "Derived Biomarker Summary Statistics") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
    print()
}
```

# Output Files {#output}

```{r}
#| label: output-summary
#| echo: false
#| results: asis

cat("## Files Generated\n\n")

cat("### Merged Files (`output_files/`)\n\n")
cat("| File | Samples | Columns |\n")
cat("|------|---------|--------|\n")
merged_files <- c("merged_standard_post_QC.csv", "merged_low_post_QC.csv", "merged_combined_post_QC.csv")
for (f in merged_files) {
  fp <- file.path(OUTPUT_DIR, f)
  if (file.exists(fp)) {
    df <- read_csv(fp, show_col_types = FALSE)
    cat("| `", f, "` | ", nrow(df), " | ", ncol(df), " |\n", sep = "")
  }
}

cat("\n### Filtered Files (`output_files/filtered/`)\n\n")
cat("| File | Samples | Columns |\n")
cat("|------|---------|--------|\n")
filtered_files <- list.files(file.path(OUTPUT_DIR, "filtered"), pattern = "\\.csv$")
for (f in filtered_files) {
  fp <- file.path(OUTPUT_DIR, "filtered", f)
  df <- read_csv(fp, show_col_types = FALSE)
  cat("| `", f, "` | ", nrow(df), " | ", ncol(df), " |\n", sep = "")
}

cat("\n### Reports\n\n")
cat("- `output_files/sample_exclusion_report.csv` - Full exclusion report\n")
if (file.exists(file.path(OUTPUT_DIR, "samples_not_in_metadata.csv"))) {
  cat("- `output_files/samples_not_in_metadata.csv` - Samples missing from metadata\n")
}
```

::: callout-tip
## Pipeline Complete!

Filtered datasets are ready for downstream analysis in `output_files/filtered/`.
:::

## Next Steps

To manually edit exclusions and re-run filtering: 1. Open `output_files/sample_exclusion_report.csv` 2. Change `exclude` column as needed 3. Re-render this document with `filter_only: true`:

``` r
quarto::quarto_render("Metadata_Merge_Pipeline.qmd",
                      execute_params = list(filter_only = TRUE))
```
