---
title: "Metadata Merge Pipeline"
subtitle: "Merge Post-QC Biomarker Data with Metadata"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
params:
  filter_only: false
  no_copy: true
  include_derived_covariates: true
  include_derived_biomarkers: false
  filter_na_covars: true
  input_dir: "input_files"   # where NPQ + metadata are read from (override for the example/ run or an alt location)
  output_dir: "output_files"   # neutral default; each run (QC_Runs/) overrides with its own dated dir
  review_dir: "review"   # optional dir holding {CDX,Ancestry}_review.csv (data-driven Step 6b) and Duplicate_review.csv (lab dup decisions)
  known_repeats_file: null   # optional path to Esther's Dup_sample_or_ind_*.xlsx; null = auto-detect in input_files/ then ../../../Datasets/
knit: |
  (function(input, ...) {
    od  <- tryCatch(rmarkdown::yaml_front_matter(input)$params$output_dir, error = function(e) NULL)
    if (is.null(od) || !nzchar(od)) od <- "output_files"
    tag <- sub("^output_files_?", "", basename(od)); if (!nzchar(tag)) tag <- basename(od)
    rmarkdown::render(input, output_file = paste0("Metadata_Merge_Pipeline_", tag, ".html"))
  })
---

<!--
CHANGELOG — Metadata_Merge_Pipeline.Rmd
  2026-06-23: NEW Step 1b — refresh the metadata `PrimaryQC` column from the latest Primary_QC results
   (retained -> PASS, triaged/bad -> Triage; samples not assayed in this batch keep prior, blank -> Pending).
   OVERWRITES in-batch values (the QC run is authoritative). The refreshed column propagates to the
   merged/filtered outputs; writes <output_dir>/metadata_PrimaryQC_refreshed.csv + reports/PrimaryQC_update.xlsx.
   Runs only in the full (non-filter-only) path when a PrimaryQC column is present.
  2026-06-17: mapping_key_Ancestry now ALWAYS written. Previously it was written only when
   <review_dir>/Ancestry_review.csv existed; runs using the built-in ancestry_config (no review CSV)
   silently produced no key even though the HTML output-files list (~L1859) always advertises it.
   Added an else branch that writes the built-in-config mapping (Group × Race × Ethnicity → Ancestry,
   n_in_batch), parallel to mapping_key_CDX which always writes when dx_config exists.
  2026-06-16 (a): APOE handling + error-report restructure.
   * Step 6d: REMOVED the swap-affected Sanger masking. APOE.geno_final now DEFAULTS TO WGS where
     available (Sanger only when WGS missing); on disagreement WGS wins. No sample-specific plate-swap
     correction lives in this general pipeline — the AD-Hispanic swap re-map belongs with the affected
     batch (QC_Runs/<run>/WGS_Midi_Remap/). Disagreements are surfaced so the swap stays recognizable.
   * APOE error reporting split: `error_APOE_integrity` = Sanger-vs-WGS *concordance* check (fires only
     when BOTH genotypes exist and disagree); `error_APOE_no_genotype` = no APOE call from either source.
     The by-sample summary counts only APOE_no_genotype (not discordance); APOE dropped from the
     missing-critical set to avoid double-counting.
   * Combined workbook: missing-data expanded into one drill-down sheet per critical field (Missing or 0
     CDX / Race / Ethnicity / Sex / Age) after missing_data_summary; redundant missing_data_by_sample and
     QC_error_summary sheets omitted from the workbook (standalone files still written). SUMMARY gained a
     blank `notes` column; error_unexpected_race_ethnicity gained a blank `Suggested_Coding` column.
  2026-06-15 (f): reports/ organized into sub-folders — `errors/` (error_* checks), `missing_data/`
   (missing_data_by_sample + summary), `duplicates/` (Duplicate_review + Longitudinal_review). The
   overviews (QC_error_report_combined / _by_sample / _summary), mapping keys, and the CDX×Case_Control
   crosstab stay at the reports/ top level. (wr() gained a `subdir` arg.)
  2026-06-15 (e): reports — ALL reports now written as .xlsx (incl. Duplicate_review / Longitudinal_review;
   the dup-override read accepts <review_dir>/Duplicate_review.xlsx, .csv still honored). Added a
   consolidated `QC_error_report_by_sample.xlsx` (one row per flagged sample: per-check flags + n_issues
   + issues) and `QC_error_summary.xlsx` (per-check flagged counts + batch totals); both also appear as
   sheets in QC_error_report_combined.xlsx.
  2026-06-15 (d): robustness — `input_dir` param (default "input_files"; lets the example/ run + alt
   locations read elsewhere). Step 6c built-in-config preview no longer counts `Ancestry` before it is
   applied (applies the config to a throwaway copy) — fixes a crash on the no-review-CSV path. Step 8
   falls back to raw `CDX` when `CDX_collapsed` is absent (covariates-off runs). Verified by rendering
   the synthetic example and the ALZ123 reproduction end-to-end.
  2026-06-15 (c): QC reports — added `metadata_source` (provenance) + `ID2` to per-sample report
   context; normalize the older `metatada_source` typo at load. New `QC_error_report_combined.xlsx`
   bundles every error/issue report into one workbook (SUMMARY sheet + one sheet per check) so the
   QC picture can be sent as a single file (per-issue files still written; mapping keys/crosstab excluded).
  2026-06-15 (b): converted Quarto .qmd -> R Markdown .Rmd for repo consistency (everything else is
   .Rmd). YAML `knit:` function auto-names the HTML after `params$output_dir` (e.g.
   Metadata_Merge_Pipeline_n2851_2026May.html) so each run's report is unique — replaces the old
   render_metadata_merge.R wrapper (removed). Quarto callouts (::: {.callout-*}) became Bootstrap
   .alert fenced divs (::: {.alert .alert-*}), which html_document styles.
   Chunk `#|` options are kept (knitr >= 1.35 supports them).
  2026-06-15 (a):
   * Duplicate + longitudinal cross-check is METADATA-DRIVEN (cross-batch): duplicates = same SAMPLE on
     >1 RUN, longitudinal = same Record_ID on >1 SAMPLE, detected from the full metadata file so they are
     flagged even when the two runs' biomarkers live in different batches (a standard check). Esther's
     list (Datasets/Dup_sample_or_ind_*.xlsx, param `known_repeats_file`) is the confirmation cross-check.
     - Duplicates: provisional keep = most-recent RUN among runs WITH biomarkers in this merge, OVERRIDABLE
       by lab via <review_dir>/Duplicate_review.csv (mark the run to keep with FINAL_keep="keep"). In a
       per-batch merge with only one run present, nothing is excluded; in a combined merge both are present.
     - Writes reports/Duplicate_review.csv + reports/Longitudinal_review.csv (cross-batch, reconciled to
       Esther's list: known-not-detected vs detected-not-listed).
     - Writes filtered/filtered_combined_duplicates.csv = BOTH runs of every duplicate that have biomarkers
       here (replicate feed for Replicate_Analysis; symmetric partner to filtered_combined_longitudinal.csv).
  2026-06-12 (propagated from the 8-plate run copy):
   * Step 6b is data-driven from `<review_dir>/{CDX,Ancestry}_review.csv` when present (edit the
     FINAL_* columns to regroup); falls back to the built-in config below when the CSVs are absent.
   * Step 6d reads APOE_WGS from the auto-detected metadata. NO WGS plate-swap correction is applied;
     swap-affected AD-Hispanic samples are masked to Sanger via
     ../../../May_2026_WGS_QC/WGS_chr1_corrected_idmap_2026Jun11.csv (replaces the archived
     2026-06-08 ..._plateswapcorrected_2026Jun.csv).
   * Step-6 chunks suppress covariate_explorer.R console side-effects and render clean kable tables.
  PROVENANCE: older rendered reports (Metadata_Merge_Pipeline.html dated 2026-06-08, output_files_n2851_*,
  output_files_n2850_*, etc.) were produced by EARLIER versions of this file on EARLIER metadata.
  They are preserved, not deleted — see README.md "Report / output provenance".
-->


```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)
library(kableExtra)

# Configuration
PRIMARY_QC_OUTPUT <- "../Primary_QC/output_files"
INPUT_DIR <- if (!is.null(params$input_dir) && nzchar(params$input_dir)) params$input_dir else "input_files"
DATASETS_DIR <- "../../../Datasets"   # source data lives outside the repo (see combine_metadata.R)
OUTPUT_DIR <- params$output_dir
NPQ_STANDARD_FILE <- "NPQ_post_QC.csv"
NPQ_LOW_FILE <- "NPQ_Low_post_QC.csv"
NPQ_DATE_PATTERN <- NULL  # Set to specific date or NULL for auto-detect

# Get parameters (set in YAML)
FILTER_ONLY              <- params$filter_only
NO_COPY                  <- params$no_copy
INCLUDE_DERIVED_COVARS   <- params$include_derived_covariates
INCLUDE_DERIVED_BIOMARKS <- params$include_derived_biomarkers
FILTER_NA_COVARS         <- params$filter_na_covars

# Covariate explorer (config-driven grouping and exploration utilities)
source("covariate_explorer.R")

# Create output directory if needed
if (!dir.exists(OUTPUT_DIR)) dir.create(OUTPUT_DIR)
if (!dir.exists(file.path(OUTPUT_DIR, "filtered"))) dir.create(file.path(OUTPUT_DIR, "filtered"))
```

# Overview

This pipeline merges post-QC biomarker data with sample metadata and applies exclusion filters to create analysis-ready datasets.

**Pipeline Steps:**

1.  **Input Files** - Verify/copy post-QC files from Primary_QC
1b. **Refresh PrimaryQC** *(if a `PrimaryQC` column is present)* — overwrite it from this run's Primary_QC triage (retained → PASS, triaged/bad → Triage; samples not in this batch keep prior, blank → Pending); writes `metadata_PrimaryQC_refreshed.csv` + `reports/PrimaryQC_update.xlsx`
2.  **Merge Data** - Merge biomarker data with metadata (APOE.geno normalized to `"a/b"` format)
2b. **Known Duplicate / Longitudinal Cross-Check** *(optional, if `Dup_sample_or_ind_*.xlsx` present)* — load the lab's independent list to cross-check Steps 3 & 8
3.  **Quality Control** - Identify duplicates, implausible values, and issues. Duplicates keep the most-recent run **provisionally**, overridable by the lab via `<review_dir>/Duplicate_review.csv`; both runs are written to `filtered/filtered_combined_duplicates.csv` for Replicate_Analysis
4.  **Apply Filters** - Create final analysis-ready datasets
5.  **Exclusion Summary** - Review filtered sample counts
6.  **Covariate Groupings** *(optional: `include_derived_covariates`, default TRUE)* — explore distributions (6a), define configs (6b), preview decisions (6c), apply + validate + filter (6d); joins WGS APOE genotypes, produces CDX_collapsed, Ancestry, APOE_WGS, APOE_WGS_norm, APOE_WGS_changed, APOE.geno_final (WGS preferred), APOE4_carrier, APOE4_count; rows excluded by default when either covariate is NA (`filter_na_covars`)
7.  **Derived Biomarkers** *(optional: `include_derived_biomarkers`, default FALSE)* - APOE4_minus_APOE, ABeta42_minus_ABeta40
8.  **Longitudinal Handling** - Flag repeat-subject visits; write the cross-sectional **default** (`filtered_combined_post_QC.csv`, one-per-subject = most recent) and a separate **longitudinal** dataset (`filtered_combined_longitudinal.csv`)
9.  **Save RDS** - Write `preprocessed_data.rds` (cross-sectional default) with data + biomarker_cols

```{r}
#| label: mode-display
#| echo: false
#| results: asis

if (FILTER_ONLY) {
  cat("::: {.alert .alert-info}\n")
  cat("**Filter-Only Mode**\n\n")
  cat("Running in **filter-only mode** - using existing merged files and exclusion report.\n")
  cat(":::\n")
}
```

```{r}
#| label: helper-functions
#| include: false

find_metadata_file <- function(input_dir) {
  all_files <- list.files(input_dir, full.names = TRUE)
  metadata_files <- all_files[grepl("metadata", basename(all_files), ignore.case = TRUE)]

  if (length(metadata_files) == 0) {
    stop("No metadata file found in ", input_dir, "\n",
         "  Expected: a file with 'metadata' in the name")
  }

  if (length(metadata_files) > 1) {
    metadata_files <- metadata_files[which.max(file.mtime(metadata_files))]
  }

  return(metadata_files)
}

# Locate the analyst-maintained known duplicate / longitudinal list
# (Esther's Dup_sample_or_ind_*.xlsx). Honors an explicit path, else auto-detects
# the most recent match in input_files/ then the external Datasets/ folder.
find_known_repeats_file <- function(explicit = NULL, dirs = c(INPUT_DIR, DATASETS_DIR)) {
  if (!is.null(explicit) && nzchar(explicit) && file.exists(explicit)) return(explicit)
  for (d in dirs) {
    if (!dir.exists(d)) next
    cand <- list.files(d, pattern = "Dup_sample_or_ind.*\\.xlsx$",
                       full.names = TRUE, ignore.case = TRUE)
    cand <- cand[!grepl("^~\\$", basename(cand))]   # skip Excel lock files
    if (length(cand)) return(cand[which.max(file.mtime(cand))])
  }
  NA_character_
}

find_latest_npq_files <- function(output_dir) {
  standard_files <- list.files(output_dir, pattern = "NPQ_.*_post_QC\\.csv$", full.names = TRUE)
  standard_files <- standard_files[!grepl("Low|triage", standard_files)]

  low_files <- list.files(output_dir, pattern = "NPQ_.*_Low_post_QC\\.csv$", full.names = TRUE)
  low_files <- low_files[!grepl("triage", low_files)]

  if (length(standard_files) == 0 || length(low_files) == 0) {
    stop("Could not find NPQ post-QC files in ", output_dir)
  }

  list(
    standard = standard_files[which.max(file.mtime(standard_files))],
    low = low_files[which.max(file.mtime(low_files))]
  )
}
```

# Step 1: Input Files {#step1}

```{r}
#| label: step1-input-files
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  # Check if NPQ files already exist
  existing_npq <- list.files(INPUT_DIR, pattern = "NPQ", ignore.case = TRUE)

  if (!NO_COPY && length(existing_npq) == 0) {
    cat("## Copying Files from Primary_QC\n\n")

    npq_files <- find_latest_npq_files(PRIMARY_QC_OUTPUT)
    standard_src <- npq_files$standard
    low_src <- npq_files$low

    cat("No NPQ files found in `input_files/` - copying from Primary_QC:\n\n")
    cat("- **Standard:** `", basename(standard_src), "`\n", sep = "")
    cat("- **Low:** `", basename(low_src), "`\n\n", sep = "")

    file.copy(standard_src, file.path(INPUT_DIR, NPQ_STANDARD_FILE), overwrite = TRUE)
    file.copy(low_src, file.path(INPUT_DIR, NPQ_LOW_FILE), overwrite = TRUE)

    cat("Files copied successfully.\n\n")
    existing_npq <- list.files(INPUT_DIR, pattern = "NPQ", ignore.case = TRUE)
  } else {
    cat("## Using Existing Input Files\n\n")
  }

  # Auto-detect metadata file
  metadata_file <- find_metadata_file(INPUT_DIR)

  cat("**Input files:**\n\n")
  cat("| File | Status |\n")
  cat("|------|--------|\n")
  for (f in existing_npq) {
    cat("| `", f, "` | ✓ Found |\n", sep = "")
  }
  cat("| `", basename(metadata_file), "` | ✓ Metadata (auto-detected) |\n", sep = "")

} else {
  cat("*Skipped in filter-only mode*\n\n")
  cat("Using existing files in `output_files/`\n")
  metadata_file <- find_metadata_file(INPUT_DIR)
}
```

```{r}
#| label: load-data
#| include: false

if (!FILTER_ONLY) {
  # Load metadata
  metadata <- read_csv(metadata_file, show_col_types = FALSE)
  # Robustness: older metadata files carry a `metatada_source` typo — normalize to `metadata_source`.
  if ("metatada_source" %in% names(metadata) && !"metadata_source" %in% names(metadata))
    metadata <- dplyr::rename(metadata, metadata_source = metatada_source)
  metadata <- metadata %>%
    rename(APOE.geno = APOE) %>%
    mutate(
      # Normalize APOE.geno to "a/b" digit format (e.g. "E3    E4" -> "3/4")
      APOE.geno = case_when(
        is.na(APOE.geno) ~ NA_character_,
        TRUE ~ paste0(str_extract(APOE.geno, "\\d"),
                      "/",
                      str_extract(APOE.geno, "\\d$"))
      ),
      Group_Race_Ethnicity = paste(Group, Race, Ethnicity, sep = "_")
    )

  # Capture metadata column names before merging (used later to identify biomarker cols)
  metadata_cols_captured <- names(metadata)

  # Load biomarker files
  biomarkers_postQC <- read_csv(file.path(INPUT_DIR, NPQ_STANDARD_FILE), show_col_types = FALSE)
  biomarkers_low_postQC <- read_csv(file.path(INPUT_DIR, NPQ_LOW_FILE), show_col_types = FALSE)

  # Clean column names
  biomarkers_postQC <- biomarkers_postQC %>%
    rename_with(~str_replace_all(., "[^A-Za-z0-9_]", "_"))
  biomarkers_low_postQC <- biomarkers_low_postQC %>%
    rename_with(~str_replace_all(., "[^A-Za-z0-9_]", "_"))
} else {
  # In filter_only mode: capture metadata cols from the metadata file directly
  meta_tmp <- read_csv(metadata_file, show_col_types = FALSE)
  meta_tmp <- meta_tmp %>% rename(APOE.geno = APOE)
  metadata_cols_captured <- c(names(meta_tmp), "Group_Race_Ethnicity")
  rm(meta_tmp)
}
```

# Step 1b: Refresh PrimaryQC from latest Primary_QC results {#step1b}

The metadata's `PrimaryQC` status column is refreshed here from **this run's** Primary_QC outputs so it
reflects the latest triage decisions (the QC run is authoritative for the samples it assayed):
**retained post-QC → `PASS`**, **triaged → `Triage`**, **pre-QC bad → `Triage`**; samples not assayed in
this batch keep their prior value (blank → `Pending`). The refreshed column propagates to the merged /
filtered outputs, and a refreshed full-metadata copy + change log are written for the lab. Runs only in
the full (non–filter-only) path and only when a `PrimaryQC` column is present.

```{r}
#| label: step1b-refresh-primaryqc
#| include: false

PRIMARYQC_REFRESHED <- FALSE
pq_changes <- NULL
if (!FILTER_ONLY && "PrimaryQC" %in% names(metadata)) {
  .rd_ids <- function(fname, col = "SampleID") {
    f <- file.path(PRIMARY_QC_OUTPUT, fname)
    if (!file.exists(f)) return(character(0))
    d <- suppressMessages(readr::read_csv(f, show_col_types = FALSE))
    if (col %in% names(d)) unique(as.character(d[[col]])) else character(0)
  }
  triaged  <- .rd_ids("samples_to_triage.csv")
  bad      <- .rd_ids("bad_samples.csv")
  retained <- unique(as.character(biomarkers_postQC$SampleID))   # cleared post-QC this run

  .sa  <- as.character(metadata$SAMPLE_ALIQUOT)
  .old <- as.character(metadata$PrimaryQC)
  metadata$PrimaryQC <- dplyr::case_when(
    .sa %in% triaged           ~ "Triage",
    .sa %in% bad               ~ "Triage",
    .sa %in% retained          ~ "PASS",
    is.na(.old) | .old == ""   ~ "Pending",
    TRUE                       ~ .old
  )
  pq_changes <- tibble::tibble(SAMPLE_ALIQUOT = .sa,
                               PrimaryQC_old = .old,
                               PrimaryQC_new = metadata$PrimaryQC) %>%
    dplyr::filter(PrimaryQC_old != PrimaryQC_new |
                  (is.na(PrimaryQC_old) & !is.na(PrimaryQC_new)))
  PRIMARYQC_REFRESHED <- TRUE
}
```

```{r}
#| label: step1b-report
#| echo: false
#| results: asis

if (!FILTER_ONLY && PRIMARYQC_REFRESHED) {
  cat("**Status after refresh (full metadata, n = ", nrow(metadata), "):**\n\n", sep = "")
  print(kable(as.data.frame(table(PrimaryQC = metadata$PrimaryQC)),
              col.names = c("PrimaryQC", "n")) %>%
          kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE))
  cat("\n**Updated ", nrow(pq_changes), " sample(s)** vs the incoming metadata", sep = "")
  if (nrow(pq_changes) > 0) {
    cat(":\n\n")
    print(kable(dplyr::count(pq_changes, PrimaryQC_old, PrimaryQC_new, name = "n"),
                caption = "PrimaryQC changes (old -> new)") %>%
            kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE))
  } else cat(" (none).\n\n")

  # Refreshed full metadata + change log, so the lab can adopt the updated PrimaryQC.
  readr::write_csv(metadata, file.path(OUTPUT_DIR, "metadata_PrimaryQC_refreshed.csv"))
  .rdir <- file.path(OUTPUT_DIR, "reports"); dir.create(.rdir, showWarnings = FALSE, recursive = TRUE)
  if (requireNamespace("writexl", quietly = TRUE))
    writexl::write_xlsx(pq_changes, file.path(.rdir, "PrimaryQC_update.xlsx"))
  else readr::write_csv(pq_changes, file.path(.rdir, "PrimaryQC_update.csv"))
  cat("\nWrote refreshed full metadata -> `", basename(OUTPUT_DIR),
      "/metadata_PrimaryQC_refreshed.csv` and change log -> `reports/PrimaryQC_update.xlsx`.\n\n", sep = "")
} else if (!FILTER_ONLY) {
  cat("*No `PrimaryQC` column in the metadata - refresh skipped.*\n\n")
}
```

# Step 2: Merge Data {#step2}

```{r}
#| label: step2-merge
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("## Merging Biomarker Data with Metadata\n\n")

  # Merge standard
  merged_standard <- metadata %>%
    inner_join(biomarkers_postQC, by = c("SAMPLE_ALIQUOT" = "SampleID"), suffix = c("_meta", "_bio")) %>%
    select(-any_of(c("Run", "Bay")))

  # Merge low
  merged_low <- metadata %>%
    inner_join(biomarkers_low_postQC, by = c("SAMPLE_ALIQUOT" = "SampleID"), suffix = c("_meta", "_bio")) %>%
    select(-any_of(c("Run", "Bay")))

  # Combined biomarkers
  standard_biomarkers <- setdiff(names(biomarkers_postQC), c("SampleID", "Run", "Bay"))
  low_biomarkers <- setdiff(names(biomarkers_low_postQC), c("SampleID", "Run", "Bay"))
  unique_to_low <- setdiff(low_biomarkers, standard_biomarkers)

  combined_biomarkers <- biomarkers_postQC %>%
    full_join(
      biomarkers_low_postQC %>% select(SampleID, all_of(unique_to_low)),
      by = "SampleID"
    )

  merged_combined <- metadata %>%
    inner_join(combined_biomarkers, by = c("SAMPLE_ALIQUOT" = "SampleID"), suffix = c("_meta", "_bio"))

  cat("| Dataset | Samples | Columns |\n")
  cat("|---------|---------|--------|\n")
  cat("| Metadata | ", nrow(metadata), " | ", ncol(metadata), " |\n", sep = "")
  cat("| Standard Biomarkers | ", nrow(biomarkers_postQC), " | ", ncol(biomarkers_postQC), " |\n", sep = "")
  cat("| Low Biomarkers | ", nrow(biomarkers_low_postQC), " | ", ncol(biomarkers_low_postQC), " |\n", sep = "")
  cat("| **Merged Standard** | ", nrow(merged_standard), " | ", ncol(merged_standard), " |\n", sep = "")
  cat("| **Merged Low** | ", nrow(merged_low), " | ", ncol(merged_low), " |\n", sep = "")
  cat("| **Merged Combined** | ", nrow(merged_combined), " | ", ncol(merged_combined), " |\n", sep = "")

} else {
  cat("*Skipped in filter-only mode*\n\n")

  # Load existing merged files
  merged_combined <- read_csv(file.path(OUTPUT_DIR, "merged_combined_post_QC.csv"), show_col_types = FALSE)
  merged_standard <- read_csv(file.path(OUTPUT_DIR, "merged_standard_post_QC.csv"), show_col_types = FALSE)
  merged_low <- read_csv(file.path(OUTPUT_DIR, "merged_low_post_QC.csv"), show_col_types = FALSE)

  cat("**Using existing merged files:**\n\n")
  cat("| Dataset | Samples | Columns |\n")
  cat("|---------|---------|--------|\n")
  cat("| Merged Standard | ", nrow(merged_standard), " | ", ncol(merged_standard), " |\n", sep = "")
  cat("| Merged Low | ", nrow(merged_low), " | ", ncol(merged_low), " |\n", sep = "")
  cat("| Merged Combined | ", nrow(merged_combined), " | ", ncol(merged_combined), " |\n", sep = "")
}
```

```{r}
#| label: samples-not-in-metadata
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("\n## Samples Not in Metadata\n\n")

  all_npq_samples <- combined_biomarkers %>% select(SampleID) %>% distinct()

  technical_rep_pattern <- "^(SC|NC|IPC|HIHG)|_(SC|NC|IPC|HIHG)_|_(SC|NC|IPC|HIHG)$|HIHG"

  samples_not_in_metadata <- all_npq_samples %>%
    filter(!SampleID %in% metadata$SAMPLE_ALIQUOT) %>%
    mutate(is_technical_rep = grepl(technical_rep_pattern, SampleID, ignore.case = TRUE))

  technical_reps <- samples_not_in_metadata %>% filter(is_technical_rep)
  unmatched_samples <- samples_not_in_metadata %>% filter(!is_technical_rep)

  if (nrow(unmatched_samples) > 0) {
    cat("::: {.alert .alert-warning}\n")
    cat("**", nrow(unmatched_samples), " Sample(s) Not in Metadata**\n\n", sep = "")
    cat("These samples are in NPQ files but NOT in metadata:\n\n")
    for (s in unmatched_samples$SampleID) {
      cat("- `", s, "`\n", sep = "")
    }
    cat("\nThese samples will NOT be included in merged output files.\n")
    cat(":::\n\n")

    write_csv(unmatched_samples %>% select(SampleID),
              file.path(OUTPUT_DIR, "samples_not_in_metadata.csv"))
  } else {
    cat("All non-technical samples in NPQ files are present in metadata. ✓\n\n")
  }

  if (nrow(technical_reps) > 0) {
    cat("*Technical replicates excluded (expected):* ", nrow(technical_reps), " samples (SC, NC, IPC, HIHG)\n", sep = "")
  }
}
```

```{r}
#| label: write-merged-files
#| include: false

if (!FILTER_ONLY) {
  write_csv(merged_standard, file.path(OUTPUT_DIR, "merged_standard_post_QC.csv"))
  write_csv(merged_low, file.path(OUTPUT_DIR, "merged_low_post_QC.csv"))
  write_csv(merged_combined, file.path(OUTPUT_DIR, "merged_combined_post_QC.csv"))
}
```

# Known Duplicate / Longitudinal Cross-Check {#known-repeats}

The lab maintains an independent list of intentional **duplicate** samples (same `SAMPLE`
run on two plates) and **longitudinal** samples (same subject `Record_ID`, different draws)
in `Dup_sample_or_ind_*.xlsx`. We load it here so Step 3 (duplicates) and Step 8
(longitudinal) can cross-check their automatic detection against it. If the file is absent
the pipeline runs exactly as before.

```{r}
#| label: known-repeats-load
#| echo: false
#| results: asis

KNOWN_REPEATS_FILE <- params$known_repeats_file
known_path         <- find_known_repeats_file(KNOWN_REPEATS_FILE)
HAVE_KNOWN_REPEATS <- !is.na(known_path) && requireNamespace("readxl", quietly = TRUE)

known_dups_df <- NULL; known_long_df <- NULL
known_dup_samples <- character(0); known_long_records <- character(0)

if (HAVE_KNOWN_REPEATS) {
  .sheets     <- readxl::excel_sheets(known_path)
  .dup_sheet  <- .sheets[grepl("dup",      .sheets, ignore.case = TRUE)][1]
  .long_sheet <- .sheets[grepl("longitud", .sheets, ignore.case = TRUE)][1]

  if (!is.na(.dup_sheet)) {
    known_dups_df <- suppressMessages(readxl::read_excel(known_path, sheet = .dup_sheet)) %>%
      mutate(SAMPLE = as.character(SAMPLE),
             SAMPLE_ALIQUOT = as.character(SAMPLE_ALIQUOT))
    known_dup_samples <- unique(known_dups_df$SAMPLE)
  }
  if (!is.na(.long_sheet)) {
    known_long_df <- suppressMessages(readxl::read_excel(known_path, sheet = .long_sheet)) %>%
      mutate(SAMPLE = as.character(SAMPLE),
             Record_ID = as.character(Record_ID))
    known_long_records <- unique(known_long_df$Record_ID)
  }

  cat(sprintf(paste0("Loaded `%s`:\n\n",
                     "- **%d** known duplicate sample IDs (%d run records)\n",
                     "- **%d** known longitudinal subjects (%d draw records)\n\n"),
              basename(known_path),
              length(known_dup_samples),  if (is.null(known_dups_df)) 0L else nrow(known_dups_df),
              length(known_long_records), if (is.null(known_long_df)) 0L else nrow(known_long_df)))
} else if (is.na(known_path)) {
  cat("*No `Dup_sample_or_ind_*.xlsx` found in `input_files/` or `Datasets/` — ",
      "duplicate/longitudinal cross-check skipped (automatic detection still runs).*\n\n", sep = "")
} else {
  cat("*Package `readxl` not installed — known duplicate/longitudinal cross-check skipped.*\n\n")
}

# ── Metadata-level (cross-batch) detection ────────────────────────────────────
# The metadata file spans ALL batches, so duplicates (same SAMPLE on >1 RUN) and longitudinal
# subjects (same Record_ID on >1 SAMPLE) are detected here even when the two runs' biomarkers were
# assayed in different batches. This is the standard, biomarker-independent check. The keep/exclude
# decision (Step 3) still operates only on the runs that have biomarkers in the current merge, so the
# pipeline behaves correctly whether run per-batch or on a combined dataset.
meta_all <- if (exists("metadata")) metadata else tryCatch(
  suppressMessages(readr::read_csv(find_metadata_file(INPUT_DIR), show_col_types = FALSE)),
  error = function(e) NULL)

meta_dups <- NULL; meta_long <- NULL
if (!is.null(meta_all) && all(c("SAMPLE", "RUN", "Record_ID") %in% names(meta_all))) {
  .ma <- meta_all %>% mutate(across(any_of(c("SAMPLE", "SAMPLE_ALIQUOT", "Record_ID", "RUN")), as.character))
  meta_dups <- .ma %>% group_by(SAMPLE)    %>% filter(dplyr::n_distinct(RUN)    > 1) %>% ungroup()
  meta_long <- .ma %>% group_by(Record_ID) %>% filter(dplyr::n_distinct(SAMPLE) > 1) %>% ungroup()
}
meta_dup_samples  <- if (is.null(meta_dups)) character(0) else unique(meta_dups$SAMPLE)
meta_long_records <- if (is.null(meta_long)) character(0) else unique(meta_long$Record_ID)

if (!is.null(meta_dups)) {
  cat(sprintf(paste0("**Metadata-level detection (all batches in the metadata file):**\n\n",
                     "- **%d** duplicate sample ID(s) — same SAMPLE on >1 RUN\n",
                     "- **%d** longitudinal subject(s) — same Record_ID on >1 SAMPLE\n\n"),
              length(meta_dup_samples), length(meta_long_records)))
}
```

# Step 3: Quality Control {#step3}

```{r}
#| label: step3-qc-duplicates
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("## Duplicate Samples\n\n")

  .REPORTS_DIR <- file.path(OUTPUT_DIR, "reports")
  dir.create(.REPORTS_DIR, showWarnings = FALSE, recursive = TRUE)
  .review_dir <- if (!is.null(params$review_dir)) params$review_dir else "review"

  # Optional lab override: <review_dir>/Duplicate_review.xlsx (or .csv) — mark the run to KEEP with
  # FINAL_keep = "keep" (per SAMPLE_ALIQUOT). Any duplicate group with at least one "keep"
  # is resolved by it (all other runs excluded); groups with no "keep" fall back to the
  # provisional rule (most-recent RUN). Blank = no decision yet.
  dup_override <- NULL
  .dup_xlsx <- file.path(.review_dir, "Duplicate_review.xlsx")
  .dup_csv  <- file.path(.review_dir, "Duplicate_review.csv")
  .dup_override_path <- if (file.exists(.dup_xlsx)) .dup_xlsx else if (file.exists(.dup_csv)) .dup_csv else NA_character_
  if (!is.na(.dup_override_path)) {
    .dro <- if (grepl("\\.xlsx$", .dup_override_path) && requireNamespace("readxl", quietly = TRUE))
              suppressMessages(readxl::read_excel(.dup_override_path))
            else suppressMessages(read_csv(.dup_override_path, show_col_types = FALSE))
    dup_override <- .dro %>%
      transmute(SAMPLE_ALIQUOT = as.character(SAMPLE_ALIQUOT),
                FINAL_keep = tolower(trimws(as.character(FINAL_keep)))) %>%
      filter(FINAL_keep %in% c("keep", "exclude"))
    if (nrow(dup_override) == 0) dup_override <- NULL
  }

  duplicate_report <- merged_combined %>%
    mutate(SAMPLE = as.character(SAMPLE),
           SAMPLE_ALIQUOT = as.character(SAMPLE_ALIQUOT)) %>%
    group_by(SAMPLE) %>%
    filter(n() > 1) %>%
    arrange(SAMPLE, RUN) %>%
    mutate(
      run_order        = row_number(),
      n_runs           = n(),
      max_run          = suppressWarnings(max(RUN)),  # suppressWarnings: harmless when 0 dup groups
      is_known_dup     = SAMPLE %in% known_dup_samples,
      provisional_keep = RUN == max_run            # default rule: keep most-recent RUN
    ) %>%
    ungroup()

  # Resolve keep/exclude: lab override (FINAL_keep) where present, else provisional.
  if (!is.null(dup_override)) {
    duplicate_report <- duplicate_report %>%
      left_join(dup_override, by = "SAMPLE_ALIQUOT") %>%
      group_by(SAMPLE) %>%
      mutate(group_has_keep = any(FINAL_keep %in% "keep")) %>%
      ungroup() %>%
      mutate(
        has_override = group_has_keep,
        keep_final   = case_when(
          FINAL_keep %in% "keep" ~ TRUE,
          group_has_keep         ~ FALSE,   # group resolved by a "keep" on another run
          TRUE                   ~ provisional_keep
        )
      )
  } else {
    duplicate_report <- duplicate_report %>%
      mutate(has_override = FALSE, keep_final = provisional_keep)
  }

  duplicate_report <- duplicate_report %>%
    mutate(
      exclude         = !keep_final,
      decision_source = if_else(has_override,
                                "lab-confirmed (Duplicate_review.csv)",
                                "provisional (most-recent RUN; pending lab review)"),
      exclude_reason  = if_else(exclude,
        paste0("Duplicate SAMPLE ", SAMPLE, " - ", decision_source),
        NA_character_)
    )

  biom_aliquots <- as.character(merged_combined$SAMPLE_ALIQUOT)

  # ── (B) Metadata-level (cross-batch) standard check + lab review template + reconciliation ──
  if (!is.null(meta_dups)) {
    meta_dup_report <- meta_dups %>%
      mutate(in_known_list       = SAMPLE %in% known_dup_samples,
             has_biomarkers_here = SAMPLE_ALIQUOT %in% biom_aliquots) %>%
      group_by(SAMPLE) %>%
      mutate(n_runs                      = dplyr::n(),
             n_runs_with_biomarkers_here = sum(has_biomarkers_here),
             .max_present = if (any(has_biomarkers_here)) max(RUN[has_biomarkers_here]) else NA_character_,
             provisional_keep = !is.na(.max_present) & RUN == .max_present) %>%
      ungroup() %>% select(-.max_present)

    n_resolvable <- meta_dup_report %>% filter(n_runs_with_biomarkers_here >= 2) %>%
                      distinct(SAMPLE) %>% nrow()
    cat("Detected **", length(meta_dup_samples),
        "** duplicate sample(s) in the metadata (same SAMPLE on >1 RUN, all batches); **",
        n_resolvable, "** have ≥2 runs with biomarkers in *this* merge (a keep is selected here) — ",
        "the rest await their other run in a combined run.\n\n", sep = "")

    if (HAVE_KNOWN_REPEATS && length(known_dup_samples) > 0) {
      .knd <- setdiff(known_dup_samples, meta_dup_samples)   # on list, not in this metadata
      .dnk <- setdiff(meta_dup_samples, known_dup_samples)   # in metadata, not on list (verify)
      cat("**Cross-check vs lab's known list:** ", length(intersect(meta_dup_samples, known_dup_samples)),
          " concordant ✓; ", length(.knd), " on list but not in this metadata; ",
          length(.dnk), " in metadata but not on list (verify).\n\n", sep = "")
      if (length(.dnk) > 0) {
        kable(meta_dup_report %>% filter(SAMPLE %in% .dnk) %>%
                select(any_of(c("SAMPLE","SAMPLE_ALIQUOT","RUN","Record_ID","CDX","Site"))),
              caption = "Metadata duplicates NOT on the lab's known list — verify") %>%
          kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>% print()
      }
    }

    .ctx_dup <- intersect(c("metadata_source","Record_ID","sex","age_at_subject","CDX",
                            "Case_Control","APOE.geno","Site"), names(meta_dup_report))
    dup_review_out <- meta_dup_report %>%
      mutate(dup_group = SAMPLE, FINAL_keep = "") %>%        # lab marks the run to keep
      select(SAMPLE, dup_group, SAMPLE_ALIQUOT, RUN, n_runs, n_runs_with_biomarkers_here,
             has_biomarkers_here, in_known_list, provisional_keep, FINAL_keep, any_of(.ctx_dup)) %>%
      arrange(SAMPLE, RUN)
    .ddir <- file.path(.REPORTS_DIR, "duplicates"); dir.create(.ddir, showWarnings = FALSE, recursive = TRUE)
    if (requireNamespace("writexl", quietly = TRUE)) writexl::write_xlsx(dup_review_out, file.path(.ddir, "Duplicate_review.xlsx")) else write_csv(dup_review_out, file.path(.ddir, "Duplicate_review.csv"), na = "")
    cat("Lab decision template (all batches) → `reports/duplicates/Duplicate_review.xlsx` ",
        "(mark the run to keep with `FINAL_keep = \"keep\"`, copy to `", .review_dir,
        "/`, re-run).\n\n", sep = "")
  } else if (nrow(duplicate_report) > 0) {
    # Fallback: metadata lacks SAMPLE/RUN/Record_ID — use the in-merge duplicates for the review file.
    meta_dup_samples <- unique(duplicate_report$SAMPLE)
    .ctx_dup <- intersect(c("Record_ID","sex","age_at_subject","CDX","Case_Control","APOE.geno","Site"),
                          names(merged_combined))
    dup_review_out <- duplicate_report %>%
      mutate(dup_group = SAMPLE, FINAL_keep = "") %>%
      select(SAMPLE, dup_group, SAMPLE_ALIQUOT, RUN, n_runs, in_known_list = is_known_dup,
             provisional_keep, FINAL_keep, decision_source, any_of(.ctx_dup)) %>%
      arrange(SAMPLE, RUN)
    .ddir <- file.path(.REPORTS_DIR, "duplicates"); dir.create(.ddir, showWarnings = FALSE, recursive = TRUE)
    if (requireNamespace("writexl", quietly = TRUE)) writexl::write_xlsx(dup_review_out, file.path(.ddir, "Duplicate_review.xlsx")) else write_csv(dup_review_out, file.path(.ddir, "Duplicate_review.csv"), na = "")
  } else {
    meta_dup_samples <- character(0)
  }

  # ── (C) Keep/exclude summary for duplicates resolvable in THIS merge ──
  if (nrow(duplicate_report) > 0) {
    n_in_merge   <- n_distinct(duplicate_report$SAMPLE)
    n_to_exclude <- sum(duplicate_report$exclude)
    n_lab        <- n_distinct(duplicate_report$SAMPLE[duplicate_report$has_override])
    cat("In this merge, **", n_in_merge, "** duplicate sample(s) have ≥2 runs with biomarkers: **",
        n_to_exclude, "** run(s) excluded; **", n_lab, "** lab-confirmed, ", n_in_merge - n_lab,
        " provisional (most-recent RUN).\n\n", sep = "")
    dup_summary <- duplicate_report %>% group_by(SAMPLE) %>%
      summarise(n_runs = dplyr::first(n_runs), is_known = dplyr::first(is_known_dup),
                runs = paste(RUN, collapse = ", "), kept_run = paste(RUN[keep_final], collapse = ", "),
                decision = dplyr::first(decision_source), .groups = "drop") %>%
      arrange(decision, SAMPLE)
    kable(dup_summary, caption = "Duplicates resolvable in this merge — kept run + decision") %>%
      kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>% print()
  } else {
    cat("No duplicate sample has >1 run with biomarkers in this merge (nothing to exclude here).\n\n")
  }

  # ── (D) Replicate feed: all runs WITH biomarkers here for metadata-duplicate samples ──
  # Symmetric partner to filtered_combined_longitudinal.csv; both runs of a duplicate are retained
  # for cross-plate concordance in Replicate_Analysis (in a per-batch merge only the present run(s) appear).
  dup_keep_lut <- duplicate_report %>% select(SAMPLE_ALIQUOT, keep_final, decision_source)
  filtered_dups <- merged_combined %>%
    mutate(SAMPLE = as.character(SAMPLE), SAMPLE_ALIQUOT = as.character(SAMPLE_ALIQUOT)) %>%
    filter(SAMPLE %in% meta_dup_samples) %>%
    left_join(dup_keep_lut, by = "SAMPLE_ALIQUOT") %>%
    mutate(dup_group = SAMPLE, provisional_keep = keep_final,
           decision_source = coalesce(decision_source, "single run in this merge")) %>%
    arrange(SAMPLE, RUN) %>%
    relocate(dup_group, provisional_keep, decision_source, .after = SAMPLE)
  write_csv(filtered_dups, file.path(OUTPUT_DIR, "filtered", "filtered_combined_duplicates.csv"))
  .n_pairs <- filtered_dups %>% count(SAMPLE) %>% filter(n >= 2) %>% nrow()
  cat("Replicate feed → `filtered/filtered_combined_duplicates.csv`: **", nrow(filtered_dups),
      "** run-row(s) across **", n_distinct(filtered_dups$SAMPLE), "** sample(s); **", .n_pairs,
      "** have ≥2 runs here (usable cross-plate replicate pairs).\n\n", sep = "")
}
```

```{r}
#| label: step3-qc-implausible
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("\n## Implausible Values\n\n")

  # Column names are weight_lb / height_inches (imperial), not weight/height — the old
  # metric keys silently skipped these checks (the `if (var %in% names(...))` guard below).
  plausible_ranges <- list(
    age_at_subject = c(18, 110),
    BMI = c(10, 80),
    weight_lb = c(60, 660),
    height_inches = c(39, 96)
  )

  implausible_list <- list()

  for (var in names(plausible_ranges)) {
    if (var %in% names(merged_combined)) {
      implausible <- merged_combined %>%
        filter(!is.na(.data[[var]]) &
               (.data[[var]] < plausible_ranges[[var]][1] |
                .data[[var]] > plausible_ranges[[var]][2])) %>%
        mutate(
          variable = var,
          value = .data[[var]],
          range = paste0("[", plausible_ranges[[var]][1], ", ", plausible_ranges[[var]][2], "]"),
          exclude_reason = paste0("Implausible ", var, ": ", value, " (expected ", range, ")")
        ) %>%
        select(SAMPLE_ALIQUOT, SAMPLE, variable, value, range, exclude_reason)

      if (nrow(implausible) > 0) {
        implausible_list[[var]] <- implausible
      }
    }
  }

  if (length(implausible_list) > 0) {
    implausible_report <- bind_rows(implausible_list)
    cat("Found **", nrow(implausible_report), "** samples with implausible values:\n\n", sep = "")

    kable(implausible_report %>% select(-exclude_reason),
          caption = "Implausible Values") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      print()

  } else {
    implausible_report <- data.frame()
    cat("No implausible values found. ✓\n")
  }
}
```

```{r}
#| label: step3-qc-race-ethnicity
#| echo: false
#| results: asis

if (!FILTER_ONLY) {
  cat("\n## Unexpected Race/Ethnicity Combinations\n\n")

  unexpected_race_eth <- merged_combined %>%
    filter(
      (grepl("BL", Group_Race_Ethnicity) & grepl("WH", Group_Race_Ethnicity)) |
      (grepl("HI", Group_Race_Ethnicity) & grepl("NH", Group_Race_Ethnicity)) |
      (grepl("AA", Group_Race_Ethnicity) & grepl("WH", Group_Race_Ethnicity)) |
      (grepl("AFDC", Group_Race_Ethnicity) & grepl("WH", Group_Race_Ethnicity)) |
      (grepl("AFDC", Group_Race_Ethnicity) & grepl("HI", Group_Race_Ethnicity)) |
      (grepl("AA", Group_Race_Ethnicity) & grepl("HI", Group_Race_Ethnicity))
    ) %>%
    mutate(
      exclude_reason = paste0("Review: Unexpected Race/Ethnicity combination: ", Group_Race_Ethnicity),
      exclude = FALSE  # Flagged for review but NOT auto-excluded
    )

  if (nrow(unexpected_race_eth) > 0) {
    cat("::: {.alert .alert-info}\n")
    cat("**Flagged for Review (Not Auto-Excluded)**\n\n")
    cat("Found **", nrow(unexpected_race_eth), "** samples with potentially unexpected Race/Ethnicity combinations.\n\n", sep = "")
    cat("These are flagged in the exclusion report but **not automatically excluded**.\n")
    cat("Review and manually set `exclude=TRUE` if needed.\n")
    cat(":::\n\n")

    kable(unexpected_race_eth %>%
            select(SAMPLE_ALIQUOT, SAMPLE, Group, Race, Ethnicity, Group_Race_Ethnicity) %>%
            distinct(),
          caption = "Unexpected Race/Ethnicity Combinations") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      print()

  } else {
    cat("No unexpected combinations found. ✓\n")
  }
}
```

```{r}
#| label: build-exclusion-report
#| include: false

if (!FILTER_ONLY) {
  # Build exclusion report
  exclusion_report <- merged_combined %>%
    select(SAMPLE_ALIQUOT, SAMPLE, RUN, Group, Race, Ethnicity, Group_Race_Ethnicity,
           Case_Control, sex, age_at_subject, CDX, APOE.geno, Site) %>%
    mutate(exclude = FALSE, exclude_reason = NA_character_)

  # Mark duplicates
  if (nrow(duplicate_report) > 0) {
    exclusion_report <- exclusion_report %>%
      left_join(
        duplicate_report %>% select(SAMPLE_ALIQUOT, exclude, exclude_reason) %>% filter(exclude),
        by = "SAMPLE_ALIQUOT", suffix = c("", "_dup")
      ) %>%
      mutate(
        exclude = coalesce(exclude_dup, exclude),
        exclude_reason = coalesce(exclude_reason_dup, exclude_reason)
      ) %>%
      select(-exclude_dup, -exclude_reason_dup)
  }

  # Mark implausible values
  if (nrow(implausible_report) > 0) {
    implausible_combined <- implausible_report %>%
      group_by(SAMPLE_ALIQUOT) %>%
      summarise(exclude_reason_implausible = paste(exclude_reason, collapse = "; "), .groups = "drop") %>%
      mutate(exclude_implausible = TRUE)

    exclusion_report <- exclusion_report %>%
      left_join(implausible_combined, by = "SAMPLE_ALIQUOT") %>%
      mutate(
        exclude_reason = case_when(
          !is.na(exclude_reason) & !is.na(exclude_reason_implausible) ~
            paste(exclude_reason, exclude_reason_implausible, sep = "; "),
          !is.na(exclude_reason_implausible) ~ exclude_reason_implausible,
          TRUE ~ exclude_reason
        ),
        exclude = exclude | coalesce(exclude_implausible, FALSE)
      ) %>%
      select(-exclude_reason_implausible, -exclude_implausible)
  }

  # Mark unexpected race/ethnicity (for review, not auto-exclude)
  if (nrow(unexpected_race_eth) > 0) {
    race_eth_combined <- unexpected_race_eth %>%
      group_by(SAMPLE_ALIQUOT) %>%
      summarise(exclude_reason_race_eth = first(exclude_reason), .groups = "drop")

    exclusion_report <- exclusion_report %>%
      left_join(race_eth_combined, by = "SAMPLE_ALIQUOT") %>%
      mutate(
        exclude_reason = case_when(
          !is.na(exclude_reason) & !is.na(exclude_reason_race_eth) ~
            paste(exclude_reason, exclude_reason_race_eth, sep = "; "),
          !is.na(exclude_reason_race_eth) ~ exclude_reason_race_eth,
          TRUE ~ exclude_reason
        )
      ) %>%
      select(-exclude_reason_race_eth)
  }

  write_csv(exclusion_report, file.path(OUTPUT_DIR, "sample_exclusion_report.csv"))
}
```

# Step 4: Exclusion Summary {#step4}

```{r}
#| label: step4-exclusion-summary
#| echo: false
#| results: asis

exclusion_report <- read_csv(file.path(OUTPUT_DIR, "sample_exclusion_report.csv"), show_col_types = FALSE)

n_total <- nrow(exclusion_report)
n_exclude <- sum(exclusion_report$exclude)
n_keep <- sum(!exclusion_report$exclude)

cat("## Summary\n\n")

cat("| Metric | Count | Percent |\n")
cat("|--------|-------|--------|\n")
cat("| Total samples | ", n_total, " | 100% |\n", sep = "")
cat("| Samples to **EXCLUDE** | ", n_exclude, " | ", round(100*n_exclude/n_total, 1), "% |\n", sep = "")
cat("| Samples to **KEEP** | ", n_keep, " | ", round(100*n_keep/n_total, 1), "% |\n", sep = "")
cat("\n")

if (n_exclude > 0) {
  cat("## Exclusion Reasons\n\n")

  reasons <- exclusion_report %>%
    filter(exclude) %>%
    count(exclude_reason, sort = TRUE)

  kable(reasons, col.names = c("Reason", "Count"),
        caption = "Samples Excluded") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
    print()
}

# Show flagged but not excluded
review_items <- exclusion_report %>%
  filter(!exclude & !is.na(exclude_reason) & exclude_reason != "")

if (nrow(review_items) > 0) {
  cat("\n## Flagged for Review (Not Excluded)\n\n")

  review_reasons <- review_items %>% count(exclude_reason, sort = TRUE)

  kable(review_reasons, col.names = c("Reason", "Count"),
        caption = "Samples Flagged for Review") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
    print()
}
```

# Step 5: Apply Filters {#step5}

```{r}
#| label: step5-apply-filters
#| echo: false
#| results: asis

cat("## Applying Exclusion Filters\n\n")

# Get samples to exclude
samples_to_exclude <- exclusion_report %>%
  filter(exclude) %>%
  pull(SAMPLE_ALIQUOT)

cat("Excluding **", length(samples_to_exclude), "** samples from merged datasets...\n\n", sep = "")

# Filter datasets
filtered_standard <- merged_standard %>%
  filter(!SAMPLE_ALIQUOT %in% samples_to_exclude)

filtered_low <- merged_low %>%
  filter(!SAMPLE_ALIQUOT %in% samples_to_exclude)

filtered_combined <- merged_combined %>%
  filter(!SAMPLE_ALIQUOT %in% samples_to_exclude)

# Write filtered files
write_csv(filtered_standard, file.path(OUTPUT_DIR, "filtered", "filtered_standard_post_QC.csv"))
write_csv(filtered_low, file.path(OUTPUT_DIR, "filtered", "filtered_low_post_QC.csv"))
write_csv(filtered_combined, file.path(OUTPUT_DIR, "filtered", "filtered_combined_post_QC.csv"))

cat("| Dataset | Before | After | Excluded |\n")
cat("|---------|--------|-------|----------|\n")
cat("| Standard | ", nrow(merged_standard), " | ", nrow(filtered_standard), " | ", nrow(merged_standard) - nrow(filtered_standard), " |\n", sep = "")
cat("| Low | ", nrow(merged_low), " | ", nrow(filtered_low), " | ", nrow(merged_low) - nrow(filtered_low), " |\n", sep = "")
cat("| Combined | ", nrow(merged_combined), " | ", nrow(filtered_combined), " | ", nrow(merged_combined) - nrow(filtered_combined), " |\n", sep = "")
```

# Step 6: Covariate Groupings {#step6}

```{r}
#| label: step6-normalize
#| include: false

if (INCLUDE_DERIVED_COVARS) {
  # Normalize ancestry columns before rule matching:
  # 1. Trim whitespace — Excel/CSV round-trips can leave trailing spaces on
  #    multi-word values like "BL WH" or "WH MU", causing silent rule mismatches.
  # 2. Group NA → "NA" string — datasets without a Group assignment (e.g.
  #    OtherProjects) store actual NA here; the rules match on the string "NA"
  #    to classify those samples by Race + Ethnicity alone.
  filtered_combined <- filtered_combined %>%
    mutate(
      across(c(Group, Race, Ethnicity), ~str_trim(as.character(.))),
      Group = replace_na(Group, "NA")
    )
}
```

## Step 6a: Explore Distributions

```{r}
#| label: step6a-explore
#| echo: false
#| results: asis
#| message: false
#| warning: false

if (INCLUDE_DERIVED_COVARS) {
  cat("Review these distributions to inform your grouping decisions in Step 6b.\n\n")

  # NOTE: explore_*() print a summary to the console as a side effect; under results='asis'
  # that pollutes the report, so capture/suppress it and render via kable instead.
  cat("### CDX (Diagnosis)\n\n")
  invisible(capture.output(suppressMessages(
    .cdx_tab <- explore_covariate(filtered_combined, "CDX"))))
  print(kable_styling(kable(.cdx_tab, col.names = c("CDX", "N", "Pct")),
                      bootstrap_options = c("striped", "condensed"), full_width = FALSE))

  cat("\n### Group × Race × Ethnicity Combinations\n\n")
  invisible(capture.output(suppressMessages(
    .gre_tab <- explore_combinations(filtered_combined, "Group", "Race", "Ethnicity"))))
  print(kable_styling(kable(.gre_tab),
                      bootstrap_options = c("striped", "condensed"), full_width = FALSE))
} else {
  cat("*Skipped (include_derived_covariates = FALSE)*\n\n")
}
```

## Step 6b: Define Grouping Configs

\*\* Pedro is checking on "Dementia" and "Cognitively Impaired" from PER. For now I am calling them Other Dementia

```{r}
#| label: step6b-configs
#| message: false
#| warning: false

if (INCLUDE_DERIVED_COVARS) {

  # ── EDIT THIS BLOCK FOR EACH NEW DATASET ───────────────────────────────────

  REVIEW_DIR <- if (is.null(params$review_dir)) "review" else params$review_dir
  .cdx_csv <- file.path(REVIEW_DIR, "CDX_review.csv")
  .anc_csv <- file.path(REVIEW_DIR, "Ancestry_review.csv")
  if (file.exists(.cdx_csv) && file.exists(.anc_csv)) {
    # DATA-DRIVEN: decisions live in the review tables' FINAL_* columns (blank = exclude)
    .cdx_rev <- read_csv(.cdx_csv, show_col_types = FALSE) %>%
      mutate(FINAL_CDX_collapsed = na_if(trimws(FINAL_CDX_collapsed), ""))
    .anc_rev <- read_csv(.anc_csv, show_col_types = FALSE) %>%
      mutate(FINAL_Ancestry = na_if(trimws(FINAL_Ancestry), ""))
    .coll <- .cdx_rev %>% filter(!is.na(FINAL_CDX_collapsed), !is.na(CDX))
    dx_config <- make_category_config(
      source_col = "CDX", target_col = "CDX_collapsed", keep = character(0),
      collapse = split(.coll$CDX, .coll$FINAL_CDX_collapsed),
      exclude  = .cdx_rev %>% filter(is.na(FINAL_CDX_collapsed), !is.na(CDX)) %>% pull(CDX) %>% as.character()
    )
    .anc_keep <- .anc_rev %>% filter(!is.na(FINAL_Ancestry))
    .anc_rules <- list()
    for (.i in seq_len(nrow(.anc_keep))) {
      .cond <- list(Group = .anc_keep$Group[.i], Race = .anc_keep$Race[.i], Ethnicity = .anc_keep$Ethnicity[.i])
      .cond <- .cond[!vapply(.cond, function(v) is.na(v) || v == "", logical(1))]
      .anc_rules <- c(.anc_rules, setNames(list(.cond), .anc_keep$FINAL_Ancestry[.i]))
    }
    ancestry_config <- make_grouping_config(
      target_col = "Ancestry", other_label = "Other", exclude_other = FILTER_NA_COVARS, rules = .anc_rules
    )
    message("Step 6b: data-driven config from ", REVIEW_DIR)

  } else {
  # BUILT-IN DEFAULTS (used when no review CSVs are present) — edit per dataset:
  dx_config <- make_category_config(
    source_col = "CDX",
    target_col = "CDX_collapsed",
    keep     = c("MCI"),
    collapse = list(
      "AD" = c("AD", "Alzheimers Disease"),
      "NCI" = c("NCI", "Non - Cognitively Impaired"),
      "Dementia_Other" = c("Dementia Undetermined","Dementia – Undetermined", "Dementia Vascular", "Dementia – Vascular", "Dementia Other", "Cognitively Impaired", "Dementia")
    ),
    exclude  = c("0", "Other", "Insufficient Data", "NA")
  )

  ancestry_config <- make_grouping_config(
    target_col    = "Ancestry",
    other_label   = "Other",
    exclude_other = FILTER_NA_COVARS,
    rules = list(
      "NH_WH" = list(Group = "NA", Race = "WH", Ethnicity= "NH"),
      "HI_WH" = list(Group = "HI", Race = "WH"),
      "HI_WH" = list(Group = "NA", Race = "WH", Ethnicity= "HI"),
      "HI_MU" = list(Group = "HI", Race = "MU"),
      "HI_MU" = list(Race = "BL WH", Ethnicity= "HI"),
      "HI_MU" = list(Race = "BL  WH", Ethnicity= "HI"),
      "HI_MU" = list(Race = "WH MU", Ethnicity= "HI"),
      "HI_MU" = list(Group = "NA", Race = "MU", Ethnicity= "HI"),
      "HI_MU" = list(Group = "NA", Race = "WH  MU", Ethnicity= "HI"),
      "HI_BL" = list(Group = "HI", Race = "BL"),
      "HI_BL" = list(Group = "AA", Ethnicity = "HI"),
      "HI_BL" = list(Group = "NA", Race = "BL", Ethnicity= "HI"),
      "AA"    = list(Group = "AA", Ethnicity = "NH", Race = "BL"),
      "AA"    = list(Group = "AA", Ethnicity = "NH", Race = "MU"),
      "AA"    = list(Group = "NA", Race = "BL", Ethnicity= "NH"),
      "AA"    = list(Group = "AA", Ethnicity = "NH", Race = "WH"),
      "AFDC"  = list(Group = "AFDC")
    )
  )
  }  # end built-in-defaults fallback (the data-driven branch sets the configs above)

  # ───────────────────────────────────────────────────────────────────────────

  invisible(capture.output(suppressMessages({
    print_config(dx_config)
    print_config(ancestry_config)
  })))
}
```

## Step 6c: Preview Decisions

```{r}
#| label: step6c-preview
#| echo: false
#| results: asis
#| message: false
#| warning: false

if (INCLUDE_DERIVED_COVARS) {
  cat("### Diagnosis (CDX → CDX_collapsed)\n\n")
  dx_summary <- summarize_category_decisions(filtered_combined, dx_config)
  kable(dx_summary$all %>% arrange(maps_to, desc(n)), caption = "CDX decision summary (sorted by maps_to)") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
    print()

  cat(sprintf("\n*Samples retained: %d of %d*\n\n",
              dx_summary$total_after, dx_summary$total_before))

  cat("### Ancestry (Group × Race × Ethnicity → Ancestry)\n\n")
  anc_summary <- summarize_grouping_decisions(filtered_combined, ancestry_config)
  kable(anc_summary$group_summary, caption = "Ancestry group summary") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
    print()

  # Explicit combo -> Ancestry mapping (parallel to the CDX raw -> collapsed table)
  cat("\n**Ancestry mapping — Group × Race × Ethnicity → Ancestry**\n\n")
  .anc_map_csv <- file.path(if (exists("REVIEW_DIR")) REVIEW_DIR else "review", "Ancestry_review.csv")
  if (file.exists(.anc_map_csv)) {
    print(kable_styling(kable(
      suppressMessages(read_csv(.anc_map_csv, show_col_types = FALSE)) %>%
        select(any_of(c("Group","Race","Ethnicity","n_in_batch","FINAL_Ancestry","unexpected_race_eth"))) %>%
        arrange(FINAL_Ancestry, desc(n_in_batch)),
      caption = "Ancestry mapping (source: Ancestry_review.csv FINAL_Ancestry; blank = excluded)"),
      bootstrap_options = c("striped", "condensed"), full_width = FALSE))
  } else {
    # No Ancestry_review.csv: preview the built-in config's mapping. `Ancestry` isn't applied to
    # filtered_combined until Step 6d, so apply it to a throwaway copy here just for the preview count.
    .anc_preview <- apply_grouping_config(filtered_combined, ancestry_config)
    print(kable_styling(kable(
      .anc_preview %>% count(Group, Race, Ethnicity, Ancestry) %>% arrange(Ancestry, desc(n)),
      caption = "Ancestry mapping (applied: Group × Race × Ethnicity → Ancestry)"),
      bootstrap_options = c("striped", "condensed"), full_width = FALSE))
  }

  if (nrow(anc_summary$unmatched) > 0) {
    cat("\n::: {.alert .alert-warning}\n")
    cat("**Unmatched rows — will be excluded**\n\n")
    kable(anc_summary$unmatched) %>%
      kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
      print()
    cat(":::\n\n")
  } else {
    cat("\n*All rows matched an ancestry rule.* ✓\n\n")
  }
}
```

## Step 6d: Apply, Validate, and Save

```{r}
#| label: step6d-apply
#| echo: false
#| results: asis
#| message: false
#| warning: false

if (INCLUDE_DERIVED_COVARS) {
  n_before <- nrow(filtered_combined)

  # Apply CDX grouping (drops excluded and unaccounted rows internally)
  filtered_combined <- apply_category_config(filtered_combined, dx_config)

  # Apply ancestry grouping (drops unmatched rows when exclude_other = FILTER_NA_COVARS)
  filtered_combined <- apply_grouping_config(filtered_combined, ancestry_config)

  # Explicit NA safety-net filter
  if (FILTER_NA_COVARS) {
    filtered_combined <- filtered_combined %>%
      filter(!is.na(CDX_collapsed), !is.na(Ancestry))
  }

  n_after <- nrow(filtered_combined)
  cat(sprintf("**Rows before grouping:** %d | **After:** %d | **Removed:** %d\n\n",
              n_before, n_after, n_before - n_after))

  # ---- Join WGS APOE genotypes ----
  # WGS is read from the auto-detected metadata. APOE.geno_final DEFAULTS TO WGS where available
  # (Sanger APOE.geno is used only when WGS is missing) — so on a Sanger-vs-WGS disagreement, WGS
  # wins. This general pipeline applies NO sample-specific plate-swap correction; the AD-Hispanic
  # WGS plate-swap disagreements are surfaced in the `error_APOE_integrity` report (step 5) so the
  # swap stays recognizable, and the sample-specific re-map lives with the affected batch
  # (e.g. QC_Runs/<run>/WGS_Midi_Remap/), not here.
  wgs_src <- read_csv(metadata_file, show_col_types = FALSE)
  if ("APOE_WGS" %in% names(wgs_src)) {
    wgs_lookup <- wgs_src %>%
      select(SAMPLE_ALIQUOT, APOE_WGS) %>%
      filter(!is.na(APOE_WGS)) %>%
      distinct() %>%
      mutate(
        APOE_WGS_norm = if_else(
          !is.na(APOE_WGS) & nchar(trimws(APOE_WGS)) == 2,
          paste0(substr(trimws(APOE_WGS), 1, 1), "/", substr(trimws(APOE_WGS), 2, 2)),
          NA_character_
        )
      )
  } else {
    wgs_lookup <- tibble(SAMPLE_ALIQUOT = character(), APOE_WGS_norm = character())
  }

  # Drop any APOE_WGS column already present so the join doesn't create .x/.y duplicates.
  filtered_combined <- filtered_combined %>%
    select(-any_of("APOE_WGS")) %>%
    left_join(wgs_lookup, by = "SAMPLE_ALIQUOT")

  # (No plate-swap masking here — WGS is the default; see header note. Disagreements are reported
  #  by error_APOE_integrity; the sample-specific re-map is run in the affected batch's folder.)

  # APOE4 carrier variables — derived from WGS-preferred genotype where available
  filtered_combined <- filtered_combined %>%
    mutate(
      # Flag samples where WGS genotype differs from original (both non-NA)
      APOE_WGS_changed = !is.na(APOE.geno) & !is.na(APOE_WGS_norm) &
                         (APOE.geno != APOE_WGS_norm),
      # Best available genotype: WGS takes priority, fall back to original
      APOE.geno_final  = coalesce(APOE_WGS_norm, APOE.geno),
      APOE4_count      = case_when(
        grepl("4.*4", APOE.geno_final) ~ 2L,
        grepl("4",    APOE.geno_final) ~ 1L,
        !is.na(APOE.geno_final)        ~ 0L,
        TRUE                           ~ NA_integer_
      ),
      # Derived from APOE4_count so NA propagates consistently
      APOE4_carrier    = APOE4_count >= 1L
    )

  cat(sprintf(
    "**WGS genotypes joined:** %d samples have WGS data | %d genotypes changed by WGS | %d samples gained a genotype via WGS\n\n",
    sum(!is.na(filtered_combined$APOE_WGS_norm)),
    sum(filtered_combined$APOE_WGS_changed, na.rm = TRUE),
    sum(is.na(filtered_combined$APOE.geno) & !is.na(filtered_combined$APOE_WGS_norm))
  ))

  cat("### Final Group Sizes (CDX_collapsed × Ancestry)\n\n")
  # validate_groups() prints to the console (mangled under asis); capture it and render a clean cross-tab.
  invisible(capture.output(suppressMessages(
    validate_groups(filtered_combined, c("CDX_collapsed", "Ancestry"), min_n = 5))))
  print(kable_styling(
    kable(filtered_combined %>%
            count(CDX_collapsed, Ancestry) %>%
            tidyr::pivot_wider(names_from = Ancestry, values_from = n, values_fill = 0),
          caption = "Final group sizes: CDX_collapsed × Ancestry"),
    bootstrap_options = c("striped", "condensed"), full_width = FALSE))

  write_csv(filtered_combined, file.path(OUTPUT_DIR, "filtered", "filtered_combined_post_QC.csv"))
  cat("\n`filtered_combined_post_QC.csv` updated with CDX_collapsed, Ancestry, APOE_WGS, APOE_WGS_norm, APOE_WGS_changed, APOE.geno_final, APOE4_carrier, APOE4_count.\n\n")

  # Sync standard and low CSVs to the final Step-6d sample set.
  # filtered_combined has been further filtered by ancestry/CDX grouping
  # (apply_grouping_config + FILTER_NA_COVARS). Standard and low were written
  # at Step 5 before this filtering; re-filter them now so all three files
  # share the same sample universe.
  passing_ids       <- filtered_combined$SAMPLE_ALIQUOT
  filtered_standard <- filtered_standard %>% filter(SAMPLE_ALIQUOT %in% passing_ids)
  filtered_low      <- filtered_low      %>% filter(SAMPLE_ALIQUOT %in% passing_ids)
  write_csv(filtered_standard, file.path(OUTPUT_DIR, "filtered", "filtered_standard_post_QC.csv"))
  write_csv(filtered_low,      file.path(OUTPUT_DIR, "filtered", "filtered_low_post_QC.csv"))
  cat(sprintf(
    "`filtered_standard_post_QC.csv` and `filtered_low_post_QC.csv` re-synced to Step-6d sample set: standard %d rows, low %d rows.\n\n",
    nrow(filtered_standard), nrow(filtered_low)
  ))

} else {
  cat("*Skipped (include_derived_covariates = FALSE)*\n\n")
}
```

# Step 7: Add Derived Biomarkers {#step7}

```{r}
#| label: step7-derived-biomarkers
#| echo: false
#| results: asis

if (INCLUDE_DERIVED_BIOMARKS) {
  cat("## Creating Derived Biomarker Columns\n\n")

  # Check that required columns exist
  required_cols <- c("APOE", "APOE4", "A_40", "A_42")
  missing_combined <- required_cols[!required_cols %in% names(filtered_combined)]

  if (length(missing_combined) > 0) {
    cat("::: {.alert .alert-warning}\n")
    cat("**Missing Source Columns**\n\n")
    cat("Missing in filtered_combined: ", paste(missing_combined, collapse = ", "), "\n")
    cat(":::\n\n")
  } else {
    # Add derived columns to filtered_combined
    filtered_combined <- filtered_combined %>%
      mutate(
        APOE4_minus_APOE = APOE4 - APOE,
        ABeta42_minus_ABeta40 = A_42 - A_40
      )

    cat("Added two derived biomarker columns:\n\n")
    cat("| Derived Biomarker | Formula |\n")
    cat("|-------------------|--------|\n")
    cat("| `APOE4_minus_APOE` | APOE4 - APOE |\n")
    cat("| `ABeta42_minus_ABeta40` | A_42 - A_40 |\n\n")

    # Overwrite filtered files with new columns
    write_csv(filtered_combined, file.path(OUTPUT_DIR, "filtered", "filtered_combined_post_QC.csv"))

    cat("Filtered files updated with derived biomarkers.\n\n")

    # Summary statistics for derived biomarkers
    cat("### Summary Statistics\n\n")

    apoe_stats <- filtered_combined %>%
      summarise(
        N = sum(!is.na(APOE4_minus_APOE)),
        Mean = round(mean(APOE4_minus_APOE, na.rm = TRUE), 3),
        SD = round(sd(APOE4_minus_APOE, na.rm = TRUE), 3),
        Min = round(min(APOE4_minus_APOE, na.rm = TRUE), 3),
        Median = round(median(APOE4_minus_APOE, na.rm = TRUE), 3),
        Max = round(max(APOE4_minus_APOE, na.rm = TRUE), 3)
      ) %>%
      mutate(Biomarker = "APOE4_minus_APOE", .before = 1)

    abeta_stats <- filtered_combined %>%
      summarise(
        N = sum(!is.na(ABeta42_minus_ABeta40)),
        Mean = round(mean(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
        SD = round(sd(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
        Min = round(min(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
        Median = round(median(ABeta42_minus_ABeta40, na.rm = TRUE), 3),
        Max = round(max(ABeta42_minus_ABeta40, na.rm = TRUE), 3)
      ) %>%
      mutate(Biomarker = "ABeta42_minus_ABeta40", .before = 1)

    derived_stats <- bind_rows(apoe_stats, abeta_stats)

    kable(derived_stats, caption = "Derived Biomarker Summary Statistics") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      print()
  }
} else {
  cat("*Skipped (include_derived_biomarkers = FALSE)*\n\n")
}
```

# Step 8: Longitudinal Handling — Cross-Sectional vs Longitudinal Datasets {#step8}

Same-sample **reruns** (same `SAMPLE` & metadata, different biomarker `RUN` — *not* the study's designed technical replicates) are already resolved in Step 2 (most recent `RUN` kept). This step handles **longitudinal repeats** — the same subject (`Record_ID`) sampled at different visits (different `SAMPLE`).

We add per-subject flags (`n_visits`, `is_repeat_subject`, `subject_visit_n` \[1 = earliest by age\], `is_most_recent`) and then write **two datasets**, so the analyst always selects the right one:

-   **`filtered_combined_post_QC.csv` — DEFAULT (cross-sectional):** one row per subject, the **most recent** visit. Use for all cross-sectional analyses (no double-counting).
-   **`filtered_combined_longitudinal.csv` — longitudinal-only:** *every* visit of subjects with ≥2 visits. Use for trajectory / conversion analyses.

`preprocessed_data.rds` is built from the cross-sectional default set.

```{r}
#| label: step8-longitudinal-datasets
#| echo: false
#| results: asis

flagged <- filtered_combined %>%
  group_by(Record_ID) %>%
  mutate(
    n_visits          = dplyr::n(),
    is_repeat_subject = n_visits > 1L,
    subject_visit_n   = rank(age_at_subject, ties.method = "first"),  # 1 = earliest (age axis)
    is_most_recent    = subject_visit_n == n_visits
  ) %>%
  ungroup()

# Two datasets
filtered_cross <- flagged %>% filter(is_most_recent)                                  # one row/subject (most recent)
filtered_long  <- flagged %>% filter(is_repeat_subject) %>% arrange(Record_ID, subject_visit_n)  # all visits, ≥2-visit subjects

write_csv(filtered_cross, file.path(OUTPUT_DIR, "filtered", "filtered_combined_post_QC.csv"))
write_csv(filtered_long,  file.path(OUTPUT_DIR, "filtered", "filtered_combined_longitudinal.csv"))

n_rep_subj <- dplyr::n_distinct(filtered_long$Record_ID)
cat(sprintf(paste0("- All post-QC samples: **%d**\n",
                   "- **Cross-sectional (DEFAULT, one-per-subject):** **%d** → `filtered_combined_post_QC.csv`\n",
                   "- **Longitudinal-only (≥2 visits, all visits):** **%d** rows / **%d** subjects → `filtered_combined_longitudinal.csv`\n\n"),
            nrow(flagged), nrow(filtered_cross), nrow(filtered_long), n_rep_subj))

# DEFAULT downstream object (drives preprocessed_data.rds) = cross-sectional set
filtered_combined <- filtered_cross

if (n_rep_subj > 0) {
  # Use the collapsed diagnosis when covariates were derived (Step 6); else fall back to raw CDX
  # (so this step also runs when include_derived_covariates = FALSE).
  .cdx_col <- if ("CDX_collapsed" %in% names(filtered_long)) "CDX_collapsed" else "CDX"
  repeat_tbl <- filtered_long %>%
    group_by(Record_ID) %>%
    summarise(
      n_visits = dplyr::first(n_visits),
      ages = paste(age_at_subject[order(subject_visit_n)], collapse = " / "),
      CDX  = paste(.data[[.cdx_col]][order(subject_visit_n)], collapse = " / "),
      cdx_changed = dplyr::n_distinct(.data[[.cdx_col]]) > 1,   # diagnosis differs across visits
      RUNs = paste(RUN[order(subject_visit_n)], collapse = " / "),
      .groups = "drop"
    ) %>% arrange(desc(cdx_changed), Record_ID)

  kable(repeat_tbl,
        caption = "Longitudinal subjects (in `filtered_combined_longitudinal.csv`); `cdx_changed` = diagnosis differs across visits") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
    print()

  cat(sprintf("\n%d of %d longitudinal subject(s) have a diagnosis change across visits.\n\n",
              sum(repeat_tbl$cdx_changed), nrow(repeat_tbl)))
}

# ── Metadata-level (cross-batch) longitudinal cross-check ──────────────────────
# Detection is METADATA-driven (same Record_ID on >1 SAMPLE across ALL batches), so a subject is
# flagged even if their visits' biomarkers are in different batches. The cross-sectional/longitudinal
# SPLIT above stays biomarker-based (only samples with biomarkers can populate a dataset).
if (!is.null(meta_long) || (HAVE_KNOWN_REPEATS && length(known_long_records) > 0)) {
  .REP <- file.path(OUTPUT_DIR, "reports"); dir.create(.REP, showWarnings = FALSE, recursive = TRUE)
  .biom_aliq   <- as.character(flagged$SAMPLE_ALIQUOT)   # all samples with biomarkers in this merge
  in_data_long <- unique(filtered_long$Record_ID)        # subjects with >=2 visits WITH biomarkers here

  meta_long_report <- if (!is.null(meta_long)) {
    meta_long %>% group_by(Record_ID) %>%
      summarise(n_samples_meta = dplyr::n_distinct(SAMPLE),
                SAMPLEs_meta    = paste(sort(unique(SAMPLE)), collapse = " / "),
                RUNs_meta       = paste(RUN[order(RUN)], collapse = " / "),
                n_visits_with_biomarkers_here = sum(SAMPLE_ALIQUOT %in% .biom_aliq),
                .groups = "drop")
  } else tibble(Record_ID = character())

  kn <- if (HAVE_KNOWN_REPEATS && !is.null(known_long_df)) {
    known_long_df %>% group_by(Record_ID) %>%
      summarise(known_n_draws  = dplyr::n(),
                known_SAMPLEs  = paste(unique(SAMPLE), collapse = " / "),
                known_RUNs     = paste(RUN, collapse = " / "),
                known_ResultQC = if ("ResultQC" %in% names(known_long_df)) paste(ResultQC, collapse = " / ") else NA_character_,
                .groups = "drop")
  } else tibble(Record_ID = character())

  det <- if (n_rep_subj > 0) {
    repeat_tbl %>% transmute(Record_ID, n_visits_in_data = n_visits, ages_in_data = ages,
                             CDX_in_data = CDX, cdx_changed, RUNs_in_data = RUNs)
  } else tibble(Record_ID = character())

  recon_long <- tibble(Record_ID = Reduce(union, list(meta_long_records, known_long_records, in_data_long))) %>%
    left_join(meta_long_report, by = "Record_ID") %>%
    left_join(kn,  by = "Record_ID") %>%
    left_join(det, by = "Record_ID") %>%
    mutate(
      in_metadata      = Record_ID %in% meta_long_records,
      in_known_list    = Record_ID %in% known_long_records,
      detected_in_data = Record_ID %in% in_data_long,
      status = case_when(
        in_metadata &  in_known_list ~ "concordant (metadata + lab list)",
        in_metadata & !in_known_list ~ "in metadata, not on lab list (verify)",
        !in_metadata & in_known_list ~ "on lab list, not in this metadata",
        TRUE                         ~ "detected in data only"
      )
    ) %>%
    arrange(status, Record_ID)

  .ldir <- file.path(.REP, "duplicates"); dir.create(.ldir, showWarnings = FALSE, recursive = TRUE)
  if (requireNamespace("writexl", quietly = TRUE)) writexl::write_xlsx(recon_long, file.path(.ldir, "Longitudinal_review.xlsx")) else write_csv(recon_long, file.path(.ldir, "Longitudinal_review.csv"), na = "")

  cat("\n**Longitudinal cross-check (metadata-level, all batches):** ",
      length(meta_long_records), " subject(s) in metadata; ",
      sum(recon_long$status == "concordant (metadata + lab list)"), " concordant with the lab list ✓; ",
      sum(recon_long$status == "in metadata, not on lab list (verify)"), " in metadata not on list; ",
      sum(recon_long$detected_in_data), " have ≥2 visits with biomarkers in *this* merge.\n",
      "Full reconciliation → `reports/duplicates/Longitudinal_review.xlsx`.\n\n", sep = "")
}
```

# Save preprocessed_data.rds {#save-rds}

```{r}
#| label: save-rds
#| include: false

# Biomarker cols = everything from NPQ files (not in original metadata, not pipeline-derived,
#   and not NPQ run-level tracking columns)
pipeline_derived_cols <- c("CDX_collapsed", "Ancestry",
                           "APOE_WGS", "APOE_WGS_norm", "APOE_WGS_changed",
                           "APOE.geno_final", "APOE4_carrier", "APOE4_count",
                           "APOE4_minus_APOE", "ABeta42_minus_ABeta40",
                           "n_visits", "is_repeat_subject", "subject_visit_n", "is_most_recent")
npq_run_cols <- c("___1", "Run", "Bay")   # NPQ instrument metadata, not protein measurements
biomarker_cols <- setdiff(
  names(filtered_combined),
  c(metadata_cols_captured, pipeline_derived_cols, npq_run_cols)
)

saveRDS(
  list(
    data           = filtered_combined,
    biomarker_cols = biomarker_cols
  ),
  file.path(OUTPUT_DIR, "preprocessed_data.rds")
)

message(sprintf("Saved preprocessed_data.rds: %d samples, %d biomarker cols",
                nrow(filtered_combined), length(biomarker_cols)))
```

# QC / Error Reports & Mapping Keys {#reports}

Per-issue error reports (with metadata context, for sending to the contributing groups) and the
grouping decisions as shareable keys. CSVs go to `<output_dir>/reports/`. Checks run on the full
batch present in metadata, so issues are surfaced even for samples that are later excluded.

```{r}
#| label: step-qc-reports
#| echo: false
#| results: asis
#| message: false
#| warning: false

if (INCLUDE_DERIVED_COVARS && exists("merged_combined")) {
  REPORTS_DIR <- file.path(OUTPUT_DIR, "reports")
  dir.create(REPORTS_DIR, showWarnings = FALSE, recursive = TRUE)
  .review_dir <- if (exists("REVIEW_DIR")) REVIEW_DIR else if (!is.null(params$review_dir)) params$review_dir else "review"

  # Robustness: normalize the `metatada_source` typo (older metadata) and guarantee the provenance/
  # identifier columns the reports carry (`metadata_source`, `ID2`) are present even if a file lacks them.
  if ("metatada_source" %in% names(merged_combined) && !"metadata_source" %in% names(merged_combined))
    merged_combined <- dplyr::rename(merged_combined, metadata_source = metatada_source)
  for (.c in c("metadata_source", "ID2")) if (!.c %in% names(merged_combined)) merged_combined[[.c]] <- NA_character_

  .miss <- function(x) is.na(x) | trimws(as.character(x)) %in% c("", "Missing", "NA", "Unknown")
  .ctx  <- intersect(c("metadata_source","SAMPLE_ALIQUOT","SAMPLE","Record_ID","ID2","Site","Group",
                       "Race","Ethnicity","sex","age_at_subject","CDX","Case_Control","APOE.geno","APOE_WGS"),
                     names(merged_combined))
  .valid_apoe <- c("2/2","2/3","2/4","3/3","3/4","4/4")
  .written <- character(0)
  .all_reports <- list()                          # collected for the single combined workbook
  .xlsx_ok <- requireNamespace("writexl", quietly = TRUE)
  # Reports are grouped into sub-folders of reports/ via `subdir` (errors/, missing_data/, duplicates/);
  # overviews + mapping keys stay at the reports/ top level (subdir = "").
  wr <- function(df, f, combine = TRUE, subdir = "") {
    base <- sub("\\.csv$", "", f)
    .dir <- file.path(REPORTS_DIR, subdir)
    if (nzchar(subdir)) dir.create(.dir, showWarnings = FALSE, recursive = TRUE)
    rel  <- if (nzchar(subdir)) file.path(subdir, base) else base
    # Reports are written as .xlsx so Excel keeps APOE genotypes ("3/4") as text (no date
    # conversion). Fall back to .csv only if the writexl package is unavailable.
    if (.xlsx_ok) {
      writexl::write_xlsx(df, file.path(.dir, paste0(base, ".xlsx")))
      .written <<- c(.written, sprintf("`%s.xlsx` — %d row(s)", rel, nrow(df)))
    } else {
      readr::write_csv(df, file.path(.dir, f), na = "")
      .written <<- c(.written, sprintf("`%s` (install writexl for .xlsx) — %d row(s)", rel, nrow(df)))
    }
    if (combine) .all_reports[[base]] <<- df      # collect error/issue reports for the combined file
  }

  # 1. Mapping keys (shareable) -----------------------------------------------
  if (exists("dx_config")) {
    .cdx_lut <- character(0)
    for (k in dx_config$keep) .cdx_lut[k] <- k
    for (tg in names(dx_config$collapse)) for (s in dx_config$collapse[[tg]]) .cdx_lut[s] <- tg
    wr(merged_combined %>% count(CDX, name = "n_in_batch") %>%
         mutate(CDX_collapsed = unname(.cdx_lut[CDX]),
                status = if_else(is.na(CDX_collapsed), "EXCLUDED", "kept")) %>%
         arrange(is.na(CDX_collapsed), desc(n_in_batch)),
       "mapping_key_CDX.csv", combine = FALSE)
  }
  if (file.exists(file.path(.review_dir, "Ancestry_review.csv"))) {
    wr(suppressMessages(read_csv(file.path(.review_dir, "Ancestry_review.csv"), show_col_types = FALSE)) %>%
         select(any_of(c("Group","Race","Ethnicity","n_in_batch","FINAL_Ancestry","unexpected_race_eth"))),
       "mapping_key_Ancestry.csv", combine = FALSE)
  } else if ("Ancestry" %in% names(merged_combined) ||
             (exists("ancestry_config") && exists("apply_grouping_config"))) {
    # No Ancestry_review.csv: write the built-in config's mapping anyway, so the shareable key always
    # exists (parallel to mapping_key_CDX, which writes whenever dx_config exists). Mirrors the display
    # branch (~L1106). `Ancestry` is already applied to merged_combined by Step 6d in normal runs;
    # apply_grouping_config() is a defensive fallback if it isn't.
    .anc_src <- if ("Ancestry" %in% names(merged_combined)) merged_combined
                else apply_grouping_config(merged_combined, ancestry_config)
    wr(.anc_src %>% count(Group, Race, Ethnicity, Ancestry, name = "n_in_batch") %>%
         arrange(Ancestry, desc(n_in_batch)),
       "mapping_key_Ancestry.csv", combine = FALSE)
  }

  # 2. Implausible values ------------------------------------------------------
  plausible <- list(age_at_subject=c(18,110), BMI=c(10,80), weight_lb=c(50,500), height_inches=c(36,90))
  .impl <- list()
  for (v in names(plausible)) if (v %in% names(merged_combined)) {
    x <- suppressWarnings(as.numeric(merged_combined[[v]]))
    hit <- which(!is.na(x) & (x < plausible[[v]][1] | x > plausible[[v]][2]))
    if (length(hit)) .impl[[v]] <- merged_combined[hit, .ctx] %>%
      mutate(variable = v, value = x[hit],
             expected_range = paste0("[", plausible[[v]][1], ", ", plausible[[v]][2], "]"))
  }
  wr(if (length(.impl)) bind_rows(.impl) else merged_combined[0, .ctx], "error_implausible_values.csv", subdir = "errors")

  # 3. Unexpected Race/Ethnicity ----------------------------------------------
  g <- merged_combined$Group_Race_Ethnicity
  unexp_idx <- which(
    (grepl("BL",g)&grepl("WH",g)) | (grepl("HI",g)&grepl("NH",g)) | (grepl("AA",g)&grepl("WH",g)) |
    (grepl("AFDC",g)&grepl("WH",g)) | (grepl("AFDC",g)&grepl("HI",g)) | (grepl("AA",g)&grepl("HI",g)))
  wr(merged_combined[unexp_idx, c(.ctx, "Group_Race_Ethnicity")] %>% distinct() %>%
       mutate(Suggested_Coding = ""),   # blank — reviewer's suggested Group_Race_Ethnicity recoding
     "error_unexpected_race_ethnicity.csv", subdir = "errors")

  # 4. Missing data ------------------------------------------------------------
  md <- merged_combined %>% transmute(metadata_source, SAMPLE_ALIQUOT, SAMPLE, Record_ID, ID2, Site,
    miss_CDX=.miss(CDX), miss_Group=.miss(Group), miss_Race=.miss(Race), miss_Ethnicity=.miss(Ethnicity),
    miss_sex=.miss(sex), miss_age=.miss(age_at_subject),
    miss_APOE_Sanger=.miss(APOE.geno), miss_APOE_WGS=.miss(APOE_WGS),
    miss_APOE_any=.miss(APOE.geno) & .miss(APOE_WGS))
  # by-sample report flags CRITICAL missingness only (WGS-missing is the norm, kept as a column).
  # APOE "no call from either source" is intentionally NOT critical here, to avoid double-counting —
  # it is reported separately as the by-sample APOE_no_genotype flag (step 8). The summary below
  # still tallies every field incl. APOE Sanger/WGS coverage and APOE_any.
  .crit <- c("miss_CDX","miss_Race","miss_Ethnicity","miss_sex","miss_age")
  wr(md %>% filter(if_any(all_of(.crit), ~ .x)), "missing_data_by_sample.csv", subdir = "missing_data")
  wr(md %>% summarise(across(starts_with("miss_"), sum)) %>%
       pivot_longer(everything(), names_to="field", values_to="n_missing") %>%
       mutate(field=sub("^miss_","",field), pct_missing=round(100*n_missing/nrow(md),1)) %>%
       arrange(desc(n_missing)), "missing_data_summary.csv", subdir = "missing_data")

  # 5. APOE integrity (concordance) — requires TWO genotypes: flag a sample only
  #    when it has BOTH a valid Sanger AND a valid WGS call that DISAGREE. Samples
  #    with one call (or none) cannot be integrity-checked and are not flagged
  #    here. (Discordance is the contested WGS plate-swap signal, also tracked in
  #    Secondary_QC/APOE.) `no_apoe_geno` (no usable call from either source) is
  #    derived here too; it feeds the by-sample summary's APOE error (step 8).
  .nrm <- function(x){ x<-trimws(as.character(x)); ifelse(!is.na(x)&nchar(x)==2, paste0(substr(x,1,1),"/",substr(x,2,2)), x) }
  ap <- merged_combined %>% mutate(
    wgs_norm     = .nrm(APOE_WGS), sanger = as.character(APOE.geno),
    sanger_ok    = !.miss(sanger)   & sanger   %in% .valid_apoe,
    wgs_ok       = !.miss(wgs_norm) & wgs_norm %in% .valid_apoe,
    discordant   = sanger_ok & wgs_ok & sanger != wgs_norm,   # both present & valid, disagree
    no_apoe_geno = !sanger_ok & !wgs_ok)                      # no usable call from either source
  wr(ap %>% filter(discordant) %>%
       select(all_of(.ctx), wgs_norm, discordant),
     "error_APOE_integrity.csv", subdir = "errors")
  # No APOE call from either source (full per-sample detail) — feeds the by-sample
  # APOE error (step 8) and gets its own sheet in the combined workbook.
  wr(ap %>% filter(no_apoe_geno) %>%
       select(all_of(.ctx), wgs_norm, sanger_ok, wgs_ok, no_apoe_geno),
     "error_APOE_no_genotype.csv", subdir = "errors")

  # 6. Clinical consistency ----------------------------------------------------
  caselike <- function(x) grepl("AD|Case|Dementia|Lewy|Cog Imp", x, ignore.case=TRUE)
  ctrllike <- function(x) grepl("NCI|Control", x, ignore.case=TRUE)
  cl <- merged_combined %>% mutate(
    .age=suppressWarnings(as.numeric(age_at_subject)), .aoo=suppressWarnings(as.numeric(AOO)),
    aoo_gt_age = !is.na(.aoo) & !is.na(.age) & .aoo > .age,
    cdx_casecontrol_conflict = (caselike(CDX)&ctrllike(Case_Control)) | (ctrllike(CDX)&caselike(Case_Control)))
  wr(cl %>% filter(aoo_gt_age | (cdx_casecontrol_conflict %in% TRUE)) %>%
       select(all_of(.ctx), AOO, Years_Onset, aoo_gt_age, cdx_casecontrol_conflict),
     "error_clinical_consistency.csv", subdir = "errors")
  wr(merged_combined %>% count(CDX, Case_Control) %>% arrange(desc(n)), "check_CDX_vs_CaseControl_crosstab.csv", combine = FALSE)

  # 7. Categorical integrity ---------------------------------------------------
  wr(merged_combined %>% mutate(
       bad_sex = !.miss(sex) & !(sex %in% c("M","F")),
       bad_ethnicity = !.miss(Ethnicity) & !(Ethnicity %in% c("HI","NH")),
       bad_race = !.miss(Race) & !grepl(" ", Race) & !(Race %in% c("BL","WH","MU","AI"))) %>%
       filter(bad_sex|bad_ethnicity|bad_race) %>%
       select(all_of(.ctx), bad_sex, bad_ethnicity, bad_race),
     "error_categorical_unexpected_levels.csv", subdir = "errors")
  wr(merged_combined %>% group_by(SAMPLE_ALIQUOT) %>% filter(n()>1) %>% ungroup() %>% select(all_of(.ctx)),
     "error_duplicate_SAMPLE_ALIQUOT.csv", subdir = "errors")

  # 8. Consolidated QC error report — BY SAMPLE + SUMMARY -----------------------
  # One row per flagged sample (which checks tripped, n_issues, issues list) and a per-check summary.
  # Derived from the per-issue reports collected above (each carries SAMPLE_ALIQUOT).
  # APOE error counted here = "no APOE call from either source" (error_APOE_no_genotype,
  # the 36). The integrity/discordance report (error_APOE_integrity) is NOT counted in
  # the by-sample summary — it needs two genotypes and is tracked separately.
  .err_checks <- c(
    implausible_value         = "error_implausible_values",
    unexpected_race_ethnicity = "error_unexpected_race_ethnicity",
    missing_critical_field    = "missing_data_by_sample",
    APOE_no_genotype          = "error_APOE_no_genotype",
    clinical_inconsistency    = "error_clinical_consistency",
    categorical_bad_level     = "error_categorical_unexpected_levels",
    duplicate_SAMPLE_ALIQUOT  = "error_duplicate_SAMPLE_ALIQUOT")
  .flag_ids <- lapply(.err_checks, function(rep) {
    d <- .all_reports[[rep]]
    if (!is.null(d) && "SAMPLE_ALIQUOT" %in% names(d)) unique(as.character(d$SAMPLE_ALIQUOT)) else character(0)
  })
  by_sample <- merged_combined %>% select(all_of(.ctx)) %>% distinct() %>%
    mutate(SAMPLE_ALIQUOT = as.character(SAMPLE_ALIQUOT))
  .flagcols <- names(.flag_ids)
  for (fl in .flagcols) by_sample[[fl]] <- by_sample$SAMPLE_ALIQUOT %in% .flag_ids[[fl]]
  .flagmat <- as.matrix(by_sample[.flagcols])
  by_sample$n_issues <- as.integer(rowSums(.flagmat))
  by_sample$issues   <- apply(.flagmat, 1, function(r) paste(.flagcols[r], collapse = "; "))
  by_sample <- by_sample %>% filter(n_issues > 0) %>%
    arrange(desc(n_issues), SAMPLE_ALIQUOT) %>%
    relocate(n_issues, issues, .after = SAMPLE_ALIQUOT)
  wr(by_sample, "QC_error_report_by_sample.csv")   # xlsx + a sheet in the combined workbook

  err_summary <- tibble(check = names(.flag_ids),
                        n_samples_flagged = vapply(.flag_ids, length, integer(1))) %>%
    arrange(desc(n_samples_flagged)) %>%
    bind_rows(tibble(check = c("— TOTAL samples in batch —", "— samples with >=1 issue —"),
                     n_samples_flagged = c(dplyr::n_distinct(merged_combined$SAMPLE_ALIQUOT), nrow(by_sample))))
  wr(err_summary, "QC_error_summary.csv")          # xlsx + a sheet in the combined workbook

  # 9. NPQ-not-in-metadata is NOT duplicated here — it already exists in the output root as
  #    samples_not_in_metadata.csv (Step 2); it is referenced in the "Files to Send" section.

  # ── Combined single-file error report (sendable as ONE file) ───────────────────
  # One workbook: SUMMARY (per-check row count + status) → per-check sheets, with the
  # missing-data block expanded into one drill-down sheet PER critical field (built
  # from merged_combined, only fields that actually have samples). The lumped
  # missing_data_by_sample and the QC_error_summary sheets are intentionally omitted
  # here — redundant (missing detail = the per-field sheets; per-check overview =
  # SUMMARY). Their standalone files are still written above.
  if (length(.all_reports) > 0) {
    .miss_specs <- list(
      c(sheet = "Missing or 0 CDX",  field = "CDX"),
      c(sheet = "Missing Race",      field = "Race"),
      c(sheet = "Missing Ethnicity", field = "Ethnicity"),
      c(sheet = "Missing Sex",       field = "sex"),
      c(sheet = "Missing Age",       field = "age_at_subject"))
    .field_sheets <- list()
    for (s in .miss_specs) {
      fld <- s[["field"]]; sht <- s[["sheet"]]; v <- merged_combined[[fld]]
      idx <- if (fld == "CDX") which(.miss(v) | trimws(as.character(v)) == "0") else which(.miss(v))
      if (length(idx)) .field_sheets[[sht]] <- merged_combined[idx, .ctx, drop = FALSE]
    }
    .pick <- function(nm) if (!is.null(.all_reports[[nm]])) setNames(list(.all_reports[[nm]]), nm) else list()
    .body <- c(
      .pick("error_implausible_values"),
      .pick("error_unexpected_race_ethnicity"),
      .pick("missing_data_summary"),
      .field_sheets,
      .pick("error_APOE_no_genotype"),
      .pick("error_APOE_integrity"),
      .pick("error_clinical_consistency"),
      .pick("error_categorical_unexpected_levels"),
      .pick("error_duplicate_SAMPLE_ALIQUOT"),
      .pick("QC_error_report_by_sample"))
    # SUMMARY = per-check overview; the per-field drill-down sheets are detail, not listed
    .chk <- setdiff(names(.body), names(.field_sheets))
    .summary <- tibble(check = .chk, n_rows = vapply(.body[.chk], nrow, integer(1))) %>%
      mutate(status = if_else(n_rows > 0, "review", "clean"), notes = "") %>%  # notes = blank, for reviewer assignments
      arrange(desc(n_rows), check)
    if (.xlsx_ok) {
      .combined <- c(list(SUMMARY = .summary), .body)
      names(.combined) <- substr(make.unique(substr(names(.combined), 1, 31), sep = "_"), 1, 31)  # Excel 31-char sheet limit
      writexl::write_xlsx(.combined, file.path(REPORTS_DIR, "QC_error_report_combined.xlsx"))
      .written <- c(.written, sprintf("`QC_error_report_combined.xlsx` — **one file**, %d sheet(s): SUMMARY + per-check + per-missing-field detail",
                                      length(.combined)))
    } else {
      .long <- bind_rows(lapply(names(.body), function(n) {
        d <- .body[[n]]; if (!nrow(d)) return(NULL)
        d %>% mutate(across(everything(), as.character), check = n)
      }))
      readr::write_csv(.summary, file.path(REPORTS_DIR, "QC_error_report_combined_summary.csv"), na = "")
      if (!is.null(.long)) readr::write_csv(relocate(.long, check), file.path(REPORTS_DIR, "QC_error_report_combined.csv"), na = "")
      .written <- c(.written, "`QC_error_report_combined.csv` + `_summary.csv` — one-file fallback (install writexl for a multi-sheet .xlsx)")
    }
  }

  cat("Reports written to `", REPORTS_DIR, "/`:\n\n", sep = "")
  for (w in .written) cat("- ", w, "\n", sep = "")
  cat("\n")
} else {
  cat("*Reports skipped (covariates off or no merged data).*\n\n")
}
```

# Output Files {#output}

```{r}
#| label: output-summary
#| echo: false
#| results: asis

cat("## Files Generated\n\n")

cat("### Merged Files (`output_files/`)\n\n")
cat("| File | Samples | Columns |\n")
cat("|------|---------|--------|\n")
merged_files <- c("merged_standard_post_QC.csv", "merged_low_post_QC.csv", "merged_combined_post_QC.csv")
for (f in merged_files) {
  fp <- file.path(OUTPUT_DIR, f)
  if (file.exists(fp)) {
    df <- read_csv(fp, show_col_types = FALSE)
    cat("| `", f, "` | ", nrow(df), " | ", ncol(df), " |\n", sep = "")
  }
}

cat("\n### Filtered Files (`output_files/filtered/`)\n\n")
cat("| File | Samples | Columns |\n")
cat("|------|---------|--------|\n")
filtered_files <- list.files(file.path(OUTPUT_DIR, "filtered"), pattern = "\\.csv$")
for (f in filtered_files) {
  fp <- file.path(OUTPUT_DIR, "filtered", f)
  df <- read_csv(fp, show_col_types = FALSE)
  cat("| `", f, "` | ", nrow(df), " | ", ncol(df), " |\n", sep = "")
}

cat("\n### Reports\n\n")
cat("- `output_files/sample_exclusion_report.csv` - Full exclusion report\n")
if (file.exists(file.path(OUTPUT_DIR, "samples_not_in_metadata.csv"))) {
  cat("- `output_files/samples_not_in_metadata.csv` - Samples missing from metadata\n")
}
if (file.exists(file.path(OUTPUT_DIR, "reports", "duplicates", "Duplicate_review.xlsx"))) {
  cat("- `output_files/reports/duplicates/Duplicate_review.xlsx` - **Lab-editable** duplicate decisions ",
      "(mark the run to keep with `FINAL_keep = \"keep\"`, copy to the review dir, re-run)\n", sep = "")
}
if (file.exists(file.path(OUTPUT_DIR, "reports", "duplicates", "Longitudinal_review.xlsx"))) {
  cat("- `output_files/reports/duplicates/Longitudinal_review.xlsx` - Longitudinal reconciliation vs lab's known list\n")
}
if (file.exists(file.path(OUTPUT_DIR, "reports", "QC_error_report_by_sample.xlsx"))) {
  cat("- `output_files/reports/QC_error_report_by_sample.xlsx` - **QC errors by sample** (one row per flagged sample) + `QC_error_summary.xlsx`\n")
}
```

::: {.alert .alert-success}
**Pipeline Complete!**

Filtered datasets are ready for downstream analysis in `output_files/filtered/`.
:::

## Next Steps

To manually edit exclusions and re-run filtering: 1. Open `output_files/sample_exclusion_report.csv` 2. Change `exclude` column as needed 3. Re-render this document with `filter_only: true`:

``` r
quarto::quarto_render("Metadata_Merge_Pipeline.qmd",
                      execute_params = list(filter_only = TRUE))
```

---

# Files to Send for Review & Verification {#send}

```{r}
#| label: files-to-send
#| echo: false
#| results: asis
#| message: false
#| warning: false

if (exists("REPORTS_DIR") && dir.exists(REPORTS_DIR)) {
  .ext  <- if (file.exists(file.path(REPORTS_DIR, "mapping_key_CDX.xlsx"))) "xlsx" else "csv"
  .rows <- function(p) {
    if (!file.exists(p)) return(NA_integer_)
    if (grepl("\\.xlsx$", p)) nrow(suppressMessages(readxl::read_excel(p)))
    else nrow(suppressMessages(readr::read_csv(p, show_col_types = FALSE)))
  }
  .rp <- function(base, sub = "") file.path(REPORTS_DIR, sub, paste0(base, ".", .ext))

  cat("Outputs below go to the contributing groups for **review / verification**. ",
      "Reports are written as `.", .ext, "`",
      if (.ext == "xlsx") " — open in Excel; APOE genotypes stay text (no date conversion)."
      else " (install the `writexl` package for Excel-safe `.xlsx`).",
      " `sample_exclusion_report.csv` is **internal** (drives filtering) and is not sent.\n\n", sep = "")

  cat("### 1. Mapping keys — confirm the groupings\n\n")
  for (b in c("mapping_key_CDX", "mapping_key_Ancestry")) {
    p <- .rp(b); if (file.exists(p)) cat(sprintf("- `reports/%s.%s` (%d rows)\n", b, .ext, .rows(p)))
  }

  cat("\n### 2. Data-issue reports — correct & verify (⚠ = has flags, ✓ = clean)\n\n")
  .comb_xlsx <- file.path(REPORTS_DIR, "QC_error_report_combined.xlsx")
  .comb_csv  <- file.path(REPORTS_DIR, "QC_error_report_combined.csv")
  if (file.exists(.comb_xlsx)) {
    cat("- 📦 **`reports/QC_error_report_combined.xlsx`** — **all checks in ONE file** (SUMMARY sheet + one sheet per check). Send this single file; the per-issue files below are the same content split out.\n")
  } else if (file.exists(.comb_csv)) {
    cat("- 📦 **`reports/QC_error_report_combined.csv`** (+ `_summary.csv`) — all checks bundled into one file.\n")
  }
  .bys <- .rp("QC_error_report_by_sample")
  if (file.exists(.bys)) cat(sprintf("- 👤 **`reports/QC_error_report_by_sample.%s`** — one row per flagged sample (checks tripped, `n_issues`, `issues`) — %d sample(s)\n", .ext, .rows(.bys)))
  .esum <- .rp("QC_error_summary")
  if (file.exists(.esum)) cat(sprintf("- 📊 `reports/QC_error_summary.%s` — per-check flagged counts + batch totals\n", .ext))

  cat("\n**Error reports** — `reports/errors/`:\n")
  for (b in c("error_implausible_values", "error_unexpected_race_ethnicity",
              "error_APOE_integrity", "error_APOE_no_genotype", "error_clinical_consistency",
              "error_categorical_unexpected_levels", "error_duplicate_SAMPLE_ALIQUOT")) {
    n <- .rows(.rp(b, "errors"))
    if (!is.na(n)) cat(sprintf("- %s `reports/errors/%s.%s` — %d flagged\n", if (n > 0) "⚠" else "✓", b, .ext, n))
  }
  cat("\n**Missing-data reports** — `reports/missing_data/`:\n")
  for (b in c("missing_data_by_sample", "missing_data_summary")) {
    n <- .rows(.rp(b, "missing_data"))
    if (!is.na(n)) cat(sprintf("- %s `reports/missing_data/%s.%s` — %d row(s)\n", if (n > 0) "⚠" else "✓", b, .ext, n))
  }
  cat("\n**Duplicate / longitudinal sample reports** — `reports/duplicates/`:\n")
  for (b in c("Duplicate_review", "Longitudinal_review")) {
    n <- .rows(.rp(b, "duplicates"))
    if (!is.na(n)) cat(sprintf("- `reports/duplicates/%s.%s` — %d row(s)\n", b, .ext, n))
  }
  nn <- .rows(file.path(OUTPUT_DIR, "samples_not_in_metadata.csv"))
  if (!is.na(nn)) cat(sprintf("\n- %s `samples_not_in_metadata.csv` (output root) — %d NPQ samples lacking metadata\n",
                              if (nn > 0) "⚠" else "✓", nn))

  cat("\n### Supporting context (not action items)\n\n")
  cat(sprintf("- `reports/mapping_key_CDX.%s`, `reports/mapping_key_Ancestry.%s` — grouping keys (top level)\n", .ext, .ext))
  cat(sprintf("- `reports/check_CDX_vs_CaseControl_crosstab.%s` — CDX × Case_Control cross-tab\n", .ext))
  cat("\n*Internal, do not send:* `sample_exclusion_report.csv` (drives sample filtering).\n\n")
} else {
  cat("*No reports/ folder (covariates off).*\n\n")
}
```
