---
title: "Primary QC of Biomarker Data for ALZ123 Combined Datasets"
author: "ERM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
knit: |
  (function(input, ...) {
    npqs <- list.files("input_files", pattern = "^NPQ.*\\.csv$", full.names = TRUE)
    npqs <- npqs[!grepl("Low|Raw|Sample_QC|triage", basename(npqs), ignore.case = TRUE)]
    tag  <- if (length(npqs)) tools::file_path_sans_ext(basename(npqs[which.max(file.mtime(npqs))])) else format(Sys.Date(), "%Y%m%d")
    rmarkdown::render(input, output_file = paste0("QC_Pipeline_Primary_Alamar_", tag, ".html"))
  })
---
# A. Overview

This protocol describes a workflow for primary **quality control (QC)** of biomarker datasets measured across multiple assay plates (Runs and Bays). 

* The protocol first identifies a set of high-quality biomarkers with low correlation as the **QC Biomarker Set**. 

* Next it uses the QC Biomarker Set to identify samples to "triage" for further inspection based on automated checks (e.g., PCA, outliers)

* It further identifies samples that should be "flagged", i.e., noted but not as bad as triage.

* All biomarkers are then reintegrated, but separated into two sets: 1) those with low detectability and/or low read count, and 2) all remaining biomarkers.

* Finally, post-QC NPQ data are written to output files, separated by the two biomarker groups as well as triaged and non-triaged samples. 

QC analyses are **agnostic** to phenotypic and covariate information.

**For a new QC run**, set the NPQ input filename in section A.3 (and optionally the title in YAML). You **no longer need to rename this file** — the `knit:` function in the YAML header auto-names the HTML after the NPQ input dataset (e.g. `QC_Pipeline_Primary_Alamar_NPQ_20251220.html`), so each run's report is unique and won't overwrite a previous one.


## A.1 Input Data

**You must specify the input NPQ filename below in section A.3**

* Example: NPQ_data <- "NPQ_20250923.csv" 

**Required Input Data Files:** Place all files in a folder called `input_files` in the working directory.

1. **Biomarker NPQ data** (`NPQ_xxxxxxxx.csv`) – Rows: biomarkers; Columns: samples; downloaded from NAS. Assign the filename to the variable `NPQ_data` below.
2. **Alamar sample QC info** (`Sample_QC.csv`) – Rows: biomarkers; Columns: annotation information (IC Median, Detactability, QC Status, etc.). Exported from NPQ_Values.xlsx.
3. **Raw count data** (`Raw_counts.csv`) – Rows: biomarkers; Columns: samples; exported from NPQ_Values.xlsx.
4. **Detectability summary** (`biomarker_detectability.csv`) – Rows: biomarkers; Columns: detectability per plate and overall.
5. **Annotation for biomarkers** (`Annotation_Targets.csv`) – Rows: samples; Columns: annotation information (LOD, NPQ values, plateID, etc.). Exported from NPQ_Values.xlsx.
6. **Replicate list** (`Replicate_list.csv`) – List of HIHG-replicate sample IDs for assay precision checks. Created by analyst.

* Previously, I included `QC_droplist.csv` to flag samples with warnings from Alamar. I'm now going to automate this and list them in the summary. 

**Output Data Files:** Analyst should create a folder `output_files` in the working directory.

1. `NPQxxx_post_QC.csv` – QC’d samples (without triaged samples) with biomarkers having **Min Detectability >50%** and **mean/median raw-read counts >500**.
2. `NPQxxx_Low_post_QC.csv` – QC'd Samples (without triaged samples) failing detectability or read count thresholds.
3. `NPQxxx_post_QC_triage.csv` – Same as above from triaged samples.
4. `NPQxxx_Low_post_QC_triage.csv` – Same as above from triaged samples.
5. Additional ancillary files for outliers, PCA coordinates, flagged samples, etc.
6. I'd like to add a pdf summary to export with final data.


## A.2 Thresholds

Adjustable QC parameters include:

| Parameter | Default | Description |
|-----------|---------|-------------|
| `PCA_SD` | 5 | Number of SDs from mean to define PCA outliers |
| `min_detectability` | 50 | Minimum detectability (%) to retain biomarker in post-QC dataset |
| `min_detectability_qc` | 98 | Minimum detectability (%) to retain biomarker in QC Biomarker dataset |
| `read_count_threshold` | 500 | Minimum mean/median raw reads per sample for QC |
| `corr_thresh` | 0.4 | Maximum correlation coefficient allowed between biomarkers in QC Biomarker dataset |
| `samp_out_thresh` | 1.5 | Multiplier for IQR to determine if a sample if an outlier from the median NPQ for a biomarker|
| `read_out_thresh` | 4 | Multiplier for IQR to determine if a sample if an outlier from the median for Total Reads|
| `FDR_threshold` | 0.01 | Significance threshold for outlier burden |
| `scaffold_mode` | `"rederive"` | QC Biomarker Set selection: `"rederive"` (per-run derive + de-correlate) or `"frozen"` (fixed 21-anchor scaffold + per-plate tripwire). Changes only the sample-triage anchors, not output biomarker content. |


## A.3 Initial Setup 

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# Input NPQ filename

NPQ_data <- "NPQ_20251220.csv" #SPECIFY THE FILE NAME CONTAINING NPQ DATA

# Create output folder if it doesn't exist
if (!dir.exists("output_files")) dir.create("output_files")

# Thresholds
PCA_SD <- 5
min_detectability <- 50
min_detectability_qc <-98
corr_thresh <- 0.4
read_count_threshold <- 500
samp_out_thresh <- 1.5
read_out_thresh <- 4
FDR_threshold <- 0.01

# QC Biomarker Set (scaffold) selection — how the markers used for SAMPLE triage are chosen.
#   "rederive" : per-run derivation (current default) — min detectability >= min_detectability_qc on every
#                plate, mean/median reads > read_count_threshold, then de-correlated (findCorrelation,
#                corr_thresh). Set size drifts with plate count.
#   "frozen"   : a fixed, pre-vetted, de-correlated anchor set (stable across batches) + a per-plate
#                tripwire. Only the triage ANCHORS change; the biomarker CONTENT of the post-QC outputs
#                is identical to "rederive" (anchors + everything else are reintegrated downstream).
scaffold_mode <- "rederive"

# Frozen scaffold: 21 markers >=98% detectable on all 50 ALZ123+8-plate+7-plate plates, de-correlated at
# |r|<=corr_thresh on the combined study samples (APOE excluded: relative-quant + Bay1 anomaly). See
# QC_Runs/Tests/QC_frozen_scaffold_06_17_26/. Used only when scaffold_mode == "frozen".
frozen_scaffold <- c("CCL2","CCL22","CCL3","CCL4","CHI3L1","CNTN2","CXCL8","DDC","FLT1","IFNG","IL10",
                     "IL5","KDR","MSLN","NPY","PDGFRB","PRDX6","SAA1","SFTPD","SMOC1","TIMP3")
# Per-plate tripwire thresholds (anchor x plate detectability %): an anchor cell in [drop, watch) is a
# WATCH (kept, flagged); below drop it is a DROP (not used as an anchor on this run, flagged).
scaffold_watch <- min_detectability_qc   # 98 — the selection bar
scaffold_drop  <- 90
```
---
  
# B. Data Formatting
This section formats the data and produces initial summaries.

Input file is: `r NPQ_data`

```{r load-and-format, message=FALSE, warning=FALSE, echo=TRUE}
library(tidyverse)

# --- Read biomarker NPQ data ---
df <- read_csv(file.path("input_files", NPQ_data), show_col_types = FALSE)

# First column = biomarker names
biomarker_names <- df[[1]]
df <- df[, -1]

# Make matrix (rows = biomarkers, cols = samples)
combined_numeric <- as.matrix(df)
rownames(combined_numeric) <- biomarker_names


# Transpose to samples in rows
combined_t <- t(combined_numeric)

# Convert to data.frame
combined_df <- as.data.frame(combined_t)

# --- Read Alamar Sample QC (per-sample QC stats + sample info) ---
# Robust to export drift: the Alamar export prepends a variable-length "metric" preamble before the
# real header row. Instead of a hardcoded skip, auto-detect the header row (the one containing
# "Sample Name") and validate that the expected columns are present, so a format change fails loudly
# with a clear message rather than cryptically downstream.
.sqc_path <- "input_files/Sample_QC.csv"
.sqc_peek <- readLines(.sqc_path, n = 40, warn = FALSE)
.sqc_skip <- which(grepl('(^|,)\\s*"?Sample Name"?\\s*(,|$)', .sqc_peek))[1] - 1L
if (is.na(.sqc_skip)) {
  .sqc_skip <- 12L
  warning("Sample_QC.csv: 'Sample Name' header not found in first 40 lines; falling back to skip=12.")
}
alamar_qc <- read_csv(.sqc_path, skip = .sqc_skip, show_col_types = FALSE)
message("Sample_QC.csv: header auto-detected at line ", .sqc_skip + 1L, " (skip=", .sqc_skip, ").")

.sqc_miss <- setdiff(c("plateID", "Sample Name", "QC Status"), names(alamar_qc))
if (length(.sqc_miss))
  stop("Sample_QC.csv is missing expected column(s): ", paste(.sqc_miss, collapse = ", "),
       ".\n  Columns found: ", paste(names(alamar_qc), collapse = ", "),
       "\n  The export format may have changed — check the header row / column names.")

# Export drift in value formatting: `IC Median`/`Detectability` may be percent STRINGS
# ("-4.6%", "95.4%") or numeric; `IC Reads`/`Reads` may carry thousands separators. Coerce to numeric
# (parse_number strips "%" and commas; genuinely numeric input passes through unchanged).
.to_num <- function(x) if (is.numeric(x)) x else readr::parse_number(as.character(x))
alamar_qc <- alamar_qc %>%
  mutate(across(any_of(c("IC Median", "Detectability", "IC Reads", "Reads")), .to_num))

# Extract Run and Bay from plateID
meta_df <- alamar_qc %>%
  mutate(
    Run = stringr::str_extract(plateID, "^[0-9]{8}-[0-9]{4}"),
    Bay = stringr::str_extract(plateID, "Bay[0-9]+")
  ) %>%
  transmute(SampleID = `Sample Name`, Run, Bay)

# Keep only samples present in combined_df
meta_df <- meta_df[meta_df$SampleID %in% rownames(combined_df), ]

# Reorder to match rownames
meta_df <- meta_df[match(rownames(combined_df), meta_df$SampleID), ]

# Add Run and Bay columns at the front
combined_df$Run <- meta_df$Run
combined_df$Bay <- meta_df$Bay
combined_df <- combined_df %>% select(Run, Bay, everything())
rownames(combined_df) <- rownames(combined_t)
```

## B.1 Initial Cleaning

### B.1.1 HIHG Replicates and Control Samples
Pull out Alamar control and HIHG replicate samples for examination in later pipeline stages.

```{r special_samples, message=FALSE, warning=FALSE, echo=TRUE}
library(dplyr)
library(tidyr)

# --- Read HIHG replicate list ---
rep <- read_csv("input_files/Replicate_list.csv", show_col_types = FALSE)
hihg_reps <- combined_df[rownames(combined_df) %in% rep[[1]], ]

combined_df <- combined_df[!rownames(combined_df) %in% rownames(hihg_reps), ]

# Identify Alamar controls and (SC, IPC, and NC) and pull out from main dataset
nc_samples   <- combined_df[grepl("^NC", rownames(combined_df)), ]
sc_samples   <- combined_df[grepl("^SC", rownames(combined_df)), ]
ipc_samples  <- combined_df[grepl("^IPC", rownames(combined_df)), ]

combined_df <- combined_df[!grepl("^(NC|IPC|SC)", rownames(combined_df)), ]

# Function to summarize counts by Run and Bay
summarize_counts <- function(df, type_name) {
  df %>%
    group_by(Run, Bay) %>%
    summarise(!!type_name := n(), .groups = "drop")
}

# Summarize each type
nc_summary   <- summarize_counts(nc_samples, "NC")
sc_summary   <- summarize_counts(sc_samples, "SC")
ipc_summary  <- summarize_counts(ipc_samples, "IPC")
hihg_summary <- summarize_counts(hihg_reps, "HIHG_Replicate")

# Combine into a single table
control_summary <- full_join(nc_summary, ipc_summary, by = c("Run","Bay")) %>%
                   full_join(sc_summary, by = c("Run","Bay")) %>%
                   full_join(hihg_summary, by = c("Run","Bay")) %>%
                   replace_na(list(NC = 0, IPC = 0, SC = 0, HIHG_Replicate = 0)) %>%
                   arrange(Run, Bay)

as.data.frame(control_summary)

#Write replicates and controls with their biomarker data to files

write_csv(hihg_reps, "output_files/hihg_reps_NPQ.csv")
write_csv(sc_samples, "output_files/sc_reps_NPQ.csv")
write_csv(nc_samples, "output_files/nc_reps_NPQ.csv")
write_csv(ipc_samples, "output_files/ipc_reps_NPQ.csv")

```
 
### B.1.2 Samples with Bad Data
Identify any samples with non-numeric entries, e.g., NA, NaN.

**Bad Samples:**

```{r bad-samples, message=FALSE, warning=FALSE, echo=TRUE}
# THIS IS A NEW SECTION. CHANGED
# Remove meta-data columns from numeric check
biom_only <- combined_df[ , -c(1,2), drop = FALSE]

# Identify rows with ONLY numeric values
is_numeric_row <- apply(biom_only, 1, function(r) all(suppressWarnings(!is.na(as.numeric(r)))))
bad_samples <- rownames(combined_df)[!is_numeric_row]
bad_rows_df <- combined_df[bad_samples, , drop = FALSE]

write_csv(bad_rows_df, "output_files/bad_samples.csv")

combined_df <- combined_df[is_numeric_row, , drop = FALSE]

print(bad_samples)

```

## B.2 Initial Summaries
  
### B.2.1 Non-replicate Sample Counts
  
```{r sample-counts, message=FALSE, warning=FALSE, echo=TRUE}
# HIHG Reps, SC, NC and IPC are already removed above

# Summarize counts by Run × Bay
nonrep_summary <- combined_df %>%
  group_by(Run, Bay) %>%
  summarise(NonRep_Samples = n(), .groups = "drop") %>%
  arrange(Run, Bay)

as.data.frame(nonrep_summary)
```

### B.2.2 Biomarker Distributions

```{r biomarker-boxplots, echo=TRUE, fig.height=6, fig.width=12}
# THIS WAS REMOVING TOO MANY COLUMNS. CHANGED
combined_df[, -(1:2)] <- lapply(combined_df[, -(1:2)], function(x) suppressWarnings(as.numeric(x)))
# Remove metadata columns
biomarker_df <- combined_df[, sapply(combined_df, is.numeric), drop = FALSE]

# Parameters for paneling
panel_size <- 28  # number of biomarkers per panel
n_biomarkers <- ncol(biomarker_df)
n_panels <- ceiling(n_biomarkers / panel_size)

# Loop through panels
for (i in seq_len(n_panels)) {
  start_col <- (i - 1) * panel_size + 1
  end_col <- min(i * panel_size, n_biomarkers)
  
  panel_matrix <- biomarker_df[, start_col:end_col]
  
  # Set up plot layout (1 panel per figure)
  par(mfrow = c(1,1), mar = c(8,4,4,2))  # adjust margins as needed
  
  boxplot(panel_matrix,
          main = paste("Biomarkers", start_col, "to", end_col),
          xlab = "Biomarker",
          ylab = "NPQ Value",
          las = 2,            # rotate x-axis labels
          col = "lightblue")
}
```
---

# C. Prep QC Biomarker Set
This is focused on generating the most stringent, high-quality biomarker dataset for QC (**QC Biomarker Set**).

## C.1 Detectability Filtering

### C.1.1 Low Detectability
Keep only biomarkers with >98% detectability on every plate for the QC Dataset. Triage others for later inspection. Importantly, some with low detectability are not necessarily bad, e.g., APOE4 and others due to polymorphisms. They are just being temporarily removed to make a clean QC Biomarker Set.

```{r detectability-filtering, message=FALSE, warning=FALSE, echo=TRUE, results='asis'}
library(tidyverse)

# Read detectability
detect <- read_csv("input_files/biomarker_detectability.csv", show_col_types = FALSE)

# Keep only the numeric plate columns and overall
plate_cols <- names(detect)[-1]  # first column is Biomarker name

# Robustness to export drift: coerce plate columns to numeric (a future export might quote them or
# add "%"), then ensure the 0-100 scale the thresholds assume (min_detectability_qc, min_detectability).
detect <- detect %>%
  mutate(across(all_of(plate_cols), ~ if (is.numeric(.x)) .x else readr::parse_number(as.character(.x))))
.det_max <- suppressWarnings(max(unlist(detect[plate_cols]), na.rm = TRUE))
if (is.finite(.det_max) && .det_max <= 1) {
  detect <- detect %>% mutate(across(all_of(plate_cols), ~ .x * 100))
  message("biomarker_detectability.csv appeared to be on a 0-1 proportion scale (max = ",
          round(.det_max, 3), "); rescaled x100 to match the 0-100 detectability thresholds.")
}

# Relative-quant metrics (Curve_Quant = "R", e.g. APOE, CRP): Alamar reports no detectability
# or LOD, so every plate value is NA. Keep them in all datasets and exempt from detectability
# filtering -- do NOT let an all-NA row pass via min()'s Inf.
rel_quant <- detect$Biomarker[rowSums(!is.na(detect[plate_cols])) == 0]

# Determine low-detectability biomarkers
low_det <- detect %>%
  pivot_longer(cols = all_of(plate_cols), names_to = "Plate", values_to = "Detectability") %>%
  group_by(Biomarker) %>%
  summarize(
    Min_Detectability = if (all(is.na(Detectability))) NA_real_ else min(Detectability, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(Min_Detectability), Min_Detectability < min_detectability_qc)

# Save removed biomarkers for record
write_csv(low_det, "output_files/biomarkers_low_detectability.csv")

# Print removed biomarkers
knitr::kable(
  low_det,
  caption = "Biomarkers Removed Due to Low Detectability (min detectability <98%)"
)

# Note relative-quant metrics retained without detectability filtering
if (length(rel_quant) > 0) {
  cat(sprintf(
    "\n> **Note -- relative-quant metrics:** %s have `Curve_Quant = \"R\"`; Alamar reports no detectability or LOD for them. They are **retained in all datasets** and exempt from detectability/LOD filtering.\n",
    paste(rel_quant, collapse = ", ")))
}

# For QC, keep only high-detectability biomarkers (rel_quant targets never entered low_det)
biomarkers_keep <- setdiff(colnames(combined_df), low_det$Biomarker)

combined_qc <- combined_df %>%
  select(Run, Bay, all_of(biomarkers_keep))
```

### C.1.2 Low Raw Counts
Keep only biomarkers with mean and median raw counts >500 on every plate for the QC Biomarker Set.  

```{r raw-count filtering, message=FALSE, warning=FALSE, echo=TRUE}
library(tidyverse)

# Load raw counts
raw_counts <- read_csv("input_files/Raw_counts.csv", show_col_types = FALSE)

# Biomarkers as rownames
raw_mat <- as.matrix(raw_counts[,-1])
rownames(raw_mat) <- raw_counts[[1]]

# Convert to samples x biomarkers to match combined_qc structure
raw_t <- t(raw_mat)

# Ensure matching sample sets between raw counts & metadata
raw_df <- as.data.frame(raw_t)
raw_df$SampleID <- rownames(raw_t)

raw_df <- raw_df %>%
  filter(SampleID %in% rownames(combined_qc)) %>%
  left_join(meta_df, by = c("SampleID" = "SampleID")) %>%
  select(SampleID, Run, Bay, everything())

#Save for later parts of the pipeline
write_csv(raw_df, "output_files/raw_df.csv")

### Compute mean & median per Biomarker per Plate ####
plate_stats <- raw_df %>%
  pivot_longer(cols = -c(Run, Bay, SampleID),
               names_to = "Biomarker",
               values_to = "RawCount") %>%
  group_by(Biomarker, Run, Bay) %>%
  summarize(
    mean_count = mean(RawCount, na.rm = TRUE),
    median_count = median(RawCount, na.rm = TRUE),
    .groups = "drop"
  )

### Identify biomarkers failing thresholds ###
low_counts <- plate_stats %>%
  group_by(Biomarker) %>%
  summarize(
    Min_Mean = min(mean_count, na.rm = TRUE),
    Min_Median = min(median_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(Min_Mean < read_count_threshold | Min_Median < read_count_threshold)

### Save for record
write_csv(low_counts, "output_files/biomarkers_low_counts.csv")

### Drop columns from QC set
biomarkers_to_drop <- low_counts$Biomarker
biomarkers_to_drop <- intersect(biomarkers_to_drop, colnames(combined_qc))
biomarkers_to_drop <- setdiff(biomarkers_to_drop, c("Run", "Bay"))

combined_qc <- combined_qc %>%
  select(-all_of(biomarkers_to_drop))

### Display formatted table in HTML report
knitr::kable(
  low_counts,
  caption = "Biomarkers Removed Due to Low Raw Counts (<500 Min Mean or Min Median)"
)

```  

## C.2 Correlation Filtering
Create a set of biomarkers with low correlation for the QC Biomarker Set

```{r correlation-filtering, message=FALSE, warning=FALSE, echo=TRUE, results='asis'}
library(caret)
library(pheatmap)

# Biomarker pool entering scaffold selection = everything that passed detectability (C.1.1) and raw-count
# (C.1.2) filtering. Both modes preserve this pool: independent_biomarkers (the triage anchors) and
# high_cor (added back in E.1) always partition `qc_pool`, so post-QC biomarker CONTENT is mode-invariant.
qc_pool <- setdiff(colnames(combined_qc), c("Run", "Bay"))
qc_pool <- qc_pool[sapply(combined_qc[qc_pool], is.numeric)]

if (identical(scaffold_mode, "frozen")) {
  # ---- Frozen scaffold + per-plate tripwire ----------------------------------------------------------
  # Anchors = frozen markers that also cleared the standard pool this run (>=98% detect, >500 reads). A
  # frozen marker below the bar simply can't anchor this batch; the tripwire below surfaces why.
  anchors  <- intersect(frozen_scaffold, qc_pool)
  not_pool <- setdiff(frozen_scaffold, anchors)   # frozen markers unavailable as anchors this run

  # Tripwire: scan every (frozen anchor x plate) detectability cell; alert on anything below the bar.
  pc <- grep("^Plate_", names(detect), value = TRUE)
  det_anch <- detect[detect$Biomarker %in% frozen_scaffold, , drop = FALSE]
  Pmat <- as.matrix(det_anch[, pc]); rownames(Pmat) <- det_anch$Biomarker
  idx  <- which(Pmat < scaffold_watch, arr.ind = TRUE)
  tripwire <- if (nrow(idx)) {
    tibble(anchor = rownames(Pmat)[idx[, 1]], plate = pc[idx[, 2]], detectability = round(Pmat[idx], 1)) %>%
      mutate(status = ifelse(detectability < scaffold_drop, "DROP + alert", "WATCH (keep + alert)")) %>%
      arrange(detectability)
  } else tibble(anchor = character(), plate = character(), detectability = numeric(), status = character())
  write_csv(tripwire, "output_files/scaffold_tripwire.csv")

  independent_biomarkers <- anchors
  high_cor               <- setdiff(qc_pool, independent_biomarkers)   # added back in E.1 -> content invariant

  combined_qc <- combined_qc %>% select(Run, Bay, all_of(independent_biomarkers))
  write.csv(independent_biomarkers, "output_files/QC_Biomarker_Set.csv")

  cat(sprintf("**Scaffold mode: FROZEN** — using **%d of %d** frozen anchors for sample triage.\n\n",
              length(independent_biomarkers), length(frozen_scaffold)))
  if (length(not_pool))
    cat(sprintf("> %d frozen marker(s) not available as anchors this run (absent or below the detectability/read bar): %s\n\n",
                length(not_pool), paste(not_pool, collapse = ", ")))
  if (nrow(tripwire)) {
    cat(sprintf("> **TRIPWIRE: %d anchor x plate cell(s) below %g%%** — WATCH = kept & flagged, DROP (<%g%%) = not anchored & flagged:\n\n",
                nrow(tripwire), scaffold_watch, scaffold_drop))
    print(knitr::kable(tripwire))
    cat("\n")
  } else {
    cat(sprintf("> Tripwire: every frozen anchor clears %g%% on every plate — no alert. ✓\n\n", scaffold_watch))
  }
  # Heatmap of the anchors actually used (confirms they remain de-correlated on this batch)
  cor_mat <- cor(combined_qc[, independent_biomarkers], use = "pairwise.complete.obs")
  pheatmap(cor_mat, main = "Heatmap of Frozen Scaffold Anchors", fontsize_row = 6, fontsize_col = 6)

} else {
  # ---- Re-derive (default): de-correlate the high-detectability / high-count pool --------------------
  samp_num <- combined_qc[, qc_pool]
  cor_mat  <- cor(samp_num, use = "pairwise.complete.obs")

  # Identify highly correlated pairs (|r| >= corr_thresh)
  high_cor <- findCorrelation(cor_mat, cutoff = corr_thresh, names = TRUE)

  # Independent biomarkers
  independent_biomarkers <- setdiff(colnames(samp_num), high_cor)

  pheatmap(cor_mat[independent_biomarkers, independent_biomarkers],
           main = "Heatmap of Independent Biomarkers",
           fontsize_row = 6, fontsize_col = 6)

  combined_qc <- combined_qc %>% select(Run, Bay, all_of(independent_biomarkers))
  write.csv(independent_biomarkers, "output_files/QC_Biomarker_Set.csv")
}

```
---

# D. Sample QC

## D.1 Alamar Sample QC
Gather the Alamar sample QC statistics and identify samples with "warning".

```{r apply-droplist, message=FALSE, warning=FALSE, echo=TRUE}
library(readr)
library(dplyr)

#alamar_qc is the Sample_QC file read in above

warn_samps <- alamar_qc %>%
  filter(grepl("warn", `QC Status`, ignore.case = TRUE)) %>%  # values are "Passed"/"Warning" (case/wording can drift)
  left_join(
    meta_df %>% select(SampleID, Run, Bay),
    by = c("Sample Name" = "SampleID")
  )

library(kableExtra)
warn_tbl <- warn_samps %>%
  arrange(Run, Bay, `Sample Name`) %>%
  transmute(`Sample Name`,
            `IC Median`   = round(`IC Median`, 3),
            Detectability = round(Detectability, 3),
            `IC Reads`,
            Well = AUTO_WELLPOSITION,
            .grp = paste(Run, Bay))
.grp_rle <- rle(warn_tbl$.grp)
warn_tbl %>%
  select(-.grp) %>%
  kbl(caption = "Alamar Sample QC Warnings (sorted, grouped by plate)") %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  pack_rows(index = setNames(.grp_rle$lengths, .grp_rle$values))

write_csv(arrange(warn_samps, Run, Bay, `Sample Name`), "output_files/Alamar_QC_warn.csv")

```

## D.2 Sample Outlier Detection 

### D.2.1 PCA-Based Outlier Check
Looking for single samples or small groups that are outliers on the first two PCs.
Using a threshold of `r PCA_SD` sd to define outliers.

```{r pca-outlier, message=FALSE, warning=FALSE, echo=TRUE}
library(tidyverse)
library(patchwork)
library(dplyr)
library(irlba)  # efficient PCA

set.seed(123)

# --- Round 1 PCA -----------------------------------------------------------
biomarker_cols <- setdiff(names(combined_qc), c("Run", "Bay"))
qc_matrix <- combined_qc[, biomarker_cols] %>% as.matrix()
qc_scaled <- scale(qc_matrix)

pca_res <- prcomp(qc_scaled, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(combined_qc),
  Run = combined_qc$Run,
  Bay = combined_qc$Bay,
  PC1 = pca_res$x[, 1],
  PC2 = pca_res$x[, 2]
)

# Identify outliers by distance

pca_df <- pca_df %>%
  group_by(Run, Bay) %>%
  mutate(
    PC_dist = sqrt(PC1^2 + PC2^2),
    mean_PCdist = mean(PC_dist, na.rm = TRUE),
    sd_PCdist   = sd(PC_dist, na.rm = TRUE),
    extreme_outlier = PC_dist > (mean_PCdist + PCA_SD * sd_PCdist)
  ) %>%
  ungroup()

extreme_samples_round1 <- pca_df$Sample[pca_df$extreme_outlier]
write_csv(pca_df, "output_files/sample_pca_coordinates_round1.csv")

# Variance explained
var_explained <- summary(pca_res)$importance["Proportion of Variance", 1:2] * 100
pc1_var <- round(var_explained[1], 1)
pc2_var <- round(var_explained[2], 1)

# Plot round 1
plot1 <- ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = Run, shape = Bay), size = 3, alpha = 0.7) +
  geom_point(
    data = filter(pca_df, extreme_outlier),
    aes(x = PC1, y = PC2),
    shape = 4, color = "black", size = 3, stroke = 0.5
  ) +
  labs(
    title = paste0("PCA Round 1 (", length(extreme_samples_round1), " outliers)"),
    x = paste0("PC1 (", pc1_var, "% variance)"),
    y = paste0("PC2 (", pc2_var, "% variance)")
  ) +
  theme_minimal() +
  theme(legend.position = "right")

#Round 1 results
print(plot1)
knitr::kable(pca_df[pca_df$extreme_outlier, ], caption = "First-round PCA outliers:")

# --- Round 2 PCA (remove round 1 outliers) ---------------------------------
if (length(extreme_samples_round1) > 0) {
  combined_qc_round2 <- combined_qc[!rownames(combined_qc) %in% extreme_samples_round1, ]

  qc_matrix2 <- combined_qc_round2[, biomarker_cols] %>% as.matrix()
  qc_scaled2 <- scale(qc_matrix2)

  pca_res2 <- prcomp(qc_scaled2, center = TRUE, scale. = TRUE)

  pca_df2 <- data.frame(
    Sample = rownames(combined_qc_round2),
    Run = combined_qc_round2$Run,
    Bay = combined_qc_round2$Bay,
    PC1 = pca_res2$x[, 1],
    PC2 = pca_res2$x[, 2]
  )
    
  pca_df2 <- pca_df2 %>%
  group_by(Run, Bay) %>%
  mutate(
    PC_dist = sqrt(PC1^2 + PC2^2),
    mean_PCdist = mean(PC_dist, na.rm = TRUE),
    sd_PCdist   = sd(PC_dist, na.rm = TRUE),
    extreme_outlier = PC_dist > (mean_PCdist + PCA_SD * sd_PCdist)
  ) %>%
  ungroup()


  # Identify new outliers
  extreme_samples_round2 <- pca_df2$Sample[pca_df2$extreme_outlier]
  write_csv(pca_df2, "output_files/sample_pca_coordinates_round2.csv")

  var_explained2 <- summary(pca_res2)$importance["Proportion of Variance", 1:2] * 100
  pc1_var2 <- round(var_explained2[1], 1)
  pc2_var2 <- round(var_explained2[2], 1)

  # Plot round 2
  plot2 <- ggplot(pca_df2, aes(x = PC1, y = PC2)) +
    geom_point(aes(color = Run, shape = Bay), size = 3, alpha = 0.7) +
    geom_point(
      data = filter(pca_df2, extreme_outlier),
      aes(x = PC1, y = PC2),
      shape = 4, color = "black", size = 3, stroke = 0.5
    ) +
    labs(
      title = paste0("PCA Round 2 (", length(extreme_samples_round2), " new outliers)"),
      x = paste0("PC1 (", pc1_var2, "% variance)"),
      y = paste0("PC2 (", pc2_var2, "% variance)")
    ) +
    theme_minimal() +
    theme(legend.position = "right")

  # Round 2
  print(plot2)
  knitr::kable(pca_df2[pca_df2$extreme_outlier, ], caption = "Second-round PCA outliers:")

} else {
  cat("\nNo first-round PCA outliers detected. Skipping second round.\n")
  plot1
}
```

I changed these to be plate-specific outliers.

### D.2.2 Sample Outlier Burden
Here we are identifying samples that have extreme outliers over biomarkers within their plate.
Compute p-value for whether outlier burden in a sample is more than expected by chance. 

```{r excessive_outliers, message=FALSE, warning=FALSE, echo=TRUE}
library(dplyr)
library(poibin)  # for Poisson-Binomial p-values

#Uses IQR method to define outliers
upper_matrix <- matrix(0, nrow = nrow(qc_matrix), ncol = ncol(qc_matrix),
                       dimnames = dimnames(qc_matrix))
lower_matrix <- upper_matrix * 0

for (rb in unique(paste(combined_qc$Run, combined_qc$Bay, sep = "_"))) {
  idx <- which(paste(combined_qc$Run, combined_qc$Bay, sep = "_") == rb)
  submat <- qc_matrix[idx, , drop = FALSE]

  # Compute Q1 and Q3 per biomarker
  Q1 <- apply(submat, 2, quantile, 0.25, na.rm = TRUE)
  Q3 <- apply(submat, 2, quantile, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  
  # Define outliers
  upper <- sweep(submat, 2, Q3 + samp_out_thresh * IQR, ">")
  lower <- sweep(submat, 2, Q1 - samp_out_thresh * IQR, "<")

  upper_matrix[idx, ] <- upper_matrix[idx, ] + upper
  lower_matrix[idx, ] <- lower_matrix[idx, ] + lower
}

# Assume upper_matrix and lower_matrix are data.frames with samples in rows, biomarkers in columns and that combined_qc has Run and Bay columns

add_outlier_stats <- function(out_mat, combined_df) {
  out_mat <- as.matrix(out_mat)
  
  out_df <- as.data.frame(out_mat) %>%
    mutate(
      SampleID = rownames(out_mat),
      Run = combined_df$Run,
      Bay = combined_df$Bay
    )

  p_biom <- colMeans(out_mat, na.rm = TRUE)

  out_df <- out_df %>%
    rowwise() %>%
    mutate(
      Total_Outliers = sum(c_across(where(is.numeric)), na.rm = TRUE),
      Expected_Outliers = sum(p_biom, na.rm = TRUE),
      P_Value = 1 - poibin::ppoibin(Total_Outliers - 1, p_biom)
    ) %>%
    ungroup() %>%
    mutate(
      FDR = p.adjust(P_Value, method = "BH")  # ← NEW COLUMN
    ) %>%
    select(SampleID, Run, Bay, Total_Outliers, Expected_Outliers,
           P_Value, FDR, everything())

  return(out_df)
}

upper_df <- add_outlier_stats(upper_matrix, combined_qc)
lower_df <- add_outlier_stats(lower_matrix, combined_qc)

# Save to CSV
write_csv(upper_df, "output_files/upper_outliers_with_stats.csv")
write_csv(lower_df, "output_files/lower_outliers_with_stats.csv")

# ---- Sample Outlier Summary (Top 10) ----

# Sort by significance (lowest P-value first)
upper_sorted <- upper_df %>% arrange(P_Value)
lower_sorted <- lower_df %>% arrange(P_Value)

# Save full sorted tables
write_csv(upper_sorted, "output_files/upper_outliers_sorted.csv")
write_csv(lower_sorted, "output_files/lower_outliers_sorted.csv")

# Select first 10 + only key columns for report
upper_top10 <- upper_sorted %>%
  select(SampleID, Run, Bay, "Total Out"=Total_Outliers, "Exp Out" = Expected_Outliers, 
         P_Value, FDR) %>%
  slice(1:10)

lower_top10 <- lower_sorted %>%
  select(SampleID, Run, Bay, "Total Out"=Total_Outliers, "Exp Out" = Expected_Outliers,
         P_Value, FDR) %>%
  slice(1:10)

# Print to RMarkdown
knitr::kable(upper_top10, caption = "Top 10 Upper Outliers (Sorted by P-value)")

knitr::kable(lower_top10, caption = "Top 10 Lower Outliers (Sorted by P-value)")
```

These plots show where the outlier samples fall for the biomarkers. Red are outliers. Blue are points for the same sample where they are not outliers. Only showing plots for samples with FDR< `r FDR_threshold`.
 
```{r visualize_excessive_outliers, message=FALSE, warning=FALSE, echo=TRUE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)

upper_sig <- upper_df$SampleID[upper_df$FDR < FDR_threshold]
lower_sig <- lower_df$SampleID[lower_df$FDR < FDR_threshold]

# List of extreme outlier SampleIDs 
extreme_samples <- c(upper_sig,lower_sig)

plot_list <- list()

plot_sample_violin <- function(sample_id) {
  
  # Get Run/Bay for title
  sample_run <- combined_qc[sample_id, "Run"]
  sample_bay <- combined_qc[sample_id, "Bay"]
  
  # Convert to long format
  long_df <- combined_qc %>%
    rownames_to_column("SampleID") %>%
    pivot_longer(
      cols = -c(SampleID, Run, Bay),
      names_to = "Biomarker",
      values_to = "Value"
    )
  
  # Create Highlight column: default "Other"
  long_df$Highlight <- "Other"
  
  # Pull the outlier flags for this sample
  outlier_flags <- upper_matrix[sample_id,]+lower_matrix[sample_id,]
  
  # Only update the Highlight for this sample
  long_df$Highlight[long_df$SampleID == sample_id] <- ifelse(
    outlier_flags[long_df$Biomarker[long_df$SampleID == sample_id]],
    "Extreme",
    "Normal"
  )
  
  # Plot
  p <- ggplot(long_df, aes(x = Biomarker, y = Value)) +
    geom_violin(fill = "gray70", alpha = 0.6, trim = FALSE) +
    geom_point(
      data = subset(long_df, Highlight != "Other"),
      aes(color = Highlight),
      size = 3
    ) +
    scale_color_manual(values = c("Extreme" = "red", "Normal" = "blue")) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
    labs(title = paste("Sample", sample_id,
                       "| Run:", sample_run, "| Bay:", sample_bay),
         x = "Biomarker",
         y = "NPQ")
  
  return(p)
}

# Loop through extreme samples
for (sample_id in extreme_samples) {
  p=plot_sample_violin(sample_id)
  print(p)
}
```

Determine samples with a significant number of outliers over biomarkers.

### D.2.3 IC-Outlier Reads
Here we are identifying samples that have extreme numbers of matching reads in aggregate within their plate.

Alamar does this with respect to the median number of reads in the IC samples within a plate: Sample IC reads relative to the median (Within +/-40% of the plate median). The IC is mCherry.
They also WARN if Sample IC reads <1000. Usually the outlier criteria would capture this, but it would not tell us if the whole plate had low reads. I added that as a conditional lower bound. 

```{r Sample_vs_IC_reads, message=FALSE, warning=FALSE, echo=TRUE}
#Look at mCherry raw reads over Run and Bay

#raw_df has the raw count data (including mCherry) for all of the samples in QC
# Ensure Run and Bay are factors
raw_df$Run <- factor(raw_df$Run)
raw_df$Bay <- factor(raw_df$Bay)

# Custom Bay colors
bay_colors <- c("Bay1" = "red", "Bay2" = "green", "Bay3" = "blue")

# The IC (mCherry) is no longer a target row in the NPQ/Raw-counts export; its per-sample
# reads now live in the Sample QC sheet as `IC Reads` (loaded above as alamar_qc). Source from
# there, restricted to the QC sample set (samples present in raw_df).
# Lower bound is -40% of the median or 1000, whichever is bigger.
df_temp <- alamar_qc %>%
  transmute(
    SampleID = `Sample Name`,
    Value    = as.numeric(`IC Reads`),
    Run      = stringr::str_extract(plateID, "^[0-9]{8}-[0-9]{4}"),
    Bay      = stringr::str_extract(plateID, "Bay[0-9]+")
  ) %>%
  filter(SampleID %in% raw_df[[1]]) %>%
  group_by(Run, Bay) %>%
  mutate(
    Median_Value = median(Value, na.rm = TRUE),
    Lower_Thresh = pmax((Median_Value * 0.6), 1000),
    Upper_Thresh = Median_Value * 1.4,  # +40%
    Outlier_Flag = Value < Lower_Thresh | Value > Upper_Thresh
  ) %>%
  ungroup()

# Plot
#I tried to highlight outliers on plot but I couldn't get them centered right
p <- ggplot(df_temp, aes(x = Run, y = Value, fill = Bay)) +
  geom_violin(alpha = 0.5, trim = FALSE, position = position_dodge(width = 0.8)) +
  geom_jitter(aes(color = Bay),
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.75, size = 1.3) +
  scale_fill_manual(values = bay_colors) +
  scale_color_manual(values = bay_colors) +
  theme_minimal() +
  labs(
    title = "Run × Bay Effects: IC Reads",
    x = "Run",
    y = "Reads"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p)

df_temp <- df_temp %>%
  rename(IC_Reads = Value)

# Print table
knitr::kable(
  df_temp %>% 
    filter(Outlier_Flag == TRUE) %>%
    select(SampleID, Run, Bay, IC_Reads) %>%
    arrange(desc(IC_Reads)),
  caption = "mCherry-Read Outliers (Sorted by Reads)"
)

IC_outlier_samples= df_temp %>% 
    filter(Outlier_Flag ==TRUE)%>%
    arrange(desc(IC_Reads)) %>%
    select(SampleID)
  
```
Note: As expected, the samples with very high reads for mCherry, have lower average NPQ than others.

### D.2.4 Extreme Matching Reads
Here we are identifying samples that have extreme numbers of total matching reads, in aggregate, within their plate. These are **flagged** samples. For the lower bound, we use a minimum count of 500,000 if the IQR bound is smaller.

```{r excessive_matching_reads, message=FALSE, warning=FALSE, echo=TRUE}
# ---- Matching-Reads QC ----
library(dplyr)
library(ggplot2)
library(readr)


# Merge matching reads info
matching_df <- alamar_qc %>%
  select(SampleID = `Sample Name`, Reads) %>%          # rename + keep Reads
  filter(SampleID %in% rownames(combined_qc)) %>%      # keep only samples in combined_qc
  left_join(                                           
    combined_qc %>%                                     # pull Run + Bay from combined_qc
      rownames_to_column("SampleID") %>%
      select(SampleID, Run, Bay),
    by = "SampleID"
  )

# ---- Outlier detection by Run × Bay ----
#IQR MAY NOT CAPTURE LOWER READ OUTLIERS. USE ALAMAR LOWER BOUND OF 500,000 if the IQR bound is too low.
#This changed from 11/06 version
# Compute per-Run/Bay medians and flag outliers

matching_outliers <- matching_df %>%
  group_by(Run, Bay) %>%
  mutate(
    Median = median(Reads, na.rm = TRUE),
    Q1     = quantile(Reads, 0.25, na.rm = TRUE),
    Q3     = quantile(Reads, 0.75, na.rm = TRUE),
    IQR    = Q3 - Q1,
    Lower_Bound = pmax(500000, Median - read_out_thresh * IQR),
    Upper_Bound = Median + read_out_thresh * IQR,
    Matching_Outlier = Reads < Lower_Bound | Reads > Upper_Bound
  ) %>%
  ungroup()

# Convert factor levels (ensure ordered)
matching_outliers$Run <- factor(matching_outliers$Run)
matching_outliers$Bay <- factor(matching_outliers$Bay, levels = c("Bay1", "Bay2", "Bay3"))

# Assign Bay colors
bay_colors <- c("Bay1" = "red", "Bay2" = "green", "Bay3" = "blue")
#bay_shapes <- c("Bay1" = 16, "Bay2" = 17, "Bay3" = 15)  # optional: circle, triangle, square

p <- ggplot(matching_outliers, aes(x = Run, y = Reads, fill = Bay)) +
  geom_violin(alpha = 0.5, trim = FALSE, position = position_dodge(width = 0.8)) +
  geom_jitter(aes(color = Bay),
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.75, size = 1.3) +
  scale_fill_manual(values = bay_colors) +
  scale_color_manual(values = bay_colors) +
  theme_minimal() +
  labs(
    title = "Run × Bay Effects: Total Reads",
    x = "Run",
    y = "Reads"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p)

read_outliers <- matching_outliers %>%
    filter(Matching_Outlier == TRUE) %>%
    arrange(desc(Reads)) %>%
    select(SampleID, Run, Bay, Lower_Bound, Upper_Bound, Reads)

# Print table
knitr::kable(
  read_outliers, 
  caption = "Matching-Read Outliers (Sorted by Reads)"
)

# Save to CSV
write_csv(
  read_outliers %>%
    arrange(Run, Bay, desc(Reads)),
  "output_files/flagged_read_outliers.csv"
)

# read_outliers are our flagged samples

```

## D.3 Sample Traige
Here we are identifying samples to be triaged based on our sample QC and comparing them to the Alamar QC.

```{r sample_triage, message=FALSE, warning=FALSE, echo=TRUE}
# --- Triage samples for removal ---

# Bad Samples

# Identify PCA extreme outliers
pca_outliers <- c(pca_df$Sample[pca_df$extreme_outlier],pca_df2$Sample[pca_df2$extreme_outlier])

# Identify samples with significant outlier burden
# Combine upper and lower matrices for total FDR per sample
upper_sig <- upper_df$SampleID[upper_df$FDR < FDR_threshold]
lower_sig <- lower_df$SampleID[lower_df$FDR < FDR_threshold]
outlier_burden_samples <- unique(c(upper_sig, lower_sig))

triage_df <- bind_rows(
  tibble(SampleID = as.character(unlist(pca_outliers)), Reason = "PCA outlier"),
  tibble(SampleID = as.character(unlist(outlier_burden_samples)), Reason = "High outlier burden (FDR)"),
  tibble(SampleID = as.character(unlist(IC_outlier_samples)), Reason = "IC outlier"),
  tibble(SampleID = as.character(unlist(bad_samples)), Reason = "Bad Data"),
)

# If a sample appears in multiple categories, collapse reasons
triage_df <- triage_df %>%
  group_by(SampleID) %>%
  summarize(Reason = paste(unique(Reason), collapse = "; "), .groups = "drop")

triage_df$AlamarWarning <- triage_df$SampleID  %in% warn_samps$`Sample Name`
# Print to RMarkdown
knitr::kable(
  triage_df, 
  caption = "TRIAGED SAMPLES"
)

warn_samps$Triage <- warn_samps$`Sample Name` %in% triage_df$SampleID
knitr::kable(
  warn_samps %>% select(`Sample Name`, `IC Median`, Detectability, `QC Status`,Triage), 
  caption = "ALAMAR WARN SAMPLES"
)

#Flagged excessive_matching_reads (not removed)
read_outliers$Triage <- read_outliers$SampleID %in% triage_df$SampleID
knitr::kable(
  read_outliers %>% select(SampleID, Run, Bay, Reads, Triage), 
  caption = "FLAGGED SAMPLES"
)
  
# Save to CSV
write_csv(triage_df, "output_files/samples_to_triage.csv")

# Remove triaged samples from main dataset
combined_qc_filtered <- combined_qc %>%
  rownames_to_column("SampleID") %>%
  filter(!SampleID %in% triage_df$SampleID) %>%
  column_to_rownames("SampleID")

#Add count summary
qc_summary <- combined_qc_filtered %>%
  group_by(Run, Bay) %>%
  summarise(combined_qc_filtered = n(), .groups = "drop") %>%
  arrange(Run, Bay)

knitr::kable(
  qc_summary, 
  caption = "Summary of post-QC counts:"
)

```

## D.4 Per-plate QC Summary
One row per plate: post-QC retained count, our triaged and flagged samples, and Alamar warnings.

```{r per_plate_summary, message=FALSE, warning=FALSE, echo=TRUE}
plate_lookup <- alamar_qc %>%
  transmute(SampleID = `Sample Name`,
            Run = stringr::str_extract(plateID, "^[0-9]{8}-[0-9]{4}"),
            Bay = stringr::str_extract(plateID, "Bay[0-9]+"))

count_by_plate <- function(ids, nm) {
  tibble(SampleID = unique(as.character(ids))) %>%
    inner_join(plate_lookup, by = "SampleID") %>%
    count(Run, Bay, name = nm)
}

plate_summary <- plate_lookup %>%
  distinct(Run, Bay) %>%
  left_join(qc_summary %>% rename(Retained_postQC = combined_qc_filtered), by = c("Run","Bay")) %>%
  left_join(count_by_plate(triage_df$SampleID,       "Triaged"),         by = c("Run","Bay")) %>%
  left_join(count_by_plate(read_outliers$SampleID,   "Flagged_Reads"),   by = c("Run","Bay")) %>%
  left_join(count_by_plate(warn_samps$`Sample Name`, "Alamar_Warnings"), by = c("Run","Bay")) %>%
  mutate(across(c(Retained_postQC, Triaged, Flagged_Reads, Alamar_Warnings),
                ~ tidyr::replace_na(., 0L))) %>%
  arrange(Run, Bay)

knitr::kable(
  plate_summary,
  caption = "Per-plate QC summary: pipeline triage/flags vs. Alamar warnings"
) %>%
  kableExtra::kable_styling(full_width = TRUE,
                            bootstrap_options = c("striped", "hover", "condensed"))

write_csv(plate_summary, "output_files/per_plate_QC_summary.csv")
```

# E. Biomarker QC
Reintegrate biomarkers and separate into groups.

## E.1 Re-integrate High-Corr Biomarkers 
Here we add back in high-correlation biomarkers that were removed for the QC Biomarker Set. These all should have good detectability and raw read counts.

```{r add_high_cor, message=FALSE, warning=FALSE, echo=TRUE}
# Add back high-correlated biomarkers to combined_qc_filtered
library(tibble)

# Ensure only rows in combined_qc_filtered are selected
high_cor_subset <- combined_df[rownames(combined_qc_filtered), high_cor, drop = FALSE]

# Add these columns to combined_qc_filtered
combined_qc_full <- cbind(combined_qc_filtered, high_cor_subset)
# Move SampleIDs to first column
combined_qc_full <- rownames_to_column(combined_qc_full, var = "SampleID")

knitr::kable(
  high_cor, 
  caption = "High-correlation biomarkers added back:"
)
dim(combined_qc_full)[2]-3 #NUMBER OF BIOMARKERS IS COLUMNS-3
```

## E.2 Low-detectability Biomarkers

### E.2.1 Re-integrate Det>50% Biomarkers 
Alamar suggests that the detectability threshold should be 50%. Here we add back in Biomarkers with minimum detectability>50% that were dropped to create the QC Biomarker set. We also require these to have raw read counts above the threshold. Those with low counts are further examined in later QC steps.

```{r add_med_det, message=FALSE, warning=FALSE, echo=TRUE}
# Add back biomarkers with >50% Min_Detectability ("medium detectability")
med_det <- low_det[low_det$Min_Detectability > min_detectability, ]

# Split into "good" and "still low count"
good_detect <- med_det[!(med_det$Biomarker %in% low_counts$Biomarker), ]
med_det_low_counts <- med_det[(med_det$Biomarker %in% low_counts$Biomarker), ]

# Extract good biomarker names
good_detect_names <- good_detect$Biomarker

# Subset combined_df to keep only good samples and these biomarkers
med_det_subset <- combined_df[rownames(combined_qc_filtered), good_detect_names, drop = FALSE]

# Add these columns back to the filtered QC dataset
combined_qc_full <- cbind(combined_qc_full, med_det_subset)
rownames(combined_qc_full) <- NULL

# Confirm dimensions
dim(combined_qc_full)[2]-3 #NUMBER OF BIOMARKERS IS COLUMNS-3

# Detectability metadata tables for each group
med_detect_table <- detect[detect$Biomarker %in% good_detect_names, ]
med_det_low_counts_table <- detect[detect$Biomarker %in% med_det_low_counts$Biomarker, ]

# View tables
knitr::kable(
  med_detect_table, 
  caption = "Medium-detectability (min det>50%) biomarkers added back:"
)
knitr::kable(
  med_det_low_counts_table, 
  caption = "Medium-detectability biomarkers with low counts to be examined below:"
)

```

### E.2.2 Re-examine Det<50% Biomarkers 
Look at these for possible re-integration, e.g., those with some expected low detectability.

```{r check_low_det, message=FALSE, warning=FALSE, echo=TRUE}
# Check biomarkers with <50% Min_Detectability to see if any should be added back in
library(kableExtra)

low_low_det <- low_det[low_det$Min_Detectability <= min_detectability, ]

# Split into "low detctability" and "low detectabiloty with low count"
low_detect <- low_low_det[!(low_low_det$Biomarker %in% low_counts$Biomarker), ]
low_det_low_counts <- low_low_det[(low_low_det$Biomarker %in% low_counts$Biomarker), ]

# Extract good biomarker names
low_detect_names <- low_detect$Biomarker

# Detectability metadata tables for each group
low_detect_table <- detect[detect$Biomarker %in% low_detect_names, ]
low_det_low_counts_table <- detect[detect$Biomarker %in% low_det_low_counts$Biomarker, ]

# View tables
knitr::kable(
  low_detect_table, 
  caption = "Low-detectability (min det>50%) biomarkers added back:"
)
knitr::kable(
  low_det_low_counts_table, 
  caption = "Low-detectability biomarkers with low counts to be examined below:"
)

# Look for bimodal distributions and extreme outliers. Note differences across plates. Also LOD for these. 

#Pull LODs
# Read annotation data
annot <- read_csv("input_files/Annotation_Targets.csv")

# Select required columns
annot_sel <- annot %>%
  select(targetName, targetDetectability, targetLOD, targetLOD_NPQ, plateID)

 # Extract Run and Bay from plateID (assuming "YYYYMMDD_BayX" or similar format)
#CHANGED THIS FOR ALZ2 TO Allow for Runs with same prefix but different plates
annot_sel <- annot_sel %>%
  mutate(Run = str_extract(plateID, "\\d{8}-\\d+"),
         Bay = str_extract(plateID, "Bay\\d+"),
         RunBay = paste0(Run, "_", Bay))

# Make wide format table: LOD per Run/Bay
lod_matrix <- annot_sel %>%
  select(targetName, RunBay, targetLOD_NPQ) %>%
  distinct() %>%   # remove any duplicates
  pivot_wider(names_from = RunBay,
              values_from = targetLOD_NPQ)

# LOD for Reads
lod_reads_matrix <- annot_sel %>%
  select(targetName, RunBay, targetLOD) %>%
  distinct() %>%   # remove any duplicates
  pivot_wider(names_from = RunBay,
              values_from = targetLOD)

# Extract non-targetName columns
cols <- colnames(lod_matrix)[-1]

# Extract run date + bay index as ordering helpers
run_date <- str_extract(cols, "\\d{8}")
bay_num <- str_extract(cols, "(?<=Bay)\\d+") |> as.numeric()

# Create order index based on date then Bay number
col_order <- order(run_date, bay_num)

# Apply reordering: keep targetName first
lod_matrix <- lod_matrix %>%
  select(targetName, all_of(cols[col_order]))

# Reorder lod_reads_matrix columns to exactly match lod_matrix
lod_reads_matrix <- lod_reads_matrix %>%
  select(all_of(colnames(lod_matrix)))

write.csv(lod_matrix,"output_files/lod_matrix.csv", row.names = TRUE)
write.csv(lod_reads_matrix,"output_files/lod_reads_matrix.csv", row.names = TRUE)

cat("Plots of Low-detectability (min det<50%) biomarkers with good read counts:\n")
# Keep only rows in combined_qc_filtered and select relevant columns
combined_low_det <- combined_df %>%
  filter(rownames(.) %in% rownames(combined_qc_filtered)) %>%
  rownames_to_column("SampleID") %>%
  select(
    SampleID,
    Run,
    Bay,
    all_of(low_detect_names)
  )

# Ensure Run and Bay are factors
combined_low_det$Run <- factor(combined_low_det$Run)
combined_low_det$Bay <- factor(combined_low_det$Bay)

# Custom Bay colors
bay_colors <- c("Bay1" = "red", "Bay2" = "green", "Bay3" = "blue")

# Identify only biomarker columns
biomarker_cols <- setdiff(names(combined_low_det), c("SampleID", "Run", "Bay"))

# Loop and create violin plots for each biomarker
for (biom in biomarker_cols) {
  
  df_temp <- combined_low_det %>%
    select(Value = all_of(biom), Run, Bay)
  
  p <- ggplot(df_temp, aes(x = Run, y = Value, fill = Bay)) +
    geom_violin(alpha = 0.5, trim = FALSE, position = position_dodge(width = 0.8)) +
    geom_jitter(aes(color = Bay), 
                position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
                alpha = 0.75, size = 1.3) +
    scale_fill_manual(values = bay_colors) +
    scale_color_manual(values = bay_colors) +
    theme_minimal() +
    labs(
      title = paste("Run × Bay Effects:", biom),
      x = "Run",
      y = biom
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p)

 # ----- Print per-biomarker LOD table -----
lod_npq_vals <- lod_matrix[lod_matrix$targetName == biom, -1, drop = FALSE]
lod_read_vals <- lod_reads_matrix[lod_reads_matrix$targetName == biom, -1, drop = FALSE]

lod_table <- rbind(
  "NPQ LOD" = lod_npq_vals,
  "Read LOD" = lod_read_vals
)

# Print LOD table
print(cbind(rownames(lod_table),lod_table)) # I tried to make this prettier but it doesn't work in a loop
}

```


## E.3 Low Raw-Count Biomarkers  
Look at these more closely in secondary QC. Many are expected, because the protein is rare or they are due to a polymorphism, e.g., APOE4, BD-MAPT.

I'm showing NPQ values here for consistency, not raw read counts. Read counts will be examined further in secondary QC.
```{r check_low_raw_counts, message=FALSE, warning=FALSE, echo=TRUE}
# Examine low-count biomarkers 

# Vector of low-count biomarker names
low_counts_vec <- low_counts[[1]]  # assuming low_counts is a data frame with one column

# Keep only rows in combined_qc_filtered and select relevant columns
combined_low_counts <- combined_df %>%
  filter(rownames(.) %in% rownames(combined_qc_filtered)) %>%
  rownames_to_column("SampleID") %>%
  select(
    SampleID,
    Run,
    Bay,
    all_of(low_counts_vec)
  )

# Ensure Run and Bay are factors
combined_low_counts$Run <- factor(combined_low_counts$Run)
combined_low_counts$Bay <- factor(combined_low_counts$Bay)

# Custom Bay colors
bay_colors <- c("Bay1" = "red", "Bay2" = "green", "Bay3" = "blue")

# Identify only biomarker columns
biomarker_cols <- setdiff(names(combined_low_counts), c("SampleID", "Run", "Bay"))

# Loop and create violin plots for each biomarker
for (biom in biomarker_cols) {
  
  df_temp <- combined_low_counts %>%
    select(Value = all_of(biom), Run, Bay)
  
  p <- ggplot(df_temp, aes(x = Run, y = Value, fill = Bay)) +
     geom_violin(alpha = 0.5, trim = FALSE, position = position_dodge(width = 0.8)) +
     geom_jitter(aes(color = Bay), 
                position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
                alpha = 0.75, size = 1.5) +
    scale_fill_manual(values = bay_colors) +
    scale_color_manual(values = bay_colors) +
    theme_minimal() +
    labs(
      title = paste("Run × Bay Effects:", biom),
      x = "Run",
      y = biom
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p)
  # Print the row from combined_low_counts corresponding to this biomarker
  print(knitr::kable(low_counts[low_counts$Biomarker == biom, , drop = FALSE],caption = "Reads"))
}

```

# F. Prepare Output  
Output the QC'd dataset with samples from "Sample QC" removed. All biomarkers are included, but separated into two files: 1) NPQxxx_post_QC.csv and 2) NPQxxx_Low_post_QC.csv. 

1) Contains all biomarkers with minimum detectability >50% across plates for QC'd samples that have mean and median raw-read counts>500 across all plates.

2) Contains all biomarkers with minimum detectability <= 50% across plates and/or mean or median raw-read counts <= 500 across all plates or QC'd samples.

3) Contains all biomarkers in 1 for triaged samples.

4) Contains all biomarkers in 2 for triaged samples. 

Low detectability and low count biomarkers are in a separate dataset for "analysis". They are not necessarily bad, but will need to be analyzed differently from continuous biomarkers.

```{r make_output, message=FALSE, warning=FALSE, echo=TRUE}
# Here we save the QC'd files
# We may want to transpose them to be like original NPQ
library(dplyr)
library(tibble)

knitr::kable(
  summarize_counts(combined_qc_full, "QC'd Samples"), 
  caption = "POST-QC SAMPLES"
)

#Output file 1

cat("QC'd Biomarkers:", dim(combined_qc_full)[2]-3, "\n")
qc_file <- file.path("output_files", sub("\\.csv$", "_post_QC.csv", NPQ_data))
write.csv(combined_qc_full, qc_file, row.names = TRUE)
cat("Saved:", qc_file, "\n")

#Output file 2

# Remove duplicated metadata columns from the second dataset
combined_low_counts_nometadata <- combined_low_counts %>%
  select(-Run, -Bay)
# Perform the join on SampleID
combined_qc_low <- combined_low_det %>%
  left_join(combined_low_counts_nometadata, by = "SampleID")

cat("QC'd Low-count/Low-detectability Biomarkers:", dim(combined_qc_low)[2]-3, "\n")
qc_file <- file.path("output_files", sub("\\.csv$", "_Low_post_QC.csv", NPQ_data))
write.csv(combined_qc_low, qc_file, row.names = FALSE)
cat("Saved:", qc_file, "\n")

# Output file 3: TRIAGE × Full QC biomarkers

# sample list
triage_samples <- triage_df$SampleID
combined_df <- rownames_to_column(combined_df, "SampleID")

triage_qc_full <- combined_df %>%
  filter(SampleID %in% triage_samples) %>%
  select(any_of(colnames(combined_qc_full)))

triage_full_file <- file.path(
  "output_files", 
  sub("\\.csv$", "_post_QC_triage.csv", NPQ_data)
)

write.csv(triage_qc_full, triage_full_file, row.names = FALSE)
cat("Saved TRIAGE full-QC biomarker file:", triage_full_file, "\n")

# Output file 4: TRIAGE × Low QC biomarkers

triage_qc_low <- combined_df %>%
  filter(SampleID %in% triage_samples)%>%
  select(any_of(colnames(combined_qc_low)))

triage_low_file <- file.path(
  "output_files", 
  sub("\\.csv$", "_Low_post_QC_triage.csv", NPQ_data)
)

write.csv(triage_qc_low, triage_low_file, row.names = FALSE)
cat("Saved TRIAGE low-QC biomarker file:", triage_low_file, "\n")

```

# G.  Notes

Combined Data: