---
title: "Primary QC of Biomarker Data for ALZ123 Combined Datasets"
author: "ERM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
knit: |
  (function(input, ...) {
    npqs <- list.files("input_files", pattern = "^NPQ.*\\.csv$", full.names = TRUE)
    npqs <- npqs[!grepl("Low|Raw|Sample_QC|triage", basename(npqs), ignore.case = TRUE)]
    tag  <- if (length(npqs)) tools::file_path_sans_ext(basename(npqs[which.max(file.mtime(npqs))])) else format(Sys.Date(), "%Y%m%d")
    rmarkdown::render(input, output_file = paste0("QC_Pipeline_Primary_Alamar_", tag, ".html"))
  })
---
# A. Overview

This protocol describes a workflow for primary **quality control (QC)** of biomarker datasets measured across multiple assay plates (Runs and Bays). 

* The protocol first identifies a set of high-quality biomarkers with low correlation as the **QC Biomarker Set**. 

* Next it uses the QC Biomarker Set to identify samples to "triage" for further inspection based on automated checks (e.g., PCA, outliers)

* It further identifies samples that should be "flagged", i.e., noted but not as bad as triage.

* All biomarkers are then reintegrated, but separated into two sets: 1) those with low detectability and/or low read count, and 2) all remaining biomarkers.

* Finally, post-QC NPQ data are written to output files, separated by the two biomarker groups as well as triaged and non-triaged samples.

QC analyses are **agnostic** to phenotypic and covariate information.

**Two sections audit the pipeline rather than the samples** (added 2026-07-30,
implementing the recommendations in `Secondary_QC/Triage_Review/`):

* **D.0** measures **whole-panel intensity** (`mean_INT`) on every well *before*
  triage. Every gate below asks a count or distance question; none asks a level
  question, so a sample that reads uniformly bright across the panel is invisible
  to all of them. `mean_INT` makes that axis visible. It is **reported, never
  gated** — it is a covariate for the analysis model, not a drop rule.

* **D.5** audits the screens themselves: does each gate fire or is it **inert**,
  is it independent of the others or a vendor metric re-derived, where does its
  cut sit on its own statistic, and is it keying on **plate position** rather
  than on the specimen. Diagnostic only — it removes nothing.

Both are why several gates below now report *headroom to the cut* alongside a
count. A gate that fired on zero wells has not been shown to be safe; it has not
been tested, and the two look identical from the count alone.

**For a new QC run**, set the NPQ input filename in section A.3 (and optionally the title in YAML). You **no longer need to rename this file** — the `knit:` function in the YAML header auto-names the HTML after the NPQ input dataset (e.g. `QC_Pipeline_Primary_Alamar_NPQ_20251220.html`), so each run's report is unique and won't overwrite a previous one.


## A.1 Input Data

**You must specify the input NPQ filename below in section A.3**

* Example: NPQ_data <- "NPQ_20250923.csv" 

**Required Input Data Files:** Place all files in a folder called `input_files` in the working directory.

1. **Biomarker NPQ data** (`NPQ_xxxxxxxx.csv`) – Rows: biomarkers; Columns: samples; downloaded from NAS. Assign the filename to the variable `NPQ_data` below.
2. **Alamar sample QC info** (`Sample_QC.csv`) – Rows: biomarkers; Columns: annotation information (IC Median, Detactability, QC Status, etc.). Exported from NPQ_Values.xlsx.
3. **Raw count data** (`Raw_counts.csv`) – Rows: biomarkers; Columns: samples; exported from NPQ_Values.xlsx.
4. **Detectability summary** (`biomarker_detectability.csv`) – Rows: biomarkers; Columns: detectability per plate and overall.
5. **Annotation for biomarkers** (`Annotation_Targets.csv`) – Rows: samples; Columns: annotation information (LOD, NPQ values, plateID, etc.). Exported from NPQ_Values.xlsx.
6. **Replicate list** (`Replicate_list.csv`) – List of HIHG-replicate sample IDs for assay precision checks. Created by analyst.

* Previously, I included `QC_droplist.csv` to flag samples with warnings from Alamar. I'm now going to automate this and list them in the summary. 

**Output Data Files:** Analyst should create a folder `output_files` in the working directory.

1. `NPQxxx_post_QC.csv` – QC’d samples (without triaged samples) with biomarkers having **Min Detectability >50%** and **mean/median raw-read counts >500**.
2. `NPQxxx_Low_post_QC.csv` – QC'd Samples (without triaged samples) failing detectability or read count thresholds.
3. `NPQxxx_post_QC_triage.csv` – Same as above from triaged samples.
4. `NPQxxx_Low_post_QC_triage.csv` – Same as above from triaged samples.
5. Additional ancillary files for outliers, PCA coordinates, flagged samples, etc.
6. I'd like to add a pdf summary to export with final data.

**Audit and covariate outputs** (D.0 and D.5 — none of these drop a sample):

| file | what it answers |
|---|---|
| `well_level_QC.csv` | **one row per well**: every gate flag, `mean_INT`, position, reads, IC. The join key for anything downstream — above all clustering by *clinical collection site*, which needs the metadata merge |
| `whole_panel_intensity.csv` | `mean_INT` / `mean_INT_z` for all wells, triaged included |
| `qc_gate_inventory.csv` | does each gate fire, or is it **INERT** (untested, not proven safe), and how much headroom to its cut |
| `qc_gate_duplication.csv` | is our IC gate the vendor's `IC Median` re-derived (Jaccard + verdict) |
| `qc_gate_overlap.csv` | pairwise Jaccard between screens — which ones are one screen reported twice |
| `triage_axis_overlap.csv` | true per-axis totals vs what a one-reason-per-well report shows |
| `triage_axis_combinations.csv` | the full Venn, flattened |
| `burden_operating_points.csv` | the complete list of `FDR_threshold` settings that exist — it is not a dial |
| `gate_cut_neighbourhood.csv` | nearest retained and nearest removed well on each axis |
| `read_gate_geometry.csv` | per plate-bay, whether the lower read bound is plate-relative or the absolute floor |
| `qc_axis_attribution.csv` | each triage class on the intensity axis (d, AUC) — **per class, never pooled** |
| `well_position_effects.csv` | per-position triage rates + permutation p |
| `well_position_effects_readflag.csv` | the same for read flags |
| `position_map_IC.csv` | median IC recovery by plate position |
| `ic_row_gradient.csv` | IC recovery down the plate, per plate-bay |


## A.2 Thresholds

Adjustable QC parameters include:

| Parameter | Default | Description |
|-----------|---------|-------------|
| `PCA_SD` | 5 | Number of SDs from mean to define PCA outliers |
| `min_detectability` | 50 | Minimum detectability (%) to retain biomarker in post-QC dataset |
| `min_detectability_qc` | 98 | Minimum detectability (%) to retain biomarker in QC Biomarker dataset |
| `read_count_threshold` | 500 | Minimum mean/median raw reads per sample for QC |
| `corr_thresh` | 0.4 | Maximum correlation coefficient allowed between biomarkers in QC Biomarker dataset |
| `samp_out_thresh` | 1.5 | Multiplier for IQR to determine if a sample if an outlier from the median NPQ for a biomarker|
| `read_out_thresh` | 4 | Multiplier for IQR to determine if a sample is a **high**-read outlier from the plate median |
| `read_floor_abs` | 500000 | Absolute low-read floor. The **only** lower read bound — see D.2.4 for why a plate-relative one cannot work |
| `FDR_threshold` | 0.01 | Significance threshold for outlier burden |
| `scaffold_mode` | `"frozen"` | QC Biomarker Set selection: `"frozen"` (fixed 21-anchor scaffold + per-plate tripwire) or `"rederive"` (per-run derive + de-correlate). Changes only the sample-triage anchors, not output biomarker content. |
| `intensity_sd_flag` | 2.5 | SDs of `mean_INT_z` for the whole-panel intensity flag. **Informational only — never triages** (D.0) |
| `ic_median_thresh` | 0.40 | Alamar's own `IC Median` bar, used in D.2.3 to test whether our IC gate is a re-derivation of it. **Overridden by the value in the export's preamble when present** |
| `n_perm` | 10000 | Permutation replicates for the plate-position tests in D.5.4 |
| `prior_region` | `NULL` | Optional plate region named *in advance* for a confirmatory contrast in D.5.4. Leave `NULL`; the layout-agnostic spatial tests always run |

**Vendor thresholds are not parameters.** `Detectability`, `IC Reads`, `Reads` and
`IC Median` bars are read from the `Sample_QC.csv` preamble at load time, and the
detectability bar is resolved against the run's own `SAMPLE_MATRIX` — so a CSF
run gets the vendor's 70% bar rather than plasma's 90%. Defaults are used only if
the preamble cannot be parsed.


## A.3 Initial Setup 

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# Input NPQ filename

NPQ_data <- "NPQ_20251220.csv" #SPECIFY THE FILE NAME CONTAINING NPQ DATA

# Create output folder if it doesn't exist
if (!dir.exists("output_files")) dir.create("output_files")

# Thresholds
PCA_SD <- 5
min_detectability <- 50
min_detectability_qc <-98
corr_thresh <- 0.4
read_count_threshold <- 500
samp_out_thresh <- 1.5
read_out_thresh <- 4
FDR_threshold <- 0.01

# Absolute low-read floor. Kept as a NAMED constant rather than buried in a pmax()
# because it is the only lower read bound this pipeline has — see D.2.4.
read_floor_abs <- 500000

# Whole-panel intensity (D.0). INFORMATIONAL. `mean_INT` is a level statistic and
# is deliberately NOT a gate: background intensity is a continuous nuisance
# covariate affecting every sample, not a population of bad wells. It belongs in
# the downstream analysis model. The flag exists to make the axis visible.
intensity_sd_flag <- 2.5

# Alamar's own IC bar (|IC Median| > 0.40 = outside +/-40% of the plate median).
# Used in D.2.3 to measure how far our IC gate is an independent check.
ic_median_thresh <- 0.40

# Permutation replicates for the plate-position tests (D.5.4). Triage counts over
# 84-96 positions leave most cells at 0 or 1, where a chi-square is not
# conservative but wrong.
n_perm <- 10000

# OPTIONAL confirmatory check of a plate region named IN ADVANCE (D.5.4).
#
# Leave NULL for a normal run. D.5.4 always runs layout-agnostic spatial tests
# (row gradient, column gradient, edge vs interior) and always reports the
# empirically worst region descriptively — none of which needs this.
#
# Set it only to re-test a region a PREVIOUS run flagged, which makes the p-value
# confirmatory rather than a search. Hardcoding a region found in this same data
# would be circular, which is why there is no default:
#   prior_region <- list(rows = c("F","G","H"), cols = 9:12)   # 50-plate frozen, 2026-07-27
prior_region <- NULL

# QC Biomarker Set (scaffold) selection — how the markers used for SAMPLE triage are chosen.
#   "frozen"   : (default) a fixed, pre-vetted, de-correlated anchor set (stable across batches) + a
#                per-plate tripwire. Set size is stable regardless of plate count.
#   "rederive" : per-run derivation (the original behavior) — min detectability >= min_detectability_qc on
#                every plate, mean/median reads > read_count_threshold, then de-correlated (findCorrelation,
#                corr_thresh). Set size drifts with plate count.
# Either way only the triage ANCHORS change; the biomarker CONTENT of the post-QC outputs is identical.
scaffold_mode <- "frozen"

# Frozen scaffold: 21 markers >=98% detectable on all 50 ALZ123+8-plate+7-plate plates, de-correlated at
# |r|<=corr_thresh on the combined study samples (APOE excluded: relative-quant + Bay1 anomaly). See
# QC_Runs/Tests/QC_frozen_scaffold_06_17_26/. Used only when scaffold_mode == "frozen".
frozen_scaffold <- c("CCL2","CCL22","CCL3","CCL4","CHI3L1","CNTN2","CXCL8","DDC","FLT1","IFNG","IL10",
                     "IL5","KDR","MSLN","NPY","PDGFRB","PRDX6","SAA1","SFTPD","SMOC1","TIMP3")
# Per-plate tripwire thresholds (anchor x plate detectability %): an anchor cell in [drop, watch) is a
# WATCH (kept, flagged); below drop it is a DROP (not used as an anchor on this run, flagged).
scaffold_watch <- min_detectability_qc   # 98 — the selection bar
scaffold_drop  <- 90

# --- Well-level read-depth screen (C.1 / D.3) ---------------------------------------------------
# A well that failed to sequence still occupies a column in the export, and Alamar's
# `targetDetectability` is computed over ALL wells. So a handful of dead wells drags EVERY target's
# per-plate detectability down by (dead / total) and can empty the QC anchor set — which then
# prevents the very sample triage that would have removed those wells (2026-09-01: 3 plates with
# a handful of dead wells on a few plates took a large combined run from 18/21 frozen anchors to 1/21,
# and 55 qualifying biomarkers down to 4).
#
# The screen is on READ DEPTH, never on detectability: detectability moves when LODs move, so
# filtering on it is unstable across lots and assay versions. The bar is the vendor's own `Reads`
# minimum, parsed from the export preamble (`vendor_thresholds$reads_min`, 500,000) — not a number
# of ours.
#
# Screened wells are NOT deleted. They ride through the pipeline and exit via the normal D.3 triage
# on a `Low read depth` axis, so they land in `samples_to_triage.csv` and the `*_post_QC_triage.csv`
# datasets with their biomarker values intact, ready for Metadata_Merge and the Secondary_QC
# Triage_Metadata / Triage_Review modules.
apply_read_floor <- TRUE

# How per-plate target detectability is obtained:
#   "recompute" : mean(NPQ >= targetLOD_NPQ) per (plate, target) over retained study-sample wells.
#                 Reproduces the vendor column to ~0.17 percentage points when no wells are screened
#                 (validated 2026-09-01, 67 plates), and is the only option that can honour the
#                 read floor, since the vendor column is precomputed over every well.
#   "vendor"    : the `targetDetectability` column as shipped, via biomarker_detectability.csv
#                 (pre-2026-09-01 behaviour; ignores apply_read_floor).
detect_source <- "recompute"
```
---
  
# B. Data Formatting
This section formats the data and produces initial summaries.

Input file is: `r NPQ_data`

```{r load-and-format, message=FALSE, warning=FALSE, echo=TRUE}
library(tidyverse)

# --- Read biomarker NPQ data ---
df <- read_csv(file.path("input_files", NPQ_data), show_col_types = FALSE)

# First column = biomarker names
biomarker_names <- df[[1]]
df <- df[, -1]

# Make matrix (rows = biomarkers, cols = samples)
combined_numeric <- as.matrix(df)
rownames(combined_numeric) <- biomarker_names


# Transpose to samples in rows
combined_t <- t(combined_numeric)

# Convert to data.frame
combined_df <- as.data.frame(combined_t)

# --- Read Alamar Sample QC (per-sample QC stats + sample info) ---
# Robust to export drift: the Alamar export prepends a variable-length "metric" preamble before the
# real header row. Instead of a hardcoded skip, auto-detect the header row (the one containing
# "Sample Name") and validate that the expected columns are present, so a format change fails loudly
# with a clear message rather than cryptically downstream.
.sqc_path <- "input_files/Sample_QC.csv"
.sqc_peek <- readLines(.sqc_path, n = 40, warn = FALSE)
.sqc_skip <- which(grepl('(^|,)\\s*"?Sample Name"?\\s*(,|$)', .sqc_peek))[1] - 1L
if (is.na(.sqc_skip)) {
  .sqc_skip <- 12L
  warning("Sample_QC.csv: 'Sample Name' header not found in first 40 lines; falling back to skip=12.")
}
alamar_qc <- read_csv(.sqc_path, skip = .sqc_skip, show_col_types = FALSE)
message("Sample_QC.csv: header auto-detected at line ", .sqc_skip + 1L, " (skip=", .sqc_skip, ").")

# --- Vendor thresholds, read from the export instead of hardcoded --------------
# That preamble we skip over is not noise: it states Alamar's own QC bars, e.g.
#   Detectability ... (Minimum threshold = plasma-90%, serum-90%, csf-70%, other-0%)
#   IC Reads      ... (Minimum threshold = 1,000)
#   Reads         ... (Minimum threshold = 500,000)
#   IC Median     ... (Within +/-40% of the plate median)
# Parsing them keeps the audit in D.5.1 correct when Alamar changes a bar, and —
# more importantly — makes the detectability bar MATRIX-SPECIFIC. A hardcoded 90%
# silently mis-scores a CSF run, where the vendor bar is 70%.
# Match on the metric NAME at the start of its own preamble row, not on the
# joined text: "reads within a sample" appears in both the `Reads` row and the
# `IC Reads` row ("...(IC) reads within a sample..."), so a blob-wide search
# returns the IC bar of 1,000 for `Reads` instead of 500,000.
.pre <- .sqc_peek[seq_len(max(1L, .sqc_skip))]
.metric_line <- function(nm) {
  i <- grep(paste0('^\\s*"?', nm, '"?\\s*,'), .pre, perl = TRUE)
  if (length(i)) .pre[i[1]] else NA_character_
}
.thresh <- function(nm, default) {
  ln <- .metric_line(nm); if (is.na(ln)) return(default)
  m <- regmatches(ln, regexpr("Minimum threshold\\s*=\\s*([0-9,]+)", ln, perl = TRUE))
  if (!length(m)) return(default)
  v <- suppressWarnings(as.numeric(gsub("[^0-9]", "", m)))
  if (is.na(v)) default else v
}
vendor_thresholds <- list(ic_reads_min = .thresh("IC Reads", 1000),
                          reads_min    = .thresh("Reads", 500000))

.icm <- .metric_line("IC Median")
vendor_thresholds$ic_median <- if (!is.na(.icm)) {
  m <- regmatches(.icm, regexpr("\\+/-\\s*([0-9.]+)\\s*%", .icm, perl = TRUE))
  if (length(m)) as.numeric(gsub("[^0-9.]", "", m)) / 100 else 0.40
} else 0.40

# Detectability bar per sample matrix, from the same row.
.dln <- .metric_line("Detectability")
.det <- if (!is.na(.dln))
  regmatches(.dln, gregexpr("([A-Za-z]+)\\s*-\\s*([0-9.]+)\\s*%", .dln, perl = TRUE))[[1]] else character(0)
vendor_thresholds$detectability_by_matrix <- if (length(.det)) {
  stats::setNames(as.numeric(sub("^.*-\\s*([0-9.]+)\\s*%$", "\\1", .det)) / 100,
                  toupper(sub("^\\s*([A-Za-z]+)\\s*-.*$", "\\1", .det)))
} else c(PLASMA = 0.90, SERUM = 0.90, CSF = 0.70, OTHER = 0)

# Resolve the bar for THIS run from its own SAMPLE_MATRIX column.
.mx <- if ("SAMPLE_MATRIX" %in% names(alamar_qc))
  toupper(names(sort(table(alamar_qc$SAMPLE_MATRIX), decreasing = TRUE))[1]) else "OTHER"
vendor_thresholds$sample_matrix <- .mx
vendor_thresholds$detectability_min <- unname(
  if (.mx %in% names(vendor_thresholds$detectability_by_matrix))
    vendor_thresholds$detectability_by_matrix[[.mx]]
  else vendor_thresholds$detectability_by_matrix[["OTHER"]])

message(sprintf("Vendor bars read from the export: matrix %s, detectability %.0f%%, IC Reads %s, Reads %s, IC Median +/-%.0f%%",
                .mx, 100 * vendor_thresholds$detectability_min,
                format(vendor_thresholds$ic_reads_min, big.mark = ","),
                format(vendor_thresholds$reads_min, big.mark = ",", scientific = FALSE),
                100 * vendor_thresholds$ic_median))

# The in-house IC gate is a re-derivation of the vendor's IC Median bar (D.2.3),
# so it should track the vendor's number rather than carry its own copy.
if (!isTRUE(all.equal(ic_median_thresh, vendor_thresholds$ic_median))) {
  message(sprintf("ic_median_thresh %.2f differs from the export's +/-%.0f%% -- using the export's.",
                  ic_median_thresh, 100 * vendor_thresholds$ic_median))
  ic_median_thresh <- vendor_thresholds$ic_median
}

.sqc_miss <- setdiff(c("plateID", "Sample Name", "QC Status"), names(alamar_qc))
if (length(.sqc_miss))
  stop("Sample_QC.csv is missing expected column(s): ", paste(.sqc_miss, collapse = ", "),
       ".\n  Columns found: ", paste(names(alamar_qc), collapse = ", "),
       "\n  The export format may have changed — check the header row / column names.")

# Export drift in value formatting: `IC Median`/`Detectability` may be percent STRINGS
# ("-4.6%", "95.4%") or numeric; `IC Reads`/`Reads` may carry thousands separators. Coerce to numeric
# (parse_number strips "%" and commas; genuinely numeric input passes through unchanged).
.to_num <- function(x) if (is.numeric(x)) x else readr::parse_number(as.character(x))
alamar_qc <- alamar_qc %>%
  mutate(across(any_of(c("IC Median", "Detectability", "IC Reads", "Reads")), .to_num))

# Extract Run and Bay from plateID
meta_df <- alamar_qc %>%
  mutate(
    Run = stringr::str_extract(plateID, "^[0-9]{8}-[0-9]{4}"),
    Bay = stringr::str_extract(plateID, "Bay[0-9]+")
  ) %>%
  transmute(SampleID = `Sample Name`, Run, Bay)

# --- Well-level read-depth screen ---------------------------------------------
# Wells below the vendor's own `Reads` floor. Identified here (before C.1) because the
# detectability denominator must exclude them; they are removed from the analysis set only later,
# by the normal D.3 triage, so their data survives into the triage outputs. See A.3.
low_read_samples <- character(0)
if (isTRUE(apply_read_floor) && "Reads" %in% names(alamar_qc)) {
  # Study samples only. Negative controls are SUPPOSED to sequence shallow — every NC well sits below
  # the floor, so sweeping them in inflates the count by one NC per plate and buries the real failures.
  # Controls are pulled out separately in B.1.1 and are not triaged here.
  low_read_tbl <- alamar_qc %>%
    { if ("Sample Type" %in% names(.)) filter(., `Sample Type` == "Sample") else . } %>%
    filter(!is.na(Reads), Reads < vendor_thresholds$reads_min) %>%
    left_join(meta_df, by = c("Sample Name" = "SampleID")) %>%
    transmute(SampleID = `Sample Name`, Run, Bay,
              Well = if ("AUTO_WELLPOSITION" %in% names(.)) AUTO_WELLPOSITION else NA_character_,
              SampleType = if ("Sample Type" %in% names(.)) `Sample Type` else NA_character_,
              Reads, `IC Reads`, `IC Median`, Detectability, `QC Status`) %>%
    arrange(Run, Bay, Well)

  low_read_samples <- low_read_tbl$SampleID
  write_csv(low_read_tbl, "output_files/low_read_wells.csv")

  message(sprintf(
    "Read-depth screen: %d well(s) below the vendor Reads floor of %s (%d flagged Warning by Alamar's own QC). Excluded from the detectability denominator; triaged in D.3 on the 'Low read depth' axis.",
    nrow(low_read_tbl), format(vendor_thresholds$reads_min, big.mark = ",", scientific = FALSE),
    sum(low_read_tbl$`QC Status` != "Passed", na.rm = TRUE)))
}

# Keep only samples present in combined_df
meta_df <- meta_df[meta_df$SampleID %in% rownames(combined_df), ]

# Reorder to match rownames
meta_df <- meta_df[match(rownames(combined_df), meta_df$SampleID), ]

# Add Run and Bay columns at the front
combined_df$Run <- meta_df$Run
combined_df$Bay <- meta_df$Bay
combined_df <- combined_df %>% select(Run, Bay, everything())
rownames(combined_df) <- rownames(combined_t)

# --- Plate position lookup -----------------------------------------------------
# Kept as a SEPARATE lookup rather than extra columns on combined_df, because the
# whole pipeline addresses biomarker columns positionally (`combined_df[, -(1:2)]`)
# and widening the metadata block would silently shift those.
#
# Position is a technical property of the plate, not phenotype, so using it here
# does not break the "QC is agnostic to phenotypic and covariate information" rule
# in A. Clustering by *clinical collection site* is a different question and needs
# the metadata merge — it lives in Secondary_QC, fed by the well-level table D.5.6
# writes. See D.5.4.
well_pos <- alamar_qc %>%
  transmute(
    SampleID = `Sample Name`,
    Run      = stringr::str_extract(plateID, "^[0-9]{8}-[0-9]{4}"),
    Bay      = stringr::str_extract(plateID, "Bay[0-9]+"),
    Well     = if ("AUTO_WELLPOSITION" %in% names(.)) as.character(AUTO_WELLPOSITION)
               else paste0(wellRow, wellCol),
    wellRow  = as.character(wellRow),
    wellCol  = suppressWarnings(as.integer(wellCol))
  ) %>%
  filter(!is.na(SampleID)) %>%
  distinct(SampleID, .keep_all = TRUE)

if (all(is.na(well_pos$Well)))
  warning("No usable well position in Sample_QC.csv (need AUTO_WELLPOSITION, or wellRow+wellCol). ",
          "The plate-position diagnostics in D.5.4 will be skipped.")
```

## B.1 Initial Cleaning

### B.1.1 HIHG Replicates and Control Samples
Pull out Alamar control and HIHG replicate samples for examination in later pipeline stages.

```{r special_samples, message=FALSE, warning=FALSE, echo=TRUE}
library(dplyr)
library(tidyr)

# --- Read HIHG replicate list ---
rep <- read_csv("input_files/Replicate_list.csv", show_col_types = FALSE)
hihg_reps <- combined_df[rownames(combined_df) %in% rep[[1]], ]

combined_df <- combined_df[!rownames(combined_df) %in% rownames(hihg_reps), ]

# Identify Alamar controls and (SC, IPC, and NC) and pull out from main dataset
nc_samples   <- combined_df[grepl("^NC", rownames(combined_df)), ]
sc_samples   <- combined_df[grepl("^SC", rownames(combined_df)), ]
ipc_samples  <- combined_df[grepl("^IPC", rownames(combined_df)), ]

combined_df <- combined_df[!grepl("^(NC|IPC|SC)", rownames(combined_df)), ]

# Function to summarize counts by Run and Bay
summarize_counts <- function(df, type_name) {
  df %>%
    group_by(Run, Bay) %>%
    summarise(!!type_name := n(), .groups = "drop")
}

# Summarize each type
nc_summary   <- summarize_counts(nc_samples, "NC")
sc_summary   <- summarize_counts(sc_samples, "SC")
ipc_summary  <- summarize_counts(ipc_samples, "IPC")
hihg_summary <- summarize_counts(hihg_reps, "HIHG_Replicate")

# Combine into a single table
control_summary <- full_join(nc_summary, ipc_summary, by = c("Run","Bay")) %>%
                   full_join(sc_summary, by = c("Run","Bay")) %>%
                   full_join(hihg_summary, by = c("Run","Bay")) %>%
                   replace_na(list(NC = 0, IPC = 0, SC = 0, HIHG_Replicate = 0)) %>%
                   arrange(Run, Bay)

as.data.frame(control_summary)

#Write replicates and controls with their biomarker data to files

write_csv(hihg_reps, "output_files/hihg_reps_NPQ.csv")
write_csv(sc_samples, "output_files/sc_reps_NPQ.csv")
write_csv(nc_samples, "output_files/nc_reps_NPQ.csv")
write_csv(ipc_samples, "output_files/ipc_reps_NPQ.csv")

```
 
### B.1.2 Samples with Bad Data
Identify any samples with non-numeric entries, e.g., NA, NaN.

**Bad Samples:**

```{r bad-samples, message=FALSE, warning=FALSE, echo=TRUE}
# THIS IS A NEW SECTION. CHANGED
# Remove meta-data columns from numeric check
biom_only <- combined_df[ , -c(1,2), drop = FALSE]

# Identify rows with ONLY numeric values
is_numeric_row <- apply(biom_only, 1, function(r) all(suppressWarnings(!is.na(as.numeric(r)))))
bad_samples <- rownames(combined_df)[!is_numeric_row]
bad_rows_df <- combined_df[bad_samples, , drop = FALSE]

write_csv(bad_rows_df, "output_files/bad_samples.csv")

combined_df <- combined_df[is_numeric_row, , drop = FALSE]

print(bad_samples)

```

## B.2 Initial Summaries
  
### B.2.1 Non-replicate Sample Counts
  
```{r sample-counts, message=FALSE, warning=FALSE, echo=TRUE}
# HIHG Reps, SC, NC and IPC are already removed above

# Summarize counts by Run × Bay
nonrep_summary <- combined_df %>%
  group_by(Run, Bay) %>%
  summarise(NonRep_Samples = n(), .groups = "drop") %>%
  arrange(Run, Bay)

as.data.frame(nonrep_summary)
```

### B.2.2 Biomarker Distributions

```{r biomarker-boxplots, echo=TRUE, fig.height=6, fig.width=12}
# THIS WAS REMOVING TOO MANY COLUMNS. CHANGED
combined_df[, -(1:2)] <- lapply(combined_df[, -(1:2)], function(x) suppressWarnings(as.numeric(x)))
# Remove metadata columns
biomarker_df <- combined_df[, sapply(combined_df, is.numeric), drop = FALSE]

# Parameters for paneling
panel_size <- 28  # number of biomarkers per panel
n_biomarkers <- ncol(biomarker_df)
n_panels <- ceiling(n_biomarkers / panel_size)

# Loop through panels
for (i in seq_len(n_panels)) {
  start_col <- (i - 1) * panel_size + 1
  end_col <- min(i * panel_size, n_biomarkers)
  
  panel_matrix <- biomarker_df[, start_col:end_col]
  
  # Set up plot layout (1 panel per figure)
  par(mfrow = c(1,1), mar = c(8,4,4,2))  # adjust margins as needed
  
  boxplot(panel_matrix,
          main = paste("Biomarkers", start_col, "to", end_col),
          xlab = "Biomarker",
          ylab = "NPQ Value",
          las = 2,            # rotate x-axis labels
          col = "lightblue")
}
```
---

# C. Prep QC Biomarker Set
This is focused on generating the most stringent, high-quality biomarker dataset for QC (**QC Biomarker Set**).

## C.1 Detectability Filtering

### C.1.1 Low Detectability
Keep only biomarkers with >98% detectability on every plate for the QC Dataset. Triage others for later inspection. Importantly, some with low detectability are not necessarily bad, e.g., APOE4 and others due to polymorphisms. They are just being temporarily removed to make a clean QC Biomarker Set.

```{r detectability-filtering, message=FALSE, warning=FALSE, echo=TRUE, results='asis'}
library(tidyverse)

# Read detectability (vendor's precomputed targetDetectability, pivoted by plate)
detect_vendor <- read_csv("input_files/biomarker_detectability.csv", show_col_types = FALSE)
detect <- detect_vendor

# --- Recompute detectability over retained wells ------------------------------------------------
# The vendor column is computed over every well, so dead wells depress every target on that plate.
# Recompute as mean(NPQ >= targetLOD_NPQ) per (plate, target) across the study-sample wells that
# clear the read floor. `combined_df` here is already study samples only (controls and HIHG
# replicates were pulled out in B.1.1). See A.3 for why the screen is on reads, not detectability.
if (identical(detect_source, "recompute")) {

  .annot <- read_csv("input_files/Annotation_Targets.csv", show_col_types = FALSE) %>%
    mutate(Run = stringr::str_extract(plateID, "^[0-9]{8}-[0-9]{4}"),
           Bay = stringr::str_extract(plateID, "Bay[0-9]+")) %>%
    select(targetName, Run, Bay, targetLOD_NPQ) %>%
    distinct()

  .biom <- setdiff(colnames(combined_df), c("Run", "Bay"))
  .keep <- setdiff(rownames(combined_df), low_read_samples)

  .long <- combined_df[.keep, , drop = FALSE] %>%
    tibble::rownames_to_column("SampleID") %>%
    pivot_longer(all_of(.biom), names_to = "targetName", values_to = "NPQ") %>%
    inner_join(.annot, by = c("targetName", "Run", "Bay")) %>%
    filter(!is.na(targetLOD_NPQ))

  .det_long <- .long %>%
    group_by(Run, Bay, targetName) %>%
    summarise(pct = 100 * mean(NPQ >= targetLOD_NPQ, na.rm = TRUE), .groups = "drop") %>%
    mutate(plate = paste(Run, Bay))

  .plates <- sort(unique(.det_long$plate))
  detect_recomp <- .det_long %>%
    select(targetName, plate, pct) %>%
    pivot_wider(names_from = plate, values_from = pct) %>%
    select(targetName, all_of(.plates)) %>%
    arrange(match(targetName, detect_vendor$Biomarker)) %>%
    rename_with(~ sprintf("Plate_%02d", seq_along(.plates)), all_of(.plates)) %>%
    rename(Biomarker = targetName)
  detect_recomp$Overall <- round(rowMeans(select(detect_recomp, starts_with("Plate_")), na.rm = TRUE), 1)

  # Carry through any target the vendor lists but the recomputation cannot score (relative-quant
  # targets have no LOD; mCherry is an internal control absent from the NPQ matrix), so the
  # rel_quant exemption below still sees them.
  .missing <- setdiff(detect_vendor$Biomarker, detect_recomp$Biomarker)
  if (length(.missing)) {
    .pad <- detect_vendor[detect_vendor$Biomarker %in% .missing, , drop = FALSE]
    .pad[, setdiff(names(.pad), "Biomarker")] <- NA_real_
    detect_recomp <- bind_rows(detect_recomp, .pad[, names(detect_recomp)])
    detect_recomp <- detect_recomp %>% arrange(match(Biomarker, detect_vendor$Biomarker))
  }

  # Validate the method against the vendor on the wells the vendor itself used (no read floor).
  .pc <- grep("^Plate_", names(detect_vendor), value = TRUE)
  .cmp <- inner_join(
    detect_vendor %>% select(Biomarker, all_of(.pc)) %>%
      pivot_longer(all_of(.pc), names_to = "Plate", values_to = "vendor"),
    detect_recomp %>% select(Biomarker, all_of(intersect(.pc, names(detect_recomp)))) %>%
      pivot_longer(-Biomarker, names_to = "Plate", values_to = "mine"),
    by = c("Biomarker", "Plate")) %>%
    filter(!is.na(vendor), !is.na(mine))

  cat(sprintf(
    "\n> **Detectability recomputed** as `mean(NPQ >= targetLOD_NPQ)` per plate x target over %d of %d study-sample wells (%d below the vendor Reads floor of %s were excluded from the denominator).\n>\n> Agreement with the vendor's shipped `targetDetectability` across %s comparable cells: mean |diff| **%.2f** pp, max **%.2f** pp.\n\n",
    length(.keep), nrow(combined_df), length(intersect(low_read_samples, rownames(combined_df))),
    format(vendor_thresholds$reads_min, big.mark = ",", scientific = FALSE),
    format(nrow(.cmp), big.mark = ","),
    mean(abs(.cmp$mine - .cmp$vendor)), max(abs(.cmp$mine - .cmp$vendor))))

  write_csv(detect_recomp, "output_files/biomarker_detectability_recomputed.csv")
  detect <- detect_recomp
}

# Keep only the numeric plate columns and overall
plate_cols <- names(detect)[-1]  # first column is Biomarker name

# Robustness to export drift: coerce plate columns to numeric (a future export might quote them or
# add "%"), then ensure the 0-100 scale the thresholds assume (min_detectability_qc, min_detectability).
detect <- detect %>%
  mutate(across(all_of(plate_cols), ~ if (is.numeric(.x)) .x else readr::parse_number(as.character(.x))))
.det_max <- suppressWarnings(max(unlist(detect[plate_cols]), na.rm = TRUE))
if (is.finite(.det_max) && .det_max <= 1) {
  detect <- detect %>% mutate(across(all_of(plate_cols), ~ .x * 100))
  message("biomarker_detectability.csv appeared to be on a 0-1 proportion scale (max = ",
          round(.det_max, 3), "); rescaled x100 to match the 0-100 detectability thresholds.")
}

# Relative-quant metrics (Curve_Quant = "R", e.g. APOE, CRP): Alamar reports no detectability
# or LOD, so every plate value is NA. Keep them in all datasets and exempt from detectability
# filtering -- do NOT let an all-NA row pass via min()'s Inf.
rel_quant <- detect$Biomarker[rowSums(!is.na(detect[plate_cols])) == 0]

# Determine low-detectability biomarkers
low_det <- detect %>%
  pivot_longer(cols = all_of(plate_cols), names_to = "Plate", values_to = "Detectability") %>%
  group_by(Biomarker) %>%
  summarize(
    Min_Detectability = if (all(is.na(Detectability))) NA_real_ else min(Detectability, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(Min_Detectability), Min_Detectability < min_detectability_qc)

# Save removed biomarkers for record
write_csv(low_det, "output_files/biomarkers_low_detectability.csv")

# Print removed biomarkers
knitr::kable(
  low_det,
  caption = "Biomarkers Removed Due to Low Detectability (min detectability <98%)"
)

# Note relative-quant metrics retained without detectability filtering
if (length(rel_quant) > 0) {
  cat(sprintf(
    "\n> **Note -- relative-quant metrics:** %s have `Curve_Quant = \"R\"`; Alamar reports no detectability or LOD for them. They are **retained in all datasets** and exempt from detectability/LOD filtering.\n",
    paste(rel_quant, collapse = ", ")))
}

# For QC, keep only high-detectability biomarkers (rel_quant targets never entered low_det)
biomarkers_keep <- setdiff(colnames(combined_df), low_det$Biomarker)

combined_qc <- combined_df %>%
  select(Run, Bay, all_of(biomarkers_keep))
```

### C.1.2 Low Raw Counts
Keep only biomarkers with mean and median raw counts >500 on every plate for the QC Biomarker Set.  

```{r raw-count filtering, message=FALSE, warning=FALSE, echo=TRUE}
library(tidyverse)

# Load raw counts
raw_counts <- read_csv("input_files/Raw_counts.csv", show_col_types = FALSE)

# Biomarkers as rownames
raw_mat <- as.matrix(raw_counts[,-1])
rownames(raw_mat) <- raw_counts[[1]]

# Convert to samples x biomarkers to match combined_qc structure
raw_t <- t(raw_mat)

# Ensure matching sample sets between raw counts & metadata
raw_df <- as.data.frame(raw_t)
raw_df$SampleID <- rownames(raw_t)

raw_df <- raw_df %>%
  filter(SampleID %in% rownames(combined_qc)) %>%
  left_join(meta_df, by = c("SampleID" = "SampleID")) %>%
  select(SampleID, Run, Bay, everything())

#Save for later parts of the pipeline
write_csv(raw_df, "output_files/raw_df.csv")

### Compute mean & median per Biomarker per Plate ####
plate_stats <- raw_df %>%
  pivot_longer(cols = -c(Run, Bay, SampleID),
               names_to = "Biomarker",
               values_to = "RawCount") %>%
  group_by(Biomarker, Run, Bay) %>%
  summarize(
    mean_count = mean(RawCount, na.rm = TRUE),
    median_count = median(RawCount, na.rm = TRUE),
    .groups = "drop"
  )

### Identify biomarkers failing thresholds ###
low_counts <- plate_stats %>%
  group_by(Biomarker) %>%
  summarize(
    Min_Mean = min(mean_count, na.rm = TRUE),
    Min_Median = min(median_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(Min_Mean < read_count_threshold | Min_Median < read_count_threshold)

### Save for record
write_csv(low_counts, "output_files/biomarkers_low_counts.csv")

### Drop columns from QC set
biomarkers_to_drop <- low_counts$Biomarker
biomarkers_to_drop <- intersect(biomarkers_to_drop, colnames(combined_qc))
biomarkers_to_drop <- setdiff(biomarkers_to_drop, c("Run", "Bay"))

combined_qc <- combined_qc %>%
  select(-all_of(biomarkers_to_drop))

### Display formatted table in HTML report
knitr::kable(
  low_counts,
  caption = "Biomarkers Removed Due to Low Raw Counts (<500 Min Mean or Min Median)"
)

```  

## C.2 Correlation Filtering
Create a set of biomarkers with low correlation for the QC Biomarker Set

```{r correlation-filtering, message=FALSE, warning=FALSE, echo=TRUE, results='asis'}
library(caret)
library(pheatmap)

# Biomarker pool entering scaffold selection = everything that passed detectability (C.1.1) and raw-count
# (C.1.2) filtering. Both modes preserve this pool: independent_biomarkers (the triage anchors) and
# high_cor (added back in E.1) always partition `qc_pool`, so post-QC biomarker CONTENT is mode-invariant.
qc_pool <- setdiff(colnames(combined_qc), c("Run", "Bay"))
qc_pool <- qc_pool[sapply(combined_qc[qc_pool], is.numeric)]

if (identical(scaffold_mode, "frozen")) {
  # ---- Frozen scaffold + per-plate tripwire ----------------------------------------------------------
  # Anchor eligibility — TWO SEPARATE GATES (WATCH tier fixed 2026-09-01):
  #   raw counts    : a frozen marker under read_count_threshold cannot anchor. Unchanged.
  #   detectability : the floor for an ANCHOR is `scaffold_drop` (90), NOT the 98% pool bar.
  #
  # This previously read `intersect(frozen_scaffold, qc_pool)`, and qc_pool had already removed every
  # biomarker under `min_detectability_qc` (98) on ANY plate. That made the tripwire's WATCH tier dead
  # code: a cell in [scaffold_drop, scaffold_watch) is documented as "kept, flagged", but had in fact
  # been dropped before the tripwire ever ran — detectability-alone filtering of the anchor set, which
  # is precisely what the frozen scaffold exists to prevent. A marker sitting just under the bar on a
  # single plate was losing anchor status for the whole run.
  #
  # A WATCH anchor sits BELOW the 98% pool bar, so it is deliberately NOT a member of `qc_pool`. It
  # anchors the sample triage; it does not change which biomarkers reach the post-QC datasets — that
  # partition stays `qc_pool` (see E.1, which reconstructs it explicitly).
  pc <- grep("^Plate_", names(detect), value = TRUE)
  .dmat <- as.matrix(detect[, pc, drop = FALSE]); rownames(.dmat) <- detect$Biomarker
  .dmin <- apply(.dmat, 1, function(r) if (all(is.na(r))) NA_real_ else min(r, na.rm = TRUE))

  .present  <- intersect(frozen_scaffold, setdiff(colnames(combined_df), c("Run", "Bay")))
  .ok_reads <- setdiff(.present, low_counts$Biomarker)          # raw-count gate: unchanged
  anchors   <- .ok_reads[!is.na(.dmin[.ok_reads]) & .dmin[.ok_reads] >= scaffold_drop]
  not_pool  <- setdiff(frozen_scaffold, anchors)   # frozen markers unavailable as anchors this run

  # Which anchors are being KEPT under the WATCH tier (would have been dropped by the old code)
  .watch_kept <- anchors[!is.na(.dmin[anchors]) & .dmin[anchors] < scaffold_watch]

  # Tripwire: scan every (frozen anchor x plate) detectability cell; alert on anything below the bar.
  pc <- grep("^Plate_", names(detect), value = TRUE)
  det_anch <- detect[detect$Biomarker %in% frozen_scaffold, , drop = FALSE]
  Pmat <- as.matrix(det_anch[, pc]); rownames(Pmat) <- det_anch$Biomarker
  idx  <- which(Pmat < scaffold_watch, arr.ind = TRUE)
  tripwire <- if (nrow(idx)) {
    tibble(anchor = rownames(Pmat)[idx[, 1]], plate = pc[idx[, 2]], detectability = round(Pmat[idx], 1)) %>%
      mutate(status = ifelse(detectability < scaffold_drop, "DROP + alert", "WATCH (keep + alert)")) %>%
      arrange(detectability)
  } else tibble(anchor = character(), plate = character(), detectability = numeric(), status = character())
  write_csv(tripwire, "output_files/scaffold_tripwire.csv")

  independent_biomarkers <- anchors
  high_cor               <- setdiff(qc_pool, independent_biomarkers)   # display only; E.1 rebuilds from qc_pool

  # Build the triage matrix from combined_df, not combined_qc: a WATCH anchor is below the 98% pool
  # bar and so is not a column of combined_qc.
  combined_qc <- combined_df %>% select(Run, Bay, all_of(independent_biomarkers))
  write.csv(independent_biomarkers, "output_files/QC_Biomarker_Set.csv")

  cat(sprintf("**Scaffold mode: FROZEN** — using **%d of %d** frozen anchors for sample triage.\n\n",
              length(independent_biomarkers), length(frozen_scaffold)))
  if (length(.watch_kept))
    cat(sprintf("> **WATCH tier active:** %d anchor(s) sit below %g%% on at least one plate but at or above the %g%% drop floor, so they are **kept and flagged** rather than dropped: %s. (Before the 2026-09-01 fix these were silently removed.)\n\n",
                length(.watch_kept), scaffold_watch, scaffold_drop,
                paste(sprintf("%s (min %.1f%%)", .watch_kept, .dmin[.watch_kept]), collapse = ", ")))
  if (length(not_pool))
    cat(sprintf("> %d frozen marker(s) not available as anchors this run (absent or below the detectability/read bar): %s\n\n",
                length(not_pool), paste(not_pool, collapse = ", ")))
  if (nrow(tripwire)) {
    cat(sprintf("> **TRIPWIRE: %d anchor x plate cell(s) below %g%%** — WATCH = kept & flagged, DROP (<%g%%) = not anchored & flagged:\n\n",
                nrow(tripwire), scaffold_watch, scaffold_drop))
    print(knitr::kable(tripwire))
    cat("\n")
  } else {
    cat(sprintf("> Tripwire: every frozen anchor clears %g%% on every plate — no alert. ✓\n\n", scaffold_watch))
  }
  # Heatmap of the anchors actually used (confirms they remain de-correlated on this batch)
  cor_mat <- cor(combined_qc[, independent_biomarkers], use = "pairwise.complete.obs")
  pheatmap(cor_mat, main = "Heatmap of Frozen Scaffold Anchors", fontsize_row = 6, fontsize_col = 6)

} else {
  # ---- Re-derive (default): de-correlate the high-detectability / high-count pool --------------------
  samp_num <- combined_qc[, qc_pool]
  cor_mat  <- cor(samp_num, use = "pairwise.complete.obs")

  # Identify highly correlated pairs (|r| >= corr_thresh)
  high_cor <- findCorrelation(cor_mat, cutoff = corr_thresh, names = TRUE)

  # Independent biomarkers
  independent_biomarkers <- setdiff(colnames(samp_num), high_cor)

  pheatmap(cor_mat[independent_biomarkers, independent_biomarkers],
           main = "Heatmap of Independent Biomarkers",
           fontsize_row = 6, fontsize_col = 6)

  combined_qc <- combined_qc %>% select(Run, Bay, all_of(independent_biomarkers))
  write.csv(independent_biomarkers, "output_files/QC_Biomarker_Set.csv")
}

```
---

# D. Sample QC

## D.0 Whole-Panel Intensity (`mean_INT`) — measured, never gated

Every screen below asks a **count** question ("on how many markers is this sample
extreme?") or a **distance** question ("how far is it from the plate centre?").
None asks a **level** question. A sample that reads moderately high on *all* 131
markers at once is therefore invisible to all of them: the outlier-burden test
counts markers outside a per-plate IQR fence in either direction, so a uniform
shift never makes any single marker extreme. That is structural, not a threshold
that can be loosened.

`mean_INT` is that missing level question. Each marker is put on a rank-based
inverse-normal scale, so every marker contributes on the same footing, and the
per-sample mean is directly readable as *how bright does this sample run across
the panel*.

> ⚠️ **This is a covariate, not a gate.** Background intensity is continuous and
> affects every sample; it is not a population of bad wells. Nothing here triages
> or flags a sample out of the dataset. `mean_INT` is computed for **all** wells
> — including those about to be triaged — written to the outputs, and intended
> for the downstream analysis model. Gating on it would remove real biological
> and technical variation that belongs in the model.

It is computed **before** triage on purpose. Run downstream, it can only ever see
retained wells, so removed wells have no intensity value and cannot be placed on
the axis at all.

```{r whole-panel-intensity, message=FALSE, warning=FALSE, echo=TRUE}
# Rank-based inverse-normal transform, per marker, over all wells present here.
# Same definition as Secondary_QC/Extremes/intensity_columns.R — there must be one
# definition of this column, or two files will disagree about the same sample.
int_scores <- function(v) {
  v  <- suppressWarnings(as.numeric(v))
  ok <- !is.na(v)
  out <- rep(NA_real_, length(v))
  n <- sum(ok)
  if (n >= 2) out[ok] <- stats::qnorm((rank(v[ok], ties.method = "average") - 0.5) / n)
  out
}

int_markers <- names(combined_df)[vapply(combined_df, is.numeric, logical(1))]
int_markers <- setdiff(int_markers, c("Run", "Bay"))

int_mat <- vapply(int_markers, function(b) int_scores(combined_df[[b]]),
                  numeric(nrow(combined_df)))
rownames(int_mat) <- rownames(combined_df)

mi <- rowMeans(int_mat, na.rm = TRUE)
mi[!is.finite(mi)] <- NA_real_
.sd_mi <- stats::sd(mi, na.rm = TRUE)

intensity_df <- tibble(
  SampleID  = rownames(combined_df),
  Run       = combined_df$Run,
  Bay       = combined_df$Bay,
  mean_INT  = round(mi, 3),
  # Round BEFORE deriving the flag, so the flag agrees with the value that is
  # actually written out (a boundary sample at z = -2.4996 otherwise prints
  # "-2.50" while not carrying the LOW flag).
  mean_INT_z = if (is.finite(.sd_mi) && .sd_mi > 0)
                 round((mi - mean(mi, na.rm = TRUE)) / .sd_mi, 2) else NA_real_
) %>%
  mutate(intensity_flag = case_when(
    is.na(mean_INT_z)                 ~ NA_character_,
    mean_INT_z >=  intensity_sd_flag  ~ "HIGH",
    mean_INT_z <= -intensity_sd_flag  ~ "LOW",
    TRUE                              ~ ""
  ))

write_csv(intensity_df, "output_files/whole_panel_intensity.csv")

cat(sprintf("mean_INT over %d wells x %d markers (pre-triage, so triaged wells are included)\n",
            nrow(intensity_df), length(int_markers)))
cat(sprintf("Informational flags at +/-%.1f SD: %d HIGH, %d LOW  — neither triages nor flags a sample.\n",
            intensity_sd_flag,
            sum(intensity_df$intensity_flag == "HIGH", na.rm = TRUE),
            sum(intensity_df$intensity_flag == "LOW",  na.rm = TRUE)))

ggplot(intensity_df, aes(x = mean_INT_z)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = c(-intensity_sd_flag, intensity_sd_flag),
             linetype = "dashed", colour = "grey30") +
  theme_minimal() +
  labs(title = "Whole-panel intensity across all wells (pre-triage)",
       subtitle = "Dashed lines are the informational flag, not a cut — nothing is removed on this axis",
       x = "mean_INT_z", y = "Wells")
```

## D.1 Alamar Sample QC
Gather the Alamar sample QC statistics and identify samples with "warning".

```{r apply-droplist, message=FALSE, warning=FALSE, echo=TRUE}
library(readr)
library(dplyr)

#alamar_qc is the Sample_QC file read in above

warn_samps <- alamar_qc %>%
  filter(grepl("warn", `QC Status`, ignore.case = TRUE)) %>%  # values are "Passed"/"Warning" (case/wording can drift)
  left_join(
    meta_df %>% select(SampleID, Run, Bay),
    by = c("Sample Name" = "SampleID")
  )

library(kableExtra)
warn_tbl <- warn_samps %>%
  arrange(Run, Bay, `Sample Name`) %>%
  transmute(`Sample Name`,
            `IC Median`   = round(`IC Median`, 3),
            Detectability = round(Detectability, 3),
            `IC Reads`,
            Well = AUTO_WELLPOSITION,
            .grp = paste(Run, Bay))
.grp_rle <- rle(warn_tbl$.grp)
warn_tbl %>%
  select(-.grp) %>%
  kbl(caption = "Alamar Sample QC Warnings (sorted, grouped by plate)") %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  pack_rows(index = setNames(.grp_rle$lengths, .grp_rle$values))

write_csv(arrange(warn_samps, Run, Bay, `Sample Name`), "output_files/Alamar_QC_warn.csv")

```

## D.2 Sample Outlier Detection 

### D.2.1 PCA-Based Outlier Check
Looking for single samples or small groups that are outliers on the first two PCs.
Using a threshold of `r PCA_SD` sd to define outliers.

```{r pca-outlier, message=FALSE, warning=FALSE, echo=TRUE}
library(tidyverse)
library(patchwork)
library(dplyr)
library(irlba)  # efficient PCA

set.seed(123)

# --- Round 1 PCA -----------------------------------------------------------
biomarker_cols <- setdiff(names(combined_qc), c("Run", "Bay"))
qc_matrix <- combined_qc[, biomarker_cols] %>% as.matrix()
qc_scaled <- scale(qc_matrix)

pca_res <- prcomp(qc_scaled, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(combined_qc),
  Run = combined_qc$Run,
  Bay = combined_qc$Bay,
  PC1 = pca_res$x[, 1],
  PC2 = pca_res$x[, 2]
)

# Identify outliers by distance

pca_df <- pca_df %>%
  group_by(Run, Bay) %>%
  mutate(
    PC_dist = sqrt(PC1^2 + PC2^2),
    mean_PCdist = mean(PC_dist, na.rm = TRUE),
    sd_PCdist   = sd(PC_dist, na.rm = TRUE),
    # The statistic the cut is actually applied to. Written out so the operating
    # point is auditable: "how close did the nearest retained well get?" cannot be
    # answered from a boolean.
    PC_z = (PC_dist - mean_PCdist) / sd_PCdist,
    extreme_outlier = PC_z > PCA_SD
  ) %>%
  ungroup()

# --- COHORT-SCALE COMPANION: a FLAG, never a drop (added 2026-09-01) -----------------------------
# PC_z above is standardised WITHIN Run x Bay. That is right for finding a lone odd well on an
# otherwise normal plate, but it SELF-MASKS a cluster: when many wells on one plate fail together
# they inflate that plate's own sd_PCdist, which is the yardstick they are then measured against.
#
# Observed on this run: 30 wells sit at PC1 < -15, all of them ~23x below the cohort median read
# depth. The per-plate screen caught 9. On the three plates where the failures clustered
# (5-11 failures each) it caught ZERO, because such a plate carries an sd_PCdist several times the
# cohort-wide value. The ones it did catch were singletons on otherwise-clean plates. Rescored
# cohort-wide, the same wells run a median PC_z of ~13.
#
# So this adds the cohort-scale score alongside, and a flag for wells extreme on it. It changes NO
# drop rule: triage still uses `extreme_outlier` only. The point is visibility - a clustered failure
# should not be able to hide behind the spread it is itself creating.
pca_df <- pca_df %>%
  mutate(PC_z_cohort   = (PC_dist - mean(PC_dist, na.rm = TRUE)) / sd(PC_dist, na.rm = TRUE),
         pca_cohort_flag = PC_z_cohort > PCA_SD,
         pca_masked_by_plate = pca_cohort_flag & !extreme_outlier)

extreme_samples_round1 <- pca_df$Sample[pca_df$extreme_outlier]
write_csv(pca_df, "output_files/sample_pca_coordinates_round1.csv")

cat(sprintf("Round 1: %d of %d wells beyond %g SD.  max PC_z = %.3f (headroom %+.3f)\n",
            length(extreme_samples_round1), nrow(pca_df), PCA_SD,
            max(pca_df$PC_z, na.rm = TRUE), max(pca_df$PC_z, na.rm = TRUE) - PCA_SD))

# Report the cohort-scale view next to it, and name the plates doing the masking.
cat(sprintf("Cohort-scale companion (FLAG ONLY, nothing dropped): %d wells beyond %g SD on the cohort scale; %d of those are NOT per-plate outliers.\n",
            sum(pca_df$pca_cohort_flag, na.rm = TRUE), PCA_SD,
            sum(pca_df$pca_masked_by_plate, na.rm = TRUE)))
if (any(pca_df$pca_masked_by_plate, na.rm = TRUE)) {
  .masked <- pca_df %>% filter(pca_masked_by_plate) %>%
    group_by(Run, Bay) %>%
    summarise(n_masked = n(), plate_sd_PCdist = round(first(sd_PCdist), 2),
              median_PC_z_plate = round(median(PC_z), 2),
              median_PC_z_cohort = round(median(PC_z_cohort), 2), .groups = "drop") %>%
    arrange(desc(n_masked))
  write_csv(.masked, "output_files/pca_cluster_flags.csv")
  cat(sprintf("  cohort-wide sd(PC_dist) = %.2f; the plates below exceed it, which is what hides their own wells:\n\n",
              sd(pca_df$PC_dist, na.rm = TRUE)))
  print(knitr::kable(.masked,
        caption = "Wells extreme cohort-wide but NOT flagged on their own plate (a clustered failure masking itself). FLAG ONLY - these are not dropped here; check whether another screen catches them."))
  cat("\n")
}

# Variance explained
var_explained <- summary(pca_res)$importance["Proportion of Variance", 1:2] * 100
pc1_var <- round(var_explained[1], 1)
pc2_var <- round(var_explained[2], 1)

# Plot round 1
plot1 <- ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = Run, shape = Bay), size = 3, alpha = 0.7) +
  geom_point(
    data = filter(pca_df, extreme_outlier),
    aes(x = PC1, y = PC2),
    shape = 4, color = "black", size = 3, stroke = 0.5
  ) +
  labs(
    title = paste0("PCA Round 1 (", length(extreme_samples_round1), " outliers)"),
    x = paste0("PC1 (", pc1_var, "% variance)"),
    y = paste0("PC2 (", pc2_var, "% variance)")
  ) +
  theme_minimal() +
  theme(legend.position = "right")

#Round 1 results
print(plot1)
knitr::kable(pca_df[pca_df$extreme_outlier, ], caption = "First-round PCA outliers:")

# --- Round 2 PCA (remove round 1 outliers) ---------------------------------
if (length(extreme_samples_round1) > 0) {
  combined_qc_round2 <- combined_qc[!rownames(combined_qc) %in% extreme_samples_round1, ]

  qc_matrix2 <- combined_qc_round2[, biomarker_cols] %>% as.matrix()
  qc_scaled2 <- scale(qc_matrix2)

  pca_res2 <- prcomp(qc_scaled2, center = TRUE, scale. = TRUE)

  pca_df2 <- data.frame(
    Sample = rownames(combined_qc_round2),
    Run = combined_qc_round2$Run,
    Bay = combined_qc_round2$Bay,
    PC1 = pca_res2$x[, 1],
    PC2 = pca_res2$x[, 2]
  )
    
  pca_df2 <- pca_df2 %>%
  group_by(Run, Bay) %>%
  mutate(
    PC_dist = sqrt(PC1^2 + PC2^2),
    mean_PCdist = mean(PC_dist, na.rm = TRUE),
    sd_PCdist   = sd(PC_dist, na.rm = TRUE),
    PC_z = (PC_dist - mean_PCdist) / sd_PCdist,
    extreme_outlier = PC_z > PCA_SD
  ) %>%
  ungroup()


  # Identify new outliers
  extreme_samples_round2 <- pca_df2$Sample[pca_df2$extreme_outlier]
  write_csv(pca_df2, "output_files/sample_pca_coordinates_round2.csv")

  # Round 2 is the pass most likely to be inert: removing round-1 outliers pulls
  # the remaining wells together, and if nothing then reaches PCA_SD the second
  # pass has added nothing on this run. Say so rather than letting a silent "0
  # new outliers" read as "the second screen confirmed the data are clean".
  .r2_max <- max(pca_df2$PC_z, na.rm = TRUE)
  cat(sprintf("Round 2: %d of %d wells beyond %g SD.  max PC_z = %.3f (headroom %+.3f)\n",
              length(extreme_samples_round2), nrow(pca_df2), PCA_SD, .r2_max, .r2_max - PCA_SD))
  if (length(extreme_samples_round2) == 0)
    cat("  -> INERT on this run: the second pass removed nothing. It is UNTESTED, not proven safe.\n")

  var_explained2 <- summary(pca_res2)$importance["Proportion of Variance", 1:2] * 100
  pc1_var2 <- round(var_explained2[1], 1)
  pc2_var2 <- round(var_explained2[2], 1)

  # Plot round 2
  plot2 <- ggplot(pca_df2, aes(x = PC1, y = PC2)) +
    geom_point(aes(color = Run, shape = Bay), size = 3, alpha = 0.7) +
    geom_point(
      data = filter(pca_df2, extreme_outlier),
      aes(x = PC1, y = PC2),
      shape = 4, color = "black", size = 3, stroke = 0.5
    ) +
    labs(
      title = paste0("PCA Round 2 (", length(extreme_samples_round2), " new outliers)"),
      x = paste0("PC1 (", pc1_var2, "% variance)"),
      y = paste0("PC2 (", pc2_var2, "% variance)")
    ) +
    theme_minimal() +
    theme(legend.position = "right")

  # Round 2
  print(plot2)
  knitr::kable(pca_df2[pca_df2$extreme_outlier, ], caption = "Second-round PCA outliers:")

} else {
  cat("\nNo first-round PCA outliers detected. Skipping second round.\n")
  # Define these anyway: D.3 and the D.5.1 inventory read them unconditionally,
  # and an undefined pca_df2 used to abort the run whenever round 1 was clean.
  pca_df2 <- pca_df[0, ]
  extreme_samples_round2 <- character(0)
  plot1
}
```

I changed these to be plate-specific outliers.

### D.2.2 Sample Outlier Burden
Here we are identifying samples that have extreme outliers over biomarkers within their plate.
Compute p-value for whether outlier burden in a sample is more than expected by chance. 

```{r excessive_outliers, message=FALSE, warning=FALSE, echo=TRUE}
library(dplyr)
library(poibin)  # for Poisson-Binomial p-values

#Uses IQR method to define outliers
upper_matrix <- matrix(0, nrow = nrow(qc_matrix), ncol = ncol(qc_matrix),
                       dimnames = dimnames(qc_matrix))
lower_matrix <- upper_matrix * 0

for (rb in unique(paste(combined_qc$Run, combined_qc$Bay, sep = "_"))) {
  idx <- which(paste(combined_qc$Run, combined_qc$Bay, sep = "_") == rb)
  submat <- qc_matrix[idx, , drop = FALSE]

  # Compute Q1 and Q3 per biomarker
  Q1 <- apply(submat, 2, quantile, 0.25, na.rm = TRUE)
  Q3 <- apply(submat, 2, quantile, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  
  # Define outliers
  upper <- sweep(submat, 2, Q3 + samp_out_thresh * IQR, ">")
  lower <- sweep(submat, 2, Q1 - samp_out_thresh * IQR, "<")

  upper_matrix[idx, ] <- upper_matrix[idx, ] + upper
  lower_matrix[idx, ] <- lower_matrix[idx, ] + lower
}

# Assume upper_matrix and lower_matrix are data.frames with samples in rows, biomarkers in columns and that combined_qc has Run and Bay columns

add_outlier_stats <- function(out_mat, combined_df) {
  out_mat <- as.matrix(out_mat)
  
  out_df <- as.data.frame(out_mat) %>%
    mutate(
      SampleID = rownames(out_mat),
      Run = combined_df$Run,
      Bay = combined_df$Bay
    )

  p_biom <- colMeans(out_mat, na.rm = TRUE)

  out_df <- out_df %>%
    rowwise() %>%
    mutate(
      Total_Outliers = sum(c_across(where(is.numeric)), na.rm = TRUE),
      Expected_Outliers = sum(p_biom, na.rm = TRUE),
      P_Value = 1 - poibin::ppoibin(Total_Outliers - 1, p_biom)
    ) %>%
    ungroup() %>%
    mutate(
      FDR = p.adjust(P_Value, method = "BH")  # ← NEW COLUMN
    ) %>%
    select(SampleID, Run, Bay, Total_Outliers, Expected_Outliers,
           P_Value, FDR, everything())

  return(out_df)
}

upper_df <- add_outlier_stats(upper_matrix, combined_qc)
lower_df <- add_outlier_stats(lower_matrix, combined_qc)

# Save to CSV
write_csv(upper_df, "output_files/upper_outliers_with_stats.csv")
write_csv(lower_df, "output_files/lower_outliers_with_stats.csv")

# ---- Sample Outlier Summary (Top 10) ----

# Sort by significance (lowest P-value first)
upper_sorted <- upper_df %>% arrange(P_Value)
lower_sorted <- lower_df %>% arrange(P_Value)

# Save full sorted tables
write_csv(upper_sorted, "output_files/upper_outliers_sorted.csv")
write_csv(lower_sorted, "output_files/lower_outliers_sorted.csv")

# Select first 10 + only key columns for report
upper_top10 <- upper_sorted %>%
  select(SampleID, Run, Bay, "Total Out"=Total_Outliers, "Exp Out" = Expected_Outliers, 
         P_Value, FDR) %>%
  slice(1:10)

lower_top10 <- lower_sorted %>%
  select(SampleID, Run, Bay, "Total Out"=Total_Outliers, "Exp Out" = Expected_Outliers,
         P_Value, FDR) %>%
  slice(1:10)

# Print to RMarkdown
knitr::kable(upper_top10, caption = "Top 10 Upper Outliers (Sorted by P-value)")

knitr::kable(lower_top10, caption = "Top 10 Lower Outliers (Sorted by P-value)")
```

#### `FDR_threshold` is not a dial

The p-value is Poisson-binomial on an **integer count** of anchor markers, so the
FDR takes only as many distinct values as there are achievable counts. Moving
`FDR_threshold` does nothing until it crosses one of them, and then it moves by a
whole anchor's worth of samples at once. The table below is the complete list of
settings that exist — anything between two rows is the same QC as the row above.

```{r burden-operating-points, message=FALSE, warning=FALSE, echo=TRUE}
# A sample is triaged if EITHER tail is significant, so the statistic the cut
# actually lands on is the per-sample minimum of the two FDRs.
burden_stat <- tibble(
  SampleID = upper_df$SampleID,
  fdr_min  = pmin(upper_df$FDR, lower_df$FDR, na.rm = TRUE),
  n_out    = upper_df$Total_Outliers + lower_df$Total_Outliers
)

n_anchor <- ncol(qc_matrix)

op_points <- burden_stat %>%
  count(fdr_min, name = "wells_here") %>%
  arrange(fdr_min) %>%
  mutate(
    # Guide, not an identity: the two tails carry different BH adjustments, so
    # the count is not a strict function of the FDR across tails.
    anchors_out_median = vapply(fdr_min, function(f)
      stats::median(burden_stat$n_out[burden_stat$fdr_min == f]), numeric(1)),
    removed_at_this_cut = cumsum(wells_here),
    is_current_cut = fdr_min < FDR_threshold &
                     dplyr::lead(fdr_min, default = Inf) >= FDR_threshold
  )

write_csv(op_points, "output_files/burden_operating_points.csv")

cat(sprintf("`FDR` takes %d distinct values over %d wells; %.1f%% of wells sit at exactly %.4g.\n",
            nrow(op_points), nrow(burden_stat),
            100 * max(op_points$wells_here) / nrow(burden_stat),
            op_points$fdr_min[which.max(op_points$wells_here)]))
cat(sprintf("Current cut FDR < %g removes %d wells, i.e. \"%.0f or more of %d anchors out\".\n",
            FDR_threshold, sum(burden_stat$fdr_min < FDR_threshold, na.rm = TRUE),
            min(burden_stat$n_out[burden_stat$fdr_min < FDR_threshold], na.rm = TRUE), n_anchor))

# Show the neighbourhood of the current cut, where any retuning would happen.
.i <- which(op_points$is_current_cut)
.win <- if (length(.i)) seq(max(1, .i - 2), min(nrow(op_points), .i + 3)) else seq_len(min(8, nrow(op_points)))
knitr::kable(
  op_points[.win, ] %>%
    transmute(FDR = signif(fdr_min, 5),
              `~anchors out` = anchors_out_median,
              `wells here` = wells_here,
              `removed at this cut` = removed_at_this_cut,
              `current` = ifelse(is_current_cut, "<-- current gate", "")),
  caption = paste0("Available operating points around FDR < ", FDR_threshold,
                   ". There is nothing to tune between adjacent rows.")
)
```

Note also what this gate **cannot** see. The burden count is a **distance**
statistic: it counts markers outside a per-plate IQR fence, in either direction.
A sample that is moderately high on every marker at once never makes any single
marker extreme, so no `FDR_threshold` recovers it. That blind spot is what D.0
measures — and the reason `mean_INT` is reported rather than added here as a
gate.

These plots show where the outlier samples fall for the biomarkers. Red are outliers. Blue are points for the same sample where they are not outliers. Only showing plots for samples with FDR< `r FDR_threshold`.
 
```{r visualize_excessive_outliers, message=FALSE, warning=FALSE, echo=TRUE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)

upper_sig <- upper_df$SampleID[upper_df$FDR < FDR_threshold]
lower_sig <- lower_df$SampleID[lower_df$FDR < FDR_threshold]

# List of extreme outlier SampleIDs 
extreme_samples <- c(upper_sig,lower_sig)

plot_list <- list()

plot_sample_violin <- function(sample_id) {
  
  # Get Run/Bay for title
  sample_run <- combined_qc[sample_id, "Run"]
  sample_bay <- combined_qc[sample_id, "Bay"]
  
  # Convert to long format
  long_df <- combined_qc %>%
    rownames_to_column("SampleID") %>%
    pivot_longer(
      cols = -c(SampleID, Run, Bay),
      names_to = "Biomarker",
      values_to = "Value"
    )
  
  # Create Highlight column: default "Other"
  long_df$Highlight <- "Other"
  
  # Pull the outlier flags for this sample
  outlier_flags <- upper_matrix[sample_id,]+lower_matrix[sample_id,]
  
  # Only update the Highlight for this sample
  long_df$Highlight[long_df$SampleID == sample_id] <- ifelse(
    outlier_flags[long_df$Biomarker[long_df$SampleID == sample_id]],
    "Extreme",
    "Normal"
  )
  
  # Plot
  p <- ggplot(long_df, aes(x = Biomarker, y = Value)) +
    geom_violin(fill = "gray70", alpha = 0.6, trim = FALSE) +
    geom_point(
      data = subset(long_df, Highlight != "Other"),
      aes(color = Highlight),
      size = 3
    ) +
    scale_color_manual(values = c("Extreme" = "red", "Normal" = "blue")) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
    labs(title = paste("Sample", sample_id,
                       "| Run:", sample_run, "| Bay:", sample_bay),
         x = "Biomarker",
         y = "NPQ")
  
  return(p)
}

# Loop through extreme samples
for (sample_id in extreme_samples) {
  p=plot_sample_violin(sample_id)
  print(p)
}
```

Determine samples with a significant number of outliers over biomarkers.

### D.2.3 IC-Outlier Reads
Here we are identifying samples that have extreme numbers of matching reads in aggregate within their plate.

Alamar does this with respect to the median number of reads in the IC samples within a plate: Sample IC reads relative to the median (Within +/-40% of the plate median). The IC is mCherry.
They also WARN if Sample IC reads <1000. Usually the outlier criteria would capture this, but it would not tell us if the whole plate had low reads. I added that as a conditional lower bound. 

```{r Sample_vs_IC_reads, message=FALSE, warning=FALSE, echo=TRUE}
#Look at mCherry raw reads over Run and Bay

#raw_df has the raw count data (including mCherry) for all of the samples in QC
# Ensure Run and Bay are factors
raw_df$Run <- factor(raw_df$Run)
raw_df$Bay <- factor(raw_df$Bay)

# Custom Bay colors
bay_colors <- c("Bay1" = "red", "Bay2" = "green", "Bay3" = "blue")

# The IC (mCherry) is no longer a target row in the NPQ/Raw-counts export; its per-sample
# reads now live in the Sample QC sheet as `IC Reads` (loaded above as alamar_qc). Source from
# there, restricted to the QC sample set (samples present in raw_df).
# Lower bound is -40% of the median or 1000, whichever is bigger.
df_temp <- alamar_qc %>%
  transmute(
    SampleID = `Sample Name`,
    Value    = as.numeric(`IC Reads`),
    Run      = stringr::str_extract(plateID, "^[0-9]{8}-[0-9]{4}"),
    Bay      = stringr::str_extract(plateID, "Bay[0-9]+")
  ) %>%
  filter(SampleID %in% raw_df[[1]]) %>%
  group_by(Run, Bay) %>%
  mutate(
    Median_Value = median(Value, na.rm = TRUE),
    Lower_Thresh = pmax((Median_Value * 0.6), 1000),
    Upper_Thresh = Median_Value * 1.4,  # +40%
    Outlier_Flag = Value < Lower_Thresh | Value > Upper_Thresh
  ) %>%
  ungroup()

# Plot
#I tried to highlight outliers on plot but I couldn't get them centered right
p <- ggplot(df_temp, aes(x = Run, y = Value, fill = Bay)) +
  geom_violin(alpha = 0.5, trim = FALSE, position = position_dodge(width = 0.8)) +
  geom_jitter(aes(color = Bay),
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.75, size = 1.3) +
  scale_fill_manual(values = bay_colors) +
  scale_color_manual(values = bay_colors) +
  theme_minimal() +
  labs(
    title = "Run × Bay Effects: IC Reads",
    x = "Run",
    y = "Reads"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p)

df_temp <- df_temp %>%
  rename(IC_Reads = Value)

# Print table
knitr::kable(
  df_temp %>% 
    filter(Outlier_Flag == TRUE) %>%
    select(SampleID, Run, Bay, IC_Reads) %>%
    arrange(desc(IC_Reads)),
  caption = "mCherry-Read Outliers (Sorted by Reads)"
)

IC_outlier_samples= df_temp %>%
    filter(Outlier_Flag ==TRUE)%>%
    arrange(desc(IC_Reads)) %>%
    select(SampleID)

```

#### Is this an independent check, or Alamar's metric recomputed?

This gate supplies the large majority of all triage, so it is worth being precise
about what it is. We test `IC Reads` against 0.6–1.4× the plate median. Alamar's
own `IC Median` column is *defined* as sample IC reads relative to the plate
median, with a ±40% bar. **Same quantity, same fence.** If the two agree closely,
this is a re-derivation of a vendor metric — a useful cross-check that the vendor
value reproduces, but **not** a second, independent line of evidence, and methods
text should not describe it as in-house QC that independently confirms Alamar's.

```{r ic-gate-duplication, message=FALSE, warning=FALSE, echo=TRUE}
# Alamar reports IC Median as a proportion (+/-0.40) in current exports, but a
# percent-scaled export (+/-40) would silently pass every well. Detect the scale
# instead of assuming it.
ic_vendor <- alamar_qc %>%
  transmute(SampleID = `Sample Name`, ic_med = suppressWarnings(as.numeric(`IC Median`))) %>%
  filter(SampleID %in% df_temp$SampleID)

.ic_scale <- if (stats::quantile(abs(ic_vendor$ic_med), 0.99, na.rm = TRUE) > 5) 100 else 1
if (.ic_scale == 100)
  message("IC Median appears to be on a 0-100 percent scale; rescaling to match ic_median_thresh.")
ic_vendor$ic_med <- ic_vendor$ic_med / .ic_scale

ours   <- df_temp$SampleID[df_temp$Outlier_Flag]
theirs <- ic_vendor$SampleID[abs(ic_vendor$ic_med) > ic_median_thresh & !is.na(ic_vendor$ic_med)]

.int <- length(intersect(ours, theirs))
.uni <- length(union(ours, theirs))
jacc <- if (.uni > 0) .int / .uni else NA_real_

ic_dup <- tibble(
  in_house_gate   = "IC Reads outside 0.6x-1.4x the Run x Bay median",
  vendor_metric   = paste0("|IC Median| > ", ic_median_thresh),
  n_in_house      = length(ours),
  n_vendor        = length(theirs),
  n_both          = .int,
  jaccard         = round(jacc, 3),
  verdict         = dplyr::case_when(
    is.na(jacc)   ~ "neither fires",
    jacc >= 0.90  ~ "DUPLICATE - the vendor metric recomputed, not an independent axis",
    jacc >= 0.50  ~ "OVERLAPPING - largely the same wells",
    TRUE          ~ "DISTINCT - the two gates see different wells")
)
write_csv(ic_dup, "output_files/qc_gate_duplication.csv")
knitr::kable(ic_dup %>% select(-in_house_gate, -vendor_metric),
             caption = "Our IC gate vs Alamar's own IC Median metric")
cat(sprintf("Verdict: %s (Jaccard %.3f)\n", ic_dup$verdict, jacc))

# The wells the two disagree on are the informative ones — usually boundary cases.
disc <- ic_vendor %>%
  filter(SampleID %in% union(setdiff(ours, theirs), setdiff(theirs, ours))) %>%
  left_join(df_temp %>% select(SampleID, Run, Bay, IC_Reads, Median_Value), by = "SampleID") %>%
  transmute(SampleID, Run, Bay, IC_Reads,
            ratio_to_plate_median = round(IC_Reads / Median_Value, 3),
            `IC Median` = round(ic_med, 4),
            called_by = ifelse(SampleID %in% ours, "ours only", "Alamar only"),
            dist_from_vendor_bar = round(abs(abs(ic_med) - ic_median_thresh), 4))
if (nrow(disc))
  knitr::kable(arrange(disc, dist_from_vendor_bar),
               caption = "Wells the two calls disagree on (nearest the bar first)")

# The stated lower guard `IC Reads < 1000`: does it ever bind, given that the
# plate-relative bound is pmax(0.6 * median, 1000)?
ic_geom <- df_temp %>%
  group_by(Run, Bay) %>%
  summarise(plate_rel_lower = 0.6 * first(Median_Value),
            floor_binds = plate_rel_lower < 1000, .groups = "drop")
cat(sprintf("IC lower bound: the 1000 floor binds in %d of %d plate-bays; %d wells fall below 1000 outright.\n",
            sum(ic_geom$floor_binds), nrow(ic_geom), sum(df_temp$IC_Reads < 1000, na.rm = TRUE)))
```
Note: As expected, the samples with very high reads for mCherry, have lower average NPQ than others.

### D.2.4 Extreme Matching Reads
Here we are identifying samples that have extreme numbers of total matching reads, in aggregate, within their plate. These are **flagged** samples, not triaged.

> **This gate is one-sided in practice, and the name hides it.** The upper bound is
> plate-relative and works. The lower bound is not plate-relative on any real run:
> total reads are strongly right-skewed *within* a plate (typically Q1 ≈ 3.6M,
> median ≈ 5.7M, Q3 ≈ 10.5M — a Q3/Q1 ratio near 2.9), so `median − 4 × IQR` is
> negative and `pmax()` collapses it to the absolute floor every time.
>
> **This is not a scale bug that a transform repairs.** Rescaling was tested on
> this run: a log2 fence leaves the lower bound below the floor in 49 of 50 bays
> *and* guts the upper side (94 flags → 3); a lower semi-IQR fence
> `median − k(median − Q1)` is negative in 48 of 50 bays at k = 4 and flags 93
> wells at k = 2; a −40% multiplicative bound, like the one the IC gate uses,
> flags 1031 wells (24.5%). The within-plate spread of total reads is simply too
> wide to support *any* lower fence that is both plate-relative and selective.
>
> So the honest description is: **a plate-relative HIGH-read flag, plus an
> absolute low-read floor.** The code below keeps `pmax()` — so behaviour is
> unchanged, and the plate-relative bound will bind on a run where it can — but
> reports the two mechanisms separately and states per plate-bay which one is
> actually holding the line.
>
> The consequence to carry forward: **a chronically low-input plate position
> cannot be caught here**, however far below its plate it sits, unless it drops
> under `read_floor_abs`. That question has to be asked across plates at the
> position level instead — which is what D.5.4 does.

```{r excessive_matching_reads, message=FALSE, warning=FALSE, echo=TRUE}
# ---- Matching-Reads QC ----
library(dplyr)
library(ggplot2)
library(readr)


# Merge matching reads info
matching_df <- alamar_qc %>%
  select(SampleID = `Sample Name`, Reads) %>%          # rename + keep Reads
  filter(SampleID %in% rownames(combined_qc)) %>%      # keep only samples in combined_qc
  left_join(                                           
    combined_qc %>%                                     # pull Run + Bay from combined_qc
      rownames_to_column("SampleID") %>%
      select(SampleID, Run, Bay),
    by = "SampleID"
  )

# ---- Outlier detection by Run × Bay ----
# The two bounds are NOT the same kind of test and are no longer computed as if
# they were. Upper: plate-relative. Lower: pmax(absolute floor, plate-relative),
# which in practice is the floor — see the note above and read_gate_geometry.csv.

matching_outliers <- matching_df %>%
  group_by(Run, Bay) %>%
  mutate(
    Median = median(Reads, na.rm = TRUE),
    Q1     = quantile(Reads, 0.25, na.rm = TRUE),
    Q3     = quantile(Reads, 0.75, na.rm = TRUE),
    IQR    = Q3 - Q1,
    Upper_Bound       = Median + read_out_thresh * IQR,
    Lower_Bound_plate = Median - read_out_thresh * IQR,   # what the fence WOULD be
    Lower_Bound       = pmax(read_floor_abs, Lower_Bound_plate),
    # Which mechanism is actually holding the lower line on this plate?
    lower_is_plate_relative = Lower_Bound_plate > read_floor_abs,
    High_Read_Plate   = Reads > Upper_Bound,
    Low_Read          = Reads < Lower_Bound,
    Matching_Outlier  = High_Read_Plate | Low_Read
  ) %>%
  ungroup()

# ---- Is the lower bound reachable at all? ----
read_gate_geometry <- matching_outliers %>%
  group_by(Run, Bay) %>%
  summarise(
    n_wells           = n(),
    median_reads      = first(Median),
    IQR_reads         = first(IQR),
    lower_bound_plate = first(Lower_Bound_plate),
    lower_bound_used  = first(Lower_Bound),
    upper_bound       = first(Upper_Bound),
    lower_is_plate_relative = first(lower_is_plate_relative),
    min_reads_here    = min(Reads, na.rm = TRUE),
    n_high            = sum(High_Read_Plate, na.rm = TRUE),
    n_low             = sum(Low_Read, na.rm = TRUE),
    .groups = "drop"
  )
write_csv(read_gate_geometry, "output_files/read_gate_geometry.csv")

cat(sprintf("Lower read bound is plate-relative in %d of %d plate-bays (the %s floor binds in the rest).\n",
            sum(read_gate_geometry$lower_is_plate_relative), nrow(read_gate_geometry),
            format(read_floor_abs, big.mark = ",", scientific = FALSE)))
cat(sprintf("Read flags: %d high (plate-relative) + %d low (%s) = %d of %d wells.\n",
            sum(matching_outliers$High_Read_Plate, na.rm = TRUE),
            sum(matching_outliers$Low_Read, na.rm = TRUE),
            ifelse(any(read_gate_geometry$lower_is_plate_relative), "mixed bound", "absolute floor only"),
            sum(matching_outliers$Matching_Outlier, na.rm = TRUE), nrow(matching_outliers)))

# Convert factor levels (ensure ordered)
matching_outliers$Run <- factor(matching_outliers$Run)
matching_outliers$Bay <- factor(matching_outliers$Bay, levels = c("Bay1", "Bay2", "Bay3"))

# Assign Bay colors
bay_colors <- c("Bay1" = "red", "Bay2" = "green", "Bay3" = "blue")
#bay_shapes <- c("Bay1" = 16, "Bay2" = 17, "Bay3" = 15)  # optional: circle, triangle, square

p <- ggplot(matching_outliers, aes(x = Run, y = Reads, fill = Bay)) +
  geom_violin(alpha = 0.5, trim = FALSE, position = position_dodge(width = 0.8)) +
  geom_jitter(aes(color = Bay),
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.75, size = 1.3) +
  scale_fill_manual(values = bay_colors) +
  scale_color_manual(values = bay_colors) +
  theme_minimal() +
  labs(
    title = "Run × Bay Effects: Total Reads",
    x = "Run",
    y = "Reads"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p)

read_outliers <- matching_outliers %>%
    filter(Matching_Outlier == TRUE) %>%
    arrange(desc(Reads)) %>%
    mutate(Direction = dplyr::case_when(
             High_Read_Plate          ~ "HIGH (plate-relative)",
             lower_is_plate_relative  ~ "LOW (plate-relative)",
             TRUE                     ~ "LOW (absolute floor)")) %>%
    select(SampleID, Run, Bay, Direction, Lower_Bound, Upper_Bound, Reads)

# Print table
knitr::kable(
  read_outliers, 
  caption = "Matching-Read Outliers (Sorted by Reads)"
)

# Save to CSV
write_csv(
  read_outliers %>%
    arrange(Run, Bay, desc(Reads)),
  "output_files/flagged_read_outliers.csv"
)

# read_outliers are our flagged samples

```

## D.3 Sample Traige
Here we are identifying samples to be triaged based on our sample QC and comparing them to the Alamar QC.

```{r sample_triage, message=FALSE, warning=FALSE, echo=TRUE}
# --- Triage samples for removal ---

# Bad Samples

# Identify PCA extreme outliers
pca_outliers <- c(pca_df$Sample[pca_df$extreme_outlier],pca_df2$Sample[pca_df2$extreme_outlier])

# Identify samples with significant outlier burden
# Combine upper and lower matrices for total FDR per sample
upper_sig <- upper_df$SampleID[upper_df$FDR < FDR_threshold]
lower_sig <- lower_df$SampleID[lower_df$FDR < FDR_threshold]
outlier_burden_samples <- unique(c(upper_sig, lower_sig))

# Wells below the vendor Reads floor (identified in B; see A.3). Restricted to the analysis set so
# control wells are not injected into the study-sample triage.
low_read_triage <- intersect(as.character(low_read_samples), rownames(combined_qc))

triage_df <- bind_rows(
  tibble(SampleID = low_read_triage, Reason = "Low read depth"),
  tibble(SampleID = as.character(unlist(pca_outliers)), Reason = "PCA outlier"),
  tibble(SampleID = as.character(unlist(outlier_burden_samples)), Reason = "High outlier burden (FDR)"),
  tibble(SampleID = as.character(unlist(IC_outlier_samples)), Reason = "IC outlier"),
  tibble(SampleID = as.character(unlist(bad_samples)), Reason = "Bad Data"),
)

# If a sample appears in multiple categories, collapse reasons
triage_df <- triage_df %>%
  group_by(SampleID) %>%
  summarize(Reason = paste(unique(Reason), collapse = "; "), .groups = "drop")

# --- Per-axis membership, alongside the collapsed reason string ----------------
# A well can exit on more than one axis. Any report that assigns ONE reason per
# well — by precedence, or by reading the first entry of `Reason` — produces a
# valid partition whose per-axis numbers are NOT axis totals. Downstream consumers
# have made exactly that mistake, undercounting the PCA and burden axes badly.
# These booleans are the axis totals; use them for "how many wells did this screen
# catch", and use `Reason`/precedence only when the parts must sum to the whole.
tri_ids  <- triage_df$SampleID
axis_ids <- list(
  tri_low_reads = low_read_triage,
  tri_IC        = as.character(unlist(IC_outlier_samples)),
  tri_PCA       = as.character(unlist(pca_outliers)),
  tri_burden    = as.character(unlist(outlier_burden_samples)),
  tri_bad_data  = as.character(unlist(bad_samples))
)
for (a in names(axis_ids)) triage_df[[a]] <- tri_ids %in% axis_ids[[a]]
triage_df$n_axes <- rowSums(as.matrix(triage_df[names(axis_ids)]))

triage_df$AlamarWarning <- triage_df$SampleID  %in% warn_samps$`Sample Name`

# Axis totals vs what a one-reason-per-well report would say, side by side.
prec <- names(axis_ids)   # precedence order: low reads > IC > PCA > burden > bad data
# `low reads` leads because it is the least ambiguous cause: a well that did not sequence has no
# usable signal, so any PCA/burden hit it also earns is a consequence, not a second finding.
sole_reason <- apply(as.matrix(triage_df[prec]), 1, function(r) prec[which(r)[1]])
triage_axis_overlap <- tibble(
  axis            = prec,
  on_axis_true    = vapply(prec, function(a) sum(triage_df[[a]]), integer(1)),
  by_precedence   = vapply(prec, function(a) sum(sole_reason == a, na.rm = TRUE), integer(1)),
  sole_axis_only  = vapply(prec, function(a)
                      sum(triage_df[[a]] & triage_df$n_axes == 1), integer(1))
) %>%
  mutate(undercount_pct = ifelse(on_axis_true > 0,
                                 round(100 * (on_axis_true - by_precedence) / on_axis_true, 1), 0))
write_csv(triage_axis_overlap, "output_files/triage_axis_overlap.csv")

triage_axis_combinations <- triage_df %>%
  count(across(all_of(prec)), name = "n_wells") %>%
  arrange(desc(n_wells))
write_csv(triage_axis_combinations, "output_files/triage_axis_combinations.csv")

knitr::kable(triage_axis_overlap,
             caption = paste0("Per-axis triage totals. `on_axis_true` is how many wells each screen caught; ",
                              "`by_precedence` is what a one-reason-per-well report shows. ",
                              sum(triage_df$n_axes > 1), " of ", nrow(triage_df),
                              " wells carry more than one axis."))
# Print to RMarkdown
knitr::kable(
  triage_df, 
  caption = "TRIAGED SAMPLES"
)

warn_samps$Triage <- warn_samps$`Sample Name` %in% triage_df$SampleID
knitr::kable(
  warn_samps %>% select(`Sample Name`, `IC Median`, Detectability, `QC Status`,Triage), 
  caption = "ALAMAR WARN SAMPLES"
)

#Flagged excessive_matching_reads (not removed)
read_outliers$Triage <- read_outliers$SampleID %in% triage_df$SampleID
knitr::kable(
  read_outliers %>% select(SampleID, Run, Bay, Reads, Triage), 
  caption = "FLAGGED SAMPLES"
)
  
# Save to CSV
write_csv(triage_df, "output_files/samples_to_triage.csv")

# Remove triaged samples from main dataset
combined_qc_filtered <- combined_qc %>%
  rownames_to_column("SampleID") %>%
  filter(!SampleID %in% triage_df$SampleID) %>%
  column_to_rownames("SampleID")

#Add count summary
qc_summary <- combined_qc_filtered %>%
  group_by(Run, Bay) %>%
  summarise(combined_qc_filtered = n(), .groups = "drop") %>%
  arrange(Run, Bay)

knitr::kable(
  qc_summary, 
  caption = "Summary of post-QC counts:"
)

```

## D.4 Per-plate QC Summary
One row per plate: post-QC retained count, our triaged and flagged samples, and Alamar warnings.

```{r per_plate_summary, message=FALSE, warning=FALSE, echo=TRUE}
plate_lookup <- alamar_qc %>%
  transmute(SampleID = `Sample Name`,
            Run = stringr::str_extract(plateID, "^[0-9]{8}-[0-9]{4}"),
            Bay = stringr::str_extract(plateID, "Bay[0-9]+"))

count_by_plate <- function(ids, nm) {
  tibble(SampleID = unique(as.character(ids))) %>%
    inner_join(plate_lookup, by = "SampleID") %>%
    count(Run, Bay, name = nm)
}

plate_summary <- plate_lookup %>%
  distinct(Run, Bay) %>%
  left_join(qc_summary %>% rename(Retained_postQC = combined_qc_filtered), by = c("Run","Bay")) %>%
  left_join(count_by_plate(triage_df$SampleID,       "Triaged"),         by = c("Run","Bay")) %>%
  left_join(count_by_plate(read_outliers$SampleID,   "Flagged_Reads"),   by = c("Run","Bay")) %>%
  left_join(count_by_plate(warn_samps$`Sample Name`, "Alamar_Warnings"), by = c("Run","Bay")) %>%
  mutate(across(c(Retained_postQC, Triaged, Flagged_Reads, Alamar_Warnings),
                ~ tidyr::replace_na(., 0L))) %>%
  arrange(Run, Bay)

knitr::kable(
  plate_summary,
  caption = "Per-plate QC summary: pipeline triage/flags vs. Alamar warnings"
) %>%
  kableExtra::kable_styling(full_width = TRUE,
                            bootstrap_options = c("striped", "hover", "condensed"))

write_csv(plate_summary, "output_files/per_plate_QC_summary.csv")
```

## D.5 QC Screen Audit

Everything above asks *what happened to the samples*. This section asks **whether
the screens are any good**: does each gate fire at all, is it independent of the
others, where does its cut sit on its own statistic, and is it keying on plate
position rather than on the specimen.

**Diagnostic only. Nothing here removes or flags a sample.**

```{r audit-setup, message=FALSE, warning=FALSE, echo=TRUE}
# The audit universe: every well that entered the screens.
audit <- tibble(SampleID = rownames(combined_qc),
                Run = combined_qc$Run, Bay = combined_qc$Bay) %>%
  mutate(plate = paste(Run, Bay)) %>%
  left_join(well_pos %>% select(SampleID, Well, wellRow, wellCol), by = "SampleID") %>%
  left_join(intensity_df %>% select(SampleID, mean_INT, mean_INT_z, intensity_flag),
            by = "SampleID") %>%
  left_join(matching_outliers %>% select(SampleID, Reads, High_Read_Plate, Low_Read),
            by = "SampleID") %>%
  left_join(df_temp %>% select(SampleID, IC_Reads, Median_Value, IC_Outlier = Outlier_Flag),
            by = "SampleID") %>%
  left_join(alamar_qc %>% transmute(SampleID = `Sample Name`,
                                    IC_Median = suppressWarnings(as.numeric(`IC Median`)) / .ic_scale,
                                    Detectability = suppressWarnings(as.numeric(Detectability)),
                                    QC_Status = `QC Status`),
            by = "SampleID") %>%
  left_join(triage_df %>% select(SampleID, Reason, tri_low_reads, tri_IC, tri_PCA, tri_burden,
                                 tri_bad_data, n_axes), by = "SampleID") %>%
  left_join(burden_stat %>% select(SampleID, fdr_min, n_out), by = "SampleID") %>%
  left_join(pca_df %>% select(SampleID = Sample, PC_z_round1 = PC_z,
                              PC_z_cohort, pca_cohort_flag, pca_masked_by_plate), by = "SampleID") %>%
  mutate(across(c(tri_low_reads, tri_IC, tri_PCA, tri_burden, tri_bad_data, High_Read_Plate, Low_Read,
                  IC_Outlier), ~ tidyr::replace_na(., FALSE)),
         triaged      = SampleID %in% triage_df$SampleID,
         read_flagged = SampleID %in% read_outliers$SampleID,
         n_axes       = tidyr::replace_na(n_axes, 0L))

# Detectability arrives as a 0-1 proportion or a 0-100 percent depending on the
# export; put it on 0-1 so the vendor bar below means one thing.
if (max(audit$Detectability, na.rm = TRUE) > 1.5) audit$Detectability <- audit$Detectability / 100

cat(sprintf("Audit universe: %d wells, %d plate-bays, %d distinct well positions.\n",
            nrow(audit), dplyr::n_distinct(audit$plate), dplyr::n_distinct(audit$Well)))
```

### D.5.1 Does each gate fire?

A gate that fires on **zero** wells has not been shown to be safe — it has not
been tested. It could be correctly finding nothing, or it could be unreachable by
construction. Those are very different, and the count alone cannot tell them
apart, so the headroom to each cut is reported next to it.

```{r gate-inventory, message=FALSE, warning=FALSE, echo=TRUE}
gate_row <- function(gate, action, fired, n, headroom = NA_character_, note = "") {
  tibble(gate, action, n_fired = sum(fired, na.rm = TRUE), n_eval = n,
         pct = round(100 * sum(fired, na.rm = TRUE) / n, 3),
         headroom, status = ifelse(sum(fired, na.rm = TRUE) == 0,
                                   "INERT - untested, not proven safe", "fires"),
         note)
}
N <- nrow(audit)
.hr <- function(x, cut) sprintf("max %.3f vs cut %.3f (%+.3f)", x, cut, x - cut)

gate_inventory <- bind_rows(
  gate_row("triage: IC outlier", "REMOVE", audit$tri_IC, N,
           sprintf("%d of %d plate-bays have the 1000 floor binding",
                   sum(ic_geom$floor_binds), nrow(ic_geom)),
           "same quantity as the vendor IC Median metric - see D.2.3"),
  gate_row("triage: PCA round 1", "REMOVE", audit$SampleID %in% extreme_samples_round1, N,
           .hr(max(pca_df$PC_z, na.rm = TRUE), PCA_SD), ""),
  gate_row("triage: PCA round 2", "REMOVE", audit$SampleID %in% extreme_samples_round2, N,
           if (nrow(pca_df2)) .hr(max(pca_df2$PC_z, na.rm = TRUE), PCA_SD) else "not run",
           "second pass, after round-1 outliers are removed"),
  gate_row("triage: outlier burden", "REMOVE", audit$tri_burden, N,
           local({
             below <- op_points$fdr_min[op_points$fdr_min <  FDR_threshold]
             above <- op_points$fdr_min[op_points$fdr_min >= FDR_threshold]
             if (!length(below) || !length(above)) "cut is outside the observed FDR range"
             else sprintf("cut sits between FDR %.4g and %.4g - nothing to tune between",
                          max(below), min(above))
           }),
           "quantised: an integer anchor count, not a dial"),
  gate_row("triage: bad data", "REMOVE", audit$tri_bad_data, N, NA_character_,
           "non-numeric NPQ entries"),
  gate_row("flag: high reads (plate-relative)", "FLAG", audit$High_Read_Plate, N,
           NA_character_, ""),
  gate_row("flag: low reads", "FLAG", audit$Low_Read, N,
           sprintf("plate-relative lower bound reachable in %d of %d bays",
                   sum(read_gate_geometry$lower_is_plate_relative), nrow(read_gate_geometry)),
           "absolute floor in practice - see D.2.4"),
  # Vendor bars come from the export's own preamble (section B), so this audit
  # stays correct if Alamar changes one, and uses the right bar for the matrix.
  gate_row(sprintf("vendor: Detectability < %.0f%%", 100 * vendor_thresholds$detectability_min),
           "vendor warning", audit$Detectability < vendor_thresholds$detectability_min, N,
           NA_character_,
           sprintf("Alamar QC Status driver; bar is for %s", vendor_thresholds$sample_matrix)),
  gate_row(sprintf("vendor: |IC Median| > %.2f", vendor_thresholds$ic_median),
           "vendor warning", abs(audit$IC_Median) > vendor_thresholds$ic_median, N,
           NA_character_, "Alamar QC Status driver"),
  gate_row(sprintf("vendor: IC Reads < %s", format(vendor_thresholds$ic_reads_min, big.mark = ",")),
           "vendor warning", audit$IC_Reads < vendor_thresholds$ic_reads_min, N,
           sprintf("lowest IC Reads seen: %s",
                   format(min(audit$IC_Reads, na.rm = TRUE), big.mark = ",")), ""),
  gate_row(sprintf("vendor: Reads < %s",
                   format(vendor_thresholds$reads_min, big.mark = ",", scientific = FALSE)),
           "vendor warning", audit$Reads < vendor_thresholds$reads_min, N,
           sprintf("lowest Reads seen: %s",
                   format(min(audit$Reads, na.rm = TRUE), big.mark = ",")), ""),
  gate_row("informational: whole-panel intensity", "NONE - reported only",
           audit$intensity_flag %in% c("HIGH", "LOW"), N,
           sprintf("range %.2f to %.2f SD", min(audit$mean_INT_z, na.rm = TRUE),
                   max(audit$mean_INT_z, na.rm = TRUE)),
           "D.0 - a covariate, never a gate")
)
write_csv(gate_inventory, "output_files/qc_gate_inventory.csv")

knitr::kable(gate_inventory, caption = "QC gate inventory: what each screen actually did") %>%
  kableExtra::kable_styling(full_width = TRUE,
                            bootstrap_options = c("striped", "hover", "condensed"))

.inert <- gate_inventory$gate[gate_inventory$n_fired == 0]
if (length(.inert))
  cat(sprintf("\nINERT on this run (%d): %s\n  These are UNTESTED. Check the headroom column before assuming they are simply finding nothing.\n",
              length(.inert), paste(.inert, collapse = "; ")))
```

### D.5.2 Are the gates independent of each other?

If two screens remove the same wells they are one screen reported twice, and the
methods text should not present them as corroborating each other.

```{r gate-overlap, message=FALSE, warning=FALSE, echo=TRUE}
axes <- c(tri_IC = "IC", tri_PCA = "PCA", tri_burden = "burden")
pairs <- utils::combn(names(axes), 2, simplify = FALSE)
gate_overlap <- bind_rows(lapply(pairs, function(p) {
  a <- audit[[p[1]]]; b <- audit[[p[2]]]
  tibble(gate_a = axes[[p[1]]], gate_b = axes[[p[2]]],
         n_a = sum(a), n_b = sum(b), n_both = sum(a & b),
         jaccard = round(sum(a & b) / max(1, sum(a | b)), 3))
}))
# The read flag is a flag, not a triage axis, but "is it just triage again?" is
# the same question.
gate_overlap <- bind_rows(gate_overlap, tibble(
  gate_a = "read flag", gate_b = "any triage",
  n_a = sum(audit$read_flagged), n_b = sum(audit$triaged),
  n_both = sum(audit$read_flagged & audit$triaged),
  jaccard = round(sum(audit$read_flagged & audit$triaged) /
                    max(1, sum(audit$read_flagged | audit$triaged)), 3)))
write_csv(gate_overlap, "output_files/qc_gate_overlap.csv")
knitr::kable(gate_overlap, caption = "Pairwise overlap between screens (Jaccard)")
```

### D.5.3 Where does each cut sit on its own statistic?

The nearest retained well and the nearest removed well on each axis. If they are
touching, the cut is splitting a continuum rather than separating two
populations, and the exact threshold is doing more work than it looks.

```{r cut-neighbourhood, message=FALSE, warning=FALSE, echo=TRUE}
cut_nbhd <- function(stat, removed, gate, cut) {
  ok <- !is.na(stat)
  if (!any(ok & removed) || !any(ok & !removed)) return(NULL)
  # Statistics here are all "larger = worse" except IC Median, handled by caller.
  tibble(gate, cut,
         nearest_removed = round(min(stat[ok & removed]), 4),
         nearest_retained = round(max(stat[ok & !removed]), 4),
         gap = round(min(stat[ok & removed]) - max(stat[ok & !removed]), 4))
}
# FDR can be exactly 0, so clamp before the log rather than carrying an Inf into
# the min/max below.
.nlp <- -log10(pmax(audit$fdr_min, .Machine$double.xmin))
# Our IC gate fires on |IC Reads / plate median - 1| > 0.40, so that ratio — not
# the vendor's IC Median — is the statistic its own cut sits on.
.ic_stat <- abs(audit$IC_Reads / audit$Median_Value - 1)
gate_cuts <- bind_rows(
  cut_nbhd(.ic_stat, audit$tri_IC, "IC (|IC reads / plate median - 1|)", 0.40),
  cut_nbhd(audit$PC_z_round1, audit$SampleID %in% extreme_samples_round1,
           "PCA round 1 (PC_z)", PCA_SD),
  cut_nbhd(.nlp, audit$tri_burden, "burden (-log10 FDR)", round(-log10(FDR_threshold), 4)),
  # Cross-axis, and labelled as such: how far our call sits from the vendor bar.
  # A negative gap here is disagreement between two gates, not a split population.
  cut_nbhd(abs(audit$IC_Median), audit$tri_IC,
           "our IC call read against the VENDOR statistic", ic_median_thresh)
)
write_csv(gate_cuts, "output_files/gate_cut_neighbourhood.csv")
knitr::kable(gate_cuts, caption = "Nearest removed vs nearest retained well on each axis. A gap at or below 0 means the cut lands inside a continuous population.")
```

### D.5.4 Is a screen keying on plate position rather than on the specimen?

A well position that is triaged far more often than the others across many plates
is a **layout or hardware** problem, not a specimen problem — and it is invisible
to every within-plate screen above, because each of those only ever compares a
well to its own plate.

The test is a permutation, not a chi-square: triage counts spread over ~96
positions leave most cells at 0 or 1, where an asymptotic test is not conservative
but wrong. Labels are shuffled **within each plate-bay**, holding each plate's
triage count fixed, so plate-level effects cannot manufacture a position signal.
The statistic is the **maximum** per-position count, which needs no further
multiple-testing correction.

> **On the word "site".** This asks about the *physical site on the plate*.
> Clustering by **clinical collection site** is a different and equally important
> question, and it cannot be asked here: Primary QC is deliberately agnostic to
> phenotype and covariates, and the site label only exists after the metadata
> merge. The well-level table in D.5.6 carries every gate flag keyed by
> `SampleID` precisely so that question can be answered downstream with a single
> join — see `Secondary_QC/Triage_Review/`.

```{r position-effects, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(123)

perm_max_position <- function(flag, pos, plate, n_perm) {
  keep <- !is.na(pos) & !is.na(flag)
  flag <- flag[keep]; pos <- pos[keep]; plate <- plate[keep]
  lev  <- sort(unique(pos))
  obs_counts <- tabulate(match(pos[flag], lev), nbins = length(lev))
  obs <- max(obs_counts)
  # Under the null, each plate's k flagged wells are a uniform random subset of
  # that plate's wells. Draw per plate, so plate composition is held fixed.
  pos_by_plate <- split(pos, plate)
  k_by_plate   <- vapply(split(flag, plate), sum, integer(1))
  null <- replicate(n_perm, {
    drawn <- unlist(Map(function(pp, k) if (k > 0) sample(pp, k) else pp[0],
                        pos_by_plate, k_by_plate), use.names = FALSE)
    max(tabulate(match(drawn, lev), nbins = length(lev)))
  })
  list(obs = obs, p = (1 + sum(null >= obs)) / (n_perm + 1),
       null_mean = mean(null),
       tab = tibble(Well = lev, n_flagged = obs_counts,
                    n_wells = as.integer(table(factor(pos, levels = lev)))) %>%
             mutate(pct = round(100 * n_flagged / n_wells, 2)) %>%
             arrange(desc(n_flagged)))
}

if (all(is.na(audit$Well))) {
  cat("No well positions available — plate-position diagnostics skipped.\n")
} else {
  pe_tri  <- perm_max_position(audit$triaged,      audit$Well, audit$plate, n_perm)
  pe_read <- perm_max_position(audit$read_flagged, audit$Well, audit$plate, n_perm)

  write_csv(pe_tri$tab,  "output_files/well_position_effects.csv")
  write_csv(pe_read$tab, "output_files/well_position_effects_readflag.csv")

  cat(sprintf("Triage by position: worst well is %s at %d of %d (%.1f%%), against %.1f%% elsewhere.\n",
              pe_tri$tab$Well[1], pe_tri$tab$n_flagged[1], pe_tri$tab$n_wells[1],
              pe_tri$tab$pct[1],
              100 * sum(pe_tri$tab$n_flagged[-1]) / sum(pe_tri$tab$n_wells[-1])))
  cat(sprintf("  permutation on the maximum: observed %d, null mean %.2f, p = %.4f\n",
              pe_tri$obs, pe_tri$null_mean, pe_tri$p))
  cat(sprintf("Read flags by position: max %d, null mean %.2f, p = %.4f\n",
              pe_read$obs, pe_read$null_mean, pe_read$p))

  knitr::kable(head(pe_tri$tab, 10),
               caption = "Most-triaged well positions across all plate-bays")

  # Which axis does the worst position exit on? A single axis points at one
  # mechanism; a spread points at a generally bad well.
  .w <- pe_tri$tab$Well[1]
  .wa <- audit %>% filter(Well == .w, triaged) %>%
    summarise(IC = sum(tri_IC), PCA = sum(tri_PCA), burden = sum(tri_burden),
              bad_data = sum(tri_bad_data))
  cat(sprintf("Position %s exits via: IC %d, PCA %d, burden %d, bad data %d\n",
              .w, .wa$IC, .wa$PCA, .wa$burden, .wa$bad_data))
}
```

#### Is there a standing spatial gradient underneath it?

A single bad position is usually the extreme of a smooth trend across the plate
rather than an isolated defect. The IC is the right quantity to look at: NPQ is
normalised to it, so a position that recovers less IC shifts every reported value
for the specimen that lands there.

```{r position-gradient, message=FALSE, warning=FALSE, echo=TRUE}
if (!all(is.na(audit$Well))) {
  pos_map <- audit %>%
    filter(!is.na(Well), !is.na(IC_Median)) %>%
    group_by(Well, wellRow, wellCol) %>%
    summarise(n = n(), median_IC_Median = round(median(IC_Median, na.rm = TRUE), 4),
              median_mean_INT = round(median(mean_INT, na.rm = TRUE), 4), .groups = "drop") %>%
    arrange(median_IC_Median)
  write_csv(pos_map, "output_files/position_map_IC.csv")

  print(
    ggplot(pos_map, aes(x = factor(wellCol), y = factor(wellRow, levels = rev(sort(unique(wellRow)))),
                        fill = median_IC_Median)) +
      geom_tile(colour = "white") +
      scale_fill_gradient2(low = "#B2182B", mid = "#F7F7F7", high = "#2166AC", midpoint = 0) +
      theme_minimal() +
      labs(title = "Median IC recovery by plate position, pooled over all plate-bays",
           subtitle = "Red = this position systematically recovers less IC than its plate",
           x = "Column", y = "Row", fill = "median\nIC Median")
  )

  # Pre-specified direction: does IC recovery fall from row A to row H? Reported
  # per plate-bay, so a single odd plate cannot carry it.
  row_grad <- audit %>%
    filter(!is.na(wellRow), !is.na(IC_Median)) %>%
    mutate(row_i = match(wellRow, LETTERS)) %>%
    group_by(Run, Bay) %>%
    filter(dplyr::n_distinct(row_i) > 2) %>%
    summarise(spearman = suppressWarnings(cor(row_i, IC_Median, method = "spearman",
                                              use = "complete.obs")),
              .groups = "drop")
  write_csv(row_grad, "output_files/ic_row_gradient.csv")
  cat(sprintf("IC recovery falls from row A to row H in %d of %d plate-bays (median rho %.2f).\n",
              sum(row_grad$spearman < 0, na.rm = TRUE), nrow(row_grad),
              median(row_grad$spearman, na.rm = TRUE)))

  # --- region contrasts, permuted within plate ------------------------------
  # One helper, used for every region test below. Shuffling the region label
  # WITHIN each plate-bay means a plate that simply ran low overall cannot
  # manufacture a contrast.
  cor_df <- audit %>% filter(!is.na(Well), !is.na(IC_Median))
  region_test <- function(inreg, label) {
    inreg <- !is.na(inreg) & inreg
    if (!any(inreg) || !any(!inreg)) return(invisible(NULL))
    obs <- median(cor_df$IC_Median[inreg], na.rm = TRUE) -
           median(cor_df$IC_Median[!inreg], na.rm = TRUE)
    ord <- order(paste(cor_df$Run, cor_df$Bay))
    v   <- cor_df$IC_Median[ord]
    grp <- split(seq_along(ord), paste(cor_df$Run, cor_df$Bay)[ord])
    lab <- inreg[ord]
    nul <- replicate(n_perm, {
      s <- logical(length(lab)); for (g in grp) s[g] <- sample(lab[g])
      median(v[s], na.rm = TRUE) - median(v[!s], na.rm = TRUE)
    })
    cat(sprintf("  %-34s %+.4f vs %+.4f elsewhere  diff %+.4f  perm p = %.4f\n",
                label, median(cor_df$IC_Median[inreg], na.rm = TRUE),
                median(cor_df$IC_Median[!inreg], na.rm = TRUE), obs,
                (1 + sum(abs(nul) >= abs(obs))) / (n_perm + 1)))
    invisible(obs)
  }

  # Layout-agnostic and always run: no region is named in advance, so these are
  # valid on any plate geometry and cannot be accused of picking their target.
  cat("Spatial structure (permuted within plate-bay):\n")
  .rl <- sort(unique(cor_df$wellRow)); .cl <- sort(unique(cor_df$wellCol))
  if (length(.rl) > 2 && length(.cl) > 2)
    region_test(cor_df$wellRow %in% .rl[c(1, length(.rl))] |
                cor_df$wellCol %in% .cl[c(1, length(.cl))], "edge vs interior")

  # Column gradient, the companion to the row gradient above.
  col_grad <- audit %>%
    filter(!is.na(wellCol), !is.na(IC_Median)) %>%
    group_by(Run, Bay) %>%
    filter(dplyr::n_distinct(wellCol) > 2) %>%
    summarise(spearman = suppressWarnings(cor(wellCol, IC_Median, method = "spearman",
                                              use = "complete.obs")), .groups = "drop")
  cat(sprintf("  IC recovery falls left-to-right in %d of %d plate-bays (median rho %.2f)\n",
              sum(col_grad$spearman < 0, na.rm = TRUE), nrow(col_grad),
              median(col_grad$spearman, na.rm = TRUE)))

  # The empirically worst quadrant, reported DESCRIPTIVELY. No p-value: the
  # region was chosen by looking, so any p-value here would be a search result
  # dressed as a test. It is a lead for the next run's `prior_region`.
  .rq <- .rl[seq_len(ceiling(length(.rl) / 2))]; .cq <- .cl[seq_len(ceiling(length(.cl) / 2))]
  quads <- list("top-left" = list(.rq, .cq),
                "top-right" = list(.rq, setdiff(.cl, .cq)),
                "bottom-left" = list(setdiff(.rl, .rq), .cq),
                "bottom-right" = list(setdiff(.rl, .rq), setdiff(.cl, .cq)))
  qd <- vapply(quads, function(q) {
    i <- cor_df$wellRow %in% q[[1]] & cor_df$wellCol %in% q[[2]]
    if (!any(i)) NA_real_ else median(cor_df$IC_Median[i], na.rm = TRUE) -
      median(cor_df$IC_Median[!i], na.rm = TRUE) }, numeric(1))
  cat(sprintf("  lowest-IC quadrant (descriptive, chosen by looking): %s at %+.4f vs rest\n",
              names(qd)[which.min(qd)], min(qd, na.rm = TRUE)))
  cat("    -> to TEST it, set prior_region in A.3 and re-run on the NEXT dataset.\n")

  # Confirmatory only if a region was named in advance (A.3 `prior_region`).
  if (!is.null(prior_region)) {
    pr <- cor_df$wellRow %in% prior_region$rows & cor_df$wellCol %in% prior_region$cols
    cat("Prior-specified region (named before this run, so this p-value is confirmatory):\n")
    region_test(pr, sprintf("rows %s x cols %s",
                            paste(prior_region$rows, collapse = ""),
                            paste(range(prior_region$cols), collapse = "-")))
  }
}
```

### D.5.5 The intensity axis against the gates

`mean_INT` (D.0) is the level statistic none of the gates compute. Reading each
triage class against it shows what each screen is actually selecting — and
whether the pipeline is quietly reshaping the intensity distribution.

**Report per class, never pooled.** The classes point in opposite directions on
this axis, so a pooled mean describes none of them.

```{r intensity-vs-gates, message=FALSE, warning=FALSE, echo=TRUE}
cohens_d <- function(x, y) {
  x <- x[!is.na(x)]; y <- y[!is.na(y)]
  if (length(x) < 2 || length(y) < 2) return(NA_real_)
  s <- sqrt(((length(x)-1)*var(x) + (length(y)-1)*var(y)) / (length(x)+length(y)-2))
  if (!is.finite(s) || s == 0) return(NA_real_)
  (mean(x) - mean(y)) / s
}
auc <- function(x, y) {   # P(random x > random y), ties at 0.5
  x <- x[!is.na(x)]; y <- y[!is.na(y)]
  if (!length(x) || !length(y)) return(NA_real_)
  r <- rank(c(x, y))
  (sum(r[seq_along(x)]) - length(x) * (length(x) + 1) / 2) / (length(x) * length(y))
}

ref <- audit$mean_INT_z[!audit$triaged & !audit$read_flagged]
classes <- list(
  "triaged (ALL - do not read this row)" = audit$triaged,
  "  IC only"        = audit$tri_IC & audit$n_axes == 1,
  "  PCA (any axis)" = audit$tri_PCA,
  "  burden (any axis)" = audit$tri_burden,
  "read-flagged HIGH" = audit$read_flagged & audit$High_Read_Plate
)
axis_attr <- bind_rows(lapply(names(classes), function(nm) {
  v <- audit$mean_INT_z[classes[[nm]]]
  tibble(class = nm, n = sum(classes[[nm]], na.rm = TRUE),
         mean_INT_z = round(mean(v, na.rm = TRUE), 3),
         cohens_d = round(cohens_d(v, ref), 3),
         AUC = round(auc(v, ref), 3))
}))
write_csv(axis_attr, "output_files/qc_axis_attribution.csv")
knitr::kable(axis_attr,
             caption = "Each triage class on the whole-panel intensity axis, against retained wells")

# Does QC reshape the intensity distribution, or only clip its tail? The moment
# skewness will move simply from deleting extreme points; a robust measure will
# not, unless the body of the distribution really changed.
bowley <- function(v) { q <- stats::quantile(v, c(.25,.5,.75), na.rm = TRUE)
  unname((q[3] + q[1] - 2*q[2]) / (q[3] - q[1])) }
m3 <- function(v) { v <- v[!is.na(v)]; mean((v - mean(v))^3) / stats::sd(v)^3 }
pre  <- audit$mean_INT_z
post <- audit$mean_INT_z[!audit$triaged]
cat(sprintf("Intensity distribution, pre-triage -> post-triage:\n  moment skew %+.3f -> %+.3f    Bowley skew %+.3f -> %+.3f    range [%.2f, %.2f] -> [%.2f, %.2f]\n",
            m3(pre), m3(post), bowley(pre), bowley(post),
            min(pre, na.rm = TRUE), max(pre, na.rm = TRUE),
            min(post, na.rm = TRUE), max(post, na.rm = TRUE)))
cat("  If Bowley barely moves while the moment skew does, QC clipped a tail and did not reshape the body.\n")
cat("  Either way the intensity axis survives QC and still belongs in the analysis model.\n")
```

#### What to do with the read-flagged wells

They are **kept**, and that is the right call — but "the flag is redundant once
you model depth" is a stronger claim than this pipeline can support, and it
should not be written into methods text as though it had been established.

Two things to carry downstream:

* **Model `log2(Reads)`, per marker.** NPQ is IC-normalised and total reads are
  close to independent of IC reads, so a pure sequencing-depth effect ought to
  divide out and leave slope 0. It does not — the depth slopes are strongly
  marker-specific, so one shared depth covariate will not absorb it identically
  across the panel.
* **"High-read" is a property of the well's neighbours.** The bound is
  plate-relative, so the same specimen run on a different plate need not be
  flagged. Treat the flag as a statement about a well, not about a subject.

### D.5.6 Well-level QC table

One row per well, carrying every gate flag, the intensity axis, plate position
and read/IC statistics. This is the join key for anything downstream that needs
to ask a question Primary QC cannot — above all **clustering by clinical
collection site**, which needs the metadata merge.

```{r well-level-table, message=FALSE, warning=FALSE, echo=TRUE}
well_qc <- audit %>%
  transmute(SampleID, Run, Bay, Well, wellRow, wellCol,
            triaged, Reason, n_axes, tri_low_reads, tri_IC, tri_PCA, tri_burden, tri_bad_data,
            read_flagged, High_Read_Plate, Low_Read,
            mean_INT, mean_INT_z, intensity_flag,
            Reads, IC_Reads, IC_ratio_to_plate_median = round(IC_Reads / Median_Value, 4),
            IC_Median, Detectability, QC_Status,
            PC_z_round1 = round(PC_z_round1, 3),
            PC_z_cohort = round(PC_z_cohort, 3), pca_cohort_flag, pca_masked_by_plate,
            burden_fdr_min = fdr_min,
            burden_n_out = n_out) %>%
  arrange(Run, Bay, Well)

write_csv(well_qc, "output_files/well_level_QC.csv")
cat(sprintf("Wrote output_files/well_level_QC.csv: %d wells x %d columns\n",
            nrow(well_qc), ncol(well_qc)))
knitr::kable(head(well_qc %>% select(SampleID, Run, Bay, Well, triaged, n_axes,
                                     read_flagged, mean_INT_z, Reads), 10),
             caption = "well_level_QC.csv (first 10 rows, subset of columns)")
```

# E. Biomarker QC
Reintegrate biomarkers and separate into groups.

## E.1 Re-integrate High-Corr Biomarkers 
Here we add back in high-correlation biomarkers that were removed for the QC Biomarker Set. These all should have good detectability and raw read counts.

```{r add_high_cor, message=FALSE, warning=FALSE, echo=TRUE}
# Add back high-correlated biomarkers to combined_qc_filtered
library(tibble)

# Rebuild the post-QC biomarker set EXPLICITLY as `qc_pool` (changed 2026-09-01).
#
# This used to cbind `high_cor` onto whatever columns combined_qc_filtered happened to carry, which
# was correct only while the anchors were guaranteed to be a subset of qc_pool. Since the WATCH-tier
# fix an anchor may sit BELOW the 98% pool bar (kept for triage, flagged), and cbind would then
# promote it out of the Low file and into the standard post-QC set — silently changing biomarker
# CONTENT, which the frozen/rederive design guarantees is mode-invariant.
#
# Stating the content set directly removes that coupling: which markers reach the post-QC datasets is
# `qc_pool` and nothing else, no matter what anchored the triage. Identical to the old result whenever
# anchors are a subset of qc_pool, which is every pre-2026-09-01 run.
combined_qc_full <- combined_df[rownames(combined_qc_filtered),
                                c("Run", "Bay", qc_pool), drop = FALSE]
stopifnot(setequal(setdiff(colnames(combined_qc_full), c("Run", "Bay")), qc_pool))
# Move SampleIDs to first column
combined_qc_full <- rownames_to_column(combined_qc_full, var = "SampleID")

knitr::kable(
  high_cor, 
  caption = "High-correlation biomarkers added back:"
)
dim(combined_qc_full)[2]-3 #NUMBER OF BIOMARKERS IS COLUMNS-3
```

## E.2 Low-detectability Biomarkers

### E.2.1 Re-integrate Det>50% Biomarkers 
Alamar suggests that the detectability threshold should be 50%. Here we add back in Biomarkers with minimum detectability>50% that were dropped to create the QC Biomarker set. We also require these to have raw read counts above the threshold. Those with low counts are further examined in later QC steps.

```{r add_med_det, message=FALSE, warning=FALSE, echo=TRUE}
# Add back biomarkers with >50% Min_Detectability ("medium detectability")
med_det <- low_det[low_det$Min_Detectability > min_detectability, ]

# Split into "good" and "still low count"
good_detect <- med_det[!(med_det$Biomarker %in% low_counts$Biomarker), ]
med_det_low_counts <- med_det[(med_det$Biomarker %in% low_counts$Biomarker), ]

# Extract good biomarker names
good_detect_names <- good_detect$Biomarker

# Subset combined_df to keep only good samples and these biomarkers
med_det_subset <- combined_df[rownames(combined_qc_filtered), good_detect_names, drop = FALSE]

# Add these columns back to the filtered QC dataset
combined_qc_full <- cbind(combined_qc_full, med_det_subset)
rownames(combined_qc_full) <- NULL

# Confirm dimensions
dim(combined_qc_full)[2]-3 #NUMBER OF BIOMARKERS IS COLUMNS-3

# Detectability metadata tables for each group
med_detect_table <- detect[detect$Biomarker %in% good_detect_names, ]
med_det_low_counts_table <- detect[detect$Biomarker %in% med_det_low_counts$Biomarker, ]

# View tables
knitr::kable(
  med_detect_table, 
  caption = "Medium-detectability (min det>50%) biomarkers added back:"
)
knitr::kable(
  med_det_low_counts_table, 
  caption = "Medium-detectability biomarkers with low counts to be examined below:"
)

```

### E.2.2 Re-examine Det<50% Biomarkers 
Look at these for possible re-integration, e.g., those with some expected low detectability.

```{r check_low_det, message=FALSE, warning=FALSE, echo=TRUE}
# Check biomarkers with <50% Min_Detectability to see if any should be added back in
library(kableExtra)

low_low_det <- low_det[low_det$Min_Detectability <= min_detectability, ]

# Split into "low detctability" and "low detectabiloty with low count"
low_detect <- low_low_det[!(low_low_det$Biomarker %in% low_counts$Biomarker), ]
low_det_low_counts <- low_low_det[(low_low_det$Biomarker %in% low_counts$Biomarker), ]

# Extract good biomarker names
low_detect_names <- low_detect$Biomarker

# Detectability metadata tables for each group
low_detect_table <- detect[detect$Biomarker %in% low_detect_names, ]
low_det_low_counts_table <- detect[detect$Biomarker %in% low_det_low_counts$Biomarker, ]

# View tables
knitr::kable(
  low_detect_table,
  caption = "Low-detectability (min det ≤50%) biomarkers with adequate read counts (routed to the Low output file):"
)
knitr::kable(
  low_det_low_counts_table, 
  caption = "Low-detectability biomarkers with low counts to be examined below:"
)

# Look for bimodal distributions and extreme outliers. Note differences across plates. Also LOD for these. 

#Pull LODs
# Read annotation data
annot <- read_csv("input_files/Annotation_Targets.csv")

# Select required columns
annot_sel <- annot %>%
  select(targetName, targetDetectability, targetLOD, targetLOD_NPQ, plateID)

 # Extract Run and Bay from plateID (assuming "YYYYMMDD_BayX" or similar format)
#CHANGED THIS FOR ALZ2 TO Allow for Runs with same prefix but different plates
annot_sel <- annot_sel %>%
  mutate(Run = str_extract(plateID, "\\d{8}-\\d+"),
         Bay = str_extract(plateID, "Bay\\d+"),
         RunBay = paste0(Run, "_", Bay))

# Make wide format table: LOD per Run/Bay
lod_matrix <- annot_sel %>%
  select(targetName, RunBay, targetLOD_NPQ) %>%
  distinct() %>%   # remove any duplicates
  pivot_wider(names_from = RunBay,
              values_from = targetLOD_NPQ)

# LOD for Reads
lod_reads_matrix <- annot_sel %>%
  select(targetName, RunBay, targetLOD) %>%
  distinct() %>%   # remove any duplicates
  pivot_wider(names_from = RunBay,
              values_from = targetLOD)

# Extract non-targetName columns
cols <- colnames(lod_matrix)[-1]

# Extract run date + bay index as ordering helpers
run_date <- str_extract(cols, "\\d{8}")
bay_num <- str_extract(cols, "(?<=Bay)\\d+") |> as.numeric()

# Create order index based on date then Bay number
col_order <- order(run_date, bay_num)

# Apply reordering: keep targetName first
lod_matrix <- lod_matrix %>%
  select(targetName, all_of(cols[col_order]))

# Reorder lod_reads_matrix columns to exactly match lod_matrix
lod_reads_matrix <- lod_reads_matrix %>%
  select(all_of(colnames(lod_matrix)))

write.csv(lod_matrix,"output_files/lod_matrix.csv", row.names = TRUE)
write.csv(lod_reads_matrix,"output_files/lod_reads_matrix.csv", row.names = TRUE)

cat("Plots of Low-detectability (min det<50%) biomarkers with good read counts:\n")
# Keep only rows in combined_qc_filtered and select relevant columns
combined_low_det <- combined_df %>%
  filter(rownames(.) %in% rownames(combined_qc_filtered)) %>%
  rownames_to_column("SampleID") %>%
  select(
    SampleID,
    Run,
    Bay,
    all_of(low_detect_names)
  )

# Ensure Run and Bay are factors
combined_low_det$Run <- factor(combined_low_det$Run)
combined_low_det$Bay <- factor(combined_low_det$Bay)

# Custom Bay colors
bay_colors <- c("Bay1" = "red", "Bay2" = "green", "Bay3" = "blue")

# Identify only biomarker columns
biomarker_cols <- setdiff(names(combined_low_det), c("SampleID", "Run", "Bay"))

# Loop and create violin plots for each biomarker
for (biom in biomarker_cols) {
  
  df_temp <- combined_low_det %>%
    select(Value = all_of(biom), Run, Bay)
  
  p <- ggplot(df_temp, aes(x = Run, y = Value, fill = Bay)) +
    geom_violin(alpha = 0.5, trim = FALSE, position = position_dodge(width = 0.8)) +
    geom_jitter(aes(color = Bay), 
                position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
                alpha = 0.75, size = 1.3) +
    scale_fill_manual(values = bay_colors) +
    scale_color_manual(values = bay_colors) +
    theme_minimal() +
    labs(
      title = paste("Run × Bay Effects:", biom),
      x = "Run",
      y = biom
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p)

 # ----- Print per-biomarker LOD table -----
lod_npq_vals <- lod_matrix[lod_matrix$targetName == biom, -1, drop = FALSE]
lod_read_vals <- lod_reads_matrix[lod_reads_matrix$targetName == biom, -1, drop = FALSE]

lod_table <- rbind(
  "NPQ LOD" = lod_npq_vals,
  "Read LOD" = lod_read_vals
)

# Print LOD table
print(cbind(rownames(lod_table),lod_table)) # I tried to make this prettier but it doesn't work in a loop
}

```


## E.3 Low Raw-Count Biomarkers  
Look at these more closely in secondary QC. Many are expected, because the protein is rare or they are due to a polymorphism, e.g., APOE4, BD-MAPT.

I'm showing NPQ values here for consistency, not raw read counts. Read counts will be examined further in secondary QC.
```{r check_low_raw_counts, message=FALSE, warning=FALSE, echo=TRUE}
# Examine low-count biomarkers 

# Vector of low-count biomarker names
low_counts_vec <- low_counts[[1]]  # assuming low_counts is a data frame with one column

# Keep only rows in combined_qc_filtered and select relevant columns
combined_low_counts <- combined_df %>%
  filter(rownames(.) %in% rownames(combined_qc_filtered)) %>%
  rownames_to_column("SampleID") %>%
  select(
    SampleID,
    Run,
    Bay,
    all_of(low_counts_vec)
  )

# Ensure Run and Bay are factors
combined_low_counts$Run <- factor(combined_low_counts$Run)
combined_low_counts$Bay <- factor(combined_low_counts$Bay)

# Custom Bay colors
bay_colors <- c("Bay1" = "red", "Bay2" = "green", "Bay3" = "blue")

# Identify only biomarker columns
biomarker_cols <- setdiff(names(combined_low_counts), c("SampleID", "Run", "Bay"))

# Loop and create violin plots for each biomarker
for (biom in biomarker_cols) {
  
  df_temp <- combined_low_counts %>%
    select(Value = all_of(biom), Run, Bay)
  
  p <- ggplot(df_temp, aes(x = Run, y = Value, fill = Bay)) +
     geom_violin(alpha = 0.5, trim = FALSE, position = position_dodge(width = 0.8)) +
     geom_jitter(aes(color = Bay), 
                position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
                alpha = 0.75, size = 1.5) +
    scale_fill_manual(values = bay_colors) +
    scale_color_manual(values = bay_colors) +
    theme_minimal() +
    labs(
      title = paste("Run × Bay Effects:", biom),
      x = "Run",
      y = biom
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p)
  # Print the row from combined_low_counts corresponding to this biomarker
  print(knitr::kable(low_counts[low_counts$Biomarker == biom, , drop = FALSE],caption = "Reads"))
}

```

# F. Prepare Output  
Output the QC'd dataset with samples from "Sample QC" removed. All biomarkers are included, but separated into two files: 1) NPQxxx_post_QC.csv and 2) NPQxxx_Low_post_QC.csv. 

1) Contains all biomarkers with minimum detectability >50% across plates for QC'd samples that have mean and median raw-read counts>500 across all plates.

2) Contains all biomarkers with minimum detectability <= 50% across plates and/or mean or median raw-read counts <= 500 across all plates or QC'd samples.

3) Contains all biomarkers in 1 for triaged samples.

4) Contains all biomarkers in 2 for triaged samples. 

Low detectability and low count biomarkers are in a separate dataset for "analysis". They are not necessarily bad, but will need to be analyzed differently from continuous biomarkers.

```{r make_output, message=FALSE, warning=FALSE, echo=TRUE}
# Here we save the QC'd files
# We may want to transpose them to be like original NPQ
library(dplyr)
library(tibble)

knitr::kable(
  summarize_counts(combined_qc_full, "QC'd Samples"), 
  caption = "POST-QC SAMPLES"
)

#Output file 1

cat("QC'd Biomarkers:", dim(combined_qc_full)[2]-3, "\n")
qc_file <- file.path("output_files", sub("\\.csv$", "_post_QC.csv", NPQ_data))
write.csv(combined_qc_full, qc_file, row.names = FALSE)  # SampleID is already a column; avoid a stray index col (matches files 2-4)
cat("Saved:", qc_file, "\n")

#Output file 2

# Remove duplicated metadata columns from the second dataset
combined_low_counts_nometadata <- combined_low_counts %>%
  select(-Run, -Bay)
# Perform the join on SampleID
combined_qc_low <- combined_low_det %>%
  left_join(combined_low_counts_nometadata, by = "SampleID")

cat("QC'd Low-count/Low-detectability Biomarkers:", dim(combined_qc_low)[2]-3, "\n")
qc_file <- file.path("output_files", sub("\\.csv$", "_Low_post_QC.csv", NPQ_data))
write.csv(combined_qc_low, qc_file, row.names = FALSE)
cat("Saved:", qc_file, "\n")

# Output file 3: TRIAGE × Full QC biomarkers

# sample list
triage_samples <- triage_df$SampleID
combined_df <- rownames_to_column(combined_df, "SampleID")

triage_qc_full <- combined_df %>%
  filter(SampleID %in% triage_samples) %>%
  select(any_of(colnames(combined_qc_full)))

triage_full_file <- file.path(
  "output_files", 
  sub("\\.csv$", "_post_QC_triage.csv", NPQ_data)
)

write.csv(triage_qc_full, triage_full_file, row.names = FALSE)
cat("Saved TRIAGE full-QC biomarker file:", triage_full_file, "\n")

# Output file 4: TRIAGE × Low QC biomarkers

triage_qc_low <- combined_df %>%
  filter(SampleID %in% triage_samples)%>%
  select(any_of(colnames(combined_qc_low)))

triage_low_file <- file.path(
  "output_files", 
  sub("\\.csv$", "_Low_post_QC_triage.csv", NPQ_data)
)

write.csv(triage_qc_low, triage_low_file, row.names = FALSE)
cat("Saved TRIAGE low-QC biomarker file:", triage_low_file, "\n")

```

# G.  Notes

## G.1 What the 2026-07-30 audit revision changed, and what it deliberately did not

Implements `Secondary_QC/Triage_Review/README.md`. **No sample's fate changed.**
Triage and flag membership are identical to the previous version; what changed is
what the pipeline measures and says about itself.

**Added**

* **D.0 `mean_INT`** — whole-panel intensity, computed pre-triage over all wells
  so removed wells are on the same scale as retained ones. Reported, never gated.
* **D.5** — the screen audit: gate inventory with INERT detection and headroom,
  gate overlap, cut neighbourhoods, plate-position permutation tests, the
  intensity axis read against each triage class, and `well_level_QC.csv`.
* **Per-axis triage booleans** (`tri_IC`, `tri_PCA`, `tri_burden`,
  `tri_bad_data`, `n_axes`) on `samples_to_triage.csv`.
* **`PC_z`** on both PCA coordinate files, so the operating point is auditable.

**Corrected in the reporting, not in the arithmetic**

* The **read gate is one-sided in practice.** Its lower bound is not
  plate-relative on any real run and cannot be made so — see D.2.4, where four
  alternative fences are shown to fail. The gate is now described and reported as
  a plate-relative *high* flag plus an absolute floor. The `pmax()` is retained,
  so behaviour is unchanged and the plate-relative bound will bind if a run ever
  supports it.
* **Our IC gate is Alamar's `IC Median` recomputed**, not an independent
  in-house check (Jaccard 0.966 on the 50-plate run). Methods text should not
  present it as a second opinion. It supplies roughly three quarters of triage.
* **`FDR_threshold` is not a dial.** It is an integer count of anchors wearing a
  p-value. D.2.2 prints the complete list of settings that exist.
* **`exit_reason`-style one-reason-per-well reports are not axis totals.**
  Precedence gives a valid partition, but it undercut the PCA axis by 37.5% and
  the burden axis by 64% on this run. Use the per-axis booleans for "how many
  wells did this screen catch".

**Deliberately not done**

* **`mean_INT` was not made a gate.** It is the statistic that would catch the
  uniformly-bright samples the burden test structurally cannot — which is exactly
  why it is tempting. But background intensity is a continuous nuisance covariate
  affecting every sample, not a population of bad wells. Gating on it would
  delete real variation. Model it instead.
* **The PCA cut was not loosened** to reach bright wells. On this run the
  brightest wells sit *below* the PCA cut on `PC_z`, so lowering it admits dim
  wells before it admits a single bright one. The instrument is not turned down;
  it is pointed elsewhere.
* **PCA round 2 was not deleted** although it fired on nothing here. It is now
  reported as INERT with its headroom, which is a statement about this run, not
  about the screen.

## G.2 Questions this pipeline cannot answer

**Clustering by clinical collection site.** Primary QC is agnostic to phenotype
and covariates by design, and the site label only exists after the metadata
merge. D.5.4 tests the *physical position on the plate*, which is a different
question that happens to share the word "site" — and which is what surfaced the
worst-position and corner-gradient findings. For collection site, join
`well_level_QC.csv` to the merged metadata; `Secondary_QC/Triage_Review/` already
does this.

**Whether the read flag is redundant given depth.** See D.5.5 — it has been shown
to be *small*, which is not the same as shown to be *nothing*.

Combined Data: