---
title: "Batch Effects Analysis for Alamar Biomarker Data"
author: "ERM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

# A. Overview

This module assesses technical batch effects (Run and Bay) in the merged biomarker dataset.
The hierarchy is: **Run → Bay (1/2/3, sequential within Run) → Sample**.

Goals:
1. Characterize plate composition (race, sex, age, site, diagnosis) by Run and Bay
2. Identify confounding between biological covariates and batch variables
3. Partition variance into Run, Bay, and biological sources per protein
4. Quantify cross-plate reproducibility from HIHG replicates
5. Make a data-driven decision about batch correction approach

Biomarkers are in log2 NPQ scale throughout (no back-transformation needed for variance
partitioning or linear models).

**For a new dataset: update the input filename in Section A.3.**

---

## A.1 Input Data

Default input: `merged_combined_post_QC.csv` from `Metadata_Merge/output_files_*/`.
This file has separate `Run` and `Bay` columns and includes both standard and low-abundance biomarkers.

Fallback: `merged_standard_post_QC.csv` from the same directory (standard biomarkers only, no separate Run/Bay — parsed from `RUN` column).

Place in `input_files/` or let the pipeline auto-copy.

## A.2 Parameters

```{r params}
# --- Adjustable parameters ---
VP_CORES       <- 4      # cores for variancePartition (increase if available)
VP_MIN_SAMPLES <- 10     # min samples per Run to include in VP
ICC_MIN_REPS   <- 2      # min replicate pairs required for ICC calculation
FDR_THRESHOLD  <- 0.05   # FDR cutoff for declaring significant batch effect
```

## A.3 Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(variancePartition)
library(BiocParallel)
library(knitr)
library(kableExtra)
library(ggplot2)
library(patchwork)

# ---- Directories ----
if (!dir.exists("input_files"))  dir.create("input_files")
if (!dir.exists("output_files")) dir.create("output_files")

# ---- Auto-copy from Metadata_Merge if needed ----
# Base = merged_combined_post_QC.csv: ALL post-QC samples incl. cross-run replicates (needed for the
# ICC reproducibility analysis — the cross-sectional filtered file drops them). The DERIVED covariates
# Ancestry + CDX_collapsed (Step 6d) live in filtered_combined and are joined on below.
input_candidates <- list.files("input_files",
                               pattern = "(merged_combined|merged_standard)_post_QC\\.csv$",
                               full.names = TRUE)

if (length(input_candidates) == 0) {
  search_dirs <- list.files(
    "../../Metadata_Merge",
    pattern = "^output_files",
    full.names = TRUE,
    include.dirs = TRUE
  )
  search_dirs <- search_dirs[order(file.mtime(search_dirs), decreasing = TRUE)]

  for (d in search_dirs) {
    src <- file.path(d, "merged_combined_post_QC.csv")
    if (file.exists(src)) { file.copy(src, "input_files/"); message("Copied merged_combined from: ", src); break }
    src <- file.path(d, "merged_standard_post_QC.csv")
    if (file.exists(src)) { file.copy(src, "input_files/"); message("Copied merged_standard from: ", src); break }
  }
  input_candidates <- list.files("input_files",
                                 pattern = "(merged_combined|merged_standard)_post_QC\\.csv$",
                                 full.names = TRUE)
}

if (length(input_candidates) == 0) {
  stop("No input file found. Place merged_combined_post_QC.csv in input_files/ or ensure Metadata_Merge outputs exist.")
}

# Prefer merged_combined if both present
if (any(grepl("merged_combined", input_candidates))) {
  input_file <- input_candidates[grepl("merged_combined", input_candidates)][1]
} else {
  input_file <- input_candidates[1]
}
message("Using: ", basename(input_file))
```

---

# B. Data Loading and Formatting

```{r load-data, message=FALSE, warning=FALSE}
raw <- read_csv(input_file, show_col_types = FALSE)

# ---- Get Run_ID and Bay ----
# merged_combined has separate Run + Bay columns -> rename Run -> Run_ID.
# merged_standard only has the combined RUN column -> parse it.
# Also handles any future file that already uses Run_ID directly.
if ("Run_ID" %in% names(raw) && "Bay" %in% names(raw)) {
  df <- raw
  message("Using existing Run_ID and Bay columns.")
} else if ("Run" %in% names(raw) && "Bay" %in% names(raw)) {
  df <- raw %>% rename(Run_ID = Run)
  message("Using existing Run and Bay columns (renamed Run -> Run_ID).")
} else {
  # Parse from combined RUN column (format: "20250711-1613_Bay1")
  df <- raw %>%
    mutate(
      Run_ID = str_extract(RUN, "^[^_]+"),     # e.g. "20250711-1613"
      Bay    = str_extract(RUN, "Bay[0-9]+")   # e.g. "Bay1"
    )
  message("Parsed Run_ID and Bay from RUN column.")
}

# ---- Derived covariates: Ancestry + CDX_collapsed (Metadata_Merge Step 6d) ----
# merged_combined carries the replicates the ICC needs but only raw Race/CDX; pull the DERIVED
# covariates from filtered_combined (one row per kept sample) and join by SAMPLE_ALIQUOT. Fall back
# to raw Race / coalesce(CDX, Case_Control) if filtered_combined is unavailable.
if (!all(c("Ancestry", "CDX_collapsed") %in% names(df))) {
  .fc <- list.files("input_files", pattern = "filtered_combined_post_QC\\.csv$", full.names = TRUE)
  if (!length(.fc)) for (d in list.files("../../Metadata_Merge", pattern = "^output_files", full.names = TRUE)) {
    .p <- file.path(d, "filtered", "filtered_combined_post_QC.csv")
    if (file.exists(.p)) { .fc <- .p; break }
  }
  if (length(.fc) && file.exists(.fc[1])) {
    .der <- suppressMessages(readr::read_csv(.fc[1], show_col_types = FALSE)) %>%
      select(SAMPLE_ALIQUOT, any_of(c("Ancestry", "CDX_collapsed"))) %>% distinct()
    df <- df %>% left_join(.der, by = "SAMPLE_ALIQUOT")
    message("Joined derived covariates (Ancestry + CDX_collapsed) from filtered_combined.")
  }
}
if (!"Ancestry" %in% names(df))      df$Ancestry      <- df$Race
if (!"CDX_collapsed" %in% names(df)) df$CDX_collapsed <- dplyr::coalesce(df$CDX, df$Case_Control)

# ---- Submission year (coarse batch level above Run) ----
# Each Run belongs to one submission year (2025 ALZ123 vs 2026 8+7-plate); year is the coarsest batch axis.
df$year <- substr(as.character(df$Run_ID), 1, 4)

# ---- Identify biomarker columns ----
# All numeric columns after the metadata block
meta_cols <- c("RUN", "SAMPLE_ALIQUOT", "SAMPLE", "Record_ID", "newsample",
               "Group", "Ethnicity", "Race", "sex", "age_at_subject",
               "CDX", "Case_Control", "AOO", "Years_Onset", "APOE.geno",
               "height_inches", "weight_lb", "BMI", "Site",
               "Country/State", "country_of_birth", "metadata_source",
               "newsamp", "Group_Race_Ethnicity", "Run_ID", "Bay", "Run",
               # derived covariates / flags from Metadata_Merge (filtered_combined) — not biomarkers
               "Ancestry", "CDX_collapsed", "APOE_WGS", "APOE_WGS_norm", "APOE_WGS_changed",
               "APOE.geno_final", "APOE4_carrier", "APOE4_count", "gp_APOE",
               "n_visits", "is_repeat_subject", "subject_visit_n", "is_most_recent",
               "ID2", "AlamarWarning", "PrimaryQC", "ResultQC",
               "___1", "APOE")

biomarker_cols <- names(df)[!(names(df) %in% meta_cols) & sapply(df, is.numeric)]

# Exclude mCherry if present (internal control, not a biomarker)
biomarker_cols <- biomarker_cols[biomarker_cols != "mCherry"]

# ---- Study samples only (exclude controls/reps by filtering on SAMPLE_ALIQUOT) ----
# Controls have prefixes NC, SC, IPC; HIHG reps are identified separately below
df_samples <- df %>%
  filter(!str_detect(SAMPLE_ALIQUOT, "^(NC|SC|IPC)"))

# ---- Identify HIHG replicates (SAMPLE appears >1 time) ----
rep_counts <- df_samples %>%
  count(SAMPLE, name = "n_runs") %>%
  filter(n_runs > 1)

df_reps    <- df_samples %>% filter(SAMPLE %in% rep_counts$SAMPLE)
df_unique  <- df_samples %>% filter(!SAMPLE %in% rep_counts$SAMPLE)

cat("Total samples (excl controls):", nrow(df_samples), "\n")
cat("Unique biological samples:    ", nrow(df_unique), "\n")
cat("Replicated biological samples:", n_distinct(df_reps$SAMPLE), "\n")
cat("Replicate observations:       ", nrow(df_reps), "\n")
cat("Runs:                         ", n_distinct(df_samples$Run_ID), "\n")
cat("Bays observed:                ", paste(sort(unique(df_samples$Bay)), collapse=", "), "\n")
cat("Biomarkers:                   ", length(biomarker_cols), "\n")
```

---

# C. Plate Composition

This section examines whether demographic and clinical variables are balanced across
Runs and Bays. Confounding between batch and biology is the primary concern.

## C.1 Sample Counts per Run x Bay

```{r composition-counts, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
count_tbl <- df_samples %>%
  count(Run_ID, Bay) %>%
  pivot_wider(names_from = Bay, values_from = n, values_fill = 0) %>%
  arrange(Run_ID)

kable(count_tbl, caption = "Sample counts per Run x Bay (study samples only)") %>%
  kable_styling(full_width = FALSE)
```

## C.2 Ancestry Distribution per Run

```{r composition-ancestry, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
anc_tbl <- df_samples %>%
  count(Run_ID, Ancestry) %>%
  group_by(Run_ID) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

ggplot(anc_tbl, aes(x = Run_ID, y = pct, fill = Ancestry)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  labs(title = "Ancestry composition per Run",
       x = "Run", y = "% of samples", fill = "Ancestry")
```

## C.3 Ancestry Distribution per Bay (across all Runs)

```{r composition-ancestry-bay, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
bay_anc <- df_samples %>%
  count(Bay, Ancestry) %>%
  group_by(Bay) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

ggplot(bay_anc, aes(x = Bay, y = pct, fill = Ancestry)) +
  geom_col() +
  theme_bw() +
  labs(title = "Ancestry composition per Bay (pooled across Runs)",
       x = "Bay", y = "% of samples", fill = "Ancestry")
```

## C.4 Sex and Age Distribution per Run

```{r composition-sex-age, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
p_sex <- df_samples %>%
  count(Run_ID, sex) %>%
  group_by(Run_ID) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ggplot(aes(x = Run_ID, y = pct, fill = sex)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  labs(title = "Sex composition per Run", x = NULL, y = "% samples", fill = "Sex")

p_age <- df_samples %>%
  ggplot(aes(x = Run_ID, y = age_at_subject)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.size = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  labs(title = "Age distribution per Run", x = "Run", y = "Age")

p_sex / p_age
```

## C.5 Site Distribution per Run

```{r composition-site, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
site_tbl <- df_samples %>%
  count(Run_ID, Site) %>%
  group_by(Run_ID) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

ggplot(site_tbl, aes(x = Run_ID, y = pct, fill = Site)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.text = element_text(size = 7)) +
  labs(title = "Site composition per Run",
       x = "Run", y = "% of samples", fill = "Site")
```

## C.6 Diagnosis (CDX_collapsed) Distribution per Run

```{r composition-dx, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
dx_tbl <- df_samples %>%
  count(Run_ID, CDX_collapsed) %>%
  group_by(Run_ID) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

ggplot(dx_tbl, aes(x = Run_ID, y = pct, fill = CDX_collapsed)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  labs(title = "Diagnosis (CDX_collapsed) composition per Run",
       x = "Run", y = "% of samples", fill = "Diagnosis")
```

## C.7 Confounding: Correlation of Batch with Covariates

Cramer's V for categorical variables, sqrt(eta-squared) for age. Values > 0.3 indicate
meaningful confounding between batch variable and covariate.

```{r confounding, message=FALSE, warning=FALSE}
cramers_v <- function(x, y) {
  tbl <- table(x, y)
  chi <- suppressWarnings(chisq.test(tbl))
  n   <- sum(tbl)
  k   <- min(nrow(tbl), ncol(tbl))
  v   <- sqrt(chi$statistic / (n * (k - 1)))
  as.numeric(v)
}

eta_sq <- function(group, value) {
  fit <- aov(value ~ group)
  ss  <- summary(fit)[[1]][["Sum Sq"]]
  ss[1] / sum(ss)
}

cov_list <- list(
  Ancestry = df_samples$Ancestry,
  sex      = df_samples$sex,
  Site     = df_samples$Site,
  CDX      = df_samples$CDX_collapsed
)

batch_vars <- list(
  Run_ID = df_samples$Run_ID,
  Bay    = df_samples$Bay,
  year   = df_samples$year
)

conf_results <- map_dfr(names(batch_vars), function(bv) {
  bvec <- batch_vars[[bv]]
  rows <- map_dfr(names(cov_list), function(cv) {
    cvec <- cov_list[[cv]]
    ok   <- !is.na(bvec) & !is.na(cvec)
    tibble(Batch_Var = bv, Covariate = cv,
           CramersV  = cramers_v(bvec[ok], cvec[ok]))
  })
  age_ok <- !is.na(bvec) & !is.na(df_samples$age_at_subject)
  bind_rows(rows, tibble(
    Batch_Var = bv, Covariate = "age",
    CramersV  = sqrt(eta_sq(bvec[age_ok], df_samples$age_at_subject[age_ok]))
  ))
}) %>%
  mutate(Concern = case_when(
    CramersV > 0.5 ~ "HIGH",
    CramersV > 0.3 ~ "moderate",
    TRUE           ~ "low"
  ))

kable(conf_results %>% arrange(desc(CramersV)), digits = 3,
      caption = "Confounding between batch variables and biological covariates") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(which(conf_results %>% arrange(desc(CramersV)) %>% pull(Concern) == "HIGH"),
           background = "#ffd6d6") %>%
  row_spec(which(conf_results %>% arrange(desc(CramersV)) %>% pull(Concern) == "moderate"),
           background = "#fff3cd")

write_csv(conf_results, "output_files/batch_covariate_confounding.csv")
```

---

# D. Variance Partitioning

Decomposes variance per protein into:
- Random effects (categorical): Ancestry, sex, Site, Diagnosis (CDX_collapsed)
- Random effects (batch): Run (between-run technical variation), Bay nested in Run (sequential/drift)
- Fixed effect: age_at_subject (continuous)
- Residual: unexplained

Model: `~ age_at_subject + (1|Ancestry) + (1|sex) + (1|Site) + (1|CDX_collapsed) + (1|Run_ID) + (1|Run_ID:Bay)`

Proteins with high Run or Bay variance fractions are candidates for batch correction.
Proteins where Run variance ~ Ancestry variance indicate confounding that correction cannot resolve.

```{r vp-prep, message=FALSE, warning=FALSE}
# Use unique samples only (no replicates) to avoid inflating residual
vp_df <- df_unique %>%
  filter(!is.na(Ancestry), !is.na(sex), !is.na(age_at_subject),
         !is.na(CDX_collapsed), !is.na(Run_ID), !is.na(Bay))

# Remove Runs with fewer than VP_MIN_SAMPLES
run_counts <- count(vp_df, Run_ID)
vp_df <- vp_df %>%
  filter(Run_ID %in% run_counts$Run_ID[run_counts$n >= VP_MIN_SAMPLES])

cat("Samples used for VP:", nrow(vp_df), "\n")
cat("Runs included:      ", n_distinct(vp_df$Run_ID), "\n")

# Biomarker matrix: proteins x samples (variancePartition convention)
expr_mat <- vp_df %>%
  select(all_of(biomarker_cols)) %>%
  as.matrix() %>%
  t()

info_df <- vp_df %>%
  select(Run_ID, Bay, Ancestry, sex, age_at_subject, Site, CDX_collapsed) %>%
  mutate(across(c(Ancestry, sex, Site, CDX_collapsed, Run_ID, Bay), as.factor)) %>%
  as.data.frame()
```

```{r vp-run, message=FALSE, warning=FALSE, fig.width=12, fig.height=6}
register(MulticoreParam(VP_CORES))

form <- ~ age_at_subject + (1|Ancestry) + (1|sex) + (1|Site) + (1|CDX_collapsed) + (1|Run_ID) + (1|Run_ID:Bay)

vp_fit <- fitExtractVarPartModel(expr_mat, form, info_df)

plotVarPart(vp_fit, main = "Variance partitioning: biological vs. batch effects")
```

```{r vp-summary, message=FALSE, warning=FALSE}
vp_df_out <- as.data.frame(vp_fit) %>%
  rownames_to_column("Biomarker") %>%
  arrange(desc(`Run_ID`))

# Threshold summary
thresh_tbl <- map_dfr(c(5, 10, 20), function(t) {
  tibble(
    Threshold    = paste0(">", t, "% Run variance"),
    N_proteins   = sum(vp_df_out$`Run_ID` * 100 > t),
    Pct_proteins = round(100 * mean(vp_df_out$`Run_ID` * 100 > t), 1)
  )
})
kable(thresh_tbl, caption = "Proteins with substantial Run-level variance") %>%
  kable_styling(full_width = FALSE)

kable(vp_df_out %>% slice_max(`Run_ID`, n = 20) %>%
        mutate(across(where(is.numeric), ~round(.x * 100, 1))),
      caption = "Top 20 proteins by % Run variance") %>%
  kable_styling(full_width = FALSE) %>% scroll_box(width = "100%")

kable(vp_df_out %>% arrange(desc(`Run_ID:Bay`)) %>% slice_head(n = 20) %>%
        mutate(across(where(is.numeric), ~round(.x * 100, 1))),
      caption = "Top 20 proteins by % Bay (within-Run) variance") %>%
  kable_styling(full_width = FALSE) %>% scroll_box(width = "100%")

write_csv(vp_df_out, "output_files/variance_partition_results.csv")
```

---

# E. Replicate-Based Reproducibility

Two independent replicate structures estimate technical reproducibility.

**E.1 — Biological duplicates** (study samples re-run on another plate). Per protein we compute an ICC
across the repeated measurements, pooled over duplicated samples (between-sample biological spread is the
"signal", technical noise the "error"):

- **Cross-Run ICC** — agreement across *different Runs*; total run-to-run technical variance. *Estimable.*
- **Within-Run ICC** — agreement across *Bays of the same Run*. Requires a sample loaded **≥2× within one
  Run**; by design the duplicates are spread *across* Runs, so **0 qualify** and this is reported as
  "Not estimable (no within-Run duplicates)". The within-plate / within-Run level is recovered instead
  from the **control pools in E.2**.

ICC > 0.9 = excellent; 0.75-0.9 = good; < 0.75 = concerning.

```{r icc-prep, message=FALSE, warning=FALSE}
library(irr)

rep_obs <- df_reps %>%
  select(SAMPLE, SAMPLE_ALIQUOT, Run_ID, Bay, all_of(biomarker_cols)) %>%
  arrange(SAMPLE, Run_ID, Bay) %>%
  group_by(SAMPLE) %>%
  mutate(n_runs_for_sample = n_distinct(Run_ID),
         any_cross_run     = n_runs_for_sample > 1) %>%
  ungroup()

within_run_reps <- rep_obs %>%
  group_by(SAMPLE, Run_ID) %>% filter(n() >= 2) %>% ungroup()

cross_run_reps <- rep_obs %>% filter(any_cross_run)

cat("Replicated samples:            ", n_distinct(rep_obs$SAMPLE), "\n")
cat("  with within-Run replicates:  ", n_distinct(within_run_reps$SAMPLE), "\n")
cat("  with cross-Run replicates:   ", n_distinct(cross_run_reps$SAMPLE), "\n")
if (n_distinct(within_run_reps$SAMPLE) == 0)
  cat("  NOTE: no biological sample is repeated within a single Run -> Within-Run ICC is not estimable\n",
      "       from duplicates (it is recovered from the control pools in E.2).\n", sep = "")
```

```{r icc-calc, message=FALSE, warning=FALSE}
calc_icc_per_protein <- function(rep_df, biomarkers, min_pairs = ICC_MIN_REPS) {
  map_dfr(biomarkers, function(bm) {
    bm_wide <- rep_df %>%
      select(SAMPLE, Run_ID, Bay, all_of(bm)) %>%
      arrange(SAMPLE, Run_ID, Bay) %>%
      group_by(SAMPLE) %>%
      mutate(obs_num = row_number()) %>%
      ungroup() %>%
      pivot_wider(id_cols = SAMPLE, names_from = obs_num,
                  values_from = all_of(bm), names_prefix = "rep_") %>%
      select(-SAMPLE) %>%
      filter(rowSums(!is.na(.)) >= min_pairs)

    if (nrow(bm_wide) < min_pairs)
      return(tibble(Biomarker = bm, ICC = NA_real_, lower = NA_real_,
                    upper = NA_real_, n_pairs = nrow(bm_wide)))

    result <- tryCatch(
      icc(bm_wide, model = "twoway", type = "agreement", unit = "average"),
      error = function(e) NULL
    )
    if (is.null(result))
      return(tibble(Biomarker = bm, ICC = NA_real_, lower = NA_real_,
                    upper = NA_real_, n_pairs = nrow(bm_wide)))

    tibble(Biomarker = bm, ICC = result$value,
           lower = result$lbound, upper = result$ubound, n_pairs = nrow(bm_wide))
  })
}

icc_within <- calc_icc_per_protein(within_run_reps, biomarker_cols)
icc_cross  <- calc_icc_per_protein(cross_run_reps,  biomarker_cols)

icc_combined <- icc_within %>%
  rename(ICC_within = ICC, lower_within = lower, upper_within = upper, n_within = n_pairs) %>%
  left_join(icc_cross %>%
              rename(ICC_cross = ICC, lower_cross = lower, upper_cross = upper, n_cross = n_pairs),
            by = "Biomarker") %>%
  mutate(
    Category_within = case_when(
      ICC_within >= 0.9  ~ "Excellent", ICC_within >= 0.75 ~ "Good",
      ICC_within >= 0.5  ~ "Moderate",  !is.na(ICC_within)  ~ "Poor",
      TRUE ~ "Not estimable (no within-Run dup)"),
    Category_cross = case_when(
      ICC_cross >= 0.9   ~ "Excellent", ICC_cross >= 0.75  ~ "Good",
      ICC_cross >= 0.5   ~ "Moderate",  !is.na(ICC_cross)   ~ "Poor",
      TRUE ~ "Insufficient data")
  )
```

```{r icc-plots, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
icc_combined %>%
  select(Biomarker, ICC_within, ICC_cross) %>%
  pivot_longer(-Biomarker, names_to = "Level", values_to = "ICC") %>%
  mutate(Level = recode(Level,
    ICC_within = "Within-Run (across Bays)", ICC_cross = "Cross-Run")) %>%
  ggplot(aes(x = ICC, fill = Level)) +
  geom_histogram(binwidth = 0.05, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = c(0.75, 0.9), linetype = "dashed", color = "red") +
  facet_wrap(~ Level) +
  theme_bw() +
  labs(title = "ICC distribution by replicate type",
       subtitle = "Red lines: 0.75 (Good) and 0.90 (Excellent) thresholds",
       x = "ICC", y = "Count of proteins") +
  theme(legend.position = "none")

icc_combined %>%
  summarise(
    `Within-Run: Excellent (>=0.9)` = sum(Category_within == "Excellent", na.rm=T),
    `Within-Run: Good (0.75-0.9)`   = sum(Category_within == "Good", na.rm=T),
    `Within-Run: Moderate/Poor`     = sum(Category_within %in% c("Moderate","Poor"), na.rm=T),
    `Cross-Run: Excellent (>=0.9)`  = sum(Category_cross  == "Excellent", na.rm=T),
    `Cross-Run: Good (0.75-0.9)`    = sum(Category_cross  == "Good", na.rm=T),
    `Cross-Run: Moderate/Poor`      = sum(Category_cross  %in% c("Moderate","Poor"), na.rm=T)
  ) %>%
  pivot_longer(everything(), names_to = "Category", values_to = "N proteins") %>%
  kable(caption = "ICC summary across proteins") %>%
  kable_styling(full_width = FALSE)

poor_icc <- icc_combined %>%
  filter(Category_cross %in% c("Moderate", "Poor") | ICC_cross < 0.75)

if (nrow(poor_icc) > 0) {
  kable(poor_icc %>% arrange(ICC_cross), digits = 3,
        caption = "Proteins with ICC_cross < 0.75") %>%
    kable_styling(full_width = FALSE)
}

write_csv(icc_combined, "output_files/icc_results.csv")
```

## E.2 Control-pool reproducibility (within-plate / within-Run / cross-Run)

The Alamar control pools are **identical material repeated on every plate** — HIHG (one pooled reference,
2 aliquots/plate), SC (3/plate) and IPC (3/plate) — so *all* of their variation is technical. A per-protein
random-effects decomposition `value ~ (1|Run) + (1|Run:Bay)` splits it into:

- **% Run** — cross-Run drift (the dominant batch axis)
- **% Bay-in-Run** — bay / sequential drift within a Run
- **% Well** — residual well-to-well noise on the same plate (within-plate precision)

For identical material a classical ICC is not meaningful (no between-subject signal), so reproducibility is
reported as variance fractions + SD (log2) + CV. Good = low Run/Bay %, low CV. This is the within-plate /
within-Run level that the biological duplicates (E.1) cannot give.

```{r e2-control-repro, message=FALSE, warning=FALSE, results='asis'}
ctrl_files <- list(
  HIHG = c("../Replicate_Analysis/output_files/hihg_reps_NPQ.csv", "../../Primary_QC/output_files/hihg_reps_NPQ.csv"),
  SC   = "../Replicate_Analysis/output_files/sc_reps_NPQ.csv",
  IPC  = "../Replicate_Analysis/output_files/ipc_reps_NPQ.csv")
read_first <- function(paths) { for (p in paths) if (file.exists(p)) return(suppressMessages(readr::read_csv(p, show_col_types = FALSE))); NULL }
ctrl_list <- purrr::compact(lapply(ctrl_files, read_first))

repro <- NULL
if (length(ctrl_list) && all(c("Run","Bay") %in% names(ctrl_list[[1]]))) {
  bm_ctrl <- Reduce(intersect, lapply(ctrl_list, function(d) { cc <- setdiff(names(d), c("SampleID","Run","Bay"))
                    cc[sapply(cc, function(x) is.numeric(d[[x]]))] }))
  bm_ctrl <- setdiff(intersect(bm_ctrl, biomarker_cols), "mCherry")

  vc_one <- function(d, b) {
    dd <- tibble(y = d[[b]], Run = factor(d$Run), Bay = factor(d$Bay)) %>% filter(is.finite(y))
    if (dplyr::n_distinct(dd$Run) < 2 || nrow(dd) < 6) return(NULL)
    fit <- tryCatch(suppressMessages(lme4::lmer(y ~ (1|Run) + (1|Run:Bay), data = dd,
                    control = lme4::lmerControl(check.conv.singular = "ignore"))), error = function(e) NULL)
    if (is.null(fit)) return(NULL)
    v <- as.data.frame(lme4::VarCorr(fit)); gv <- function(g){ x <- v$vcov[v$grp==g]; if (length(x)) x else 0 }
    vr <- gv("Run"); vb <- gv("Run:Bay"); vw <- gv("Residual"); tot <- vr+vb+vw
    if (tot <= 0) return(NULL)
    lin <- 2^dd$y
    tibble(Biomarker = b, pct_Run = 100*vr/tot, pct_Bay_in_Run = 100*vb/tot, pct_Well = 100*vw/tot,
           SD_within_plate = sqrt(vw), SD_within_run = sqrt(vb+vw), SD_cross_run = sqrt(tot),
           CV_pct = 100*sd(lin, na.rm = TRUE)/mean(lin, na.rm = TRUE))
  }
  repro <- purrr::map_dfr(names(ctrl_list), function(ct)
    purrr::map_dfr(bm_ctrl, ~vc_one(ctrl_list[[ct]], .x)) %>% mutate(Control = ct, .before = 1))
  write_csv(repro, "output_files/control_reproducibility.csv")

  print(repro %>% group_by(Control) %>%
    summarise(n_biomarkers = n(), `median %Run` = round(median(pct_Run),1),
              `median %Bay` = round(median(pct_Bay_in_Run),1), `median %Well` = round(median(pct_Well),1),
              `median SD cross-run (log2)` = round(median(SD_cross_run),2),
              `median CV%` = round(median(CV_pct),1), .groups = "drop") %>%
    kable(caption = "Control-pool variance decomposition (identical material; median across biomarkers). Low Run/Bay % and low CV = good.") %>%
    kable_styling(full_width = FALSE))

  print(repro %>% filter(Control == "HIHG") %>% arrange(desc(pct_Run)) %>% head(15) %>%
    transmute(Biomarker, `%Run` = round(pct_Run,1), `%Bay` = round(pct_Bay_in_Run,1),
              `%Well` = round(pct_Well,1), `SD cross-run` = round(SD_cross_run,2), `CV%` = round(CV_pct,1)) %>%
    kable(caption = "Most run-unstable proteins in the HIHG pool (identical material) — batch-correction priority") %>%
    kable_styling(full_width = FALSE) %>% scroll_box(width = "100%"))
} else {
  cat("Control replicate files not found (Replicate_Analysis/output_files or Primary_QC/output_files); skipping E.2.\n")
}
```

---

# F. Formal Per-Protein Batch Effect Test

Linear model per protein: `NPQ ~ Run_ID + Bay + Ancestry + sex + age_at_subject + Site + CDX_collapsed`

Run and Bay are tested by **nested-model F-tests** — each batch term as the gain over a model
containing all biological covariates plus the *other* batch term (`anova(reduced, full)`).
FDR correction across proteins within each term.

**Why not `Anova(type="III")`:** Site is partly nested in Run (some sites appear in a single Run),
so the full fixed-effect design is rank-deficient and `car::Anova(type="III")` errors on the aliased
coefficients (this previously NA'd every protein). The nested F-test compares residual sums of squares
and is immune to rank deficiency. Because Run and Site are confounded, the Run test is *conditional on
Site* (conservative). **Limitation:** Run_ID is fixed — low power with many Runs. Interpret alongside
variance partitioning (Section D) and the control-pool variance components (Section E.2).

```{r formal-test, message=FALSE, warning=FALSE}
library(car)

test_df <- df_unique %>%
  filter(!is.na(Ancestry), !is.na(sex), !is.na(age_at_subject),
         !is.na(CDX_collapsed), !is.na(Run_ID), !is.na(Bay)) %>%
  mutate(Run_fac = as.factor(Run_ID))

batch_test_results <- map_dfr(biomarker_cols, function(bm) {
  y  <- test_df[[bm]]
  ok <- complete.cases(test_df[, c("Ancestry","sex","age_at_subject","CDX_collapsed","Site","Bay")]) & !is.na(y)
  if (sum(ok) < 20)
    return(tibble(Biomarker = bm, p_Run = NA_real_, p_Bay = NA_real_))
  d <- test_df[ok, ]; d$.y <- y[ok]

  # Nested-model F-tests (robust to the Run/Site rank-deficiency that breaks Anova type III).
  # Each batch term = its gain over a model with all bio covariates + the OTHER batch term.
  fit_full  <- tryCatch(lm(.y ~ Run_fac + Bay + Ancestry + sex + age_at_subject + Site + CDX_collapsed, data = d), error = function(e) NULL)
  fit_noRun <- tryCatch(lm(.y ~           Bay + Ancestry + sex + age_at_subject + Site + CDX_collapsed, data = d), error = function(e) NULL)
  fit_noBay <- tryCatch(lm(.y ~ Run_fac +       Ancestry + sex + age_at_subject + Site + CDX_collapsed, data = d), error = function(e) NULL)
  if (is.null(fit_full)) return(tibble(Biomarker = bm, p_Run = NA_real_, p_Bay = NA_real_))

  p_run <- if (!is.null(fit_noRun)) tryCatch(anova(fit_noRun, fit_full)[["Pr(>F)"]][2], error = function(e) NA_real_) else NA_real_
  p_bay <- if (!is.null(fit_noBay)) tryCatch(anova(fit_noBay, fit_full)[["Pr(>F)"]][2], error = function(e) NA_real_) else NA_real_
  tibble(Biomarker = bm, p_Run = p_run, p_Bay = p_bay)
})

batch_test_results <- batch_test_results %>%
  mutate(FDR_Run = p.adjust(p_Run, "BH"),
         FDR_Bay = p.adjust(p_Bay, "BH"),
         Sig_Run = FDR_Run < FDR_THRESHOLD,
         Sig_Bay = FDR_Bay < FDR_THRESHOLD) %>%
  arrange(FDR_Run)

cat("Proteins with significant Run effect (FDR <", FDR_THRESHOLD, "):",
    sum(batch_test_results$Sig_Run, na.rm=TRUE), "\n")
cat("Proteins with significant Bay effect (FDR <", FDR_THRESHOLD, "):",
    sum(batch_test_results$Sig_Bay, na.rm=TRUE), "\n")

batch_test_results %>%
  filter(Sig_Run | Sig_Bay) %>% head(30) %>%
  kable(digits = 4, caption = "Proteins with significant Run or Bay effect (FDR corrected)") %>%
  kable_styling(full_width = FALSE)

write_csv(batch_test_results, "output_files/batch_effect_test_results.csv")
```

---

# G. Submission-year (2025 vs 2026) batch structure

The 50-plate set spans two submission years — **2025** (ALZ123) and **2026** (8+7-plate). Each Run belongs
to one year, so the strong Run effects above are partly a **year-level** shift. This section isolates it by
clustering the **HIHG control** (the *same* sample on every plate — any structure is purely technical) and
decomposing each protein's variance by year.

```{r year-cluster, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
year_pc1_r2 <- NA_real_; n_year_sig <- NA_integer_; plates_split <- NA_real_; year_res <- NULL
.hihg <- "../../Primary_QC/output_files/hihg_reps_NPQ.csv"
if (file.exists(.hihg)) {
  H <- suppressMessages(read_csv(.hihg, show_col_types = FALSE))
  bc <- intersect(biomarker_cols, names(H))
  if (!length(bc)) bc <- setdiff(names(H), c("Run","Bay"))[sapply(setdiff(names(H), c("Run","Bay")), function(c) is.numeric(H[[c]]))]
  H$year <- substr(as.character(H$Run), 1, 4)
  M <- as.matrix(H[, bc]); M <- M[, apply(M, 2, function(c) sd(c, na.rm = TRUE) > 0)]; M[is.na(M)] <- 0
  pca <- prcomp(M, scale. = TRUE); ve <- 100 * summary(pca)$importance[2, 1:2]
  sc  <- data.frame(PC1 = pca$x[,1], PC2 = pca$x[,2], year = H$year)
  ss  <- summary(aov(PC1 ~ year, sc))[[1]]; year_pc1_r2 <- ss[1,"Sum Sq"] / sum(ss[,"Sum Sq"])
  print(ggplot(sc, aes(PC1, PC2, color = year)) + geom_point(size = 2, alpha = .8) + stat_ellipse() +
        theme_bw() + labs(title = "HIHG control PCA by submission year (same sample, every plate)",
                          subtitle = sprintf("PC1 %.0f%% var; year explains %.0f%% of PC1", ve[1], 100*year_pc1_r2)))
  ggsave("output_files/hihg_pca_by_year.png", width = 7, height = 5, dpi = 110)
  H$plate <- paste0(H$Run, "_", H$Bay)
  pm <- H %>% group_by(plate, year) %>% summarise(across(all_of(colnames(M)), ~mean(.x, na.rm = TRUE)), .groups = "drop")
  PM <- scale(as.matrix(pm[, colnames(M)])); PM[is.na(PM)] <- 0
  tabp <- table(cluster = cutree(hclust(dist(PM), "ward.D2"), 2), year = pm$year)
  plates_split <- sum(apply(tabp, 1, max)) / sum(tabp)
  cat(sprintf("HIHG PC1 explained by year: %.0f%%  |  2-cluster plate split matches year: %.0f%% of %d plates\n\n",
              100*year_pc1_r2, 100*plates_split, sum(tabp)))
  print(knitr::kable(as.data.frame.matrix(tabp), caption = "Per-plate 2-cluster cut vs submission year"))
}
```

```{r year-per-protein, message=FALSE, warning=FALSE}
if (file.exists(.hihg)) {
  year_res <- purrr::map_dfr(bc, function(b) {
    x25 <- H[[b]][H$year == "2025"]; x26 <- H[[b]][H$year == "2026"]
    x25 <- x25[is.finite(x25)]; x26 <- x26[is.finite(x26)]
    if (length(x25) < 3 || length(x26) < 3) return(NULL)
    ss <- summary(aov(H[[b]] ~ H$year))[[1]]
    tt <- suppressWarnings(t.test(x26, x25))
    tibble(Biomarker = b, pct_var_year = 100 * ss[1,"Sum Sq"] / sum(ss[,"Sum Sq"]),
           log2_diff_26_25 = unname(diff(rev(tt$estimate))), p = tt$p.value)
  })
  year_res$FDR <- p.adjust(year_res$p, "BH"); n_year_sig <- sum(year_res$FDR < 0.05, na.rm = TRUE)
  write_csv(year_res %>% arrange(desc(pct_var_year)), "output_files/year_effect_per_protein.csv")
  cat(sprintf("Proteins with a significant 2025-vs-2026 shift in the HIHG control (FDR < 0.05): %d / %d\n\n",
              n_year_sig, nrow(year_res)))
  year_res %>% arrange(desc(pct_var_year)) %>% head(12) %>%
    mutate(across(c(pct_var_year, log2_diff_26_25), ~round(., 2)), FDR = signif(FDR, 2)) %>%
    kable(caption = "Top proteins by % HIHG variance explained by submission year") %>%
    kable_styling(full_width = FALSE)
}
```

## G.2 Control-pool PCA (HIHG / SC / IPC) — backup batch view

PCA of each control pool (identical material), one point per well, colored by submission year. Any structure
is purely technical, so clean year separation = batch. Wells that sit off their *own* year-cluster are
candidate problem plates. Per-well PC1–4 scores are written to `pc_scores_controls.csv`.

```{r g2-control-pca, message=FALSE, warning=FALSE, fig.width=13, fig.height=4.5, results='asis'}
read_first2 <- function(paths) { for (p in paths) if (file.exists(p)) return(suppressMessages(readr::read_csv(p, show_col_types = FALSE))); NULL }
ctrl_pca_files <- list(
  HIHG = c("../Replicate_Analysis/output_files/hihg_reps_NPQ.csv", "../../Primary_QC/output_files/hihg_reps_NPQ.csv"),
  SC   = "../Replicate_Analysis/output_files/sc_reps_NPQ.csv",
  IPC  = "../Replicate_Analysis/output_files/ipc_reps_NPQ.csv")
.cl <- purrr::compact(lapply(ctrl_pca_files, read_first2))
rz <- function(x) { m <- mad(x); if (m == 0) return(rep(0, length(x))); abs((x - median(x))/(1.4826*m)) }
ctrl_scores <- NULL
if (length(.cl) && all(c("Run","Bay") %in% names(.cl[[1]]))) {
  do_pca <- function(d, label) {
    d$year <- substr(as.character(d$Run), 1, 4)
    bmc <- setdiff(names(d), c("SampleID","Run","Bay","year")); bmc <- bmc[sapply(bmc, function(x) is.numeric(d[[x]]))]
    bmc <- setdiff(bmc, "mCherry")
    M <- as.matrix(d[, bmc]); M <- M[, apply(M, 2, function(c) sd(c, na.rm = TRUE) > 0), drop = FALSE]; M[is.na(M)] <- 0
    p <- prcomp(M, scale. = TRUE); ve <- 100*summary(p)$importance[2, 1:4]
    sc <- as.data.frame(p$x[, 1:4]); names(sc) <- paste0("PC", 1:4)
    list(out = tibble(Control = label, SampleID = d$SampleID, Run = d$Run, Bay = d$Bay, year = d$year,
                      plate = paste0(d$Run, "_", d$Bay)) %>% bind_cols(sc), ve = ve)
  }
  pl <- lapply(names(.cl), function(ct) do_pca(.cl[[ct]], ct)); names(pl) <- names(.cl)
  ctrl_scores <- bind_rows(lapply(pl, `[[`, "out")) %>% mutate(across(starts_with("PC"), ~round(.x, 3))) %>%
    group_by(Control, year) %>% mutate(is_outlier_in_year = rz(PC1) > 4 | rz(PC2) > 4) %>% ungroup() %>%
    mutate(pos = sub("_[0-9]{8}-.*$", "", SampleID),
           pos = ifelse(grepl("^(SC|IPC|NC)_[0-9]", pos), pos, Control))   # SC_1/2/3, IPC_1/2/3; HIHG is one pool
  FLAGGED_RUNS <- ctrl_scores %>% filter(Control == "HIHG", is_outlier_in_year) %>% distinct(Run) %>% pull(Run)
  ctrl_scores <- ctrl_scores %>% mutate(flag = ifelse(Run %in% FLAGGED_RUNS, Run, "other plates"))
  write_csv(ctrl_scores, "output_files/pc_scores_controls.csv")

  yr_cols <- c("2025" = "#1b9e77", "2026" = "#d95f02")
  plots <- lapply(names(pl), function(ct) {
    d <- pl[[ct]]$out; ve <- pl[[ct]]$ve
    ggplot(d, aes(PC1, PC2, color = year)) + geom_point(size = 1.8, alpha = .85) +
      stat_ellipse(aes(group = year), linewidth = .4) + scale_color_manual(values = yr_cols) + theme_bw() +
      labs(title = paste0(ct, " pool"), subtitle = sprintf("PC1 %.0f%% / PC2 %.0f%%", ve[1], ve[2]), x = "PC1", y = "PC2", color = "Year")
  })
  .gctrl <- patchwork::wrap_plots(plots, nrow = 1) + patchwork::plot_annotation(
    title = "Control-pool PCA by submission year — identical material, so all separation is technical/batch")
  ggsave("output_files/pca_controls_by_year.png", .gctrl, width = 13, height = 4.5, dpi = 120)
  print(.gctrl)

  print(ctrl_scores %>% filter(is_outlier_in_year) %>% distinct(Control, plate, year) %>% arrange(Control, plate) %>%
    kable(caption = "Control wells off their own year-cluster (candidate problem plates)") %>% kable_styling(full_width = FALSE))
} else {
  cat("Control replicate files not found; skipping G.2.\n")
}
```

Same control PCAs **colored multiple ways** — by year (batch), by Bay, and by position (SC_1/2/3,
IPC_1/2/3) or, for the single-pool HIHG, by flagged plate. IPC clusters by *position* with no year
separation (the stable normalization anchor); HIHG/SC separate by year (batch).

**Reminder — the three IPC clusters are three distinct control pool *mixes* (IPC_1/2/3), one per plate
position: the expected pattern, not reagent lots.** k=3 clustering maps cleanly onto position (diagonal
cross-tab) and **every run carries all three** (19/21 runs) — by-design within-plate mixes, not run/time
(lot) blocks. The three share a near-identical overall mean (≈13.3 log2) but their per-protein profiles are
**anti-correlated** (r ≈ −0.2 to −0.6; they differ most in Aβ38/Aβ40/pTau-217/PTN) → genuinely different
compositions, as expected for three reference mixes.

A targeted **within-year lot scan** of the HIHG/SC pools found **no sustained reagent/lot step beyond the
2025↔2026 year shift**: within-year, run-to-run variation is mostly noise, and the only discrete within-2025
break is the Nov-24/25 degradation already flagged in G.3 (two runs, not a sustained block). So the batch
axis to adjust for remains **submission year** (plus the flagged plates).

```{r g2b-control-multicolor, message=FALSE, warning=FALSE, fig.width=14, fig.height=11}
if (!is.null(ctrl_scores)) {
  yrc <- c("2025" = "#1b9e77", "2026" = "#d95f02")
  flagc <- setNames(c("grey75", scales::hue_pal()(max(1, length(FLAGGED_RUNS)))), c("other plates", FLAGGED_RUNS))
  cpanel <- function(d, color_by, ttl) {
    g <- ggplot(d, aes(PC1, PC2, color = .data[[color_by]])) + geom_point(size = 1.4, alpha = .8) +
      theme_bw(base_size = 9) + labs(title = ttl, x = "PC1", y = "PC2", color = color_by) +
      theme(legend.key.size = unit(.25, "cm"), legend.text = element_text(size = 6),
            legend.title = element_text(size = 7), plot.title = element_text(size = 9))
    if (color_by == "year") g <- g + scale_color_manual(values = yrc)
    if (color_by == "flag") g <- g + scale_color_manual(values = flagc)
    g
  }
  rows <- lapply(unique(ctrl_scores$Control), function(ct) {
    d <- ctrl_scores %>% filter(Control == ct)
    has_pos <- dplyr::n_distinct(d$pos) > 1
    third <- if (has_pos) "pos" else "flag"; tlab <- if (has_pos) "position" else "flagged plates"
    cpanel(d, "year", paste0(ct, " — by year")) | cpanel(d, "Bay", paste0(ct, " — by Bay")) | cpanel(d, third, paste0(ct, " — by ", tlab))
  })
  .gmc <- patchwork::wrap_plots(rows, ncol = 1) +
    patchwork::plot_annotation(title = "Control-pool PCAs colored multiple ways (identical material; all structure is technical)")
  ggsave("output_files/pca_controls_multicolor.png", .gmc, width = 14, height = 11, dpi = 120)
  print(.gmc)
}
```

## G.3 Candidate problem plates — Bay-sequential drift (`20251124-1407`, `20251125-1332`)

The HIHG control's 4 extreme wells are both Bay3 wells of two consecutive-day runs,
**`20251124-1407_Bay3`** and **`20251125-1332_Bay3`**. The shift is a *within-run sequential Bay drift*
(Bay1 → Bay2 → Bay3 monotonic in the identical control), **not** a Bay3-wide effect — other 2025 Bay3
plates are normal. It is **plate-wide, not control-only**: study samples on these plates show the same
gradient (CD40LG drops ≈4 log2 / 16× from Bay1 to Bay3, NEFH follows; PTN/Aβ42/pTau-217 move in the
control's direction). `20251124-1407` is **also** the APOE off-band plate (APOE §C) — two QC modules
independently flag it.

**Do these need to be treated as a batch?** The shift is real and plate-wide, so these plates are *not*
exchangeable with the rest — but the principled handling is to **model the hierarchy (year → Run →
Run:Bay) as random effects** (which already absorbs a per-plate/bay shift) rather than invent a one-off
batch label, **plus flag `20251124-1407` / `20251125-1332` for a leave-out sensitivity analysis** on the
affected markers (CD40LG, NEFH, PTN, Aβ42, pTau-217). Validate any correction on the HIHG control (its
within-plate variance on these plates should shrink afterward). A blanket whole-plate correction
calibrated on the control would *over-correct* markers where study samples drift less than the control
(e.g. PTN: ≈1 log2 in the control vs ≈0.1 in study samples).

---

# H. Sample-level PCA (outlier identification & structure)

PCA of all study samples (biomarker matrix, scaled, mean-imputed). Per-sample PC1–4 are written to
`pc_scores_per_sample.csv` (sorted most-extreme first, with robust-z outlier flags) so individual outliers
can be named. The association table shows **which variable each PC captures** — on this data PC1–PC2 is
largely outlier/intensity-driven and the ancestry/site/year structure lives on **PC3**, so the featured plot
is PC2 vs PC3.

```{r h-study-pca, message=FALSE, warning=FALSE, results='asis'}
pca_study <- df_samples
Mst <- as.matrix(pca_study[, biomarker_cols])
keepc <- apply(Mst, 2, function(c) sum(is.finite(c)) > 0 && sd(c, na.rm = TRUE) > 0); Mst <- Mst[, keepc, drop = FALSE]
cm <- colMeans(Mst, na.rm = TRUE); for (j in seq_len(ncol(Mst))) { na <- !is.finite(Mst[, j]); if (any(na)) Mst[na, j] <- cm[j] }
pst <- prcomp(Mst, scale. = TRUE); ve_st <- 100*summary(pst)$importance[2, 1:4]
sc_st <- as.data.frame(pst$x[, 1:4]); names(sc_st) <- paste0("PC", 1:4)
ps <- pca_study %>%
  select(SAMPLE_ALIQUOT, SAMPLE, Run_ID, Bay, year,
         any_of(c("Site","Ancestry","sex","age_at_subject","CDX_collapsed","metadata_source"))) %>%
  bind_cols(sc_st)
rzc <- function(x) { m <- mad(x); if (m == 0) return(rep(0, length(x))); (x - median(x))/(1.4826*m) }
for (pc in paste0("PC", 1:4)) ps[[paste0("rz_", pc)]] <- round(rzc(ps[[pc]]), 2)
ps$is_outlier <- with(ps, abs(rz_PC1) > 5 | abs(rz_PC2) > 5 | abs(rz_PC3) > 5 | abs(rz_PC4) > 5)
ps <- ps %>% mutate(across(starts_with("PC"), ~round(.x, 3))) %>% arrange(desc(sqrt(PC1^2 + PC2^2 + PC3^2 + PC4^2)))
write_csv(ps, "output_files/pc_scores_per_sample.csv")

r2f <- function(pc, v) { ok <- !is.na(ps[[v]]); s <- summary(aov(ps[[pc]][ok] ~ factor(ps[[v]][ok])))[[1]]; round(100*s[1,"Sum Sq"]/sum(s[,"Sum Sq"]), 1) }
assoc <- expand.grid(PC = paste0("PC", 1:4), var = intersect(c("year","Ancestry","Site","CDX_collapsed"), names(ps)), stringsAsFactors = FALSE)
assoc$R2_pct <- mapply(r2f, assoc$PC, assoc$var); assoc <- tidyr::pivot_wider(assoc, names_from = var, values_from = R2_pct)
write_csv(assoc, "output_files/pc_variable_association.csv")
print(kable(assoc, caption = "% variance in each study PC explained by each variable (one at a time)") %>% kable_styling(full_width = FALSE))
cat(sprintf("\n\n*Study PCA variance explained: PC1 %.1f%%, PC2 %.1f%%, PC3 %.1f%%, PC4 %.1f%%.*\n\n", ve_st[1], ve_st[2], ve_st[3], ve_st[4]))
cat(sprintf("*Outliers (|robust z| > 5 on any PC1-4): %d — see `pc_scores_per_sample.csv` (sorted most-extreme first).*\n\n", sum(ps$is_outlier)))
```

```{r h-study-matrix, message=FALSE, warning=FALSE, fig.width=13, fig.height=8}
anc_cols <- c(HI_WH = "#2166AC", HI_MU = "#74ADD1", HI_BL = "#92C5DE", AA = "#D6604D", AFDC = "#8B0000")
labp <- function(p) sprintf("%s (%.1f%%)", p, ve_st[as.integer(substr(p, 3, 3))])
prs <- list(c("PC1","PC2"), c("PC1","PC3"), c("PC1","PC4"), c("PC2","PC3"), c("PC2","PC4"), c("PC3","PC4"))
pan <- function(xy, legend = FALSE) ggplot(ps %>% filter(!is.na(Ancestry)), aes(.data[[xy[1]]], .data[[xy[2]]], color = Ancestry)) +
  geom_point(size = .4, alpha = .4) + stat_ellipse(linewidth = .4) + scale_color_manual(values = anc_cols, na.value = "grey75") +
  theme_bw(base_size = 9) + labs(x = labp(xy[1]), y = labp(xy[2])) + theme(legend.position = if (legend) "right" else "none")
.gmat <- patchwork::wrap_plots(c(lapply(prs[-1], pan), list(pan(prs[[1]], TRUE))), ncol = 3) +
  patchwork::plot_annotation(title = "Study-sample PCA scatter-matrix (PC1-PC4), colored by ancestry")
ggsave("output_files/pca_study_matrix.png", .gmat, width = 13, height = 8, dpi = 120)
print(.gmat)
```

```{r h-study-pc2pc3, message=FALSE, warning=FALSE, fig.width=13, fig.height=10}
yr_cols <- c("2025" = "#1b9e77", "2026" = "#d95f02")
featc <- function(fill, cols = NULL, legend = TRUE, ttl) {
  d <- ps %>% filter(!is.na(.data[[fill]]))
  g <- ggplot(d, aes(PC2, PC3, color = .data[[fill]])) + geom_point(size = .5, alpha = .45) + stat_ellipse(linewidth = .5) +
    theme_bw(base_size = 10) + labs(title = ttl, x = labp("PC2"), y = labp("PC3"), color = fill) +
    theme(legend.position = if (legend) "right" else "none", legend.key.size = unit(.3, "cm"), legend.text = element_text(size = 7))
  if (!is.null(cols)) g <- g + scale_color_manual(values = cols, na.value = "grey75")
  g
}
top_out <- ps %>% slice_max(sqrt(PC1^2 + PC2^2 + PC3^2 + PC4^2), n = 12)
pol <- ggplot(ps, aes(PC2, PC3)) + geom_point(size = .4, alpha = .3, color = "grey60") +
  geom_point(data = top_out, color = "red", size = 1.5) +
  ggrepel::geom_text_repel(data = top_out, aes(label = SAMPLE_ALIQUOT), size = 2.3, max.overlaps = 20) +
  theme_bw(base_size = 10) + labs(title = "12 most extreme (labeled)", x = labp("PC2"), y = labp("PC3"))
.gpc23 <- (featc("Ancestry", anc_cols, TRUE, "by ancestry") | featc("year", yr_cols, TRUE, "by year")) /
      (featc("Site", NULL, FALSE, "by site") | pol) +
      patchwork::plot_annotation(title = "Study-sample PCA - PC2 vs PC3 (the ancestry/site/year plane)")
ggsave("output_files/pca_study_PC2PC3.png", .gpc23, width = 13, height = 10, dpi = 120)
print(.gpc23)
```

---

# I. Summary and Decision

```{r decision, message=FALSE, warning=FALSE}
n_sig_run      <- sum(batch_test_results$Sig_Run, na.rm=TRUE)
n_sig_bay      <- sum(batch_test_results$Sig_Bay, na.rm=TRUE)
n_proteins     <- length(biomarker_cols)
median_run_var <- median(vp_df_out$`Run_ID`, na.rm=TRUE) * 100
median_bay_var <- median(vp_df_out$`Run_ID:Bay`, na.rm=TRUE) * 100
max_run_var    <- max(vp_df_out$`Run_ID`, na.rm=TRUE) * 100
high_conf      <- conf_results %>% filter(Concern == "HIGH")
n_poor_icc     <- nrow(poor_icc)

cat("=== BATCH EFFECT SUMMARY ===\n\n")
cat(sprintf("Formal test:        %d/%d proteins significant Run effect (FDR < %.2f)\n",
            n_sig_run, n_proteins, FDR_THRESHOLD))
cat(sprintf("                    %d/%d proteins significant Bay effect\n\n",
            n_sig_bay, n_proteins))
cat(sprintf("Variance partition: median Run = %.1f%%, max Run = %.1f%%, median Bay = %.1f%%\n\n",
            median_run_var, max_run_var, median_bay_var))
cat(sprintf("ICC:                %d proteins with cross-Run ICC < 0.75\n\n", n_poor_icc))

if (nrow(high_conf) > 0) {
  cat("HIGH confounding (Cramers V > 0.5):\n")
  print(high_conf %>% select(Batch_Var, Covariate, CramersV))
  cat("\n")
}

cat("=== RECOMMENDATION ===\n\n")
if (n_sig_run == 0 && median_run_var < 5) {
  cat("Batch effects MINIMAL.\n")
  cat("-> Include Run_ID as a fixed covariate in downstream models.\n")
} else if (median_run_var < 10 && nrow(high_conf) == 0) {
  cat("Batch effects MODERATE, not strongly confounded with biology.\n")
  cat("-> Include Run_ID as random effect (lme4) in downstream models.\n")
  cat("-> OR apply ComBat (sva) for methods that cannot use random effects.\n")
} else if (nrow(high_conf) > 0) {
  cat("CAUTION: High confounding detected. ComBat may remove biological signal.\n")
  cat("-> For association testing: use Run_ID as random effect (lme4) -- preferred.\n")
  cat("-> For PCA/clustering: ComBat with bio covariates as protected, interpret with caution.\n")
  cat("-> Run sensitivity analysis with and without correction.\n")
} else {
  cat("Batch effects SUBSTANTIAL.\n")
  cat("-> Apply ComBat (sva::ComBat) with Ancestry + sex + age as protected covariates.\n")
  cat("-> Validate correction using HIHG replicate ICC post-correction.\n")
}

cat("\n=== WHAT TO ADJUST FOR (this dataset) ===\n")
if (!is.na(year_pc1_r2)) {
  cat(sprintf("- SUBMISSION YEAR (2025 vs 2026): the dominant batch axis (HIHG PC1 %.0f%% year; %d proteins shift;\n",
              100*year_pc1_r2, n_year_sig))
  cat(sprintf("  unsupervised clustering splits %.0f%% of plates by year). Adjust for YEAR (or Run nested in year)\n",
              100*plates_split))
  cat("  in EVERY cross-year analysis, or ComBat by year; anchor/validate the correction on the HIHG control.\n")
}
.site_conf <- suppressWarnings(max(conf_results$CramersV[conf_results$Covariate == "Site"], na.rm = TRUE))
if (is.finite(.site_conf) && .site_conf > 0.3)
  cat(sprintf("- SITE: moderate confounding with Run (Cramers V %.2f). Include Site as a COVARIATE; do not 'correct'\n  it away (site is partly design/biology).\n", .site_conf))
cat("- Keep biological covariates (Ancestry, sex, age, diagnosis) PROTECTED -- never regress them out.\n")

cat("\n=== OUTPUT FILES ===\n")
cat("batch_covariate_confounding.csv  - Cramers V between batch (incl. year) and covariates\n")
cat("variance_partition_results.csv   - Per-protein variance fractions\n")
cat("icc_results.csv                  - Per-protein ICC (biological duplicates): cross-Run + within-Run (not estimable)\n")
cat("control_reproducibility.csv      - Per-protein variance components (Run/Bay/well) + SD + CV for HIHG/SC/IPC pools (E.2)\n")
cat("year_effect_per_protein.csv      - Per-protein 2025-vs-2026 shift (HIHG control); hihg_pca_by_year.png\n")
cat("batch_effect_test_results.csv    - Per-protein p/FDR for Run and Bay terms (nested F-test)\n")
cat("pc_scores_controls.csv           - Per control well PC1-4 + position + off-year-cluster flag (G.2)\n")
cat("pca_controls_by_year.png / pca_controls_multicolor.png - control PCAs (by year; and by year/Bay/position) (G.2)\n")
cat("pc_scores_per_sample.csv         - Per study sample PC1-4 (+ robust-z outlier flags), sorted most-extreme (H)\n")
cat("pc_variable_association.csv      - % variance of each study PC explained by year/ancestry/site/diagnosis (H)\n")
```

---

# Dataset Notes

```{r notes, echo=FALSE, results='asis'}
notes_file <- "notes.md"
if (file.exists(notes_file)) {
  cat(paste(readLines(notes_file, warn = FALSE), collapse = "\n"))
} else {
  cat("*No notes file found. Create `notes.md` for dataset-specific observations.*\n")
}
```
