---
title: "Batch Effects Analysis for Alamar Biomarker Data"
author: "ERM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

# A. Overview

This module assesses technical batch effects (Run and Bay) in the merged biomarker dataset.
The hierarchy is: **Run → Bay (1/2/3, sequential within Run) → Sample**.

Goals:
1. Characterize plate composition (race, sex, age, site, diagnosis) by Run and Bay
2. Identify confounding between biological covariates and batch variables
3. Partition variance into Run, Bay, and biological sources per protein
4. Quantify cross-plate reproducibility from HIHG replicates
5. Make a data-driven decision about batch correction approach

Biomarkers are in log2 NPQ scale throughout (no back-transformation needed for variance
partitioning or linear models).

**For a new dataset: update the input filename in Section A.3.**

---

## A.1 Input Data

Default input: `merged_combined_post_QC.csv` from `Metadata_Merge/output_files_*/`.
This file has separate `Run` and `Bay` columns and includes both standard and low-abundance biomarkers.

Fallback: `merged_standard_post_QC.csv` from the same directory (standard biomarkers only, no separate Run/Bay — parsed from `RUN` column).

Place in `input_files/` or let the pipeline auto-copy.

## A.2 Parameters

```{r params}
# --- Adjustable parameters ---
VP_CORES       <- 4      # cores for variancePartition (increase if available)
VP_MIN_SAMPLES <- 10     # min samples per Run to include in VP
ICC_MIN_REPS   <- 2      # min replicate pairs required for ICC calculation
FDR_THRESHOLD  <- 0.05   # FDR cutoff for declaring significant batch effect
```

## A.3 Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(variancePartition)
library(BiocParallel)
library(knitr)
library(kableExtra)
library(ggplot2)
library(patchwork)

# ---- Directories ----
if (!dir.exists("input_files"))  dir.create("input_files")
if (!dir.exists("output_files")) dir.create("output_files")

# ---- Auto-copy from Metadata_Merge if needed ----
# Default: merged_combined_post_QC.csv (has separate Run + Bay, all biomarkers)
# Fallback: merged_standard_post_QC.csv (combined RUN column, standard biomarkers only)
input_candidates <- list.files("input_files",
                               pattern = "(merged_combined|merged_standard)_post_QC\\.csv$",
                               full.names = TRUE)

if (length(input_candidates) == 0) {
  search_dirs <- list.files(
    "../../Metadata_Merge",
    pattern = "^output_files",
    full.names = TRUE,
    include.dirs = TRUE
  )
  search_dirs <- search_dirs[order(file.mtime(search_dirs), decreasing = TRUE)]

  for (d in search_dirs) {
    src <- file.path(d, "merged_combined_post_QC.csv")
    if (file.exists(src)) {
      file.copy(src, "input_files/")
      message("Copied merged_combined from: ", src)
      break
    }
    src <- file.path(d, "merged_standard_post_QC.csv")
    if (file.exists(src)) {
      file.copy(src, "input_files/")
      message("Copied merged_standard from: ", src)
      break
    }
  }
  input_candidates <- list.files("input_files",
                                 pattern = "(merged_combined|merged_standard)_post_QC\\.csv$",
                                 full.names = TRUE)
}

if (length(input_candidates) == 0) {
  stop("No input file found. Place merged_combined_post_QC.csv in input_files/ or ensure Metadata_Merge outputs exist.")
}

# Prefer merged_combined if both present
if (any(grepl("merged_combined", input_candidates))) {
  input_file <- input_candidates[grepl("merged_combined", input_candidates)][1]
} else {
  input_file <- input_candidates[1]
}
message("Using: ", basename(input_file))
```

---

# B. Data Loading and Formatting

```{r load-data, message=FALSE, warning=FALSE}
raw <- read_csv(input_file, show_col_types = FALSE)

# ---- Get Run_ID and Bay ----
# merged_combined has separate Run + Bay columns -> rename Run -> Run_ID.
# merged_standard only has the combined RUN column -> parse it.
# Also handles any future file that already uses Run_ID directly.
if ("Run_ID" %in% names(raw) && "Bay" %in% names(raw)) {
  df <- raw
  message("Using existing Run_ID and Bay columns.")
} else if ("Run" %in% names(raw) && "Bay" %in% names(raw)) {
  df <- raw %>% rename(Run_ID = Run)
  message("Using existing Run and Bay columns (renamed Run -> Run_ID).")
} else {
  # Parse from combined RUN column (format: "20250711-1613_Bay1")
  df <- raw %>%
    mutate(
      Run_ID = str_extract(RUN, "^[^_]+"),     # e.g. "20250711-1613"
      Bay    = str_extract(RUN, "Bay[0-9]+")   # e.g. "Bay1"
    )
  message("Parsed Run_ID and Bay from RUN column.")
}

# ---- Identify biomarker columns ----
# All numeric columns after the metadata block
meta_cols <- c("RUN", "SAMPLE_ALIQUOT", "SAMPLE", "Record_ID", "newsample",
               "Group", "Ethnicity", "Race", "sex", "age_at_subject",
               "CDX", "Case_Control", "AOO", "Years_Onset", "APOE.geno",
               "height_inches", "weight_lb", "BMI", "Site",
               "Country/State", "country_of_birth", "metadata_source",
               "newsamp", "Group_Race_Ethnicity", "Run_ID", "Bay", "Run",
               "___1", "APOE")

biomarker_cols <- names(df)[!(names(df) %in% meta_cols) & sapply(df, is.numeric)]

# Exclude mCherry if present (internal control, not a biomarker)
biomarker_cols <- biomarker_cols[biomarker_cols != "mCherry"]

# ---- Study samples only (exclude controls/reps by filtering on SAMPLE_ALIQUOT) ----
# Controls have prefixes NC, SC, IPC; HIHG reps are identified separately below
df_samples <- df %>%
  filter(!str_detect(SAMPLE_ALIQUOT, "^(NC|SC|IPC)"))

# ---- Identify HIHG replicates (SAMPLE appears >1 time) ----
rep_counts <- df_samples %>%
  count(SAMPLE, name = "n_runs") %>%
  filter(n_runs > 1)

df_reps    <- df_samples %>% filter(SAMPLE %in% rep_counts$SAMPLE)
df_unique  <- df_samples %>% filter(!SAMPLE %in% rep_counts$SAMPLE)

cat("Total samples (excl controls):", nrow(df_samples), "\n")
cat("Unique biological samples:    ", nrow(df_unique), "\n")
cat("Replicated biological samples:", n_distinct(df_reps$SAMPLE), "\n")
cat("Replicate observations:       ", nrow(df_reps), "\n")
cat("Runs:                         ", n_distinct(df_samples$Run_ID), "\n")
cat("Bays observed:                ", paste(sort(unique(df_samples$Bay)), collapse=", "), "\n")
cat("Biomarkers:                   ", length(biomarker_cols), "\n")
```

---

# C. Plate Composition

This section examines whether demographic and clinical variables are balanced across
Runs and Bays. Confounding between batch and biology is the primary concern.

## C.1 Sample Counts per Run x Bay

```{r composition-counts, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
count_tbl <- df_samples %>%
  count(Run_ID, Bay) %>%
  pivot_wider(names_from = Bay, values_from = n, values_fill = 0) %>%
  arrange(Run_ID)

kable(count_tbl, caption = "Sample counts per Run x Bay (study samples only)") %>%
  kable_styling(full_width = FALSE)
```

## C.2 Race Distribution per Run

```{r composition-race, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
race_tbl <- df_samples %>%
  count(Run_ID, Race) %>%
  group_by(Run_ID) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

ggplot(race_tbl, aes(x = Run_ID, y = pct, fill = Race)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  labs(title = "Race composition per Run",
       x = "Run", y = "% of samples", fill = "Race")
```

## C.3 Race Distribution per Bay (across all Runs)

```{r composition-race-bay, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
bay_race <- df_samples %>%
  count(Bay, Race) %>%
  group_by(Bay) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

ggplot(bay_race, aes(x = Bay, y = pct, fill = Race)) +
  geom_col() +
  theme_bw() +
  labs(title = "Race composition per Bay (pooled across Runs)",
       x = "Bay", y = "% of samples", fill = "Race")
```

## C.4 Sex and Age Distribution per Run

```{r composition-sex-age, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
p_sex <- df_samples %>%
  count(Run_ID, sex) %>%
  group_by(Run_ID) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ggplot(aes(x = Run_ID, y = pct, fill = sex)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  labs(title = "Sex composition per Run", x = NULL, y = "% samples", fill = "Sex")

p_age <- df_samples %>%
  ggplot(aes(x = Run_ID, y = age_at_subject)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.size = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  labs(title = "Age distribution per Run", x = "Run", y = "Age")

p_sex / p_age
```

## C.5 Site Distribution per Run

```{r composition-site, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
site_tbl <- df_samples %>%
  count(Run_ID, Site) %>%
  group_by(Run_ID) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

ggplot(site_tbl, aes(x = Run_ID, y = pct, fill = Site)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.text = element_text(size = 7)) +
  labs(title = "Site composition per Run",
       x = "Run", y = "% of samples", fill = "Site")
```

## C.6 Diagnosis Distribution per Run

```{r composition-dx, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
dx_tbl <- df_samples %>%
  mutate(Dx = coalesce(CDX, Case_Control)) %>%
  count(Run_ID, Dx) %>%
  group_by(Run_ID) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

ggplot(dx_tbl, aes(x = Run_ID, y = pct, fill = Dx)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  labs(title = "Diagnosis composition per Run",
       x = "Run", y = "% of samples", fill = "Diagnosis")
```

## C.7 Confounding: Correlation of Batch with Covariates

Cramer's V for categorical variables, sqrt(eta-squared) for age. Values > 0.3 indicate
meaningful confounding between batch variable and covariate.

```{r confounding, message=FALSE, warning=FALSE}
cramers_v <- function(x, y) {
  tbl <- table(x, y)
  chi <- suppressWarnings(chisq.test(tbl))
  n   <- sum(tbl)
  k   <- min(nrow(tbl), ncol(tbl))
  v   <- sqrt(chi$statistic / (n * (k - 1)))
  as.numeric(v)
}

eta_sq <- function(group, value) {
  fit <- aov(value ~ group)
  ss  <- summary(fit)[[1]][["Sum Sq"]]
  ss[1] / sum(ss)
}

cov_list <- list(
  Race  = df_samples$Race,
  sex   = df_samples$sex,
  Site  = df_samples$Site,
  Dx    = coalesce(df_samples$CDX, df_samples$Case_Control)
)

batch_vars <- list(
  Run_ID = df_samples$Run_ID,
  Bay    = df_samples$Bay
)

conf_results <- map_dfr(names(batch_vars), function(bv) {
  bvec <- batch_vars[[bv]]
  rows <- map_dfr(names(cov_list), function(cv) {
    cvec <- cov_list[[cv]]
    ok   <- !is.na(bvec) & !is.na(cvec)
    tibble(Batch_Var = bv, Covariate = cv,
           CramersV  = cramers_v(bvec[ok], cvec[ok]))
  })
  age_ok <- !is.na(bvec) & !is.na(df_samples$age_at_subject)
  bind_rows(rows, tibble(
    Batch_Var = bv, Covariate = "age",
    CramersV  = sqrt(eta_sq(bvec[age_ok], df_samples$age_at_subject[age_ok]))
  ))
}) %>%
  mutate(Concern = case_when(
    CramersV > 0.5 ~ "HIGH",
    CramersV > 0.3 ~ "moderate",
    TRUE           ~ "low"
  ))

kable(conf_results %>% arrange(desc(CramersV)), digits = 3,
      caption = "Confounding between batch variables and biological covariates") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(which(conf_results %>% arrange(desc(CramersV)) %>% pull(Concern) == "HIGH"),
           background = "#ffd6d6") %>%
  row_spec(which(conf_results %>% arrange(desc(CramersV)) %>% pull(Concern) == "moderate"),
           background = "#fff3cd")

write_csv(conf_results, "output_files/batch_covariate_confounding.csv")
```

---

# D. Variance Partitioning

Decomposes variance per protein into:
- Fixed effects: Race, sex, age, Site, Diagnosis
- Random effects: Run (between-run technical variation), Bay nested in Run (sequential/drift)
- Residual: unexplained

Model: `~ Race + sex + age_at_subject + Site + Dx + (1|Run_ID) + (1|Run_ID:Bay)`

Proteins with high Run or Bay variance fractions are candidates for batch correction.
Proteins where Run variance ~ Race variance indicate confounding that correction cannot resolve.

```{r vp-prep, message=FALSE, warning=FALSE}
# Use unique samples only (no replicates) to avoid inflating residual
vp_df <- df_unique %>%
  filter(!is.na(Race), !is.na(sex), !is.na(age_at_subject),
         !is.na(Run_ID), !is.na(Bay)) %>%
  mutate(Dx = coalesce(CDX, Case_Control))

# Remove Runs with fewer than VP_MIN_SAMPLES
run_counts <- count(vp_df, Run_ID)
vp_df <- vp_df %>%
  filter(Run_ID %in% run_counts$Run_ID[run_counts$n >= VP_MIN_SAMPLES])

cat("Samples used for VP:", nrow(vp_df), "\n")
cat("Runs included:      ", n_distinct(vp_df$Run_ID), "\n")

# Biomarker matrix: proteins x samples (variancePartition convention)
expr_mat <- vp_df %>%
  select(all_of(biomarker_cols)) %>%
  as.matrix() %>%
  t()

info_df <- vp_df %>%
  select(Run_ID, Bay, Race, sex, age_at_subject, Site, Dx) %>%
  mutate(across(c(Race, sex, Site, Dx, Run_ID, Bay), as.factor)) %>%
  as.data.frame()
```

```{r vp-run, message=FALSE, warning=FALSE, fig.width=12, fig.height=6}
register(MulticoreParam(VP_CORES))

form <- ~ Race + sex + age_at_subject + Site + Dx + (1|Run_ID) + (1|Run_ID:Bay)

vp_fit <- fitExtractVarPartModel(expr_mat, form, info_df)

plotVarPart(vp_fit, main = "Variance partitioning: biological vs. batch effects")
```

```{r vp-summary, message=FALSE, warning=FALSE}
vp_df_out <- as.data.frame(vp_fit) %>%
  rownames_to_column("Biomarker") %>%
  arrange(desc(`Run_ID`))

# Threshold summary
thresh_tbl <- map_dfr(c(5, 10, 20), function(t) {
  tibble(
    Threshold    = paste0(">", t, "% Run variance"),
    N_proteins   = sum(vp_df_out$`Run_ID` * 100 > t),
    Pct_proteins = round(100 * mean(vp_df_out$`Run_ID` * 100 > t), 1)
  )
})
kable(thresh_tbl, caption = "Proteins with substantial Run-level variance") %>%
  kable_styling(full_width = FALSE)

kable(vp_df_out %>% slice_max(`Run_ID`, n = 20) %>%
        mutate(across(where(is.numeric), ~round(.x * 100, 1))),
      caption = "Top 20 proteins by % Run variance") %>%
  kable_styling(full_width = FALSE) %>% scroll_box(width = "100%")

kable(vp_df_out %>% arrange(desc(`Run_ID:Bay`)) %>% slice_head(n = 20) %>%
        mutate(across(where(is.numeric), ~round(.x * 100, 1))),
      caption = "Top 20 proteins by % Bay (within-Run) variance") %>%
  kable_styling(full_width = FALSE) %>% scroll_box(width = "100%")

write_csv(vp_df_out, "output_files/variance_partition_results.csv")
```

---

# E. Replicate-Based ICC Analysis

Uses HIHG biological replicates (same sample run on multiple plates) to estimate:
- **Within-Run ICC**: reproducibility across Bays within the same Run
- **Cross-Run ICC**: reproducibility across different Runs (total technical variance)

ICC > 0.9 = excellent; 0.75-0.9 = good; < 0.75 = concerning.

```{r icc-prep, message=FALSE, warning=FALSE}
library(irr)

rep_obs <- df_reps %>%
  select(SAMPLE, SAMPLE_ALIQUOT, Run_ID, Bay, all_of(biomarker_cols)) %>%
  arrange(SAMPLE, Run_ID, Bay) %>%
  group_by(SAMPLE) %>%
  mutate(n_runs_for_sample = n_distinct(Run_ID),
         any_cross_run     = n_runs_for_sample > 1) %>%
  ungroup()

within_run_reps <- rep_obs %>%
  group_by(SAMPLE, Run_ID) %>% filter(n() >= 2) %>% ungroup()

cross_run_reps <- rep_obs %>% filter(any_cross_run)

cat("Replicated samples:            ", n_distinct(rep_obs$SAMPLE), "\n")
cat("  with within-Run replicates:  ", n_distinct(within_run_reps$SAMPLE), "\n")
cat("  with cross-Run replicates:   ", n_distinct(cross_run_reps$SAMPLE), "\n")
```

```{r icc-calc, message=FALSE, warning=FALSE}
calc_icc_per_protein <- function(rep_df, biomarkers, min_pairs = ICC_MIN_REPS) {
  map_dfr(biomarkers, function(bm) {
    bm_wide <- rep_df %>%
      select(SAMPLE, Run_ID, Bay, all_of(bm)) %>%
      arrange(SAMPLE, Run_ID, Bay) %>%
      group_by(SAMPLE) %>%
      mutate(obs_num = row_number()) %>%
      ungroup() %>%
      pivot_wider(id_cols = SAMPLE, names_from = obs_num,
                  values_from = all_of(bm), names_prefix = "rep_") %>%
      select(-SAMPLE) %>%
      filter(rowSums(!is.na(.)) >= min_pairs)

    if (nrow(bm_wide) < min_pairs)
      return(tibble(Biomarker = bm, ICC = NA_real_, lower = NA_real_,
                    upper = NA_real_, n_pairs = nrow(bm_wide)))

    result <- tryCatch(
      icc(bm_wide, model = "twoway", type = "agreement", unit = "average"),
      error = function(e) NULL
    )
    if (is.null(result))
      return(tibble(Biomarker = bm, ICC = NA_real_, lower = NA_real_,
                    upper = NA_real_, n_pairs = nrow(bm_wide)))

    tibble(Biomarker = bm, ICC = result$value,
           lower = result$lbound, upper = result$ubound, n_pairs = nrow(bm_wide))
  })
}

icc_within <- calc_icc_per_protein(within_run_reps, biomarker_cols)
icc_cross  <- calc_icc_per_protein(cross_run_reps,  biomarker_cols)

icc_combined <- icc_within %>%
  rename(ICC_within = ICC, lower_within = lower, upper_within = upper, n_within = n_pairs) %>%
  left_join(icc_cross %>%
              rename(ICC_cross = ICC, lower_cross = lower, upper_cross = upper, n_cross = n_pairs),
            by = "Biomarker") %>%
  mutate(
    Category_within = case_when(
      ICC_within >= 0.9  ~ "Excellent", ICC_within >= 0.75 ~ "Good",
      ICC_within >= 0.5  ~ "Moderate",  !is.na(ICC_within)  ~ "Poor",
      TRUE ~ "Insufficient data"),
    Category_cross = case_when(
      ICC_cross >= 0.9   ~ "Excellent", ICC_cross >= 0.75  ~ "Good",
      ICC_cross >= 0.5   ~ "Moderate",  !is.na(ICC_cross)   ~ "Poor",
      TRUE ~ "Insufficient data")
  )
```

```{r icc-plots, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
icc_combined %>%
  select(Biomarker, ICC_within, ICC_cross) %>%
  pivot_longer(-Biomarker, names_to = "Level", values_to = "ICC") %>%
  mutate(Level = recode(Level,
    ICC_within = "Within-Run (across Bays)", ICC_cross = "Cross-Run")) %>%
  ggplot(aes(x = ICC, fill = Level)) +
  geom_histogram(binwidth = 0.05, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = c(0.75, 0.9), linetype = "dashed", color = "red") +
  facet_wrap(~ Level) +
  theme_bw() +
  labs(title = "ICC distribution by replicate type",
       subtitle = "Red lines: 0.75 (Good) and 0.90 (Excellent) thresholds",
       x = "ICC", y = "Count of proteins") +
  theme(legend.position = "none")

icc_combined %>%
  summarise(
    `Within-Run: Excellent (>=0.9)` = sum(Category_within == "Excellent", na.rm=T),
    `Within-Run: Good (0.75-0.9)`   = sum(Category_within == "Good", na.rm=T),
    `Within-Run: Moderate/Poor`     = sum(Category_within %in% c("Moderate","Poor"), na.rm=T),
    `Cross-Run: Excellent (>=0.9)`  = sum(Category_cross  == "Excellent", na.rm=T),
    `Cross-Run: Good (0.75-0.9)`    = sum(Category_cross  == "Good", na.rm=T),
    `Cross-Run: Moderate/Poor`      = sum(Category_cross  %in% c("Moderate","Poor"), na.rm=T)
  ) %>%
  pivot_longer(everything(), names_to = "Category", values_to = "N proteins") %>%
  kable(caption = "ICC summary across proteins") %>%
  kable_styling(full_width = FALSE)

poor_icc <- icc_combined %>%
  filter(Category_cross %in% c("Moderate", "Poor") | ICC_cross < 0.75)

if (nrow(poor_icc) > 0) {
  kable(poor_icc %>% arrange(ICC_cross), digits = 3,
        caption = "Proteins with ICC_cross < 0.75") %>%
    kable_styling(full_width = FALSE)
}

write_csv(icc_combined, "output_files/icc_results.csv")
```

---

# F. Formal Per-Protein Batch Effect Test

Linear model per protein: `NPQ ~ Run_ID + Bay + Race + sex + age_at_subject + Site + Dx`

F-test (Type III) for Run_ID and Bay terms after adjusting for biological covariates.
FDR correction across proteins within each term.

**Limitation:** Treats Run_ID as fixed — low power with many Runs. Interpret alongside variance partitioning.

```{r formal-test, message=FALSE, warning=FALSE}
library(car)

test_df <- df_unique %>%
  filter(!is.na(Race), !is.na(sex), !is.na(age_at_subject),
         !is.na(Run_ID), !is.na(Bay)) %>%
  mutate(Dx     = coalesce(CDX, Case_Control),
         Run_fac = as.factor(Run_ID)) %>%
  filter(!is.na(Dx))

batch_test_results <- map_dfr(biomarker_cols, function(bm) {
  y  <- test_df[[bm]]
  ok <- complete.cases(test_df[, c("Race","sex","age_at_subject","Dx")]) & !is.na(y)
  if (sum(ok) < 20)
    return(tibble(Biomarker = bm, p_Run = NA_real_, p_Bay = NA_real_))

  fit <- tryCatch(
    lm(y[ok] ~ Run_fac + Bay + Race + sex + age_at_subject + Site + Dx,
       data = test_df[ok, ]),
    error = function(e) NULL)
  if (is.null(fit)) return(tibble(Biomarker = bm, p_Run = NA_real_, p_Bay = NA_real_))

  aov_fit <- tryCatch(Anova(fit, type = "III"), error = function(e) NULL)
  if (is.null(aov_fit)) return(tibble(Biomarker = bm, p_Run = NA_real_, p_Bay = NA_real_))

  aov_df <- as.data.frame(aov_fit)
  tibble(
    Biomarker = bm,
    p_Run = aov_df["Run_fac", "Pr(>F)"],
    p_Bay = if ("Bay" %in% rownames(aov_df)) aov_df["Bay", "Pr(>F)"] else NA_real_
  )
})

batch_test_results <- batch_test_results %>%
  mutate(FDR_Run = p.adjust(p_Run, "BH"),
         FDR_Bay = p.adjust(p_Bay, "BH"),
         Sig_Run = FDR_Run < FDR_THRESHOLD,
         Sig_Bay = FDR_Bay < FDR_THRESHOLD) %>%
  arrange(FDR_Run)

cat("Proteins with significant Run effect (FDR <", FDR_THRESHOLD, "):",
    sum(batch_test_results$Sig_Run, na.rm=TRUE), "\n")
cat("Proteins with significant Bay effect (FDR <", FDR_THRESHOLD, "):",
    sum(batch_test_results$Sig_Bay, na.rm=TRUE), "\n")

batch_test_results %>%
  filter(Sig_Run | Sig_Bay) %>% head(30) %>%
  kable(digits = 4, caption = "Proteins with significant Run or Bay effect (FDR corrected)") %>%
  kable_styling(full_width = FALSE)

write_csv(batch_test_results, "output_files/batch_effect_test_results.csv")
```

---

# G. Summary and Decision

```{r decision, message=FALSE, warning=FALSE}
n_sig_run      <- sum(batch_test_results$Sig_Run, na.rm=TRUE)
n_sig_bay      <- sum(batch_test_results$Sig_Bay, na.rm=TRUE)
n_proteins     <- length(biomarker_cols)
median_run_var <- median(vp_df_out$`Run_ID`, na.rm=TRUE) * 100
median_bay_var <- median(vp_df_out$`Run_ID:Bay`, na.rm=TRUE) * 100
max_run_var    <- max(vp_df_out$`Run_ID`, na.rm=TRUE) * 100
high_conf      <- conf_results %>% filter(Concern == "HIGH")
n_poor_icc     <- nrow(poor_icc)

cat("=== BATCH EFFECT SUMMARY ===\n\n")
cat(sprintf("Formal test:        %d/%d proteins significant Run effect (FDR < %.2f)\n",
            n_sig_run, n_proteins, FDR_THRESHOLD))
cat(sprintf("                    %d/%d proteins significant Bay effect\n\n",
            n_sig_bay, n_proteins))
cat(sprintf("Variance partition: median Run = %.1f%%, max Run = %.1f%%, median Bay = %.1f%%\n\n",
            median_run_var, max_run_var, median_bay_var))
cat(sprintf("ICC:                %d proteins with cross-Run ICC < 0.75\n\n", n_poor_icc))

if (nrow(high_conf) > 0) {
  cat("HIGH confounding (Cramers V > 0.5):\n")
  print(high_conf %>% select(Batch_Var, Covariate, CramersV))
  cat("\n")
}

cat("=== RECOMMENDATION ===\n\n")
if (n_sig_run == 0 && median_run_var < 5) {
  cat("Batch effects MINIMAL.\n")
  cat("-> Include Run_ID as a fixed covariate in downstream models.\n")
} else if (median_run_var < 10 && nrow(high_conf) == 0) {
  cat("Batch effects MODERATE, not strongly confounded with biology.\n")
  cat("-> Include Run_ID as random effect (lme4) in downstream models.\n")
  cat("-> OR apply ComBat (sva) for methods that cannot use random effects.\n")
} else if (nrow(high_conf) > 0) {
  cat("CAUTION: High confounding detected. ComBat may remove biological signal.\n")
  cat("-> For association testing: use Run_ID as random effect (lme4) -- preferred.\n")
  cat("-> For PCA/clustering: ComBat with bio covariates as protected, interpret with caution.\n")
  cat("-> Run sensitivity analysis with and without correction.\n")
} else {
  cat("Batch effects SUBSTANTIAL.\n")
  cat("-> Apply ComBat (sva::ComBat) with Race + sex + age as protected covariates.\n")
  cat("-> Validate correction using HIHG replicate ICC post-correction.\n")
}

cat("\n=== OUTPUT FILES ===\n")
cat("batch_covariate_confounding.csv  - Cramers V between batch and covariates\n")
cat("variance_partition_results.csv   - Per-protein variance fractions\n")
cat("icc_results.csv                  - Per-protein ICC within-Run and cross-Run\n")
cat("batch_effect_test_results.csv    - Per-protein p/FDR for Run and Bay terms\n")
```

---

# Dataset Notes

```{r notes, echo=FALSE, results='asis'}
notes_file <- "notes.md"
if (file.exists(notes_file)) {
  cat(paste(readLines(notes_file, warn = FALSE), collapse = "\n"))
} else {
  cat("*No notes file found. Create `notes.md` for dataset-specific observations.*\n")
}
```
