---
title: "MCI Progression Analysis: Biomarker Classification and Clustering"
author: "ERM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

# A. Overview

This report performs end-to-end analysis of biomarker data to characterize MCI (Mild Cognitive Impairment) progression risk. The pipeline includes:

1. **Covariate exploration** -- inspect distributions of CDX, Group, Race, Ethnicity
2. **Covariate grouping** -- configure and apply diagnosis and ancestry groupings
3. **Preprocessing** -- APOE4 status, missing data filtering, minimum sample filtering
4. **Composite scores** -- ATN framework, brain-derived enrichment, inflammation PCs
5. **Elastic net classification** -- NCI vs AD by ancestry group (with covariates)
6. **MCI risk prediction** -- apply NCI/AD model to MCI samples
7. **Supervised MCI+AD vs NCI** -- feature selection for cognitive impairment
8. **MCI clustering** -- unsupervised subtyping of MCI samples with NCI/AD overlay

**To modify groupings:** Edit the configs in Section B.2 and re-knit.

## A.1 Input Data

**Data Source:** Post-QC merged biomarker + metadata dataset.

## A.2 Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(glmnet)
library(pROC)
library(cluster)
library(factoextra)
library(ggplot2)
library(patchwork)
library(pheatmap)
library(knitr)

# Source modules
script_dir <- here::here()
source(file.path(script_dir, "00_data_loading.R"))
source(file.path(script_dir, "covariate_explorer.R"))
source(file.path(script_dir, "01_preprocessing.R"))
source(file.path(script_dir, "02_composite_scores.R"))
source(file.path(script_dir, "03_elastic_net.R"))
source(file.path(script_dir, "04_clustering.R"))
source(file.path(script_dir, "05_visualization.R"))

# Create output folder
output_dir <- file.path(script_dir, "output_files")
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

set.seed(123)
```

---

# B. Covariate Exploration and Grouping

## B.1 Load Raw Data

```{r load-raw-data, message=FALSE, warning=FALSE}
data_path <- file.path(script_dir, "input_files", "filtered_combined_post_QC_mod.csv")
dat_list <- load_alamar_data(data_path)
dat_raw <- dat_list$data

cat("**Total samples:**", nrow(dat_raw), "\n\n")
cat("**Total biomarkers:**", dat_list$n_biomarkers, "\n\n")
```

### B.1.1 CDX Distribution

```{r explore-cdx, results='asis'}
cdx_counts <- dat_raw %>%
  count(CDX, name = "n") %>%
  arrange(desc(n)) %>%
  mutate(pct = round(n / sum(n) * 100, 1))
kable(cdx_counts, caption = "CDX (Diagnosis) Distribution")
```

### B.1.2 Group, Race, Ethnicity Distributions

```{r explore-group-race-eth, results='asis'}
group_counts <- dat_raw %>%
  count(Group, name = "n") %>%
  arrange(desc(n)) %>%
  mutate(pct = round(n / sum(n) * 100, 1))
kable(group_counts, caption = "Group Distribution")

race_counts <- dat_raw %>%
  count(Race, name = "n") %>%
  arrange(desc(n)) %>%
  mutate(pct = round(n / sum(n) * 100, 1))
kable(race_counts, caption = "Race Distribution")

eth_counts <- dat_raw %>%
  count(Ethnicity, name = "n") %>%
  arrange(desc(n)) %>%
  mutate(pct = round(n / sum(n) * 100, 1))
kable(eth_counts, caption = "Ethnicity Distribution")
```

### B.1.3 Cross-Tabulations

```{r cross-tabs, results='asis'}
ct_cdx_group <- dat_raw %>%
  count(CDX, Group) %>%
  pivot_wider(names_from = Group, values_from = n, values_fill = 0) %>%
  mutate(Total = rowSums(across(where(is.numeric)))) %>%
  arrange(desc(Total))
kable(ct_cdx_group, caption = "CDX x Group")

ct_group_eth <- dat_raw %>%
  count(Group, Ethnicity) %>%
  pivot_wider(names_from = Ethnicity, values_from = n, values_fill = 0) %>%
  mutate(Total = rowSums(across(where(is.numeric))))
kable(ct_group_eth, caption = "Group x Ethnicity")

ct_group_race <- dat_raw %>%
  count(Group, Race) %>%
  pivot_wider(names_from = Race, values_from = n, values_fill = 0) %>%
  mutate(Total = rowSums(across(where(is.numeric))))
kable(ct_group_race, caption = "Group x Race")
```

### B.1.4 All Group x Race x Ethnicity Combinations

```{r all-combos, results='asis'}
combos <- dat_raw %>%
  count(Group, Race, Ethnicity, name = "n") %>%
  arrange(desc(n)) %>%
  mutate(pct = round(n / sum(n) * 100, 1))
kable(combos, caption = "All Group x Race x Ethnicity Combinations")
```

### B.1.5 CDX x Group_Race_Ethnicity

```{r cdx-gre, results='asis'}
ct_cdx_gre <- dat_raw %>%
  count(CDX, Group_Race_Ethnicity) %>%
  pivot_wider(names_from = Group_Race_Ethnicity, values_from = n, values_fill = 0) %>%
  mutate(Total = rowSums(across(where(is.numeric)))) %>%
  arrange(desc(Total))
kable(ct_cdx_gre, caption = "CDX x Group_Race_Ethnicity")
```

---

## B.2 Configure Groupings

Edit the configs below to change how diagnoses and ancestry groups are defined, then re-knit.

```{r configure-groupings}
# --- Diagnosis grouping ---
# keep:     levels to retain as-is in CDX_collapsed
# collapse: named list where name = new label, value = source levels to merge
# exclude:  levels to drop entirely (rows removed)
dx_config <- list(
  source_col = "CDX",
  target_col = "CDX_collapsed",
  keep = c("NCI", "MCI", "AD"),
  collapse = list(
    "Dementia_Other" = c("Dementia Undetermined", "Dementia Vascular", "Dementia Other")
  ),
  exclude = c("0", "Other", "Insufficient Data")
)

# --- Ancestry grouping ---
# rules:  evaluated in order, first match wins
#         each rule is a list of column=value conditions (ALL must match)
# other_label:   label for unmatched rows
# exclude_other: if TRUE, drop unmatched rows
ancestry_config <- list(
  target_col = "Ancestry",
  rules = list(
    "HI_WH" = list(Group = "HI", Race = "WH"),
    "HI_MU" = list(Group = "HI", Race = "MU"),
    "HI_MU" = list(Group = "HI", Race = "BL WH"),
    "HI_BL" = list(Group = "HI", Race = "BL"),
    "HI_BL" = list(Group = "AA", Ethnicity = "HI"),
    "AA"    = list(Group = "AA", Ethnicity = "NH", Race = "BL"),
    "AFDC"  = list(Group = "AFDC")
  ),
  other_label = "Other",
  exclude_other = TRUE
)
```

---

## B.3 Diagnosis Grouping Decisions

```{r dx-decisions, results='asis'}
dx_summary <- summarize_category_decisions(dat_raw, dx_config)

cat("**Before:**", dx_summary$total_before, "samples\n\n")
cat("**After:**", dx_summary$total_after, "samples\n\n")

kable(dx_summary$all,
      caption = "Diagnosis Grouping Decisions: What happened to each CDX level",
      col.names = c("Original CDX", "N", "%", "Action", "Maps To"))
```

### Excluded CDX Levels

```{r dx-excluded, results='asis'}
if (nrow(dx_summary$excluded) > 0) {
  kable(dx_summary$excluded %>% select(1:3),
        caption = "Samples REMOVED by CDX exclusion",
        col.names = c("CDX", "N", "%"))
  cat(sprintf("\n**Total excluded:** %d samples\n\n", sum(dx_summary$excluded$n)))
} else {
  cat("No CDX levels were excluded.\n\n")
}
```

### Collapsed CDX Levels

```{r dx-collapsed, results='asis'}
if (nrow(dx_summary$collapsed) > 0) {
  kable(dx_summary$collapsed %>% select(1:3, maps_to),
        caption = "CDX levels COLLAPSED into new categories",
        col.names = c("Original CDX", "N", "%", "Collapsed To"))
} else {
  cat("No CDX levels were collapsed.\n\n")
}
```

---

## B.4 Ancestry Grouping Decisions

```{r ancestry-decisions, results='asis', message=FALSE}
# Use data after CDX filtering for ancestry decisions
dat_after_dx <- apply_category_config(dat_raw, dx_config)
anc_summary <- summarize_grouping_decisions(dat_after_dx, ancestry_config)

cat("**Total matched:**", anc_summary$total_matched, "samples\n\n")
cat("**Total unmatched (excluded):**", anc_summary$total_unmatched, "samples\n\n")

kable(anc_summary$group_summary,
      caption = "Ancestry Group Assignments",
      col.names = c("Ancestry", "N", "%"))
```

### Detailed Rule Matches

```{r ancestry-matches, results='asis'}
kable(anc_summary$rule_matches,
      caption = "How source columns mapped to Ancestry groups")
```

### Unmatched Rows (Excluded)

```{r ancestry-unmatched, results='asis'}
if (nrow(anc_summary$unmatched) > 0) {
  kable(anc_summary$unmatched,
        caption = "Rows excluded: matched no ancestry rule")
} else {
  cat("All rows matched an ancestry rule.\n\n")
}
```

---

## B.5 Apply Groupings

```{r apply-groupings, message=FALSE}
dat_processed <- dat_raw
dat_processed <- apply_category_config(dat_processed, dx_config)
dat_processed <- apply_grouping_config(dat_processed, ancestry_config)
```

## B.6 Validate Final Groups

```{r validate-groups, results='asis', message=FALSE}
final_ct <- dat_processed %>%
  count(CDX_collapsed, Ancestry) %>%
  pivot_wider(names_from = Ancestry, values_from = n, values_fill = 0) %>%
  mutate(Total = rowSums(across(where(is.numeric)))) %>%
  arrange(desc(Total))
kable(final_ct, caption = "Final CDX_collapsed x Ancestry Counts")

cat(sprintf("\n**Total samples after grouping:** %d\n\n", nrow(dat_processed)))
```

---

# C. Preprocessing

```{r preprocessing, message=FALSE}
dat_list$data <- dat_processed

dat_list <- dat_list %>%
  create_apoe4_status() %>%
  filter_missing_biomarkers(max_missing_pct = 0.2) %>%
  filter_min_samples(min_n = 5)

# Save preprocessed data
saveRDS(dat_list, file.path(output_dir, "preprocessed_data.rds"))
```

```{r preproc-summary, results='asis'}
cat(sprintf("**Final dataset:** %d samples x %d biomarkers\n\n",
            nrow(dat_list$data), length(dat_list$biomarker_cols)))

apoe_ct <- dat_list$data %>%
  count(APOE4_carrier, name = "n") %>%
  mutate(pct = round(n / sum(n) * 100, 1))
kable(apoe_ct, caption = "APOE4 Carrier Status")

final_groups <- dat_list$data %>%
  count(CDX_collapsed, Ancestry) %>%
  pivot_wider(names_from = Ancestry, values_from = n, values_fill = 0) %>%
  mutate(Total = rowSums(across(where(is.numeric))))
kable(final_groups, caption = "Final Sample Counts After Preprocessing")
```

---

# D. Composite Scores

This section creates multi-biomarker composite scores that summarize key biological pathways. All individual biomarkers are NPQ (log2-transformed) values from the Alamar platform. Arithmetic on the log2 scale has natural biological meaning: differences = log2 fold-changes, so `A_42 - A_40` = log2(Abeta42/Abeta40).

Individual-level composite score patterns and AD-likeness analysis are in D.9 and D.10 below.

```{r composite-scores, message=FALSE}
dat_list <- create_all_composite_scores(dat_list)
saveRDS(dat_list, file.path(output_dir, "data_with_composite_scores.rds"))
```

## D.1 Score Definitions

| Score | What it measures | Input markers | How it's computed | Interpretation |
|-------|-----------------|---------------|-------------------|----------------|
| **A_score** | Amyloid pathology | A_42, A_40 | A_42 - A_40 on log2 scale = log2(Abeta42/Abeta40) | **Lower = more amyloid pathology** (reduced Abeta42/40 is a hallmark of AD) |
| **T_score** | Tau pathology | BD_pTau_181/217/231, pTau_181/217/231 | Mean of z-scored phospho-tau markers | **Higher = more tau pathology** |
| **N_score** | Neurodegeneration | NEFL, NEFH, GFAP, NRGN, SNAP25 | Mean of z-scored neurodegeneration markers | **Higher = more neurodegeneration** |
| **BD_pTau217_ratio** | Brain-derived tau enrichment | BD_pTau_217 - pTau_217 | Log2 difference = log2(BD/peripheral fold-change) | **Higher = more brain-derived signal** relative to peripheral |
| **BD_pTau181_ratio** | Brain-derived tau enrichment | BD_pTau_181 - pTau_181 | Same as above for pTau181 | Same interpretation |
| **BD_pTau231_ratio** | Brain-derived tau enrichment | BD_pTau_231 - pTau_231 | Same as above for pTau231 | Same interpretation |
| **Proteinopathy_burden** | Multi-protein pathology | A_score, BD_pTau_217, pTDP43_409, Oligo_SNCA | Mean of z-scored markers | **Higher = more aggregate proteinopathy** |

**Note on z-scoring:** T_score and N_score z-score each marker before averaging because the individual log2 NPQ values have different dynamic ranges (e.g., NEFH median ~17.7 vs SNAP25 median ~13.1). Without z-scoring, markers with higher absolute values would dominate the composite.

## D.2 Score Summary Table

```{r composite-summary, results='asis'}
score_cols <- c("A_score", "T_score", "N_score",
                "BD_pTau217_ratio", "BD_pTau181_ratio", "BD_pTau231_ratio",
                "Proteinopathy_burden")
available_scores <- intersect(score_cols, names(dat_list$data))

if (length(available_scores) > 0) {
  score_summary <- dat_list$data %>%
    select(all_of(available_scores)) %>%
    summarize(across(everything(), list(
      n = ~sum(!is.na(.x)),
      mean = ~mean(.x, na.rm = TRUE),
      sd = ~sd(.x, na.rm = TRUE),
      median = ~median(.x, na.rm = TRUE),
      min = ~min(.x, na.rm = TRUE),
      max = ~max(.x, na.rm = TRUE),
      missing = ~sum(is.na(.x))
    ))) %>%
    pivot_longer(everything(),
                 names_to = c("score", ".value"),
                 names_pattern = "(.+)_(n|mean|sd|median|min|max|missing)") %>%
    mutate(across(where(is.numeric), ~round(.x, 3)))
  kable(score_summary, caption = "Composite Score Summary Statistics")
}
```

## D.3 ATN Scores by Diagnosis

These plots show how the ATN scores separate diagnostic groups. In an effective biomarker panel, AD should show lower A_score (more amyloid pathology) and higher T_score/N_score (more tau and neurodegeneration).

```{r atn-by-diagnosis, fig.width=12, fig.height=5, warning=FALSE}
atn_cols <- intersect(c("A_score", "T_score", "N_score"), names(dat_list$data))

if (length(atn_cols) > 0) {
  atn_long <- dat_list$data %>%
    filter(CDX_collapsed %in% c("NCI", "MCI", "AD")) %>%
    select(CDX_collapsed, all_of(atn_cols)) %>%
    pivot_longer(all_of(atn_cols), names_to = "Score", values_to = "Value") %>%
    filter(is.finite(Value))

  ggplot(atn_long, aes(x = CDX_collapsed, y = Value, fill = CDX_collapsed)) +
    geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
    facet_wrap(~Score, scales = "free_y", nrow = 1) +
    scale_fill_manual(values = c("NCI" = "#4DAF4A", "MCI" = "#FF7F00", "AD" = "#E41A1C")) +
    labs(title = "ATN Scores by Diagnosis",
         subtitle = "A_score: lower = more pathology | T/N_score: higher = more pathology",
         x = NULL, y = "Score Value") +
    theme_bw() +
    theme(legend.position = "none")
}
```

## D.4 Brain-Derived Enrichment by Diagnosis

BD ratios measure how much signal comes from brain-derived exosomes vs. peripheral sources. Higher values indicate greater brain-specific signal. These tend to be elevated in AD.

```{r bd-ratios-by-diagnosis, fig.width=12, fig.height=5, warning=FALSE}
bd_cols <- intersect(c("BD_pTau217_ratio", "BD_pTau181_ratio", "BD_pTau231_ratio"),
                     names(dat_list$data))

if (length(bd_cols) > 0) {
  bd_long <- dat_list$data %>%
    filter(CDX_collapsed %in% c("NCI", "MCI", "AD")) %>%
    select(CDX_collapsed, all_of(bd_cols)) %>%
    pivot_longer(all_of(bd_cols), names_to = "Ratio", values_to = "Value") %>%
    filter(is.finite(Value))

  ggplot(bd_long, aes(x = CDX_collapsed, y = Value, fill = CDX_collapsed)) +
    geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
    facet_wrap(~Ratio, scales = "free_y", nrow = 1) +
    scale_fill_manual(values = c("NCI" = "#4DAF4A", "MCI" = "#FF7F00", "AD" = "#E41A1C")) +
    labs(title = "Brain-Derived pTau Enrichment by Diagnosis",
         subtitle = "Higher = more brain-derived signal relative to peripheral",
         x = NULL, y = "BD - Peripheral (log2 scale)") +
    theme_bw() +
    theme(legend.position = "none")
}
```

## D.5 ATN Scores by Diagnosis and Ancestry

```{r atn-by-ancestry, fig.width=12, fig.height=8, warning=FALSE}
if (length(atn_cols) > 0) {
  atn_anc_long <- dat_list$data %>%
    filter(CDX_collapsed %in% c("NCI", "MCI", "AD")) %>%
    select(CDX_collapsed, Ancestry, all_of(atn_cols)) %>%
    pivot_longer(all_of(atn_cols), names_to = "Score", values_to = "Value") %>%
    filter(is.finite(Value))

  ggplot(atn_anc_long, aes(x = CDX_collapsed, y = Value, fill = CDX_collapsed)) +
    geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
    facet_grid(Score ~ Ancestry, scales = "free_y") +
    scale_fill_manual(values = c("NCI" = "#4DAF4A", "MCI" = "#FF7F00", "AD" = "#E41A1C")) +
    labs(title = "ATN Scores by Diagnosis and Ancestry",
         x = NULL, y = "Score Value") +
    theme_bw() +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## D.6 Proteinopathy Burden

The proteinopathy burden score combines z-scored amyloid, tau, TDP-43, and alpha-synuclein markers into a single multi-proteinopathy index. Higher values indicate greater overall protein pathology burden.

```{r proteinopathy-by-diagnosis, fig.width=8, fig.height=5, warning=FALSE}
if ("Proteinopathy_burden" %in% names(dat_list$data)) {
  proto_dat <- dat_list$data %>%
    filter(CDX_collapsed %in% c("NCI", "MCI", "AD"), is.finite(Proteinopathy_burden))

  ggplot(proto_dat, aes(x = CDX_collapsed, y = Proteinopathy_burden, fill = CDX_collapsed)) +
    geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
    geom_jitter(width = 0.15, alpha = 0.15, size = 0.5) +
    scale_fill_manual(values = c("NCI" = "#4DAF4A", "MCI" = "#FF7F00", "AD" = "#E41A1C")) +
    labs(title = "Multi-Proteinopathy Burden by Diagnosis",
         subtitle = "Mean of z-scored A_score, BD_pTau_217, pTDP43_409, Oligo_SNCA",
         x = NULL, y = "Proteinopathy Burden (z-score units)") +
    theme_bw() +
    theme(legend.position = "none")
}
```

## D.7 Score Correlation Matrix

```{r score-correlations, fig.width=8, fig.height=7, warning=FALSE}
all_score_cols <- intersect(c("A_score", "T_score", "N_score",
                              "BD_pTau217_ratio", "BD_pTau181_ratio", "BD_pTau231_ratio",
                              "Proteinopathy_burden",
                              "Inflammation_PC1", "Inflammation_PC2", "Inflammation_PC3"),
                            names(dat_list$data))

if (length(all_score_cols) >= 3) {
  score_mat <- dat_list$data %>%
    select(all_of(all_score_cols)) %>%
    filter(if_all(everything(), is.finite))

  cor_mat <- cor(score_mat, use = "pairwise.complete.obs")

  pheatmap(cor_mat,
           display_numbers = TRUE,
           number_format = "%.2f",
           fontsize_number = 8,
           main = "Composite Score Correlations",
           color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(50),
           breaks = seq(-1, 1, length.out = 51))
}
```

## D.8 Export ATN Scores

Export composite scores with sample IDs and all covariates to a CSV file for downstream use.

```{r export-atn, results='asis'}
# Identify available covariate columns
export_covariate_cols <- c("Record_ID", "CDX_collapsed", "Ancestry", "Group", "Race",
                           "Ethnicity", "sex", "age_at_subject", "APOE.geno",
                           "APOE4_carrier", "APOE4_count", "BMI", "Site",
                           "Case_Control", "AOO", "Years_Onset")
export_score_cols <- c("A_score", "T_score", "N_score",
                       "BD_pTau217_ratio", "BD_pTau181_ratio", "BD_pTau231_ratio",
                       "Proteinopathy_burden",
                       "Inflammation_PC1", "Inflammation_PC2", "Inflammation_PC3")

avail_cov <- intersect(export_covariate_cols, names(dat_list$data))
avail_scores <- intersect(export_score_cols, names(dat_list$data))

atn_export <- dat_list$data %>%
  select(all_of(avail_cov), all_of(avail_scores))

atn_export_path <- file.path(output_dir, "ATN_scores_with_covariates.csv")
write_csv(atn_export, atn_export_path)

cat(sprintf("**Exported:** %d samples x %d columns to `%s`\n\n",
            nrow(atn_export), ncol(atn_export), basename(atn_export_path)))
cat(sprintf("**Covariate columns:** %s\n\n", paste(avail_cov, collapse = ", ")))
cat(sprintf("**Score columns:** %s\n\n", paste(avail_scores, collapse = ", ")))

kable(head(atn_export, 10), caption = "First 10 rows of exported ATN scores", digits = 3)
```

## D.9 Composite Score Patterns Across All Diagnosis Groups

This section shows individual-level composite score distributions for all available diagnosis groups, including non-AD dementias. Each point represents one individual, overlaid on violin plots to show the full distribution shape.

```{r composite-patterns-all-dx, fig.width=14, fig.height=10, warning=FALSE}
score_cols_for_profiles <- c("A_score", "T_score", "N_score", "Proteinopathy_burden")
available_profile_scores <- intersect(score_cols_for_profiles, names(dat_list$data))

if (length(available_profile_scores) > 0) {
  # Include all diagnosis groups that exist
  all_dx_groups <- unique(dat_list$data$CDX_collapsed)
  all_dx_groups <- all_dx_groups[!is.na(all_dx_groups)]

  profile_long <- dat_list$data %>%
    filter(CDX_collapsed %in% all_dx_groups) %>%
    select(Record_ID, CDX_collapsed, all_of(available_profile_scores)) %>%
    pivot_longer(all_of(available_profile_scores), names_to = "Score", values_to = "Value") %>%
    filter(is.finite(Value))

  # Order diagnosis groups by clinical severity
  dx_order <- c("NCI", "MCI", "AD", intersect(
    setdiff(all_dx_groups, c("NCI", "MCI", "AD")), all_dx_groups
  ))
  profile_long$CDX_collapsed <- factor(profile_long$CDX_collapsed, levels = dx_order)

  # Color palette -- fixed for NCI/MCI/AD, generate extras for others
  base_colors <- c("NCI" = "#4DAF4A", "MCI" = "#FF7F00", "AD" = "#E41A1C")
  other_groups <- setdiff(dx_order, names(base_colors))
  if (length(other_groups) > 0) {
    extra_palette <- c("#984EA3", "#A65628", "#F781BF", "#999999", "#377EB8")
    extra_colors <- setNames(extra_palette[seq_along(other_groups)], other_groups)
    dx_colors <- c(base_colors, extra_colors)
  } else {
    dx_colors <- base_colors
  }

  ggplot(profile_long, aes(x = CDX_collapsed, y = Value, color = CDX_collapsed)) +
    geom_violin(aes(fill = CDX_collapsed), alpha = 0.15, color = NA) +
    geom_jitter(width = 0.2, alpha = 0.3, size = 0.5) +
    facet_wrap(~Score, scales = "free_y", nrow = 1) +
    scale_color_manual(values = dx_colors) +
    scale_fill_manual(values = dx_colors) +
    labs(title = "Individual Composite Score Distributions by Diagnosis",
         subtitle = "Each point = one individual | A_score: lower = more pathology | T/N/Proteinopathy: higher = more pathology",
         x = NULL, y = "Score Value") +
    theme_bw() +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## D.10 AD-Likeness Score

To quantify how AD-like each individual's biomarker profile is, we compute a composite "AD-likeness" score:

1. Z-score each composite score relative to the **NCI distribution** (mean and SD from NCI group)
2. Flip the sign of A_score (since lower A_score = more amyloid pathology = more AD-like)
3. Average the z-scores into a single AD-likeness metric

Higher AD-likeness = biomarker profile more similar to AD than NCI.

```{r ad-likeness, fig.width=12, fig.height=6, results='asis', warning=FALSE}
ad_likeness_scores <- c("A_score", "T_score", "N_score", "Proteinopathy_burden")
avail_al <- intersect(ad_likeness_scores, names(dat_list$data))

if (length(avail_al) >= 2) {
  # Compute NCI reference statistics
  nci_ref <- dat_list$data %>%
    filter(CDX_collapsed == "NCI") %>%
    summarize(across(all_of(avail_al), list(
      mean = ~mean(.x, na.rm = TRUE),
      sd = ~sd(.x, na.rm = TRUE)
    )))

  # Z-score all individuals relative to NCI
  dat_likeness <- dat_list$data %>%
    filter(!is.na(CDX_collapsed))

  for (sc in avail_al) {
    nci_mean <- nci_ref[[paste0(sc, "_mean")]]
    nci_sd <- nci_ref[[paste0(sc, "_sd")]]
    z_col <- paste0(sc, "_z")
    if (!is.na(nci_sd) && nci_sd > 0) {
      dat_likeness[[z_col]] <- (dat_likeness[[sc]] - nci_mean) / nci_sd
    } else {
      dat_likeness[[z_col]] <- NA_real_
    }
  }

  # Flip A_score_z sign: lower A_score = more AD-like, so negate the z-score
  if ("A_score_z" %in% names(dat_likeness)) {
    dat_likeness$A_score_z <- -dat_likeness$A_score_z
  }

  # Average z-scores into AD-likeness
  z_cols <- paste0(avail_al, "_z")
  z_cols <- intersect(z_cols, names(dat_likeness))
  dat_likeness$AD_likeness <- rowMeans(dat_likeness[, z_cols, drop = FALSE], na.rm = TRUE)

  # Order diagnosis groups
  dx_order_al <- c("NCI", "MCI", "AD", setdiff(unique(dat_likeness$CDX_collapsed), c("NCI", "MCI", "AD")))
  dat_likeness$CDX_collapsed <- factor(dat_likeness$CDX_collapsed, levels = dx_order_al)

  # Plot AD-likeness by diagnosis group
  dat_plot_al <- dat_likeness %>% filter(is.finite(AD_likeness))

  p_al <- ggplot(dat_plot_al, aes(x = CDX_collapsed, y = AD_likeness, fill = CDX_collapsed)) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.25, size = 0.5) +
    scale_fill_manual(values = dx_colors) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    annotate("text", x = 0.5, y = -0.3, label = "NCI-like", hjust = 0, size = 3, color = "gray40") +
    annotate("text", x = 0.5, y = 0.3, label = "AD-like", hjust = 0, size = 3, color = "gray40") +
    labs(title = "AD-Likeness Score by Diagnosis Group",
         subtitle = "Composite z-score relative to NCI | Higher = more AD-like biomarker profile",
         x = NULL, y = "AD-Likeness Score (z-score units)") +
    theme_bw() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1))
  print(p_al)

  # Summary statistics by group
  al_summary <- dat_plot_al %>%
    group_by(CDX_collapsed) %>%
    summarize(
      N = n(),
      Mean_AD_likeness = round(mean(AD_likeness, na.rm = TRUE), 3),
      SD = round(sd(AD_likeness, na.rm = TRUE), 3),
      Median = round(median(AD_likeness, na.rm = TRUE), 3),
      Pct_above_AD_median = round(
        mean(AD_likeness > median(dat_plot_al$AD_likeness[dat_plot_al$CDX_collapsed == "AD"], na.rm = TRUE),
             na.rm = TRUE) * 100, 1),
      .groups = "drop"
    )
  kable(al_summary,
        caption = "AD-Likeness by Diagnosis Group (Pct_above_AD_median = % of group with AD-likeness above the AD group median)")

  # Show most AD-like non-AD/non-MCI individuals
  non_ad_groups <- setdiff(levels(dat_likeness$CDX_collapsed), c("NCI", "MCI", "AD"))
  if (length(non_ad_groups) > 0) {
    cat("\n\n### Most AD-like Non-AD Dementia Individuals\n\n")
    top_ad_like <- dat_plot_al %>%
      filter(CDX_collapsed %in% non_ad_groups) %>%
      arrange(desc(AD_likeness)) %>%
      select(Record_ID, CDX_collapsed, any_of(c("Ancestry", "age_at_subject", "APOE4_carrier")),
             AD_likeness) %>%
      head(15) %>%
      mutate(AD_likeness = round(AD_likeness, 3))
    kable(top_ad_like, caption = "Top 15 Most AD-like Non-AD Dementia Individuals")
  }
}
```

---

# E. Elastic Net Classification (NCI vs AD)

This section fits elastic net models to classify NCI vs AD, using biomarkers plus covariates (sex, age, APOE genotype). The model uses cross-validation to select regularization strength and reports AUC performance.

```{r elastic-net-config, results='asis', message=FALSE}
# Covariates to include alongside biomarkers
enet_covariates <- c("sex", "age_at_subject", "APOE.geno")
cat(sprintf("**Covariates included:** %s\n\n", paste(enet_covariates, collapse = ", ")))
```

```{r elastic-net, results='asis'}
# Derive ancestry groups from data
available_ancestries <- unique(dat_list$data$Ancestry)
available_ancestries <- available_ancestries[available_ancestries != "Other"]

elastic_net_results <- run_elastic_net_all_ancestries(
  dat_list,
  ancestries = c("All", available_ancestries),
  alpha = 0.5,
  include_covariates = enet_covariates
)

saveRDS(elastic_net_results, file.path(output_dir, "elastic_net_results.rds"))
```

## E.1 Model Performance

```{r enet-performance, results='asis'}
perf_df <- map_dfr(names(elastic_net_results), function(anc) {
  r <- elastic_net_results[[anc]]
  if (!is.null(r)) {
    tibble(
      Ancestry = anc,
      N = r$n_samples,
      Features_Total = r$n_features_total,
      AUC = round(as.numeric(r$auc), 3),
      AUC_95_CI = sprintf("%.3f - %.3f", r$auc_ci[1], r$auc_ci[3]),
      Biomarkers_Selected = r$n_biomarkers_selected,
      Covariates_Selected = r$n_covariates_selected
    )
  }
})
kable(perf_df, caption = "Elastic Net Performance: NCI vs AD by Ancestry (with covariates)")
```

## E.2 ROC Curves

```{r roc-curves, fig.width=10, fig.height=8}
plot_roc_by_ancestry(elastic_net_results)
```

## E.3 Top Features by Ancestry

Biomarkers are shown at full opacity; covariates (age, sex, APOE) are shown faded. Red bars indicate features elevated in AD; blue bars indicate features elevated in NCI.

```{r top-biomarkers, fig.width=12, fig.height=10}
plot_top_biomarkers(elastic_net_results, top_n = 20)
```

## E.4 Selected Covariates Detail

```{r covariate-detail, results='asis'}
cov_detail <- map_dfr(names(elastic_net_results), function(anc) {
  r <- elastic_net_results[[anc]]
  if (!is.null(r) && nrow(r$coefficients) > 0) {
    r$coefficients %>%
      filter(is_covariate) %>%
      mutate(Ancestry = anc) %>%
      select(Ancestry, Feature = biomarker, Coefficient = coefficient)
  }
})

if (nrow(cov_detail) > 0) {
  cov_detail <- cov_detail %>% mutate(Coefficient = round(Coefficient, 4))
  kable(cov_detail, caption = "Covariate Coefficients Selected by Elastic Net")
} else {
  cat("No covariates were selected by the elastic net (all coefficients penalized to zero).\n\n")
}
```

## E.5 Biomarker Comparison Across Ancestries

```{r biomarker-comparison, results='asis'}
biomarker_comparison <- compare_biomarkers_across_ancestries(elastic_net_results)
write_csv(biomarker_comparison,
          file.path(output_dir, "biomarkers_comparison_across_ancestries.csv"))

universal <- biomarker_comparison %>%
  filter(n_ancestries_selected == length(elastic_net_results))

if (nrow(universal) > 0) {
  kable(universal %>% select(biomarker, n_ancestries_selected),
        caption = "Universal Features (selected in ALL ancestry groups)")
} else {
  cat("No features were selected across all ancestry groups.\n\n")
}

kable(head(biomarker_comparison, 30),
      caption = "Top 30 Features by Number of Ancestries Selected",
      digits = 4)
```

---

# F. MCI Progression Risk Prediction

This section applies the NCI-vs-AD elastic net models from Section E to score MCI samples on an AD-risk continuum. Each MCI sample receives a predicted probability of having an AD-like biomarker profile. Higher scores indicate biomarker profiles more similar to AD than NCI.

The predictions below use the **pan-ancestry ("All") model** -- the elastic net trained on all ancestry groups combined from Section E. Per-ancestry predictions using ancestry-specific models are shown in F.4.

```{r mci-predictions}
mci_predictions_all <- predict_mci_risk(
  dat_list,
  elastic_net_results[["All"]],
  ancestry_group = "All"
)

write_csv(mci_predictions_all,
          file.path(output_dir, "mci_ad_risk_predictions.csv"))
```

## F.1 Risk Score Distribution

```{r risk-summary, results='asis'}
if (!is.null(mci_predictions_all)) {
  risk_ct <- mci_predictions_all %>%
    count(risk_category, name = "n") %>%
    mutate(pct = round(n / sum(n) * 100, 1))
  kable(risk_ct, caption = "MCI Risk Category Distribution (All Ancestries Model)")

  risk_by_anc <- mci_predictions_all %>%
    count(Ancestry, risk_category) %>%
    pivot_wider(names_from = risk_category, values_from = n, values_fill = 0)
  kable(risk_by_anc, caption = "Risk Category by Ancestry (All Ancestries Model)")
}
```

## F.2 Risk Scores by Ancestry

```{r risk-by-ancestry, fig.width=10, fig.height=6}
if (!is.null(mci_predictions_all)) {
  plot_mci_risk_scores(mci_predictions_all, group_by = "Ancestry")
}
```

## F.3 Risk Scores by APOE4

```{r risk-by-apoe4, fig.width=8, fig.height=6}
if (!is.null(mci_predictions_all) &&
    "APOE4_carrier" %in% names(mci_predictions_all)) {
  plot_mci_risk_scores(mci_predictions_all, group_by = "APOE4_carrier")
}
```

## F.4 Ancestry-Specific MCI Risk Predictions

The pan-ancestry model above may miss ancestry-specific biomarker effects. Here we apply each ancestry-specific elastic net model (from Section E) to predict AD risk for MCI samples within that ancestry group.

```{r mci-predictions-by-ancestry, results='asis'}
ancestry_models <- setdiff(names(elastic_net_results), "All")
ancestry_predictions <- list()

for (anc in ancestry_models) {
  if (!is.null(elastic_net_results[[anc]])) {
    pred <- tryCatch(
      predict_mci_risk(dat_list, elastic_net_results[[anc]], ancestry_group = anc),
      error = function(e) { warning(sprintf("Prediction failed for %s: %s", anc, e$message)); NULL }
    )
    if (!is.null(pred) && nrow(pred) > 0) {
      ancestry_predictions[[anc]] <- pred
      cat(sprintf("\n### %s (N = %d MCI samples)\n\n", anc, nrow(pred)))
      risk_ct_anc <- pred %>%
        count(risk_category, name = "n") %>%
        mutate(pct = round(n / sum(n) * 100, 1))
      print(kable(risk_ct_anc, caption = sprintf("Risk Categories: %s Model", anc)))
      cat("\n")
    }
  }
}

if (length(ancestry_predictions) == 0) {
  cat("No ancestry-specific models had sufficient MCI samples for prediction.\n\n")
}
```

---

# G. Supervised Feature Selection: MCI+AD vs NCI

This section takes a different approach from the NCI-vs-AD classification in Section E. Instead of distinguishing AD from NCI, it asks: **which biomarkers and covariates best separate cognitively impaired individuals (MCI + AD combined) from cognitively normal (NCI)?** This uses the same elastic net framework as Section E but with a different outcome definition.

This is relevant because:

- Features that separate MCI+AD from NCI may include early markers of cognitive decline (not just late-stage AD pathology)
- Comparing these features with the NCI-vs-AD model highlights which biomarkers are specific to advanced disease vs. general cognitive impairment
- This analysis is independent of the ATN framework groupings

TODO: Try this stratified by ancestry. List shared features.

```{r mci-ad-vs-nci}
mci_ad_vs_nci_result <- run_mci_ad_vs_nci(
  dat_list,
  alpha = 0.5,
  include_covariates = enet_covariates
)

saveRDS(mci_ad_vs_nci_result, file.path(output_dir, "mci_ad_vs_nci_results.rds"))
```

## G.1 Model Performance

```{r mci-ad-nci-perf, results='asis'}
cat(sprintf("**N samples:** %d\n\n", mci_ad_vs_nci_result$n_samples))
cat(sprintf("**AUC:** %.3f (95%% CI: %.3f - %.3f)\n\n",
            as.numeric(mci_ad_vs_nci_result$auc),
            mci_ad_vs_nci_result$auc_ci[1],
            mci_ad_vs_nci_result$auc_ci[3]))
cat(sprintf("**Total features:** %d | **Selected:** %d\n\n",
            mci_ad_vs_nci_result$n_features_total,
            nrow(mci_ad_vs_nci_result$coefficients)))

kable(mci_ad_vs_nci_result$score_by_diagnosis,
      caption = "Predicted MCI+AD Probability by Actual Diagnosis",
      digits = 3)
```

## G.2 ROC Curve

```{r mci-ad-nci-roc, fig.width=7, fig.height=6}
roc_obj <- mci_ad_vs_nci_result$roc
roc_df <- data.frame(
  fpr = 1 - roc_obj$specificities,
  tpr = roc_obj$sensitivities
)

ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "#984EA3", linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  annotate("text", x = 0.6, y = 0.2,
           label = sprintf("AUC = %.3f", as.numeric(mci_ad_vs_nci_result$auc)),
           size = 5, color = "#984EA3") +
  labs(title = "ROC Curve: MCI+AD vs NCI",
       subtitle = "Elastic net with biomarkers + covariates",
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_bw()
```

## G.3 Top Features

```{r mci-ad-nci-features, results='asis', fig.width=10, fig.height=8}
top_feats <- mci_ad_vs_nci_result$coefficients

if (nrow(top_feats) > 0) {
  top_feats_display <- top_feats %>%
    mutate(
      Type = ifelse(is_covariate, "Covariate", "Biomarker"),
      Direction = ifelse(coefficient > 0, "Higher in MCI+AD", "Higher in NCI"),
      coefficient = round(coefficient, 4)
    ) %>%
    select(Feature = feature, Coefficient = coefficient, Type, Direction)

  kable(head(top_feats_display, 30),
        caption = "Top 30 Selected Features (MCI+AD vs NCI)")

  # Plot
  plot_df <- top_feats %>%
    slice_max(abs(coefficient), n = 25) %>%
    mutate(feature_type = ifelse(is_covariate, "Covariate", "Biomarker"))

  ggplot(plot_df, aes(x = reorder(feature, coefficient),
                       y = coefficient,
                       fill = coefficient > 0,
                       alpha = feature_type)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#E41A1C", "FALSE" = "#377EB8"),
                      labels = c("TRUE" = "Higher in MCI+AD", "FALSE" = "Higher in NCI"),
                      name = "Direction") +
    scale_alpha_manual(values = c("Biomarker" = 1, "Covariate" = 0.5),
                       name = "Feature Type") +
    labs(title = "Top 25 Features: MCI+AD vs NCI",
         subtitle = "Elastic net coefficients | Faded = covariate",
         x = "Feature", y = "Coefficient") +
    theme_bw() +
    theme(legend.position = "bottom")
} else {
  cat("No features selected by the model.\n\n")
}
```

## G.4 Predicted Score Distribution by Diagnosis

```{r mci-ad-nci-scores, fig.width=8, fig.height=6}
score_df <- data.frame(
  CDX_collapsed = mci_ad_vs_nci_result$diagnosis,
  pred_score = mci_ad_vs_nci_result$predictions
)
score_df$CDX_collapsed <- factor(score_df$CDX_collapsed, levels = c("NCI", "MCI", "AD"))

ggplot(score_df, aes(x = CDX_collapsed, y = pred_score, fill = CDX_collapsed)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  geom_jitter(width = 0.15, alpha = 0.15, size = 0.5) +
  scale_fill_manual(values = c("NCI" = "#4DAF4A", "MCI" = "#FF7F00", "AD" = "#E41A1C")) +
  labs(title = "Predicted MCI+AD Score by Actual Diagnosis",
       subtitle = "Higher score = more likely to be cognitively impaired",
       x = NULL, y = "Predicted P(MCI+AD)") +
  theme_bw() +
  theme(legend.position = "none")
```

## G.5 Comparison with NCI-vs-AD Features

```{r compare-models, results='asis'}
# Compare which features are selected by each model
if (!is.null(elastic_net_results[["All"]])) {
  nci_ad_feats <- elastic_net_results[["All"]]$coefficients$biomarker
  mci_ad_nci_feats <- mci_ad_vs_nci_result$coefficients$feature

  shared <- intersect(nci_ad_feats, mci_ad_nci_feats)
  only_nci_ad <- setdiff(nci_ad_feats, mci_ad_nci_feats)
  only_mci_ad_nci <- setdiff(mci_ad_nci_feats, nci_ad_feats)

  cat(sprintf("**NCI vs AD model:** %d features selected\n\n", length(nci_ad_feats)))
  cat(sprintf("**MCI+AD vs NCI model:** %d features selected\n\n", length(mci_ad_nci_feats)))
  cat(sprintf("**Shared:** %d features\n\n", length(shared)))

  if (length(shared) > 0) {
    kable(data.frame(Shared_Features = sort(shared)),
          caption = "Features Selected by Both Models")
  }

  if (length(only_mci_ad_nci) > 0) {
    cat(sprintf("\n\n**Unique to MCI+AD vs NCI** (%d features): %s\n\n",
                length(only_mci_ad_nci), paste(sort(only_mci_ad_nci), collapse = ", ")))
  }

  if (length(only_nci_ad) > 0) {
    cat(sprintf("\n\n**Unique to NCI vs AD** (%d features): %s\n\n",
                length(only_nci_ad), paste(sort(only_nci_ad), collapse = ", ")))
  }
}
```

---

# H. MCI Clustering (Unsupervised Discovery)

This section takes a complementary **unsupervised** approach to the supervised analyses above. Rather than using pre-defined diagnostic labels (NCI/AD), it discovers natural subgroups among MCI samples based on their biomarker profiles using k-means clustering. This can reveal MCI subtypes that may have different progression trajectories.

The clustering uses the top 50 most variable biomarkers plus key covariates (age, sex, APOE4 carrier status) as features, with gap statistics to determine the optimal number of clusters. All features are equally weighted after standardization.

```{r mci-clustering, results= 'asis'}
cluster_result <- cluster_mci_samples(
  dat_list,
  n_clusters = "auto",
  biomarker_subset = NULL,
  top_n = 50,
  include_covariates = c("age_at_subject", "sex", "APOE4_carrier")
)

saveRDS(cluster_result, file.path(output_dir, "mci_clustering_results.rds"))
```

## H.1 Cluster Summary

```{r cluster-summary, results='asis'}
cat(sprintf("**Optimal clusters:** %d\n\n", cluster_result$n_clusters))
cat(sprintf("**Average silhouette width:** %.3f\n\n", cluster_result$avg_silhouette))
cat(sprintf("**Biomarkers used:** %d (top most variable)\n\n", length(cluster_result$biomarkers_used)))
if (length(cluster_result$covariate_names) > 0) {
  cat(sprintf("**Covariates included:** %s\n\n", paste(cluster_result$covariate_names, collapse = ", ")))
  cat(sprintf("**Total features:** %d (%d biomarkers + %d covariate features)\n\n",
              length(cluster_result$biomarkers_used) + length(cluster_result$covariate_names),
              length(cluster_result$biomarkers_used), length(cluster_result$covariate_names)))
}

cluster_sizes <- as.data.frame(table(cluster_result$cluster_assignments))
names(cluster_sizes) <- c("Cluster", "N")
kable(cluster_sizes, caption = "Cluster Sizes")
```

## H.2 Cluster Profiles

```{r cluster-profiles, results='asis'}
cluster_profiles <- profile_mci_clusters(cluster_result, dat_list)
write_csv(cluster_profiles$biomarker_profiles,
          file.path(output_dir, "mci_cluster_profiles.csv"))

if (!is.null(cluster_profiles$age_profile)) {
  kable(cluster_profiles$age_profile,
        caption = "Age by Cluster", digits = 1)
}

if (!is.null(cluster_profiles$apoe_profile)) {
  apoe_wide <- cluster_profiles$apoe_profile %>%
    select(cluster, APOE4_carrier, n, pct) %>%
    mutate(label = sprintf("%d (%.1f%%)", n, pct)) %>%
    select(cluster, APOE4_carrier, label) %>%
    pivot_wider(names_from = APOE4_carrier, values_from = label, values_fill = "0 (0.0%)")
  kable(apoe_wide, caption = "APOE4 Carrier Status by Cluster")
}

if (!is.null(cluster_profiles$ancestry_profile)) {
  anc_wide <- cluster_profiles$ancestry_profile %>%
    select(cluster, Ancestry, n, pct) %>%
    mutate(label = sprintf("%d (%.1f%%)", n, pct)) %>%
    select(cluster, Ancestry, label) %>%
    pivot_wider(names_from = Ancestry, values_from = label, values_fill = "0 (0.0%)")
  kable(anc_wide, caption = "Ancestry by Cluster")
}
```

## H.3 Distance to AD Centroid

How similar is each MCI cluster to the AD biomarker profile? Euclidean distances are computed in the same scaled space used for clustering. Smaller distance = more AD-like biomarker profile.

```{r cluster-distances, results='asis'}
cluster_distances <- calculate_cluster_ad_distance(dat_list, cluster_result, "All")
write_csv(cluster_distances, file.path(output_dir, "cluster_distances_to_AD.csv"))
kable(cluster_distances, caption = "MCI Cluster Distance to AD Centroid (closer = more AD-like)",
      digits = 3)
```

## H.4 PCA of MCI Clusters (with NCI and AD overlay)

PCA is computed on MCI samples, then NCI and AD samples are projected into the same space. This shows how MCI clusters relate to the healthy (NCI) and disease (AD) populations.

```{r cluster-pca, fig.width=10, fig.height=8}
cluster_result <- add_dim_reduction(cluster_result, dat_list = dat_list,
                                     method = "pca", n_pcs = 5,
                                     project_all = TRUE)
saveRDS(cluster_result, file.path(output_dir, "mci_clustering_with_dimred.rds"))

# Plot 1: MCI clusters only
plot_mci_clusters(cluster_result, color_by = "cluster")
```

```{r cluster-all-dx, fig.width=10, fig.height=8}
# Plot 2: All diagnoses in same PCA space
plot_all_diagnoses_pca(cluster_result)
```

```{r cluster-ancestry, fig.width=10, fig.height=8}
if ("Ancestry" %in% names(cluster_result$data)) {
  plot_mci_clusters(cluster_result, color_by = "Ancestry")
}
```

## H.5 Higher PCs

PC1 and PC2 capture the most variance, but additional PCs may reveal structure not visible in the first two dimensions.

```{r higher-pcs, fig.width=10, fig.height=8, warning=FALSE}
# PC3 vs PC4
if ("PC3" %in% names(cluster_result$data) && "PC4" %in% names(cluster_result$data)) {
  print(plot_pc_pair(cluster_result, pc_x = 3, pc_y = 4, color_by = "cluster"))
}
```

```{r pca-variance, results='asis'}
if (!is.null(cluster_result$var_explained)) {
  ve_df <- data.frame(
    PC = paste0("PC", 1:length(cluster_result$var_explained)),
    Variance_Explained_Pct = round(cluster_result$var_explained, 2),
    Cumulative_Pct = round(cumsum(cluster_result$var_explained), 2)
  )
  kable(ve_df, caption = "Variance Explained by Principal Components")
}
```

## H.6 PC-Ancestry Correlation

A concern with PCA on biomarkers is that the first PCs may capture population structure (ancestry differences) rather than disease-relevant variation. To check this, we compute eta-squared (proportion of variance in each PC explained by ancestry group) using one-way ANOVA.

```{r pc-ancestry-correlation, results='asis'}
if (!is.null(cluster_result$pca_result) && "Ancestry" %in% names(cluster_result$data)) {
  pc_cols <- paste0("PC", 1:min(5, length(cluster_result$var_explained)))
  available_pcs <- intersect(pc_cols, names(cluster_result$data))

  if (length(available_pcs) > 0) {
    ancestry_eta2 <- sapply(available_pcs, function(pc) {
      fit <- aov(as.formula(paste(pc, "~ Ancestry")), data = cluster_result$data)
      ss <- summary(fit)[[1]]
      ss["Ancestry", "Sum Sq"] / sum(ss[, "Sum Sq"])
    })

    eta2_df <- data.frame(
      PC = available_pcs,
      Variance_Explained_Total = round(cluster_result$var_explained[1:length(available_pcs)], 1),
      Ancestry_eta2 = round(ancestry_eta2, 3),
      Pct_of_PC_from_Ancestry = round(ancestry_eta2 * 100, 1)
    )
    print(kable(eta2_df,
          caption = "How much of each PC is explained by Ancestry (eta-squared from ANOVA)"))

    cat("\n\n**Interpretation:** High eta-squared (>0.1) indicates that a PC largely captures ancestry ",
        "differences rather than disease-relevant variation. If PC1/PC2 are ancestry-driven, ",
        "the unsupervised clustering above may be grouping by ancestry rather than by disease subtype. ",
        "The supervised PCA approach below addresses this by defining axes from the AD/NCI contrast.\n\n")
  }
} else {
  cat("PC-ancestry analysis skipped (PCA or Ancestry not available).\n\n")
}
```

## H.7 Supervised PCA: AD/NCI-Defined Space

The unsupervised PCA above (H.4) computes principal components from MCI samples alone. A limitation is that the resulting axes maximize variance among MCI samples, which may be driven by ancestry or batch effects rather than disease-relevant biology.

**Supervised PCA** takes a different approach: PCA axes are defined by the **AD + NCI reference** samples. MCI samples are then projected into this space. This means each PC axis captures variation that distinguishes the known disease groups. MCI position along these axes directly reflects AD-likeness.

```{r supervised-pca, results='asis', fig.width=12, fig.height=10, warning=FALSE}
sup_pca <- supervised_pca_clustering(dat_list, cluster_result,
                                      n_pcs = 10, n_clusters = "auto")

saveRDS(sup_pca, file.path(output_dir, "supervised_pca_results.rds"))

cat(sprintf("**Reference samples:** %d (NCI + AD)\n\n",
            sum(sup_pca$projected$CDX_collapsed != "MCI")))
cat(sprintf("**MCI re-clustered in supervised space:** %d clusters\n\n", sup_pca$n_clusters))
cat(sprintf("**PCs retained:** %d\n\n", sup_pca$n_pcs))
```

### H.7.1 MCI in AD/NCI PCA Space (PC1 vs PC2)

```{r sup-pca-scatter, fig.width=12, fig.height=8, warning=FALSE}
# All samples: NCI, MCI, AD in supervised PCA space
proj <- sup_pca$projected
proj$CDX_collapsed <- factor(proj$CDX_collapsed, levels = c("NCI", "MCI", "AD"))

# NCI and AD as background, MCI highlighted
p_bg <- ggplot() +
  # NCI and AD as transparent background
  geom_point(data = proj %>% filter(CDX_collapsed != "MCI"),
             aes(x = PC1, y = PC2, color = CDX_collapsed),
             alpha = 0.15, size = 1) +
  # MCI colored by supervised cluster
  geom_point(data = sup_pca$mci_projected,
             aes(x = PC1, y = PC2, color = factor(supervised_cluster)),
             alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("NCI" = "#4DAF4A", "AD" = "#E41A1C",
                                 "1" = "#377EB8", "2" = "#FF7F00",
                                 "3" = "#984EA3", "4" = "#A65628",
                                 "5" = "#F781BF", "6" = "#999999"),
                      name = "Group / Cluster") +
  labs(title = "Supervised PCA: MCI Clusters in AD/NCI-Defined Space",
       subtitle = "PCA axes defined by AD + NCI samples | MCI projected and re-clustered",
       x = sprintf("PC1 (%.1f%% var)", sup_pca$var_explained[1]),
       y = sprintf("PC2 (%.1f%% var)", sup_pca$var_explained[2])) +
  theme_bw()
print(p_bg)
```

### H.7.2 MCI Cluster Position Along the NCI-to-AD Axis

For each PC, we compute where each MCI cluster falls as a fraction of the NCI-to-AD distance (0 = at NCI centroid, 1 = at AD centroid, >1 = beyond AD).

```{r sup-pca-ad-fraction, results='asis', fig.width=10, fig.height=6}
ad_frac <- sup_pca$ad_fraction
pc_cols <- paste0("PC", 1:min(5, sup_pca$n_pcs))

# Reshape to long format for table
frac_long <- ad_frac %>%
  select(supervised_cluster, all_of(pc_cols)) %>%
  pivot_longer(all_of(pc_cols), names_to = "PC", values_to = "Fraction_NCI_to_AD") %>%
  mutate(Fraction_NCI_to_AD = round(Fraction_NCI_to_AD, 3))

frac_wide <- frac_long %>%
  pivot_wider(names_from = PC, values_from = Fraction_NCI_to_AD)

kable(frac_wide,
      caption = "MCI Cluster Position: Fraction of NCI-to-AD Distance per PC (0=NCI, 1=AD)")

# Bar plot of AD-fraction for PC1
if ("PC1" %in% pc_cols) {
  frac_pc1 <- ad_frac %>%
    select(supervised_cluster, all_of(pc_cols[1:min(3, length(pc_cols))])) %>%
    pivot_longer(-supervised_cluster, names_to = "PC", values_to = "Fraction")

  p_frac <- ggplot(frac_pc1, aes(x = factor(supervised_cluster), y = Fraction, fill = PC)) +
    geom_col(position = "dodge", alpha = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "#4DAF4A", linewidth = 0.5) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "#E41A1C", linewidth = 0.5) +
    annotate("text", x = 0.5, y = -0.05, label = "NCI", color = "#4DAF4A", size = 3, hjust = 0) +
    annotate("text", x = 0.5, y = 1.05, label = "AD", color = "#E41A1C", size = 3, hjust = 0) +
    scale_fill_brewer(palette = "Set2") +
    labs(title = "MCI Cluster Position Relative to NCI and AD",
         subtitle = "Fraction of NCI-to-AD distance on each PC | Closer to 1 = more AD-like",
         x = "MCI Cluster (in supervised PCA space)",
         y = "Position (0 = NCI centroid, 1 = AD centroid)") +
    theme_bw()
  print(p_frac)
}
```

### H.7.3 Variance Explained by Ancestry in Supervised PCA

```{r sup-pca-ancestry-check, results='asis'}
if ("Ancestry" %in% names(sup_pca$projected)) {
  sup_pc_cols <- paste0("PC", 1:min(5, sup_pca$n_pcs))
  mci_only <- sup_pca$mci_projected

  sup_eta2 <- sapply(sup_pc_cols, function(pc) {
    if (pc %in% names(mci_only) && "Ancestry" %in% names(mci_only)) {
      fit <- tryCatch(
        aov(as.formula(paste(pc, "~ Ancestry")), data = mci_only),
        error = function(e) NULL
      )
      if (!is.null(fit)) {
        ss <- summary(fit)[[1]]
        return(ss["Ancestry", "Sum Sq"] / sum(ss[, "Sum Sq"]))
      }
    }
    return(NA_real_)
  })

  eta2_sup_df <- data.frame(
    PC = sup_pc_cols,
    Supervised_Var_Explained = round(sup_pca$var_explained[1:length(sup_pc_cols)], 1),
    Ancestry_eta2_MCI = round(sup_eta2, 3)
  )
  print(kable(eta2_sup_df,
        caption = "Ancestry influence on supervised PCs (among MCI samples only)"))

  cat("\n\n**Compare with H.6:** If ancestry eta-squared is lower in the supervised PCA than in ",
      "the unsupervised PCA, the disease-defined axes are less confounded by population structure.\n\n")
}
```

## H.8 Cluster Biomarker Heatmap

```{r cluster-heatmap, fig.width=10, fig.height=12}
plot_cluster_heatmap(cluster_result, top_n = 30)
```

---

# I. Summary

```{r final-summary, results='asis'}
summary_stats <- list(
  total_samples = nrow(dat_list$data),
  n_biomarkers = length(dat_list$biomarker_cols),
  diagnosis_counts = table(dat_list$data$CDX_collapsed),
  ancestry_counts = table(dat_list$data$Ancestry),
  elastic_net_aucs = sapply(elastic_net_results, function(x) {
    if (!is.null(x$auc)) round(as.numeric(x$auc), 3) else NA
  }),
  n_mci_clusters = cluster_result$n_clusters,
  avg_silhouette = round(cluster_result$avg_silhouette, 3),
  mci_risk_summary = if (!is.null(mci_predictions_all)) summary(mci_predictions_all$AD_risk_score) else NULL,
  mci_ad_vs_nci_auc = round(as.numeric(mci_ad_vs_nci_result$auc), 3)
)

capture.output(summary_stats, file = file.path(output_dir, "analysis_summary.txt"))

summary_df <- data.frame(
  Metric = c("Total samples", "Biomarkers", "MCI clusters",
             "Avg silhouette", "MCI+AD vs NCI AUC",
             paste0("AUC (", names(summary_stats$elastic_net_aucs), ")")),
  Value = c(summary_stats$total_samples, summary_stats$n_biomarkers,
            summary_stats$n_mci_clusters, summary_stats$avg_silhouette,
            summary_stats$mci_ad_vs_nci_auc,
            summary_stats$elastic_net_aucs)
)
kable(summary_df, caption = "Analysis Summary")

cat(sprintf("\n\n**Results saved to:** `%s`\n", output_dir))
```
