---
title: "APOE"
format: html
editor: visual
---

::: {.callout-important title="Summary for molecular colleagues — APOE Sanger vs WGS discordance: RESOLVED (WGS plate swap)"}

**Resolution (June 2026).** The Sanger–WGS APOE discordance documented below is
**not** an ancestry-specific WGS-calling problem, and **not** biology. It was a
**physical sample mix-up during WGS plate handling**, confined to three of the eight
plates on the AD-Hispanic (`pU5.ADhisp`) WGS set:

- **Plate 2** was loaded **180° rotated** (every well `w` swapped with well `97−w`);
- **Plates 3 and 7** were **swapped** well-for-well.

Plates 1, 4, 5, 6, 8 were clean. The DNA and sequence are good — only the **labels**
were wrong, so ~261 Hispanic samples carried another subject's WGS APOE genotype.

**How we know.** A genome-wide **chromosome-X fingerprint** (~16,000 sites,
independent of APOE) showed discordant samples disagree at ~50% of sites — a
*different person*, not a miscall. The swaps are clean pairwise involutions,
**plate-quantized** (plates 2/3/7 = 100% discordant, the others 0%), and the
corrective geometry derived purely from well position matched the chrX genetics
**147 / 147**. Re-deriving APOE after the relabel restores agreement with Sanger
from 47% → **99.6%**.

**Why it masqueraded as ancestry.** The affected plates *were* the AD-Hispanic
batch, so 100% of the mislabeled samples are Hispanic, and the apparent "admixture
gradient" (Colombia/Peru high → USA/Puerto Rico low) was just **plate composition** —
which countries' samples happened to load onto plates 2/3/7. Within any bad plate
the error rate is ~100% regardless of country of birth. Every ancestry-targeted
read of the discordance below (the "random-but-Hispanic-specific" mechanisms, the
admixture gradient, the 0-of-284 test) was this artifact.

**What we did.** The combined metadata's `APOE_WGS` was deterministically relabeled
for the affected samples (`Metadata_Merge/apply_wgs_plate_swap_correction.R`; audit
log `Datasets/WGS_plateswap_metadata_changes_2026Jun.csv`), and `APOE.geno_final`
was regenerated by the Metadata Merge pipeline. **Post-correction the Sanger-vs-WGS
discordant set drops from 137 to 12** — and those 12 are genuine *same-person* assay
miscalls (no plate structure), exactly the residual the chrX analysis predicted.

**Recommendation (reversed).** `APOE.geno_final` (WGS-priority) is **trustworthy
again** — the earlier advice to prefer Sanger over short-read WGS for this admixed
cohort **no longer applies**. Full diagnosis, figure, and per-sample corrected ID
map: `../../../May_2026_WGS_QC/WGS_Plate_Swap_Summary.md`.

*(Unchanged: the biomarker **assay** anomaly on run `20251124-1407_Bay1` — ~38%
APOE4-vs-genotype mismatch among otherwise-concordant samples, **ancestry-blind**
(AA ~41%, Hispanic ~36%), driven by the **APOE4 assay alone** losing dynamic range
on that bay — is a separate Alamar issue, unrelated to the plate swap, and still
stands. See `Biomarker_batch_mismatch_flags.xlsx`.)*

:::

::: {.callout-note title="About everything below — the May-2026 investigation that led to the diagnosis"}

The sections that follow are the **pre-diagnosis investigation** (May 2026),
retained as the record that ultimately pointed to a block-level sample mix-up. They
were written under the *working hypothesis* that the discordance was an
ancestry-specific WGS phenomenon — the hypothesis the plate-swap finding above
overturned. **They now run on the corrected metadata**, so the discordant set is the
~12 residual rather than 137, and the "ancestry gradients" they were built to explain
**flatten to ~0** — which is itself positive confirmation that the signal was the
plate swap, not biology. The original pre-correction figures and flag tables are
archived in `output_files_pre_plateswap_2026Jun/`. Two findings in here are
*independent* of the plate swap and remain valid: the `20251124-1407_Bay1` biomarker
assay anomaly, and the check that APOE proteins do **not** read lower in
African-ancestry samples.

:::

```{r}
#| echo: false

library(ggplot2)
library(dplyr)
library(ggrepel)
library(writexl)
library(patchwork)

# ---- Read data ----
filtered_combined_post_QC <- read.csv(
  "../../Metadata_Merge/output_files/filtered/filtered_combined_post_QC.csv"
)

# ---- Load WGS APOE lookup and join (if not already present from Metadata_Merge) ----
if (!"APOE_WGS" %in% names(filtered_combined_post_QC)) {
  wgs_lookup <- read.csv("U19_Alamar_metadata_2026Jan16_with_WGSAPOE.csv") %>%
    select(SAMPLE_ALIQUOT, APOE_WGS) %>%
    distinct()
  filtered_combined_post_QC <- filtered_combined_post_QC %>%
    left_join(wgs_lookup, by = "SAMPLE_ALIQUOT")
}

# ---- Clean APOE genotype ----
apoe_clean <- filtered_combined_post_QC$APOE.geno |>
  trimws() |>
  gsub("/", "", x = _) |>     # remove slash
  gsub("\\s+", "", x = _)     # remove spaces

apoe_clean[apoe_clean == ""] <- NA

# biologically meaningful genotype order
apoe_levels <- c("22","23","33","24","34","44")

filtered_combined_post_QC$APOEgeno <- factor(
  apoe_clean,
  levels = apoe_levels
)

# QC check
str(filtered_combined_post_QC$APOEgeno)
table(filtered_combined_post_QC$APOEgeno, useNA = "ifany")


# ---- Create APOE4 minus APOE variable ----
filtered_combined_post_QC <- filtered_combined_post_QC %>%
  mutate(APOE4_minus_APOE = APOE4 - APOE)

# ---- Pretty ggplot: APOE4_minus_APOE by genotype ----
p <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno), !is.na(APOE4_minus_APOE)) %>%
  ggplot(aes(x = APOEgeno, y = APOE4_minus_APOE, fill = APOEgeno)) +
  geom_violin(
    scale = "width",
    alpha = 0.6,
    width = 0.65
  ) +
  stat_summary(fun = median, geom = "crossbar", width = 0.35,
               linewidth = 0.3, colour = "grey20") +
  geom_jitter(
    width = 0.15,
    size = 1.5,
    alpha = 0.5
  ) +
  labs(
    x = "APOE Genotype",
    y = "APOE4 − APOE",
    title = "APOE4 − APOE by APOE Genotype"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

p



# ---- 1. Compute genotype-specific bounds ----
apoe_bounds <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno), !is.na(APOE4_minus_APOE)) %>%
  group_by(APOEgeno) %>%
  summarise(
    Q1 = quantile(APOE4_minus_APOE, 0.25, na.rm = TRUE),
    Q3 = quantile(APOE4_minus_APOE, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    lower = Q1 - 3 * IQR,
    upper = Q3 + 3 * IQR,
    .groups = "drop"
  )

# ---- 2. Identify extreme outliers ----
apoe_outliers <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno), !is.na(APOE4_minus_APOE)) %>%
  left_join(apoe_bounds, by = "APOEgeno") %>%
  filter(APOE4_minus_APOE < lower | APOE4_minus_APOE > upper)

# ---- 3. QC summary ----
cat("\nExtreme outliers by APOE genotype:\n")
print(table(apoe_outliers$APOEgeno))

cat("\nTotal extreme outliers:\n")
print(nrow(apoe_outliers))

# ---- Compute genotype medians ----
apoe_medians <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno), !is.na(APOE4_minus_APOE)) %>%
  group_by(APOEgeno) %>%
  summarise(
    genotype_median = median(APOE4_minus_APOE, na.rm = TRUE),
    .groups = "drop"
  )

# ---- Create ranked QC table ----
apoe_qc_ranked <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno), !is.na(APOE4_minus_APOE)) %>%
  left_join(apoe_medians, by = "APOEgeno") %>%
  mutate(
    deviation = APOE4_minus_APOE - genotype_median,
    abs_deviation = abs(deviation)
  ) %>%
  arrange(desc(abs_deviation))

# ---- 4. Write outliers to file ----
write_xlsx(
  apoe_outliers,
  "APOE4_minus_APOE_extreme_outliers_by_genotype.xlsx"
)
write_xlsx(
  apoe_qc_ranked,
  "Outliers_Ranked.xlsx"
)

# ---- 5. Add outliers to plot ----
p_outliers <- p +
  geom_point(
    data = apoe_outliers,
    aes(x = APOEgeno, y = APOE4_minus_APOE),
    color = "red",
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_text_repel(
    data = apoe_outliers,
    aes(
      x = APOEgeno,
      y = APOE4_minus_APOE,
      label = SAMPLE_ALIQUOT   # <-- replace with your ID column
    ),
    size = 3,
    color = "red",
    max.overlaps = 20,
    inherit.aes = FALSE
  )

p_outliers
```

# WGS APOE Comparison

```{r}
#| echo: false
#| label: wgs-concordance
#| results: asis

# ---- Build WGS genotype factor (same levels as original) ----
filtered_combined_post_QC <- filtered_combined_post_QC %>%
  mutate(APOEgeno_WGS = factor(trimws(APOE_WGS), levels = apoe_levels))

# ---- Concordance summary ----
concordance_df <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno), !is.na(APOEgeno_WGS)) %>%
  mutate(
    orig_digits = gsub("/", "", as.character(APOEgeno)),
    concordant  = orig_digits == trimws(APOE_WGS)
  )

n_total   <- nrow(filtered_combined_post_QC)
n_wgs     <- sum(!is.na(filtered_combined_post_QC$APOEgeno_WGS))
n_gained  <- sum(is.na(filtered_combined_post_QC$APOEgeno) &
                   !is.na(filtered_combined_post_QC$APOEgeno_WGS))
conc_rate <- mean(concordance_df$concordant, na.rm = TRUE)
n_discord <- sum(!concordance_df$concordant, na.rm = TRUE)

cat(sprintf("- Samples with WGS genotype: **%d** / %d (%.1f%%)\n",
            n_wgs, n_total, 100 * n_wgs / n_total))
cat(sprintf("- Samples gaining genotype from WGS (were NA in original): **%d**\n", n_gained))
cat(sprintf("- Concordance rate (where both available): **%.1f%%** (%d/%d)\n",
            100 * conc_rate, sum(concordance_df$concordant), nrow(concordance_df)))
cat(sprintf("- Discordant samples: **%d**\n\n", n_discord))

cat("### Genotype Cross-Tabulation (Original vs WGS)\n\n")
ct <- table(Original = concordance_df$orig_digits, WGS = concordance_df$APOE_WGS)
print(knitr::kable(ct))
```

```{r}
#| echo: false
#| label: wgs-discordant-table
#| results: asis

discordant_df <- concordance_df %>%
  filter(!concordant) %>%
  select(SAMPLE_ALIQUOT, CDX, APOE.geno, APOE_WGS, APOE4_minus_APOE) %>%
  arrange(APOE.geno, APOE_WGS)

cat("### Discordant Samples\n\n")
print(knitr::kable(
  discordant_df,
  col.names = c("Sample Aliquot", "Diagnosis", "Original", "WGS", "APOE4 \u2212 APOE"),
  digits = 3,
  caption = "Samples where original and WGS APOE genotypes differ"
))
```

```{r}
#| echo: false
#| label: wgs-boxplot

# ---- APOE4_minus_APOE by WGS genotype ----
p_wgs <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno_WGS), !is.na(APOE4_minus_APOE)) %>%
  ggplot(aes(x = APOEgeno_WGS, y = APOE4_minus_APOE, fill = APOEgeno_WGS)) +
  geom_violin(scale = "width", alpha = 0.6, width = 0.65) +
  stat_summary(fun = median, geom = "crossbar", width = 0.35, linewidth = 0.3, colour = "grey20") +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.5) +
  labs(
    x     = "APOE Genotype (WGS)",
    y     = "APOE4 \u2212 APOE",
    title = "APOE4 \u2212 APOE by WGS Genotype"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5))

p_wgs
```

```{r}
#| echo: false
#| label: wgs-discordant-plots

# ---- Highlight discordant samples on both plots ----
discordant_pts <- concordance_df %>% filter(!concordant)

p_orig_discord <- p +
  geom_point(
    data        = discordant_pts,
    aes(x = APOEgeno, y = APOE4_minus_APOE),
    color       = "purple", size = 2.5, inherit.aes = FALSE
  ) +
  geom_text_repel(
    data        = discordant_pts,
    aes(x = APOEgeno, y = APOE4_minus_APOE,
        label = paste0(SAMPLE_ALIQUOT, "\n(WGS:", APOE_WGS, ")")),
    size        = 2.5, color = "purple", max.overlaps = 20, inherit.aes = FALSE
  ) +
  labs(title = "Original Genotype \u2014 Discordant Samples in Purple")

p_wgs_discord <- p_wgs +
  geom_point(
    data        = discordant_pts,
    aes(x = APOEgeno_WGS, y = APOE4_minus_APOE),
    color       = "purple", size = 2.5, inherit.aes = FALSE
  ) +
  geom_text_repel(
    data        = discordant_pts,
    aes(x = APOEgeno_WGS, y = APOE4_minus_APOE,
        label = paste0(SAMPLE_ALIQUOT, "\n(orig:", orig_digits, ")")),
    size        = 2.5, color = "purple", max.overlaps = 20, inherit.aes = FALSE
  ) +
  labs(title = "WGS Genotype \u2014 Discordant Samples in Purple")

p_orig_discord
p_wgs_discord
```

```{r}
#| echo: false
#| label: wgs-outliers
#| results: asis

# ---- Re-run outlier detection using WGS genotype ----
wgs_bounds <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno_WGS), !is.na(APOE4_minus_APOE)) %>%
  group_by(APOEgeno_WGS) %>%
  summarise(
    Q1    = quantile(APOE4_minus_APOE, 0.25, na.rm = TRUE),
    Q3    = quantile(APOE4_minus_APOE, 0.75, na.rm = TRUE),
    IQR   = Q3 - Q1,
    lower = Q1 - 3 * IQR,
    upper = Q3 + 3 * IQR,
    .groups = "drop"
  )

apoe_wgs_outliers <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno_WGS), !is.na(APOE4_minus_APOE)) %>%
  left_join(wgs_bounds, by = "APOEgeno_WGS") %>%
  filter(APOE4_minus_APOE < lower | APOE4_minus_APOE > upper)

orig_ids     <- apoe_outliers$SAMPLE_ALIQUOT
wgs_ids      <- apoe_wgs_outliers$SAMPLE_ALIQUOT
resolved     <- setdiff(orig_ids, wgs_ids)
new_outliers <- setdiff(wgs_ids, orig_ids)
persisted    <- intersect(orig_ids, wgs_ids)

cat("### WGS Outlier Comparison (3\u00d7IQR)\n\n")
cat("| | Count |\n|---|---|\n")
cat(sprintf("| Original genotype outliers | %d |\n", length(orig_ids)))
cat(sprintf("| WGS genotype outliers | %d |\n", length(wgs_ids)))
cat(sprintf("| **Resolved by WGS re-assignment** | **%d** |\n", length(resolved)))
cat(sprintf("| New outliers introduced | %d |\n", length(new_outliers)))
cat(sprintf("| Persistent (both methods) | %d |\n\n", length(persisted)))

if (length(resolved) > 0) {
  cat("**Resolved outliers (no longer extreme under WGS genotype):**\n")
  cat(paste("-", resolved, collapse = "\n"), "\n\n")
}
if (length(new_outliers) > 0) {
  cat("**New outliers introduced under WGS genotype:**\n")
  cat(paste("-", new_outliers, collapse = "\n"), "\n\n")
}
if (length(persisted) > 0) {
  cat("**Persistent outliers (flagged by both methods \u2014 likely true biology):**\n")
  cat(paste("-", persisted, collapse = "\n"), "\n\n")
}

write_xlsx(apoe_wgs_outliers, "APOE4_minus_APOE_WGS_outliers.xlsx")
cat("*WGS-based outliers written to `APOE4_minus_APOE_WGS_outliers.xlsx`*\n")
```

# APOE.geno_final — Primary Analysis

```{r}
#| echo: false
#| label: final-geno-setup
#| results: asis

# ---- Build APOEgeno_final factor (same cleaning as APOEgeno) ----
final_clean <- filtered_combined_post_QC$APOE.geno_final |>
  trimws() |>
  gsub("/", "", x = _) |>
  gsub("\\s+", "", x = _)
final_clean[final_clean == ""] <- NA

filtered_combined_post_QC$APOEgeno_final <- factor(final_clean, levels = apoe_levels)

n_total         <- nrow(filtered_combined_post_QC)
n_final         <- sum(!is.na(filtered_combined_post_QC$APOEgeno_final))
n_wgs_corrected <- sum(filtered_combined_post_QC$APOE_WGS_changed, na.rm = TRUE)
n_orig_only     <- sum(!is.na(filtered_combined_post_QC$APOEgeno_final) &
                          is.na(filtered_combined_post_QC$APOEgeno_WGS))

cat(sprintf("- Total samples: **%d**\n", n_total))
cat(sprintf("- Samples with APOE.geno_final: **%d** (%.1f%%)\n",
            n_final, 100 * n_final / n_total))
cat(sprintf("- Genotypes from WGS (corrected or new): **%d**\n",
            sum(!is.na(filtered_combined_post_QC$APOEgeno_WGS))))
cat(sprintf("- Genotypes corrected by WGS: **%d**\n", n_wgs_corrected))
cat(sprintf("- Genotypes from original only (no WGS): **%d**\n\n", n_orig_only))

cat("### Genotype distribution (APOE.geno_final)\n\n")
geno_final_tbl <- as.data.frame(table(Genotype = filtered_combined_post_QC$APOEgeno_final, useNA = "ifany"))
names(geno_final_tbl) <- c("Genotype", "N")
print(knitr::kable(geno_final_tbl, row.names = FALSE))
```

```{r}
#| echo: false
#| label: final-geno-boxplot

# ---- APOE4_minus_APOE by APOE.geno_final ----
p_final <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno_final), !is.na(APOE4_minus_APOE)) %>%
  ggplot(aes(x = APOEgeno_final, y = APOE4_minus_APOE, fill = APOEgeno_final)) +
  geom_violin(scale = "width", alpha = 0.6, width = 0.65) +
  stat_summary(fun = median, geom = "crossbar", width = 0.35, linewidth = 0.3, colour = "grey20") +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.5) +
  labs(
    x     = "APOE Genotype (final)",
    y     = "APOE4 \u2212 APOE",
    title = "APOE4 \u2212 APOE by APOE.geno_final\n(WGS-corrected where available, original otherwise)"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5))

# ---- Highlight WGS-corrected samples in orange ----
wgs_changed_pts <- filtered_combined_post_QC %>%
  filter(APOE_WGS_changed == TRUE, !is.na(APOEgeno_final), !is.na(APOE4_minus_APOE)) %>%
  mutate(orig_digits = gsub("/", "", as.character(APOEgeno)))

p_final_annotated <- p_final +
  geom_point(
    data        = wgs_changed_pts,
    aes(x = APOEgeno_final, y = APOE4_minus_APOE),
    color       = "darkorange", size = 2.5, inherit.aes = FALSE
  ) +
  geom_text_repel(
    data        = wgs_changed_pts,
    aes(x = APOEgeno_final, y = APOE4_minus_APOE,
        label = paste0(SAMPLE_ALIQUOT, "\n(", orig_digits, "\u2192", trimws(APOE_WGS), ")")),
    size        = 2.5, color = "darkorange", max.overlaps = 20, inherit.aes = FALSE
  ) +
  labs(title = "APOE4 \u2212 APOE by APOE.geno_final\nWGS-corrected samples in orange (orig\u2192WGS)")

p_final_annotated
```

```{r}
#| echo: false
#| label: final-geno-outliers
#| results: asis

# ---- Outlier detection using APOE.geno_final ----
final_bounds <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno_final), !is.na(APOE4_minus_APOE)) %>%
  group_by(APOEgeno_final) %>%
  summarise(
    Q1    = quantile(APOE4_minus_APOE, 0.25, na.rm = TRUE),
    Q3    = quantile(APOE4_minus_APOE, 0.75, na.rm = TRUE),
    IQR   = Q3 - Q1,
    lower = Q1 - 3 * IQR,
    upper = Q3 + 3 * IQR,
    .groups = "drop"
  )

apoe_final_outliers <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno_final), !is.na(APOE4_minus_APOE)) %>%
  left_join(final_bounds, by = "APOEgeno_final") %>%
  filter(APOE4_minus_APOE < lower | APOE4_minus_APOE > upper)

# ---- Compare to original outliers ----
orig_ids     <- apoe_outliers$SAMPLE_ALIQUOT
final_ids    <- apoe_final_outliers$SAMPLE_ALIQUOT
resolved     <- setdiff(orig_ids, final_ids)
new_outliers <- setdiff(final_ids, orig_ids)
persisted    <- intersect(orig_ids, final_ids)

# ---- Annotate outlier sets with WGS status ----
outlier_meta <- filtered_combined_post_QC %>%
  select(SAMPLE_ALIQUOT, APOEgeno, APOEgeno_WGS, APOEgeno_final, APOE_WGS_changed) %>%
  mutate(has_wgs = !is.na(APOEgeno_WGS))

# Resolved: split into WGS-corrected vs. shifted IQR boundary (no WGS change)
resolved_meta    <- outlier_meta %>% filter(SAMPLE_ALIQUOT %in% resolved)
resolved_by_wgs  <- resolved_meta %>% filter(APOE_WGS_changed == TRUE)  %>% pull(SAMPLE_ALIQUOT)
resolved_no_wgs  <- resolved_meta %>% filter(APOE_WGS_changed != TRUE | is.na(APOE_WGS_changed)) %>% pull(SAMPLE_ALIQUOT)

# Persistent: three sub-categories
persisted_meta      <- outlier_meta %>% filter(SAMPLE_ALIQUOT %in% persisted)
p_wgs_agreed        <- persisted_meta %>% filter(has_wgs, APOE_WGS_changed != TRUE) %>% pull(SAMPLE_ALIQUOT)
p_wgs_still_outlier <- persisted_meta %>% filter(APOE_WGS_changed == TRUE)           %>% pull(SAMPLE_ALIQUOT)
p_no_wgs            <- persisted_meta %>% filter(!has_wgs)                            %>% pull(SAMPLE_ALIQUOT)

cat("### APOE.geno_final Outlier Summary (3\u00d7IQR)\n\n")
cat("| | Count |\n|---|---|\n")
cat(sprintf("| Original genotype outliers | %d |\n", length(orig_ids)))
cat(sprintf("| APOE.geno_final outliers | %d |\n", length(final_ids)))
cat(sprintf("| Resolved total | %d |\n", length(resolved)))
cat(sprintf("| \u00a0\u00a0 \u2514 resolved by WGS genotype correction | %d |\n", length(resolved_by_wgs)))
cat(sprintf("| \u00a0\u00a0 \u2514 resolved for other reasons (IQR shift, no WGS change) | %d |\n", length(resolved_no_wgs)))
cat(sprintf("| New outliers introduced | %d |\n", length(new_outliers)))
cat(sprintf("| Persistent total | %d |\n", length(persisted)))
cat(sprintf("| \u00a0\u00a0 \u2514 WGS confirms genotype, still outlier | %d |\n", length(p_wgs_agreed)))
cat(sprintf("| \u00a0\u00a0 \u2514 WGS corrected genotype but still outlier | %d |\n", length(p_wgs_still_outlier)))
cat(sprintf("| \u00a0\u00a0 \u2514 no WGS yet (trivially persistent \u2014 genotype unchanged) | %d |\n\n", length(p_no_wgs)))

if (length(resolved) > 0) {
  cat("**Resolved outliers:**\n\n")
  if (length(resolved_by_wgs) > 0) {
    cat(sprintf("  *Resolved by WGS genotype correction (%d) \u2014 WGS moved them to a better-fitting group:*\n",
                length(resolved_by_wgs)))
    cat(paste("  -", resolved_by_wgs, collapse = "\n"), "\n\n")
  }
  if (length(resolved_no_wgs) > 0) {
    cat(sprintf("  *Resolved without WGS change (%d) \u2014 IQR boundary shift or no WGS data:*\n",
                length(resolved_no_wgs)))
    cat(paste("  -", resolved_no_wgs, collapse = "\n"), "\n\n")
  }
}

if (length(new_outliers) > 0) {
  cat("**New outliers under APOE.geno_final:**\n")
  cat(paste("-", new_outliers, collapse = "\n"), "\n\n")
}

if (length(persisted) > 0) {
  cat("**Persistent outliers \u2014 sub-categories:**\n\n")
  if (length(p_wgs_agreed) > 0) {
    cat(sprintf("  *WGS confirms genotype, still extreme (%d) \u2014 strongest signal for true biology or QC concern:*\n",
                length(p_wgs_agreed)))
    cat(paste("  -", p_wgs_agreed, collapse = "\n"), "\n\n")
  }
  if (length(p_wgs_still_outlier) > 0) {
    cat(sprintf("  *WGS corrected genotype but still outlier (%d) \u2014 investigate further:*\n",
                length(p_wgs_still_outlier)))
    cat(paste("  -", p_wgs_still_outlier, collapse = "\n"), "\n\n")
  }
  if (length(p_no_wgs) > 0) {
    cat(sprintf("  *No WGS data yet (%d) \u2014 trivially persistent (genotype unchanged); flag for future WGS follow-up:*\n",
                length(p_no_wgs)))
    cat(paste("  -", p_no_wgs, collapse = "\n"), "\n\n")
  }
}

# ---- Ranked table ----
final_medians <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno_final), !is.na(APOE4_minus_APOE)) %>%
  group_by(APOEgeno_final) %>%
  summarise(genotype_median = median(APOE4_minus_APOE, na.rm = TRUE), .groups = "drop")

apoe_final_ranked <- filtered_combined_post_QC %>%
  filter(!is.na(APOEgeno_final), !is.na(APOE4_minus_APOE)) %>%
  left_join(final_medians, by = "APOEgeno_final") %>%
  mutate(deviation = APOE4_minus_APOE - genotype_median,
         abs_deviation = abs(deviation)) %>%
  arrange(desc(abs_deviation))

# ---- Overlay outliers on p_final ----
p_final_outliers <- p_final +
  geom_point(
    data = apoe_final_outliers,
    aes(x = APOEgeno_final, y = APOE4_minus_APOE),
    color = "red", size = 2, inherit.aes = FALSE
  ) +
  geom_text_repel(
    data = apoe_final_outliers,
    aes(x = APOEgeno_final, y = APOE4_minus_APOE, label = SAMPLE_ALIQUOT),
    size = 3, color = "red", max.overlaps = 20, inherit.aes = FALSE
  ) +
  labs(title = "APOE4 \u2212 APOE by APOE.geno_final \u2014 Extreme Outliers in Red")

p_final_outliers

# ---- Write Excel ----
write_xlsx(apoe_final_outliers, "APOE4_minus_APOE_final_outliers.xlsx")
write_xlsx(apoe_final_ranked,   "Outliers_Ranked_Final.xlsx")
cat("*APOE.geno_final outliers written to `APOE4_minus_APOE_final_outliers.xlsx` and `Outliers_Ranked_Final.xlsx`*\n")
```

# Sanger vs WGS: Which Agrees with the Biomarker?

`APOE4 − APOE` is essentially a molecular readout of **ε4 dosage** (the APOE4 assay
detects the ε4 isoform). Within concordant samples the three dosage groups are cleanly
separated, so for any sample where the original **Sanger** genotype (`APOE.geno`) and the
**WGS** genotype (`APOE_WGS`) disagree, we can ask which call the protein supports. The
biomarker reads ε4 *count* only — it cannot adjudicate ε2-vs-ε3 differences.

```{r}
#| echo: false
#| label: sw-setup

e4_count <- function(g) ifelse(is.na(g) | g == "", NA_integer_,
                               lengths(regmatches(g, gregexpr("4", g))))

sw <- filtered_combined_post_QC %>%
  mutate(
    sanger    = dplyr::na_if(gsub("[/ ]", "", trimws(APOE.geno)), ""),
    wgs       = dplyr::na_if(gsub("[/ ]", "", trimws(APOE_WGS)),  ""),
    samp_num  = suppressWarnings(as.numeric(sub("-.*", "", SAMPLE_ALIQUOT))),
    e4_sanger = e4_count(dplyr::na_if(gsub("[/ ]", "", trimws(APOE.geno)), "")),
    e4_wgs    = e4_count(dplyr::na_if(gsub("[/ ]", "", trimws(APOE_WGS)),  "")),
    call_category = dplyr::case_when(
      !is.na(sanger) & !is.na(wgs) & sanger == wgs ~ "Both agree",
      !is.na(sanger) & !is.na(wgs) & sanger != wgs ~ "Both disagree",
      !is.na(sanger) &  is.na(wgs)                 ~ "Sanger only",
       is.na(sanger) & !is.na(wgs)                 ~ "WGS only",
      TRUE                                          ~ "Neither"
    )
  )

# Reference ε4-dosage medians AND spreads from high-confidence concordant samples
ref_med <- sw %>%
  filter(call_category == "Both agree", !is.na(APOE4_minus_APOE)) %>%
  group_by(e4 = e4_sanger) %>%
  summarise(med_dm = median(APOE4_minus_APOE, na.rm = TRUE),
            iqr_dm = IQR(APOE4_minus_APOE, na.rm = TRUE), .groups = "drop") %>%
  arrange(e4)
med_lookup <- setNames(ref_med$med_dm, as.character(ref_med$e4))
iqr_lookup <- setNames(ref_med$iqr_dm, as.character(ref_med$e4))

# "Fits neither band" threshold: a sample whose APOE4−APOE is farther than
# DEV_MULT × (band IQR) from EVERY reference band fits no expected ε4 dosage —
# a candidate sample switch / assay problem independent of which genotype is right.
DEV_MULT <- 3

# Biomarker-implied ε4 dosage = nearest reference median
bio_e4 <- function(x) {
  out <- rep(NA_integer_, length(x))
  ok  <- !is.na(x)
  if (any(ok)) {
    dmat <- abs(outer(x[ok], ref_med$med_dm, "-"))
    out[ok] <- ref_med$e4[max.col(-dmat, ties.method = "first")]
  }
  out
}

sw <- sw %>%
  mutate(
    e4_bio     = bio_e4(APOE4_minus_APOE),
    dev_sanger = abs(APOE4_minus_APOE - med_lookup[as.character(e4_sanger)]),
    dev_wgs    = abs(APOE4_minus_APOE - med_lookup[as.character(e4_wgs)]),
    # deviation from the NEAREST band (e4_bio) and that band's IQR
    dev_nearest = abs(APOE4_minus_APOE - med_lookup[as.character(e4_bio)]),
    fits_neither = !is.na(dev_nearest) & !is.na(e4_bio) &
                   dev_nearest > DEV_MULT * iqr_lookup[as.character(e4_bio)],
    biomarker_favors = dplyr::case_when(
      is.na(dev_sanger) | is.na(dev_wgs) ~ NA_character_,
      e4_sanger == e4_wgs                ~ "tie (same ε4)",
      dev_sanger < dev_wgs               ~ "Sanger",
      dev_wgs    < dev_sanger            ~ "WGS",
      TRUE                                ~ "tie"
    )
  )

sw_both <- sw %>% filter(call_category %in% c("Both agree", "Both disagree"))
sw_disc <- sw %>% filter(call_category == "Both disagree")
```

## Does Sanger or WGS fit the biomarker?

```{r}
#| echo: false
#| label: sw-fit-test
#| results: asis

cat("Reference ε4-dosage medians of APOE4−APOE (concordant samples):\n\n")
print(knitr::kable(ref_med[, c("e4", "med_dm")],
                   col.names = c("ε4 alleles", "median APOE4−APOE"), digits = 2))

cat(sprintf("\n\n**Among %d discordant samples, which call does the biomarker support?**\n\n",
            nrow(sw_disc)))
ft <- with(sw_disc, table(
  `Comparison`       = ifelse(e4_sanger != e4_wgs, "ε4 dosage differs", "ε2/ε3 only (same ε4)"),
  `Biomarker favors` = biomarker_favors))
print(knitr::kable(as.data.frame.matrix(ft)))

n_san <- sum(sw_disc$biomarker_favors == "Sanger", na.rm = TRUE)
n_wgs <- sum(sw_disc$biomarker_favors == "WGS",    na.rm = TRUE)
cat(sprintf(
  "\n\nWhere the two calls imply a different ε4 dosage, the biomarker matches **Sanger in %d** vs **WGS in %d**. The remaining discordances are ε2/ε3 differences the biomarker cannot resolve.\n\n",
  n_san, n_wgs))
```

```{r}
#| echo: false
#| label: sw-category-plot
#| fig-width: 9
#| fig-height: 6

cat_levels <- c("Both agree", "Both disagree", "Sanger only", "WGS only")
cat_cols   <- c("Both agree" = "grey75", "Both disagree" = "#E41A1C",
                "Sanger only" = "#377EB8", "WGS only" = "#4DAF4A")

plot_df <- sw %>%
  filter(!is.na(sanger), !is.na(APOE4_minus_APOE)) %>%
  mutate(sanger        = factor(sanger, levels = apoe_levels),
         call_category = factor(call_category, levels = cat_levels))

ggplot(plot_df, aes(x = sanger, y = APOE4_minus_APOE)) +
  geom_jitter(aes(color = call_category, size = call_category, alpha = call_category),
              width = 0.22, height = 0) +
  scale_color_manual(values = cat_cols, drop = FALSE) +
  scale_size_manual(values = c("Both agree" = 1, "Both disagree" = 2.4,
                               "Sanger only" = 1.5, "WGS only" = 1.5), drop = FALSE) +
  scale_alpha_manual(values = c("Both agree" = 0.3, "Both disagree" = 0.95,
                                "Sanger only" = 0.6, "WGS only" = 0.6), drop = FALSE) +
  labs(x = "Sanger (original) genotype", y = "APOE4 − APOE",
       color = "Call category", size = "Call category", alpha = "Call category",
       title = "APOE4 − APOE by Sanger genotype, colored by Sanger/WGS agreement",
       subtitle = "Discordant points (red) sitting in a band that matches their Sanger genotype indicate WGS is the wrong call") +
  theme_classic(base_size = 13)
```

## Samples that fit neither ε4 band

Independent of which genotype is "right," a sample whose `APOE4 − APOE` sits more
than `r DEV_MULT`× the band IQR from **every** reference ε4 band fits no expected
dosage — a candidate sample switch or assay problem. These are flagged across all
dual-typed samples (not just discordant ones).

```{r}
#| echo: false
#| label: sw-fits-neither
#| results: asis

cat("### Reference band spreads (IQR) used for the threshold\n\n")
print(knitr::kable(ref_med, col.names = c("ε4 alleles", "median APOE4−APOE", "IQR"), digits = 2))

fn <- sw %>%
  filter(call_category %in% c("Both agree", "Both disagree", "Sanger only", "WGS only"),
         !is.na(fits_neither))
cat(sprintf("\n\n**%d** of %d typed samples fit neither ε4 band (> %d× band IQR from the nearest band).\n\n",
            sum(fn$fits_neither), nrow(fn), DEV_MULT))
print(knitr::kable(
  fn %>% group_by(`Call category` = call_category) %>%
    summarise(n = n(), fits_neither = sum(fits_neither),
              pct = round(100 * mean(fits_neither), 1), .groups = "drop")))
cat("\n\nThese are exported in the flag tables below (`fits_neither = TRUE`).\n\n")
```

## Which genotype is most error-prone?

```{r}
#| echo: false
#| label: sw-worst-genotype
#| results: asis

cat("### Discordance by Sanger genotype\n\n")
wg <- sw_both %>%
  mutate(Sanger = factor(sanger, levels = apoe_levels)) %>%
  group_by(Sanger) %>%
  summarise(
    n_dual   = n(),
    n_disc   = sum(call_category == "Both disagree"),
    pct_disc = round(100 * n_disc / n_dual, 1),
    biomarker_backs_Sanger = sum(call_category == "Both disagree" & biomarker_favors == "Sanger", na.rm = TRUE),
    biomarker_backs_WGS    = sum(call_category == "Both disagree" & biomarker_favors == "WGS",    na.rm = TRUE),
    .groups = "drop") %>%
  filter(!is.na(Sanger))
print(knitr::kable(wg))

cat("\n\n### Discordance confusion matrix (Sanger → WGS), discordant only\n\n")
cm <- table(Sanger = sw_disc$sanger, WGS = sw_disc$wgs)
print(knitr::kable(as.data.frame.matrix(cm)))
cat("\n\nThe 33↔34 boundary (one ε4 allele) dominates — the hardest single call and the bulk of discordance.\n\n")

# Direction of error: does WGS systematically UNDER-call the ε4 (non-reference)
# allele, as European-reference mapping bias would predict, or are errors symmetric?
cat("### Direction of the ε4 error (reference-bias test)\n\n")
dir_df <- sw_disc %>% filter(!is.na(e4_sanger), !is.na(e4_wgs), e4_sanger != e4_wgs)
n_under <- sum(dir_df$e4_wgs < dir_df$e4_sanger)   # WGS calls fewer ε4 than Sanger
n_over  <- sum(dir_df$e4_wgs > dir_df$e4_sanger)   # WGS calls more  ε4 than Sanger
cat(sprintf("Among the %d discordants whose ε4 *dosage* differs, WGS calls **fewer** ε4 alleles than Sanger in **%d** and **more** in **%d**.\n\n",
            nrow(dir_df), n_under, n_over))
print(knitr::kable(
  dir_df %>% count(`Sanger ε4` = e4_sanger, `WGS ε4` = e4_wgs, name = "n") %>%
    arrange(desc(n))))
cat("\n\nThe split is essentially symmetric and the rs429358 flips run both ways, so this is **not** a one-directional reference-allele dropout — it looks like unreliable calling at a hard locus (and/or sample mis-linkage), not a systematic European-reference bias.\n\n")
```

## Do the wrong WGS calls look random? (mix-up test)

::: {.callout-note appearance="simple"}
**Superseded / confirmed.** This test correctly inferred a **block-level sample
mix-up** ("~30% of labels randomized") rather than directional miscalling — the
WGS plate-swap diagnosis confirmed exactly that. Its open question, mis-linkage (1)
vs admixed-calling-noise (2), is now resolved in favor of **(1)**: the plate-2
rotation and 3↔7 swap. On the corrected data below the discordant pool is the ~12
residual, so the statistics are degenerate; see the pre-correction version in
`output_files_pre_plateswap_2026Jun/`.
:::

If a plate's labels were scrambled (a sample / manifest mix-up), each wrong WGS
genotype should look like an independent **random draw** from the population
genotype distribution, *independent of the true genotype*. We test that against
the observed confusion matrix, and compare a genotype-distance statistic: structured
single-SNP **miscalling** would suppress two-allele-substitution ("distance-2")
errors, whereas random reassignment would not.

```{r}
#| echo: false
#| label: sw-mixup-test
#| results: asis

set.seed(1)
genos5 <- c("23", "24", "33", "34", "44")
pool <- sw %>%
  filter(call_category %in% c("Both agree", "Both disagree"),
         (if ("Ethnicity" %in% names(.)) Ethnicity == "HI" else TRUE),
         sanger %in% genos5, wgs %in% genos5)
pi5  <- prop.table(table(factor(pool$sanger, levels = genos5)))
disc5 <- pool %>% filter(sanger != wgs)
D5   <- nrow(disc5)

O <- table(Sanger = factor(disc5$sanger, genos5), WGS = factor(disc5$wgs, genos5))
Dh <- rowSums(O)
E  <- matrix(0, 5, 5, dimnames = list(genos5, genos5))
for (h in genos5) { if (Dh[h] == 0) next
  pr <- pi5; pr[h] <- 0; pr <- pr / sum(pr); E[h, ] <- Dh[h] * pr }

offd <- which(row(O) != col(O)); Ev <- as.vector(E)[offd]; keep <- Ev > 0
Gstat <- function(v) 2 * sum(ifelse(v[keep] > 0, v[keep] * log(v[keep] / Ev[keep]), 0))
G_obs <- Gstat(as.vector(O)[offd])

gdist <- function(a, b) { aa <- sort(as.integer(strsplit(a, "")[[1]]))
  bb <- as.integer(strsplit(b, "")[[1]])
  min(sum(aa != sort(bb)), sum(aa != sort(rev(bb)))) }
dist_mat <- outer(genos5, genos5, Vectorize(gdist)); dimnames(dist_mat) <- list(genos5, genos5)
obs_d2 <- sum(O * (dist_mat == 2)) / D5

B <- 10000; sim_G <- numeric(B); sim_d2 <- numeric(B); th <- disc5$sanger
for (b in 1:B) {
  sw5 <- character(D5)
  for (h in unique(th)) { idx <- which(th == h)
    pr <- pi5; pr[h] <- 0; pr <- pr / sum(pr)
    sw5[idx] <- sample(genos5, length(idx), TRUE, pr) }
  Os <- table(factor(th, genos5), factor(sw5, genos5))
  sim_G[b]  <- Gstat(as.vector(Os)[offd])
  sim_d2[b] <- sum(Os * (dist_mat == 2)) / D5
}
p_G <- mean(sim_G >= G_obs)
r_max <- 1 - sum(pi5^2); f_hat <- (D5 / nrow(pool)) / r_max

cat("### Observed vs expected confusion under random reassignment\n\n")
side <- cbind(as.data.frame.matrix(O), `|` = "|", as.data.frame.matrix(round(E, 1)))
print(knitr::kable(side, caption = "Left: observed (Sanger→WGS). Right: expected if WGS genotypes were random draws."))

cat(sprintf("\n\n**Monte Carlo goodness-of-fit to the random-mix-up model: p = %.2f** (B = %d). ", p_G, B))
cat("A large p-value means the discordant calls are statistically *indistinguishable from random reassignment* — no structure beyond random draws from the population genotype frequencies.\n\n")
cat(sprintf("**Distance-2 (two-allele-substitution) error share:** observed %.1f%% vs random-null %.1f%% [%.1f–%.1f%%]. ",
            100*obs_d2, 100*mean(sim_d2), 100*quantile(sim_d2,.025), 100*quantile(sim_d2,.975)))
cat("Structured single-SNP miscalling would push this *below* the band; it sits inside it, so reference-bias miscalling is ruled out.\n\n")
cat(sprintf("Under partial random permutation of the pool, the observed rate implies **~%.0f%%** of WGS labels would be randomized — a large, block-level effect, not a handful of isolated miscalls.\n\n", 100*f_hat))
cat("**Caveat:** a *random-looking* pattern is produced both by (a) scrambled sample/label assignment and (b) calls so poor for these admixed samples that the output is effectively noise. This test cannot separate those two — but it decisively excludes structured, directional miscalling. The per-sample read depth / genotype-quality pull (see top summary) distinguishes (a) from (b).\n\n")
```

## Global scramble or group-specific? The 0-of-284 test

::: {.callout-note appearance="simple"}
**Superseded.** This test rightly excluded a *delivery-wide* scramble, but read the
Hispanic-only concentration as an **ancestry-targeted** process. The real reason the
errors were Hispanic-only is **plate composition**: the mislabeled plates (2/3/7)
were the AD-Hispanic batch, so the non-Hispanic samples — which sat on clean plates —
were correctly 0/284. It is plate-confined, not ancestry-targeted. On corrected data
both groups are now ~0.
:::

A *random-looking* error pattern within the Hispanic pool is consistent with a
sample/manifest mix-up — but is the **whole delivery** scrambled, or only the
admixed samples? If a plate's labels were globally swapped, every ancestry on
that delivery should be hit equally. We test this on the **newly-added** WGS
genotypes (the n2871 May-2026 delivery), split by ancestry.

```{r}
#| echo: false
#| label: sw-global-vs-group
#| results: asis
#| fig-width: 9
#| fig-height: 5

# Identify newly-added WGS using the prior (Jan-2026) WGS lookup
old_wgs <- tryCatch(
  read.csv("U19_Alamar_metadata_2026Jan16_with_WGSAPOE.csv"),
  error = function(e) NULL)

if (!is.null(old_wgs)) {
  old_col <- intersect(c("APOE_WGS", "APOE_WGS_norm", "WGS_APOE"), names(old_wgs))[1]
  old_ids <- old_wgs %>% filter(!is.na(.data[[old_col]])) %>%
    pull(SAMPLE_ALIQUOT) %>% unique()

  na <- sw %>%
    filter(call_category %in% c("Both agree", "Both disagree"),
           !(SAMPLE_ALIQUOT %in% old_ids)) %>%
    mutate(disc = call_category == "Both disagree")

  by_eth <- na %>% group_by(Ethnicity) %>%
    summarise(n = n(), discordant = sum(disc),
              pct = round(100 * mean(disc), 1), .groups = "drop") %>%
    arrange(desc(pct))
  cat("### Newly-added WGS discordance by ethnicity\n\n")
  print(knitr::kable(by_eth))

  hi_rate <- with(na, sum(disc[Ethnicity == "HI"]) / sum(Ethnicity == "HI"))
  n_nh    <- sum(na$Ethnicity == "NH")
  obs_nh  <- sum(na$disc[na$Ethnicity == "NH"])
  cat(sprintf(paste0("\n\n**The decisive test.** A delivery-wide random scramble ",
    "would hit non-Hispanic samples at the same rate. At the Hispanic rate ",
    "(%.1f%%), the %d non-Hispanic newly-added samples should yield **~%.0f ",
    "discordant** — observed: **%d**. A global lab swap is excluded; the errors ",
    "are ancestry-targeted.\n\n"),
    100 * hi_rate, n_nh, hi_rate * n_nh, obs_nh))

  # ── Admixture gradient plot: discordance by country of birth ──────────────
  if ("country_of_birth" %in% names(na)) {
    grad <- na %>%
      group_by(country = country_of_birth) %>%
      summarise(n = n(), discordant = sum(disc),
                pct = round(100 * mean(disc), 1), .groups = "drop") %>%
      filter(n >= 10, !is.na(country)) %>%
      arrange(pct)

    grad <- grad %>%
      mutate(country = factor(country, levels = country),
             grp = case_when(
               country %in% c("USA", "JAM", "HTI") ~ "USA / Afro-Caribbean",
               TRUE ~ "Latin-American (admixed)"))

    p_grad <- ggplot(grad, aes(x = country, y = pct, fill = grp)) +
      geom_col(width = 0.7) +
      geom_text(aes(label = sprintf("%.0f%% (%d/%d)", pct, discordant, n)),
                hjust = -0.1, size = 3.1) +
      coord_flip() +
      scale_fill_manual(values = c("Latin-American (admixed)" = "#D55E00",
                                   "USA / Afro-Caribbean" = "#0072B2")) +
      expand_limits(y = max(grad$pct) * 1.18) +
      labs(title = "(Pre-correction) Apparent admixture gradient — an artifact of plate composition",
           subtitle = "Pre-diagnosis framing. The mislabeled WGS plates were the AD-Hispanic batch; on corrected data this gradient is flat.",
           x = "Country of birth (n ≥ 10)", y = "Sanger-vs-WGS discordance (%)",
           fill = NULL) +
      theme_bw(base_size = 11) + theme(legend.position = "top")
    print(p_grad)
  }
} else {
  cat("*(Prior WGS lookup `U19_Alamar_metadata_2026Jan16_with_WGSAPOE.csv` not found; skipping newly-added split.)*\n\n")
}
```

The gradient is the signature of **ancestry-specific calling failure**: WGS APOE
accuracy degrades smoothly as Amerindian/admixed ancestry increases (Colombia,
Peru, Mexico at the top; USA-born and Afro-Caribbean at 0%), rather than failing
uniformly as a manifest swap would. Whether the residual is pure calling noise or
a mix-up *confined* to the admixed subset is still settled by the DP/GQ pull — but
a generic, delivery-wide lab swap is off the table.

## (Superseded) Apparent ancestry association — explained by plate composition

::: {.callout-note appearance="simple"}
**Superseded.** The discordance is **not** driven by ancestry. It is concentrated in
Hispanic samples only because the three mislabeled WGS plates (2/3/7) *were* the
AD-Hispanic batch — plate composition, not biology or "hard-to-call admixed
haplotypes." The tables below recompute on corrected data and are now ~flat across
every ancestry/origin grouping. Retained as the record of the (mistaken) working
hypothesis; see `../../../May_2026_WGS_QC/WGS_Plate_Swap_Summary.md`.
:::

The biomarker RUN date is the **assay** date — it cannot *cause* two genotyping
methods (Sanger, WGS) to disagree with each other. The driver was originally read as
**ancestry**, since Sanger–WGS discordance was overwhelmingly concentrated in
**Hispanic / admixed** samples. We now know that concentration was an artifact of the
WGS plate swap (those subjects loaded onto plates 2/3/7), not a property of ancestry
or of the biomarker assay. The "Oct-2025 run" correlation remains a confound — the
Hispanic cohort was WGS-typed and biomarker-run together.

```{r}
#| echo: false
#| label: sw-ancestry-driver
#| results: asis

disc_rate_by <- function(df, col) {
  if (!col %in% names(df)) return(NULL)
  df %>%
    mutate(.lvl = .data[[col]]) %>%
    group_by(.lvl) %>%
    summarise(n_dual = n(),
              n_disc = sum(call_category == "Both disagree"),
              pct_disc = round(100 * n_disc / n_dual, 1), .groups = "drop") %>%
    arrange(desc(pct_disc)) %>%
    rename(!!col := .lvl)
}

for (col in c("Ethnicity", "Ancestry", "Race", "Site",
              "Country.State", "country_of_birth", "CDX", "metatada_source")) {
  tb <- disc_rate_by(sw_both, col)
  if (!is.null(tb)) {
    cat(sprintf("### Discordance rate by %s\n\n", col))
    # country_of_birth has a long tail; show only levels with >=5 dual-typed
    if (col == "country_of_birth") tb <- dplyr::filter(tb, n_dual >= 5)
    print(knitr::kable(tb))
    cat("\n\n")
  }
}
cat("**Study/cohort:** every discordant sample is `metatada_source = U19_2026Apr30` (the READD/Miami WGS additions; `ID2` present only for these). The discordance is therefore confined to the U19 WGS-typed study, and *within* it tracks Hispanic-Caribbean origin (PRI/DOM/CUB) — i.e. an ancestry-stratified WGS calling problem, not a study-wide genotype swap.\n\n")
```

```{r}
#| echo: false
#| label: sw-confound-check
#| results: asis

# RUN looks predictive only because Oct-2025 runs are Hispanic-heavy. Holding
# ancestry fixed, the "batch" signal should collapse in non-Hispanic samples.
if ("Ethnicity" %in% names(sw_both)) {
  era_tbl <- sw_both %>%
    filter(!is.na(Ethnicity)) %>%
    mutate(era = ifelse(grepl("^202510", RUN), "Oct-2025 runs", "other runs")) %>%
    group_by(Ethnicity, era) %>%
    summarise(n_dual = n(), n_disc = sum(call_category == "Both disagree"),
              pct_disc = round(100 * n_disc / n_dual, 1), .groups = "drop")
  cat("### RUN era stratified by ancestry (the confound, exposed)\n\n")
  print(knitr::kable(era_tbl))
  cat("\n\nAmong **non-Hispanic** samples the Oct-2025 runs are ~as clean as any other run, so the run correlation is a confound for the underlying ancestry effect; the residual elevation within Hispanic samples points to a specific WGS *delivery* batch, not the biomarker assay.\n\n")
}
```

```{r}
#| echo: false
#| label: sw-ancestry-plot
#| fig-width: 8
#| fig-height: 5

if ("Ethnicity" %in% names(sw_both)) {
  ap <- sw_both %>%
    filter(!is.na(Ethnicity)) %>%
    group_by(Ethnicity) %>%
    summarise(pct_disc = 100 * mean(call_category == "Both disagree"),
              n_dual = n(), .groups = "drop")
  ggplot(ap, aes(x = reorder(Ethnicity, pct_disc), y = pct_disc)) +
    geom_col(fill = "#984EA3") +
    geom_text(aes(label = sprintf("%.1f%% (n=%d)", pct_disc, n_dual)), hjust = -0.05) +
    coord_flip() +
    expand_limits(y = max(ap$pct_disc) * 1.25) +
    labs(x = NULL, y = "% discordant (Sanger vs WGS)",
         title = "(Pre-correction) Discordance concentrated in Hispanic samples — a plate artifact",
         subtitle = "Pre-diagnosis framing; the mislabeled plates were the AD-Hispanic batch. On corrected data this is ~flat.") +
    theme_classic(base_size = 12)
}
```

## Three-way arbitration (Sanger × WGS × Biomarker)

```{r}
#| echo: false
#| label: sw-three-way
#| results: asis

cat("Biomarker reads ε4 *dosage* only; pattern = which sources agree on ε4 count.\n\n")
tw <- sw_both %>%
  filter(!is.na(e4_sanger), !is.na(e4_wgs), !is.na(e4_bio)) %>%
  mutate(pattern = dplyr::case_when(
    e4_sanger == e4_wgs & e4_wgs == e4_bio        ~ "All three agree",
    e4_sanger == e4_bio & e4_sanger != e4_wgs     ~ "Sanger + Biomarker (WGS off)",
    e4_wgs    == e4_bio & e4_wgs != e4_sanger     ~ "WGS + Biomarker (Sanger off)",
    e4_sanger == e4_wgs & e4_wgs != e4_bio        ~ "Sanger + WGS agree, Biomarker off",
    TRUE                                           ~ "All three differ"
  )) %>%
  count(pattern, name = "n") %>%
  arrange(desc(n))
print(knitr::kable(tw))
```

## The genuine biomarker batch effect (concordant samples)

A *biomarker* batch problem shows up as the protein disagreeing with a genotype
**both** methods agree on — not as Sanger-vs-WGS discordance. Restricting to
concordant ("Both agree") samples, where the genotype is trustworthy, we ask how
often the biomarker-implied ε4 dosage (`e4_bio`) misses the agreed genotype, by
RUN and Bay.

```{r}
#| echo: false
#| label: sw-biomarker-batch
#| results: asis

conc <- sw %>%
  filter(call_category == "Both agree", !is.na(APOE4_minus_APOE), !is.na(e4_bio)) %>%
  mutate(bio_mismatch = e4_bio != e4_sanger)

cat(sprintf("Concordant samples with biomarker data: **%d** | biomarker-vs-genotype mismatches: **%d** (%.1f%%).\n\n",
            nrow(conc), sum(conc$bio_mismatch, na.rm = TRUE),
            100 * mean(conc$bio_mismatch, na.rm = TRUE)))

cat("### Biomarker-vs-genotype mismatch by RUN (top 12 by mismatch count)\n\n")
br <- conc %>%
  group_by(RUN) %>%
  summarise(n = n(), mismatch = sum(bio_mismatch, na.rm = TRUE),
            pct = round(100 * mean(bio_mismatch, na.rm = TRUE), 1), .groups = "drop") %>%
  arrange(desc(mismatch)) %>% head(12)
print(knitr::kable(br))

if ("Bay" %in% names(conc)) {
  cat("\n\n### Biomarker-vs-genotype mismatch by Bay\n\n")
  bb <- conc %>%
    group_by(Bay) %>%
    summarise(n = n(), mismatch = sum(bio_mismatch, na.rm = TRUE),
              pct = round(100 * mean(bio_mismatch, na.rm = TRUE), 1), .groups = "drop") %>%
    arrange(desc(pct))
  print(knitr::kable(bb))
}

cat("\n\nOne run, **20251124-1407_Bay1**, stands far above the rest (~38% mismatch) with a mild Bay1 gradient — a localized biomarker-assay anomaly, entirely separate from the ancestry-driven genotype discordance above.\n\n")

# Per-sample export of the biomarker-batch outliers for follow-up
bio_batch_flagged <- conc %>%
  filter(bio_mismatch) %>%
  select(SAMPLE_ALIQUOT, SAMPLE, RUN, any_of("Bay"), CDX, any_of("ID2"),
         any_of(c("Ethnicity", "Ancestry", "Race", "Site",
                  "Country.State", "country_of_birth", "metatada_source")),
         sanger, e4_agreed = e4_sanger, e4_bio,
         APOE4_minus_APOE, dev_nearest, fits_neither) %>%
  arrange(RUN, SAMPLE_ALIQUOT)
write_xlsx(bio_batch_flagged, "Biomarker_batch_mismatch_flags.xlsx")
cat(sprintf("*Per-sample biomarker-batch mismatches written to `Biomarker_batch_mismatch_flags.xlsx` (%d samples).*\n\n", nrow(bio_batch_flagged)))
```

### What is happening on `20251124-1407_Bay1`?

The mismatch is computed from `APOE4 − APOE`, so it must come from one of the two
proteins drifting. We plot the per-run mismatch rate (to confirm the run is a true
outlier), then ask which protein breaks and whether the rest of the panel is
affected.

```{r}
#| echo: false
#| label: sw-batch-run-plot
#| fig-width: 10
#| fig-height: 5.5

TARGET <- "20251124-1407_Bay1"

run_mm <- conc %>% group_by(RUN) %>%
  summarise(n = n(), mismatch = sum(bio_mismatch),
            pct = 100 * mean(bio_mismatch), .groups = "drop") %>%
  filter(n >= 15) %>% arrange(pct) %>%
  mutate(RUN = factor(RUN, levels = RUN), hl = RUN == TARGET)

pRun <- ggplot(run_mm, aes(RUN, pct, fill = hl)) +
  geom_col(width = 0.75) +
  geom_text(data = subset(run_mm, hl),
            aes(label = sprintf("%.0f%% (%d/%d)", pct, mismatch, n)),
            hjust = -0.1, size = 3.2, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c(`FALSE` = "grey70", `TRUE` = "#E41A1C"), guide = "none") +
  expand_limits(y = max(run_mm$pct) * 1.15) +
  labs(title = "Mismatch rate by run (concordant samples)",
       subtitle = paste0(TARGET, " stands far above every other run (n ≥ 15)."),
       x = NULL, y = "Mismatch rate (%)") +
  theme_bw(base_size = 10) + theme(axis.text.y = element_text(size = 6))

bay_mm <- conc %>% filter(!is.na(Bay)) %>% group_by(Bay) %>%
  summarise(n = n(), mismatch = sum(bio_mismatch),
            pct = 100 * mean(bio_mismatch), .groups = "drop")
pBay <- ggplot(bay_mm, aes(Bay, pct, fill = Bay)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%% (%d/%d)", pct, mismatch, n)),
            vjust = -0.3, size = 3) +
  expand_limits(y = max(bay_mm$pct) * 1.3) +
  labs(title = "Mild Bay gradient (all runs pooled)", x = NULL, y = "Mismatch (%)") +
  theme_bw(base_size = 10) + theme(legend.position = "none")

print(pRun + pBay + patchwork::plot_layout(widths = c(2.2, 1)))
```

```{r}
#| echo: false
#| label: sw-batch-mechanism
#| results: asis
#| fig-width: 9
#| fig-height: 5

tg <- sw %>% filter(!is.na(e4_sanger)) %>%
  mutate(grp = ifelse(RUN == TARGET, TARGET, "All other runs"))

# Dynamic range of APOE4 by carrier status: target vs others
rng <- tg %>% mutate(carrier = ifelse(e4_sanger > 0, "carrier", "non-carrier")) %>%
  group_by(carrier, grp) %>%
  summarise(n = n(), APOE4_med = round(median(APOE4, na.rm = TRUE), 2),
            APOE4_IQR = round(IQR(APOE4, na.rm = TRUE), 2),
            APOE_med = round(median(APOE, na.rm = TRUE), 2),
            APOE_IQR = round(IQR(APOE, na.rm = TRUE), 2), .groups = "drop") %>%
  arrange(carrier, desc(grp))
cat("### APOE4 vs APOE spread on the target run\n\n")
print(knitr::kable(rng))

cat("\n\n**APOE4 loses its dynamic range on this run; APOE is unaffected.** Among",
    "carriers, APOE4's IQR balloons (~1 → ~10 log2 units) and carriers scatter down",
    "into the non-carrier range, while non-carrier APOE4 drifts up off the floor. The",
    "carrier/non-carrier separation that the ε4 inference relies on is erased. Total",
    "APOE (the denominator) stays tight in both groups.\n\n")

ggplot(tg, aes(factor(e4_sanger), APOE4, colour = grp)) +
  geom_violin(scale = "width", position = position_dodge(0.8), width = 0.65, alpha = 0.4) +
  stat_summary(fun = median, geom = "crossbar", width = 0.4, linewidth = 0.3,
               position = position_dodge(0.8), show.legend = FALSE) +
  geom_point(position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.8),
             size = 0.9, alpha = 0.5) +
  scale_colour_manual(values = setNames(c("#E41A1C", "grey55"),
                                        c(TARGET, "All other runs"))) +
  labs(title = "APOE4 by ε4 dosage: the assay loses dynamic range on this run",
       subtitle = "Red boxes (target run) span almost the full range; carriers crash toward non-carrier levels.",
       x = "ε4 allele count (genotype)", y = "APOE4 (log2 NPQ)", colour = NULL) +
  theme_bw(base_size = 11) + theme(legend.position = "top")
```

```{r}
#| echo: false
#| label: sw-batch-panel
#| results: asis
#| fig-width: 9
#| fig-height: 5

# Is the whole panel off, or just APOE4? Residualize each biomarker on ε4 band
# (removes APOE4's genotype-driven bimodality) then compare within-run spread.
meta_drop <- c("APOE4_count", "APOE4_carrier", "APOE_WGS", "APOE_WGS_norm",
               "APOE_WGS_changed", "APOE.geno_final", "BMI", "height_inches",
               "weight_lb", "age_at_subject", "AOO", "Years_Onset", "samp_num")
num_cols <- names(sw)[sapply(sw, is.numeric)]
bm <- setdiff(num_cols, c(meta_drop, "APOE4_minus_APOE", "dev_sanger", "dev_wgs",
                          "dev_nearest", "e4_sanger", "e4_wgs", "e4_bio"))
dr <- sw %>% filter(!is.na(e4_sanger))
is_t <- dr$RUN == TARGET
vr <- sapply(bm, function(c) {
  x <- dr[[c]]; r <- x
  for (g in unique(dr$e4_sanger)) { i <- dr$e4_sanger == g; r[i] <- x[i] - median(x[i], na.rm = TRUE) }
  IQR(r[is_t], na.rm = TRUE) / IQR(r[!is_t], na.rm = TRUE)
})
vrdf <- tibble(bm = bm, ratio = vr) %>% filter(is.finite(ratio), ratio > 0) %>%
  arrange(desc(ratio)) %>% mutate(rank = row_number(), hl = bm %in% c("APOE4", "APOE"))
lab <- vrdf %>% filter(hl)

cat(sprintf("Across the %d-protein panel, APOE4 has the single most inflated within-run variance (%.1f×), ranking **#%d of %d**; APOE is unremarkable (%.1f×, rank %d). The rest of the panel hugs 1.0, so this is **not** a plate-wide failure.\n\n",
            nrow(vrdf), vrdf$ratio[vrdf$bm == "APOE4"], vrdf$rank[vrdf$bm == "APOE4"],
            nrow(vrdf), vrdf$ratio[vrdf$bm == "APOE"], vrdf$rank[vrdf$bm == "APOE"]))

ggplot(vrdf, aes(rank, ratio, colour = hl)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey60") +
  geom_point(size = 1.6, alpha = 0.75) +
  ggrepel::geom_text_repel(data = lab, aes(label = bm), size = 3.4,
                           fontface = "bold", box.padding = 0.6) +
  scale_colour_manual(values = c(`FALSE` = "grey65", `TRUE` = "#E41A1C"), guide = "none") +
  labs(title = "Within-run spread vs the rest of the panel",
       subtitle = "ε4-band-residualized IQR ratio (target / others). APOE4 is the lone outlier.",
       x = "Biomarker (ranked by variance ratio)", y = "IQR ratio (target / others)") +
  theme_bw(base_size = 11)

cat("\n\n**Interpretation.** A single Alamar assay (the APOE4 / ε4-specific reagent",
    "or its panel position) malfunctioned on `20251124-1407_Bay1` — its readout went",
    "noisy and lost dynamic range, while total APOE and the other ~130 proteins were",
    "fine. That is why the genotypes are untouched and the effect is ancestry-blind:",
    "it is a per-target assay failure, not a sample mix-up or a plate-wide drift. The",
    "fix is to flag/repeat APOE4 for that bay; the other proteins on it remain usable.\n\n")
```

## Per-run APOE4 assay QC — a standing screen {#sec-apoe4-run-qc}

::: {.callout-tip appearance="simple"}
**Why this exists.** The `20251124-1407_Bay1` APOE4 failure was *invisible* to the QC
we already run: Alamar's sample-level auto-QC flagged a different bay (Bay2) of that
run, and the pooled control replicates (IPC/SC/NC/HIHG in `Primary_QC` /
`Secondary_QC/Replicate_Analysis`) read **APOE4 normally** on Bay1 — the failure spares
the control wells and hits ~1/3 of the *sample* wells. The check below is what catches
it: for every run-bay, how far its genotyped samples' APOE4 sits from the expected
ε4-dosage band. It runs automatically on every freeze. Deep-dive + evidence:
`../../Secondary_QC/Batch_Effects/investigate_20251124_Bay1.R`.
:::

```{r}
#| echo: false
#| label: per-run-apoe4-qc
#| results: asis

# Genotype-anchored per-run APOE4 assay QC (added 2026-06). Uses APOE.geno_final's
# ε4 dosage (APOE4_count) as the anchor; a healthy assay keeps each sample's APOE4
# near its ε4-band median. Run-bays with many off-band samples or an inflated
# carrier-band spread have an unreliable APOE4 target.
OFF_PCT_THRESH  <- 15     # flag a run-bay if >15% of genotyped samples are off-band
IQR_MULT_THRESH <- 3      # ...or carrier-band APOE4 IQR > 3x the median run-bay

qc_dat <- filtered_combined_post_QC %>%
  filter(!is.na(APOE4_count), !is.na(APOE4))
band_md <- qc_dat %>% group_by(APOE4_count) %>%
  summarise(md = median(APOE4), .groups = "drop")
qc_dat <- qc_dat %>%
  mutate(apoe4_off = abs(APOE4 - band_md$md[match(APOE4_count, band_md$APOE4_count)]) > 2)

run_qc <- qc_dat %>%
  group_by(RUN) %>%
  summarise(n = n(), carriers = sum(APOE4_count >= 1),
            APOE4_carrier_IQR = round(IQR(APOE4[APOE4_count >= 1], na.rm = TRUE), 2),
            off_band_pct = round(100 * mean(apoe4_off), 1), .groups = "drop") %>%
  filter(n >= 15)
med_iqr <- median(run_qc$APOE4_carrier_IQR, na.rm = TRUE)
run_qc <- run_qc %>%
  mutate(FLAG = off_band_pct > OFF_PCT_THRESH |
                APOE4_carrier_IQR > IQR_MULT_THRESH * med_iqr) %>%
  arrange(desc(off_band_pct))

cat(sprintf("Screened **%d** run-bays (n ≥ 15 genotyped). Median off-band **%.1f%%**, median carrier-band APOE4 IQR **%.2f**. Flag rule: off-band > %d%% **or** carrier IQR > %d× median.\n\n",
            nrow(run_qc), median(run_qc$off_band_pct), med_iqr, OFF_PCT_THRESH, IQR_MULT_THRESH))

flagged <- run_qc %>% filter(FLAG)
if (nrow(flagged) > 0) {
  cat(sprintf("**%d run-bay(s) FLAGGED — treat APOE4 (and APOE4 − APOE) as unreliable there:**\n\n", nrow(flagged)))
  print(knitr::kable(flagged,
        col.names = c("Run-Bay","n","carriers","APOE4 carrier IQR","off-band %","FLAG")))
} else cat("No run-bays flagged.\n\n")

cat("\n\nWorst run-bays on the screen:\n\n")
print(knitr::kable(head(run_qc, 8),
      col.names = c("Run-Bay","n","carriers","APOE4 carrier IQR","off-band %","FLAG")))
cat("\n\n*Action for a flagged run-bay: exclude its APOE4 / APOE4−APOE from analysis (the broken subset can't be separated from the rest); total APOE and the other proteins remain usable. This is an assay issue only — genotypes are unaffected.*\n\n")
```

## The November plates: African ancestry or a plate effect?

::: {.callout-note appearance="simple"}
**Partly superseded.** The *genotype-discordance* angle here is moot — the November
plates were never part of the mislabeled WGS set (the all-African plate is
Sanger-only; 1124-Bay1 had no discordant pairs). The **biology check in this section
remains valid and independent of the plate swap**: APOE proteins do **not** read
lower in African-ancestry samples (if anything slightly higher in carrier bands),
and the ε4-dosage separation is preserved across ancestry.
:::

The two November runs were flagged as "all Africans," raising the question of
whether the proteins legitimately read differently in African-ancestry samples
rather than the genotypes being wrong. Two clarifications from the data come
first:

- **`20251125-1332`** is the genuinely all-African plate (246 AFDC: Tanzania,
  Nigeria, Uganda), but it is **Sanger-only — it carries no WGS calls**, so it
  has no genotype discordance to explain.
- **`20251124-1407`** is mostly **Hispanic (HI_MU)**; only its Bay1 carries WGS,
  and that Bay1 is the biomarker-batch outlier from the previous section.

So the genotype-discordance story does not touch the African plate. The only
question left there is biological: do the APOE proteins plot lower for Africans?
We answer with the trustworthy **Sanger** genotype (no WGS required), so the
full African plate is included.

```{r}
#| echo: false
#| label: sw-african-plate
#| results: asis
#| fig-width: 11
#| fig-height: 4.8

afr <- sw %>%
  filter(!is.na(sanger), !is.na(APOE4_minus_APOE), !is.na(e4_sanger)) %>%
  mutate(
    nov       = grepl("^20251124|^20251125", RUN),
    afr_plate = grepl("^20251125", RUN),
    african   = Ancestry %in% c("AFDC", "AA"),
    grp = dplyr::case_when(
      afr_plate          ~ "African plate (20251125)",
      african & !nov     ~ "African, other plates",
      !african & !nov    ~ "Non-African, other plates",
      TRUE               ~ "Other November"),
    e4_lab = paste0(e4_sanger, " ε4")
  )

cat("### Median APOE4 − APOE by ε4 band and group (Sanger genotype)\n\n")
mtab <- afr %>%
  group_by(`ε4` = e4_sanger, group = grp) %>%
  summarise(n = n(), median = round(median(APOE4_minus_APOE), 2), .groups = "drop") %>%
  arrange(`ε4`, group)
print(knitr::kable(mtab))

band_meds <- afr %>% group_by(e4_lab) %>%
  summarise(med = median(APOE4_minus_APOE), .groups = "drop")

p_afr <- ggplot(afr, aes(x = grp, y = APOE4_minus_APOE, colour = grp)) +
  geom_hline(data = band_meds, aes(yintercept = med),
             linetype = "dashed", colour = "grey50") +
  geom_violin(scale = "width", width = 0.6, alpha = 0.6) +
  stat_summary(fun = median, geom = "crossbar", width = 0.32, linewidth = 0.3, colour = "grey20") +
  geom_jitter(width = 0.18, size = 0.9, alpha = 0.45) +
  facet_wrap(~ e4_lab, scales = "free_y") +
  labs(title = "APOE4 − APOE by ε4 dosage: is the African plate lower?",
       subtitle = "Dashed line = overall band median. African plate is NOT lower — slightly higher in carrier bands.",
       x = NULL, y = "APOE4 − APOE (log2)") +
  theme_bw(base_size = 11) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
print(p_afr)
```

```{r}
#| echo: false
#| label: sw-african-pooled
#| fig-width: 8.5
#| fig-height: 5

afr_pool <- afr %>% mutate(anc = ifelse(african, "African (AFDC/AA)", "Non-African"))
ggplot(afr_pool, aes(x = factor(e4_sanger), y = APOE4_minus_APOE, colour = anc)) +
  geom_violin(scale = "width", position = position_dodge(0.8), width = 0.6, alpha = 0.4) +
  stat_summary(fun = median, geom = "crossbar", width = 0.38, linewidth = 0.3,
               position = position_dodge(0.8), show.legend = FALSE) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
             size = 0.8, alpha = 0.4) +
  labs(title = "Pooled across all plates: African vs non-African by ε4 band",
       subtitle = "ε4-dosage separation is preserved in both groups; band medians nearly overlap.",
       x = "ε4 allele count (Sanger genotype)", y = "APOE4 − APOE (log2)",
       colour = NULL) +
  theme_bw(base_size = 11) + theme(legend.position = "top")
```

```{r}
#| echo: false
#| label: sw-november-hispanic
#| results: asis

# Is the November-1124 Bay1 biomarker anomaly driven by its Hispanic samples?
bay1 <- sw %>%
  filter(RUN == "20251124-1407_Bay1", call_category == "Both agree",
         !is.na(APOE4_minus_APOE), !is.na(e4_bio)) %>%
  mutate(bio_mismatch = e4_bio != e4_sanger,
         grp = ifelse(Ethnicity == "HI", "Hispanic", "Non-Hispanic (AA)"))

cat("### Is the November-1124 Bay1 anomaly a Hispanic effect?\n\n")
b1 <- bay1 %>%
  group_by(group = grp) %>%
  summarise(n = n(), mismatch = sum(bio_mismatch),
            pct = round(100 * mean(bio_mismatch), 1), .groups = "drop")
print(knitr::kable(b1))

cat("\n\n**No.** Removing the Hispanic samples does not clear the bay — the",
    "biomarker–genotype mismatch is essentially identical in its non-Hispanic",
    "(AA) samples (~41%) and Hispanic samples (~36%). The Bay1 anomaly is an",
    "ancestry-blind plate/assay effect, not a Hispanic or African signal.\n\n")
```

**Reading these results.**

- **Where are the HI_MU from?** Predominantly the Dominican Republic, Peru, and
  Puerto Rico (then Ecuador, Colombia, Cuba, Mexico) — admixed Caribbean and
  Andean Latin-American populations, exactly the ancestries where short-read WGS
  APOE calling is hardest.
- **Are the HI_MU in the "randomly-mixed-up" WGS set?** Yes — they are the
  *largest* discordant group (76 of 137 discordant samples; HI_WH 43, HI_BL 15,
  AA 3). The random-reassignment finding is driven mainly by HI_MU.
- **Do biomarkers plot lower for Africans?** No. The all-African plate is not
  lower; in the ε4-carrier bands it reads slightly *higher* (ε4=1: −1.71 vs
  −2.66 non-African; ε4=2: +0.09 vs −1.05), and African samples on other plates
  track non-African almost exactly. The ε4-dosage signal that underpins the
  biomarker arbitration is fully preserved across ancestry.
- **If we remove the HI, is the November plate OK?** The genotype discordance was
  never on the November plates (the African plate has no WGS; 1124-Bay1 had zero
  discordant pairs). The remaining 1124-Bay1 *biomarker* anomaly is not Hispanic-
  driven — it hits AA (~41%) and HI (~36%) equally — so dropping HI does not fix
  it. It is a localized assay/plate effect.

## Sample-switch hunt (adjacent sample numbers)

::: {.callout-note appearance="simple"}
**Superseded / confirmed in principle.** This hunt for adjacency-clustered, pairwise
WGS-label swaps was the right instinct — the true mix-up *was* a structured
plate-position swap (plate-2 180° rotation; plates 3↔7), resolved in
`../../../May_2026_WGS_QC/` from well geometry + the chrX fingerprint rather than from
biomarker-adjacency. On the corrected data the discordant set is the ~12 residual, so
little remains to cluster here.
:::

```{r}
#| echo: false
#| label: sw-switch-hunt
#| results: asis

adj <- sw_disc %>%
  filter(!is.na(samp_num)) %>%
  arrange(RUN, samp_num) %>%
  group_by(RUN) %>%
  mutate(gap_prev = samp_num - lag(samp_num),
         gap_next = lead(samp_num) - samp_num,
         adjacent_discordant = (!is.na(gap_prev) & gap_prev <= 2) |
                               (!is.na(gap_next) & gap_next <= 2)) %>%
  ungroup()

n_adj <- sum(adj$adjacent_discordant, na.rm = TRUE)
cat(sprintf("**%d** discordant samples have another discordant sample within 2 SAMPLE numbers in the same run — the signature of a plate/manifest mix-up rather than isolated genotyping miscalls.\n\n", n_adj))

# WGS-label crossover test: among adjacent discordant pairs, does swapping the two
# WGS calls make both match the biomarker-implied dosage?
pairs <- adj %>%
  arrange(RUN, samp_num) %>%
  group_by(RUN) %>%
  mutate(nb_gap = lead(samp_num) - samp_num,
         crossover = !is.na(nb_gap) & nb_gap <= 2 &
                     e4_wgs == lead(e4_bio) & lead(e4_wgs) == e4_bio &
                     e4_bio != lead(e4_bio)) %>%
  ungroup()

sx <- which(pairs$crossover)
if (length(sx) > 0) {
  sp <- data.frame(
    RUN      = pairs$RUN[sx],
    Sample_A = pairs$SAMPLE_ALIQUOT[sx],
    A_Sanger = pairs$sanger[sx], A_WGS = pairs$wgs[sx], A_bio_e4 = pairs$e4_bio[sx],
    Sample_B = pairs$SAMPLE_ALIQUOT[sx + 1],
    B_Sanger = pairs$sanger[sx + 1], B_WGS = pairs$wgs[sx + 1], B_bio_e4 = pairs$e4_bio[sx + 1]
  )
  cat(sprintf("\n**%d candidate WGS-label crossover pair(s)** — swapping the two WGS calls reconciles both with the biomarker:\n\n", nrow(sp)))
  print(knitr::kable(sp))
} else {
  cat("\nNo clean pairwise WGS-label crossovers detected; the adjacency clustering above suggests longer block shifts rather than isolated 2-sample swaps — review the flagged list below by run.\n\n")
}
```

## Flagged samples for follow-up

```{r}
#| echo: false
#| label: sw-flag-export
#| results: asis

flagged <- sw_disc %>%
  left_join(adj %>% select(SAMPLE_ALIQUOT, adjacent_discordant), by = "SAMPLE_ALIQUOT") %>%
  mutate(flag_category = dplyr::case_when(
    fits_neither                     ~ "Fits neither ε4 band (possible switch / assay)",
    e4_sanger == e4_wgs              ~ "Cannot adjudicate (ε2/ε3 only)",
    biomarker_favors == "Sanger"     ~ "Likely WGS error (biomarker backs Sanger)",
    biomarker_favors == "WGS"        ~ "Likely Sanger error (biomarker backs WGS)",
    TRUE                             ~ "Biomarker ambiguous"
  )) %>%
  select(SAMPLE_ALIQUOT, SAMPLE, RUN, CDX, any_of("ID2"),
         any_of(c("Ethnicity", "Ancestry", "Race", "Site",
                  "Country.State", "country_of_birth", "metatada_source")),
         sanger, wgs, e4_sanger, e4_wgs, e4_bio,
         APOE4_minus_APOE, dev_sanger, dev_wgs, dev_nearest,
         fits_neither, biomarker_favors, adjacent_discordant, flag_category) %>%
  arrange(flag_category, RUN, SAMPLE_ALIQUOT)

cat("### Flag summary\n\n")
print(knitr::kable(as.data.frame(table(`Flag category` = flagged$flag_category))))

write_xlsx(flagged, "Sanger_vs_WGS_biomarker_flags.xlsx")
cat("\n\n*Per-sample flags written to `Sanger_vs_WGS_biomarker_flags.xlsx` (",
    nrow(flagged), " discordant samples).*\n\n", sep = "")
```
