---
title: "APOE per-batch QC"
subtitle: "Sanger/WGS concordance · protein-vs-genotype outliers · per-run APOE4 assay screen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
params:
  merged_csv: "../../Metadata_Merge/output_files/filtered/filtered_combined_post_QC.csv"
  output_dir: "output_files"
  ref_bands: ""   # optional manual band override (Section A) for the no-clean-bay edge case; default "" = self-contained two-pass
knit: |
  (function(input, ...) {
    od  <- tryCatch(rmarkdown::yaml_front_matter(input)$params$output_dir, error = function(e) NULL)
    if (is.null(od) || !nzchar(od)) od <- "output_files"
    tag <- sub("^output_files_?", "", basename(od)); if (!nzchar(tag)) tag <- "run"
    rmarkdown::render(input, output_file = paste0("APOE_perbatch_QC_", tag, ".html"))
  })
---

<!--
Per-batch APOE QC. The routine, batch-agnostic subset extracted (2026-06-16) from the May-2026
investigation APOE_geno_protein.qmd (archived under _archive_APOE_pre_perbatch_2026Jun16/). Three checks,
all on one batch's Metadata_Merge output (filtered_combined_post_QC.csv):
  A. APOE4-protein-vs-genotype outliers   (APOE4 - APOE, genotype-anchored)
  B. Sanger-vs-WGS genotype concordance   (disagreements are EXPECTED under WGS-default — they flag the
     plate swap; the sample-specific re-map lives in the affected batch's WGS_Midi_Remap/ folder)
  C. Per-run APOE4 assay QC screen         (catches Bay1-type APOE4-channel failures)
Run via the YAML knit() or: rmarkdown::render("APOE_perbatch_QC.Rmd",
  params = list(merged_csv = "<...>/filtered_combined_post_QC.csv", output_dir = "output_files_<tag>"))
-->

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ggplot2); library(dplyr); library(writexl); library(knitr)
OUT <- params$output_dir
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)

dat <- read.csv(params$merged_csv, check.names = FALSE)
stopifnot(all(c("SAMPLE_ALIQUOT", "APOE.geno", "APOE4", "APOE") %in% names(dat)))

apoe_levels <- c("22","23","33","24","34","44")
.clean <- function(x) { y <- gsub("\\s|/", "", trimws(as.character(x))); y[y == ""] <- NA; y }
has_wgs <- "APOE_WGS" %in% names(dat) && any(!is.na(dat$APOE_WGS))
# WGS-default genotype for banding (the stated policy; matches the APOE4_count dosage used in C/D).
# Prefer the remapped APOE.geno_final; else coalesce(WGS, Sanger); else raw Sanger. (Section B below
# still reads raw APOE.geno directly for the Sanger-vs-WGS comparison.)
.geno_band <- if ("APOE.geno_final" %in% names(dat)) dat$APOE.geno_final else
              if (has_wgs) dplyr::coalesce(dat$APOE_WGS, dat$APOE.geno) else dat$APOE.geno
dat$APOEgeno         <- factor(.clean(.geno_band), levels = apoe_levels)
dat$APOE4_minus_APOE <- dat$APOE4 - dat$APOE

# plate = run-bay (RUN already encodes Bay). Drives the by-plate facet/lollipop figure heights.
has_run     <- "RUN" %in% names(dat) && any(!is.na(dat$RUN))
n_plates    <- if (has_run) dplyr::n_distinct(dat$RUN[!is.na(dat$RUN)]) else 0L
fig_h_facet <- max(4.5, ceiling(max(n_plates, 1) / 4) * 2.3)   # by-plate genotype facets
fig_h_lolli <- max(3,   n_plates * 0.26 + 1.2)                 # per-plate off-band lollipop
```

This report runs on **`r basename(params$merged_csv)`** (`r nrow(dat)` samples). APOE genotype policy is
**WGS-default**: `APOE.geno_final` prefers WGS, Sanger fills missing; Sanger-vs-WGS disagreements are
reported (Section B) — a cluster of them is the WGS plate-swap signal, corrected by the batch's own
`WGS_Midi_Remap/` re-map, not here.

# A. APOE4 − APOE protein-vs-genotype outliers

`APOE4 − APOE` should sit in a tight, genotype-specific band (ε4 dosage drives the APOE4 channel).
Samples far from their genotype's band are candidate genotype/label or assay problems.
Bands are **per-plate** (each run-bay vs. itself): plates legitimately drift a little, and this is QC —
not genotype-calling — so we absorb that and flag only **big within-plate** deviations. **Important:** a
broadly-shifted or scrambled plate would normally **self-band** (its inflated within-genotype spread
swallows its own outliers), so to prevent that a **high-spread plate** (within-genotype IQR ≫ the typical
plate's) is forced to use the **clean per-genotype reference band** instead of its own — so its scattered
points *are* flagged here (e.g. the Bay1 APOE4-scramble plate surfaces as the most outliers, not the
fewest). The faceted view below overlays a **pooled reference band** so a shifted plate is visible at a
glance, and Section C's per-run screen catches plate-level shifts independently. Genotype = the WGS-default
call (`APOE.geno_final`), so known Sanger errors that WGS corrected are not re-flagged here.

```{r outliers}
#| results: asis
#| fig-width: 8
#| fig-height: 5
plotdat <- dat %>% filter(!is.na(APOEgeno), !is.na(APOE4_minus_APOE))

# BANDING (each run-bay x genotype). Catch an individual off-band sample while (a) NOT over-flagging a
# tight cluster and (b) NOT being fooled by a SCRAMBLED plate that self-bands. Strategy:
#  - a plate whose within-genotype IQR is >> the typical plate's IQR is HIGH-SPREAD / problematic (e.g.
#    the Bay1 APOE4 scramble: its 33-band runs ~13x normal width);
#  - the per-genotype CLEAN reference band is built from the NORMAL-spread plates only (the scramble's
#    wild values don't inflate the reference);
#  - a normal plate uses its OWN band but with the width FLOORED at the clean reference (a tight cluster
#    can't make a sliver band); a high-spread or tiny-n plate uses the CLEAN reference band directly, so
#    its scattered points are flagged against the normal expectation.
MIN_N_BAND    <- 8   # min run x genotype n to trust a plate's own spread
SPREAD_FACTOR <- 3   # plate IQR > SPREAD_FACTOR x typical (median) plate IQR  =>  high-spread / problematic

gkey <- if (has_run) c("RUN", "APOEgeno") else "APOEgeno"
grp  <- plotdat %>% group_by(across(all_of(gkey))) %>%
  summarise(n_grp = n(), q1 = quantile(APOE4_minus_APOE,.25), q3 = quantile(APOE4_minus_APOE,.75),
            med = median(APOE4_minus_APOE), iqr = IQR(APOE4_minus_APOE), .groups = "drop")
if (has_run) {
  typ <- grp %>% filter(n_grp >= MIN_N_BAND) %>% group_by(APOEgeno) %>%
    summarise(typ_iqr = median(iqr), .groups = "drop")
  grp <- grp %>% left_join(typ, by = "APOEgeno") %>%
    mutate(high_spread = n_grp >= MIN_N_BAND & !is.na(typ_iqr) & iqr > SPREAD_FACTOR * typ_iqr)
  clean <- plotdat %>% semi_join(grp %>% filter(n_grp >= MIN_N_BAND, !high_spread), by = gkey) %>%
    group_by(APOEgeno) %>%
    summarise(c_lo = quantile(APOE4_minus_APOE,.25) - 3*IQR(APOE4_minus_APOE),
              c_hi = quantile(APOE4_minus_APOE,.75) + 3*IQR(APOE4_minus_APOE),
              c_med = median(APOE4_minus_APOE), .groups = "drop")
} else { grp <- grp %>% mutate(typ_iqr = NA_real_, high_spread = FALSE); clean <- tibble(APOEgeno = factor()) }

# pooled per-genotype band = fallback when a genotype has no clean plate
ref_pool <- plotdat %>% group_by(APOEgeno) %>%
  summarise(p_lo = quantile(APOE4_minus_APOE,.25) - 3*IQR(APOE4_minus_APOE),
            p_hi = quantile(APOE4_minus_APOE,.75) + 3*IQR(APOE4_minus_APOE),
            p_med = median(APOE4_minus_APOE), .groups = "drop")
# the per-genotype reference band actually used (clean where available, else pooled) — also the plot overlay
ref_band <- ref_pool %>% left_join(clean, by = "APOEgeno") %>%
  transmute(APOEgeno, lower = coalesce(c_lo, p_lo), upper = coalesce(c_hi, p_hi),
            genotype_median = coalesce(c_med, p_med))

bounds <- grp %>% left_join(ref_band, by = "APOEgeno") %>%
  rename(ref_lo = lower, ref_hi = upper, ref_med = genotype_median) %>%
  mutate(ref_half = (ref_hi - ref_lo) / 2,
         use_plate = n_grp >= MIN_N_BAND & !high_spread,
         lower = if_else(use_plate, pmin(q1 - 3*iqr, ref_med - ref_half), ref_lo),  # own band, width floored
         upper = if_else(use_plate, pmax(q3 + 3*iqr, ref_med + ref_half), ref_hi),
         genotype_median = if_else(use_plate, med, ref_med)) %>%
  select(all_of(gkey), n_grp, high_spread, lower, upper, genotype_median)

outliers <- plotdat %>% left_join(bounds, by = gkey) %>%
  filter(!is.na(lower) & (APOE4_minus_APOE < lower | APOE4_minus_APOE > upper))
ranked <- plotdat %>% left_join(bounds %>% select(all_of(gkey), genotype_median), by = gkey) %>%
  mutate(deviation = APOE4_minus_APOE - genotype_median, abs_deviation = abs(deviation)) %>%
  arrange(desc(abs_deviation))

# high-spread (problematic) plates — cross-checks the Section C screen
high_spread_plates <- if (has_run) grp %>% filter(high_spread) %>%
    transmute(RUN, genotype = APOEgeno, n = n_grp, plate_IQR = round(iqr,1),
              typical_IQR = round(typ_iqr,1), ratio = round(iqr/typ_iqr,1)) %>% arrange(desc(ratio)) else grp[0,]
if (has_run && nrow(high_spread_plates)) write_xlsx(high_spread_plates, file.path(OUT, "high_spread_plates.xlsx"))
cat(if (has_run)
  sprintf("Outlier bands are **per-plate** (each run-bay vs. itself; QC, not genotype-calling), but a plate whose within-genotype spread is >%d× the typical plate is treated as **high-spread/problematic** and is scored against a **clean reference band** (built from normal-spread plates) instead of self-banding — so a scrambled plate's off-band samples are flagged here, not hidden. Normal plates keep their own band with the width floored at the clean reference (tight clusters can't over-flag).\n\n", SPREAD_FACTOR)
  else "Per-genotype bands (single batch — no `RUN` column).\n\n")

cat(sprintf("- Genotyped samples with an APOE4 − APOE value: **%d**\n", nrow(plotdat)))
cat(sprintf("- Extreme outliers (outside their assigned 3×IQR band): **%d**\n", nrow(outliers)))
if (has_run) cat(sprintf("- High-spread / problematic (run-bay × genotype) bands re-scored against the clean reference: **%d**\n", nrow(high_spread_plates)))
cat("\n")
if (nrow(outliers)) print(kable(as.data.frame(table(genotype = outliers$APOEgeno)),
                                col.names = c("Genotype","n outliers")))
if (has_run && nrow(high_spread_plates)) {
  cat("\n**High-spread plates** (within-genotype IQR ≫ typical — scored against the clean band; cross-check vs Section C):\n\n")
  print(kable(high_spread_plates, col.names = c("Run-Bay","Genotype","n","plate IQR","typical IQR","ratio")))
  cat("\n")
}

# one jittered point layer (outliers coloured, not redrawn) -> each sample appears exactly ONCE (no
# doubling); violins scaled by COUNT so rare genotypes stay narrow; clean reference band overlaid.
pj <- plotdat %>% mutate(.outlier = SAMPLE_ALIQUOT %in% outliers$SAMPLE_ALIQUOT) %>% arrange(.outlier)
ggplot(pj, aes(APOEgeno, APOE4_minus_APOE)) +
  geom_violin(aes(fill = APOEgeno), scale = "count", alpha = .4, width = .9, colour = NA) +
  geom_crossbar(data = ref_band, aes(x = APOEgeno, y = genotype_median, ymin = lower, ymax = upper),
                inherit.aes = FALSE, width = .55, fill = NA, colour = "steelblue",
                linetype = "dashed", linewidth = .35) +
  geom_point(aes(colour = .outlier, size = .outlier),
             position = position_jitter(width = .15, height = 0, seed = 42), alpha = .55) +
  scale_colour_manual(values = c(`FALSE` = "grey45", `TRUE` = "red"), guide = "none") +
  scale_size_manual(  values = c(`FALSE` = 1.1,     `TRUE` = 2.3),   guide = "none") +
  labs(x = "APOE genotype", y = "APOE4 − APOE",
       title = "APOE4 − APOE by genotype (red = extreme outlier; blue dashed = clean reference band)") +
  theme_classic(base_size = 13) + theme(legend.position = "none", plot.title = element_text(face = "bold", hjust = .5))

write_xlsx(outliers, file.path(OUT, "APOE4_minus_APOE_extreme_outliers.xlsx"))
write_xlsx(ranked,   file.path(OUT, "Outliers_Ranked.xlsx"))
cat(sprintf("\n\nWrote `Outliers_Ranked.xlsx` (%d) and `APOE4_minus_APOE_extreme_outliers.xlsx` (%d) to `%s/`.\n\n",
            nrow(ranked), nrow(outliers), OUT))
```

## Same view, by plate

Each plate (run-bay) faceted, with the **clean per-genotype reference band** (blue dashed) drawn identically
in every facet — a plate whose points push past the band, or sit shifted, stands out at a glance.

```{r outliers-by-plate, fig.width=10, fig.height=fig_h_facet}
if (!has_run) {
  cat("*No `RUN` column — by-plate view skipped.*\n")
} else {
  pb <- plotdat %>%                          # red = outside its assigned band (matches §A)
    left_join(bounds, by = gkey) %>%
    mutate(outlier = !is.na(lower) & (APOE4_minus_APOE < lower | APOE4_minus_APOE > upper))
  ggplot(pb, aes(APOEgeno, APOE4_minus_APOE)) +
    # clean reference band (drawn identically in every facet): a scrambled plate's points spill past it
    # AND are now flagged red (its inflated own-band is overridden by the clean band).
    geom_crossbar(data = ref_band, aes(x = APOEgeno, y = genotype_median, ymin = lower, ymax = upper),
                  inherit.aes = FALSE, width = .65, fill = NA, colour = "steelblue",
                  linetype = "dashed", linewidth = .3) +
    geom_jitter(aes(colour = outlier), width = .15, size = .8, alpha = .6) +
    scale_colour_manual(values = c(`FALSE` = "grey35", `TRUE` = "red"), guide = "none") +
    facet_wrap(~ RUN, ncol = 4) +
    labs(x = "APOE genotype", y = "APOE4 − APOE",
         title = "APOE4 − APOE by genotype, per plate  (blue dashed = clean reference band; red = outlier)") +
    theme_bw(base_size = 9) +
    theme(plot.title = element_text(face = "bold", hjust = .5),
          panel.grid.minor = element_blank())
}
```

# B. Sanger vs WGS genotype concordance

```{r concordance}
#| results: asis
if (!has_wgs) {
  cat("*No `APOE_WGS` calls in this batch — concordance check skipped.*\n\n")
} else {
  dat$APOEgeno_WGS <- factor(.clean(dat$APOE_WGS), levels = apoe_levels)
  apoe_sanger      <- factor(.clean(dat$APOE.geno), levels = apoe_levels)   # raw Sanger (B compares Sanger vs WGS)
  conc <- dat %>% mutate(orig = .clean(APOE.geno), wgs = .clean(APOE_WGS)) %>%
    filter(!is.na(orig), !is.na(wgs)) %>% mutate(concordant = orig == wgs)
  n_wgs    <- sum(!is.na(dat$APOEgeno_WGS))
  n_gained <- sum(is.na(apoe_sanger) & !is.na(dat$APOEgeno_WGS))
  cat(sprintf("- Samples with a WGS genotype: **%d** / %d (%.1f%%)\n", n_wgs, nrow(dat), 100*n_wgs/nrow(dat)))
  cat(sprintf("- Gained a genotype from WGS (Sanger was NA): **%d**\n", n_gained))
  cat(sprintf("- Concordance where both available: **%.1f%%** (%d/%d)\n", 100*mean(conc$concordant),
              sum(conc$concordant), nrow(conc)))
  cat(sprintf("- **Discordant: %d** (expected to cluster if a plate swap is present — see WGS_Midi_Remap/)\n\n",
              sum(!conc$concordant)))
  cat("### Cross-tabulation (Sanger rows × WGS cols)\n\n")
  print(kable(table(Sanger = conc$orig, WGS = conc$wgs)))

  discord <- conc %>% filter(!concordant) %>%
    select(any_of(c("SAMPLE_ALIQUOT","SAMPLE","RUN","CDX","Group","Race","Ethnicity")),
           APOE.geno, APOE_WGS, APOE4_minus_APOE) %>% arrange(APOE.geno, APOE_WGS)
  write_xlsx(discord, file.path(OUT, "APOE_Sanger_vs_WGS_discordant.xlsx"))
  cat(sprintf("\n### Discordant samples (%d) — written to `APOE_Sanger_vs_WGS_discordant.xlsx`\n\n", nrow(discord)))
  if (nrow(discord)) print(kable(head(discord, 40), digits = 3))
}
```

# C. Per-run APOE4 assay QC screen

Genotype-anchored: for each run-bay, the share of genotyped samples whose APOE4 sits off its ε4-dosage
band. Catches APOE4-channel failures (e.g. `20251124-1407_Bay1`) that control replicates and sample-level
auto-QC miss. **A flagged run-bay → drop its APOE4 / APOE4 − APOE (genotypes unaffected; total APOE and
the other proteins remain usable).**

```{r per-run-apoe4}
#| results: asis
OFF_PCT_THRESH  <- 15   # flag if >15% of genotyped samples off-band
IQR_MULT_THRESH <- 3    # ...or carrier-band APOE4 IQR > 3x the median run-bay
if (!all(c("APOE4_count","RUN") %in% names(dat)) || sum(!is.na(dat$APOE4_count)) == 0) {
  cat("*`APOE4_count` / `RUN` not available — per-run APOE4 screen skipped.*\n\n")
} else {
  qc <- dat %>% filter(!is.na(APOE4_count), !is.na(APOE4))
  band_md <- qc %>% group_by(APOE4_count) %>% summarise(md = median(APOE4), .groups = "drop")
  qc <- qc %>% mutate(off = abs(APOE4 - band_md$md[match(APOE4_count, band_md$APOE4_count)]) > 2)
  run_qc <- qc %>% group_by(RUN) %>%
    summarise(n = n(), carriers = sum(APOE4_count >= 1),
              APOE4_carrier_IQR = round(IQR(APOE4[APOE4_count >= 1], na.rm = TRUE), 2),
              off_band_pct = round(100*mean(off), 1), .groups = "drop") %>%
    filter(n >= 15)
  med_iqr <- median(run_qc$APOE4_carrier_IQR, na.rm = TRUE)
  run_qc <- run_qc %>%
    mutate(FLAG = off_band_pct > OFF_PCT_THRESH | APOE4_carrier_IQR > IQR_MULT_THRESH * med_iqr) %>%
    arrange(desc(off_band_pct))

  cat(sprintf("Screened **%d** run-bays (n ≥ 15 genotyped). Median off-band **%.1f%%**, median carrier-band IQR **%.2f**. Flag: off-band > %d%% **or** carrier IQR > %d× median.\n\n",
              nrow(run_qc), median(run_qc$off_band_pct), med_iqr, OFF_PCT_THRESH, IQR_MULT_THRESH))
  flagged <- run_qc %>% filter(FLAG)
  if (nrow(flagged)) {
    cat(sprintf("**%d run-bay(s) FLAGGED — treat APOE4 (and APOE4 − APOE) as unreliable there:**\n\n", nrow(flagged)))
    print(kable(flagged, col.names = c("Run-Bay","n","carriers","APOE4 carrier IQR","off-band %","FLAG")))
  } else cat("**No run-bays flagged.**\n\n")
  cat("\nWorst run-bays on the screen:\n\n")
  print(kable(head(run_qc, 8), col.names = c("Run-Bay","n","carriers","APOE4 carrier IQR","off-band %","FLAG")))
  write_xlsx(run_qc, file.path(OUT, "APOE4_per_run_QC.xlsx"))
  cat(sprintf("\n\nWrote `APOE4_per_run_QC.xlsx` (%d run-bays) to `%s/`.\n\n", nrow(run_qc), OUT))
}
```

```{r per-run-apoe4-plot, fig.width=8, fig.height=fig_h_lolli}
if (exists("run_qc")) {
  ggplot(run_qc, aes(x = reorder(RUN, off_band_pct), y = off_band_pct, colour = FLAG)) +
    geom_segment(aes(xend = RUN, yend = 0), linewidth = .4) +
    geom_point(size = 2.2) +
    geom_hline(yintercept = OFF_PCT_THRESH, linetype = "dashed", colour = "red") +
    coord_flip() +
    scale_colour_manual(values = c(`FALSE` = "grey45", `TRUE` = "red"), guide = "none") +
    labs(x = NULL, y = "off-band % of genotyped samples",
         title = sprintf("Per-plate APOE4 off-band %%  (dashed = %d%% flag cutoff; red = flagged)", OFF_PCT_THRESH)) +
    theme_bw(base_size = 10) + theme(plot.title = element_text(face = "bold", hjust = .5),
                                     panel.grid.minor = element_blank())
}
```

# D. Adjacent-accession reciprocal swap candidates

Two samples with **nearby accession numbers in the same run-bay** whose APOE4 readings **reciprocally
cross** (each looks like the other's ε4 dosage) and where **swapping fixes both** are candidate
neighbour-tube swaps (accessioning / layout slips). This is a standing version of the ALZ123
side-investigation (`investigate_outlier_swaps.R`). Confirm a candidate by checking source-rack
positions and re-genotyping / re-assaying APOE4 — prioritise pairs whose genotypes are Sanger==WGS confirmed.

```{r swap-detector, results='asis'}
SWAP_GAP <- 5   # accession gap to call two samples "neighbours"
if (!all(c("APOE4_count","RUN","SAMPLE") %in% names(dat)) || sum(!is.na(dat$APOE4_count)) == 0) {
  cat("*APOE4_count / RUN / SAMPLE not available - swap scan skipped.*\n\n")
} else {
  ng <- function(g){ g <- gsub("[/ ]","",trimws(as.character(g))); ifelse(g==""|is.na(g)|nchar(g)!=2, NA, g) }
  sw <- dat
  sw$snum <- suppressWarnings(as.numeric(sw$SAMPLE))
  sw$gs <- ng(sw$APOE.geno); sw$gw <- if ("APOE_WGS" %in% names(sw)) ng(sw$APOE_WGS) else NA
  flagged_runs <- if (exists("run_qc")) run_qc$RUN[run_qc$FLAG] else character(0)
  # ε4-dosage bands from a clean reference (exclude bays the per-run screen flagged)
  refb <- sw[!is.na(sw$APOE4_count) & !is.na(sw$APOE4_minus_APOE) & !(sw$RUN %in% flagged_runs), ]
  bb <- refb %>% group_by(APOE4_count) %>%
    summarise(lo = quantile(APOE4_minus_APOE,.25) - 3*IQR(APOE4_minus_APOE),
              hi = quantile(APOE4_minus_APOE,.75) + 3*IQR(APOE4_minus_APOE),
              md = median(APOE4_minus_APOE), .groups = "drop")
  LO<-setNames(bb$lo,bb$APOE4_count); HI<-setNames(bb$hi,bb$APOE4_count); MD<-setNames(bb$md,bb$APOE4_count)
  inb  <- function(v,e) !is.na(v) & !is.na(e) & v >= LO[as.character(e)] & v <= HI[as.character(e)]
  sw$impl <- vapply(sw$APOE4_minus_APOE, function(v) if (is.na(v)) NA_integer_ else
                    as.integer(names(MD)[which.min(abs(v - MD))]), integer(1))
  cand <- sw[!is.na(sw$snum) & !is.na(sw$APOE4_count) & !is.na(sw$impl) & !is.na(sw$APOE4_minus_APOE), ]
  res <- list()
  for (r in unique(cand$RUN)) {
    s <- cand[cand$RUN == r, ]; s <- s[order(s$snum), ]; if (nrow(s) < 2) next
    for (i in 1:(nrow(s)-1)) for (j in (i+1):nrow(s)) {
      if (s$snum[j] - s$snum[i] > SWAP_GAP) next
      if (s$APOE4_count[i] != s$APOE4_count[j] &&
          s$impl[i] == s$APOE4_count[j] && s$impl[j] == s$APOE4_count[i] &&
          inb(s$APOE4_minus_APOE[j], s$APOE4_count[i]) && inb(s$APOE4_minus_APOE[i], s$APOE4_count[j])) {
        res[[length(res)+1]] <- data.frame(
          RUN = r, A = s$SAMPLE[i], B = s$SAMPLE[j], acc_gap = s$snum[j] - s$snum[i],
          Site_A = if ("Site" %in% names(s)) as.character(s$Site[i]) else NA,
          Site_B = if ("Site" %in% names(s)) as.character(s$Site[j]) else NA,
          Sanger_A = s$gs[i], WGS_A = s$gw[i], Sanger_B = s$gs[j], WGS_B = s$gw[j],
          e4_A = s$APOE4_count[i], implies_A = s$impl[i], e4_B = s$APOE4_count[j], implies_B = s$impl[j],
          geno_confirmed = (!is.na(s$gw[i]) && s$gs[i]==s$gw[i]) && (!is.na(s$gw[j]) && s$gs[j]==s$gw[j]),
          in_flagged_bay = r %in% flagged_runs, stringsAsFactors = FALSE)
      }
    }
  }
  swaps <- if (length(res)) do.call(rbind, res) else data.frame()
  if (nrow(swaps)) {
    swaps <- swaps[order(swaps$in_flagged_bay, swaps$acc_gap), ]
    cat(sprintf("**%d candidate swap pair(s)** — same run-bay, accession gap <= %d, reciprocal e4 cross, swapping fixes both.\n\n",
                nrow(swaps), SWAP_GAP))
    print(knitr::kable(swaps, row.names = FALSE))
    cat("\n\n*`geno_confirmed = TRUE`: Sanger and WGS agree for both samples, so the genotype is certain and it is the",
        "sample/protein that crossed. `in_flagged_bay = TRUE`: the pair sits in a bay already flagged by the per-run",
        "APOE4 screen (Section C) — interpret as part of that bay's scramble, not an isolated swap.*\n\n")
    writexl::write_xlsx(swaps, file.path(OUT, "APOE_swap_candidates.xlsx"))
    cat(sprintf("Wrote `APOE_swap_candidates.xlsx` (%d) to `%s/`.\n\n", nrow(swaps), OUT))
  } else cat("No adjacent-accession reciprocal swap candidates detected.\n\n")
}
```
